---
title: "Testing Compost Maturity"
runningheader: "Compost Maturity" # only for pdf output
subtitle: "RTricks: Compost Maturity" # only for html output
author: "Marc Los Huertos"
date: "`r Sys.Date()`"
output:
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_html: default
bibliography: skeleton.bib
link-citations: yes
---

```{r setup, include=FALSE}
library(tufte)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
```

# Introduction

##Broader Impacts

Research on compost maturity has significant implications across agriculture, environmental management, waste reduction, and public health. In agriculture, mature compost enhances soil structure, fertility, and water retention, boosting crop yields and reducing reliance on synthetic fertilizers and pesticides. It also supports climate-resilient farming by improving soil carbon sequestration and drought tolerance, promoting sustainable practices in changing environmental conditions.

In waste management, optimizing compost maturity diverts organic waste from landfills, reducing methane emissions and contributing to a circular economy by transforming waste into valuable resources. Efficient composting processes stabilize organic matter more quickly, recovering nutrients while enhancing waste system sustainability.

Environmentally, mature compost minimizes greenhouse gas emissions, prevents nutrient leaching, and protects water quality. It rehabilitates degraded lands, restores ecological balance, and replaces synthetic fertilizers, mitigating the impacts of fossil fuel-based agricultural inputs. These benefits make compost a key tool for environmental restoration and emission reductions.

Public health also benefits from compost maturity. Properly matured compost eliminates pathogens, reduces odor, and limits volatile organic compound emissions, improving air quality and safety. In agriculture, it minimizes contamination risks, supporting healthier food systems and reducing public health concerns

# Procedures

## Methods to Measure Compost Maturity Using Solvita’s Paddles

The Solvita Compost Maturity Test is a widely recognized tool for assessing compost stability and maturity. It provides a straightforward, reliable method for evaluating the quality of compost by measuring the biological activity and carbon dioxide (CO~2) and ammonia (NH~3) emissions. Here's how to use Solvita paddles to measure compost maturity:

1. Prepare the Compost Sample

Collect Representative Compost: Take a composite sample from various parts of the compost pile to ensure it accurately represents the batch. Avoid sections with excess moisture or contamination.

Moisture Adjustment: Ensure the compost is at an appropriate moisture level (approximately 50% moisture content). To test, the compost should feel damp but not release water when squeezed.

2. Set Up the Test

Select the Testing Container: Use a sealed jar or a Solvita-approved container to create a controlled environment for gas emissions.

Place Compost in the Jar: Fill the container with the appropriate amount of compost (usually 100-150 grams), leaving space for air circulation.

Insert the Solvita Paddle: Place the paddle carefully in the jar, ensuring it does not come into direct contact with the compost. The paddle should sit securely to allow accurate gas measurement.

3. Conduct the Test

Seal the Container: Close the container tightly to prevent outside air from affecting the readings.

Incubation Period: Allow the container to sit undisturbed for 4-24 hours at room temperature. The duration depends on the specific test protocol and the compost's expected activity level.

Color Change Observation: After the incubation period, examine the Solvita paddle for color changes. Each paddle has a color-coded scale corresponding to CO~2 and NH~3 levels, indicating compost maturity.

4. Interpret the Results

CO~2 and NH~3 Scales: Match the paddle colors to the provided Solvita chart to determine CO~2 respiration and NH~3 release levels. Lower emissions indicate greater stability and maturity.

Maturity Index: Combine the CO~2 and NH~3 results to calculate the Solvita Compost Maturity Index, typically ranging from 1 (very immature) to 8 (fully mature).

Maturity Assessment: Compost with a Solvita score of 6 or higher is generally considered mature and suitable for most agricultural applications.

5. Document and Repeat

Record Data: Document the test results, including the compost batch, sample date, moisture level, and Solvita score.

Re-test if Needed: For borderline results or when composting conditions change significantly, repeat the test to ensure consistency.


## Notes and Best Practices

Consistency: Always follow the same sampling and testing protocols to maintain accuracy.

Supplementary Tests: Pair Solvita testing with other assessments (e.g., temperature, pH, or C/N ratio) for a comprehensive evaluation of compost quality.

Safety: Handle compost samples with gloves and work in a well-ventilated area to avoid exposure to gases.

By following these steps, the Solvita paddle test provides an efficient way to measure compost maturity, helping to ensure its suitability for agricultural, landscaping, or restoration use.

# Analysis of Variance: Testing Multiple Treatments

## What is ANOVA?

ANOVA (Analysis of Variance) is a statistical method used to compare the means of three or more groups to determine if at least one group mean is significantly different from the others. It tests the null hypothesis that all group means are equal against the alternative hypothesis that at least one group mean is different. ANOVA is widely used in experiments and observational studies to analyze the impact of categorical independent variables on a continuous dependent variable.

## Types of ANOVA

- One-Way ANOVA: Compares means across one categorical independent variable with multiple levels (e.g., different fertilizers on plant growth).

- Two-Way ANOVA: Compares means across two independent variables and evaluates their interaction effects (e.g., fertilizers and watering frequency on plant growth).

- Repeated Measures ANOVA: Used when the same subjects are measured multiple times under different conditions.

## Assumptions of ANOVA
1. The dependent variable is continuous and normally distributed.

2. The independent variable(s) consist of two or more categorical, independent groups.

3. Homogeneity of variances (equal variance across groups).

4. Observations are independent.

## Coding in R

```{r}
# Example dataset: Plant growth across different fertilizers
data <- data.frame(
  Fertilizer = rep(c("A", "B", "C"), each = 10),
  Growth = c(rnorm(10, mean = 20, sd = 2), 
             rnorm(10, mean = 25, sd = 2), 
             rnorm(10, mean = 30, sd = 2))
)
head(data) # Display the first few rows of the data
```





