---
title: "Group 3: Storm Events and Water Impairments"
author: "Marc Los Huertos"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Correlation Tests

A correlation test in R is used to evaluate the strength and direction of a relationship between two numeric variables. The most common correlation test is Pearson’s correlation, but others include Spearman’s and Kendall’s if the data isn’t normally distributed.

Here’s a simple example using Pearson’s correlation with the built-in cor.test() function in R.


See [Correlation Test Tutorial](https://www.sthda.com/english/wiki/correlation-test-between-two-variables-in-r)

### An example

```{r}
# Sample data
height <- c(160, 165, 170, 175, 180)
weight <- c(55, 60, 65, 70, 75)

# Perform Pearson correlation test
cor.test(height, weight, method = "pearson")

```

cor: 0.9984 → Strong positive correlation

p-value: 0.000635 → Very small, so the result is statistically significant

95% CI: [0.9771, 0.9999] → We're 95% confident the true correlation lies in this interval.


## Hypotheses

## Data Set
```{r}

group3.csv = "/home/mwl04747/RTricks/00_Project_Group_Demos/Group3_FakData.csv"
riohonda = read.csv(group3.csv)
head(riohonda)
names(riohonda) = c("ID", "Date", "Phosphate", "Ammonia")

data2.csv = "/home/mwl04747/RTricks/00_Project_Group_Demos/Group3_FakeData2.csv"
data2 = read.csv(data2.csv); 

names(data2) = c("ID", "Distance", "Date", "Phosphate", "Ammonia",  "Nitrate")





```


```{r}
#install.packages("ggpubr")

library(ggpubr)
library(ggplot2)


```
## Summary Stats

```{r}
summary(riohonda)
summary(data2)

```



## Hypothesis Tests
```{r}

cor(riohonda$Ammonia, riohonda$Phosphate,  method = "pearson")

# Create correlation matrix



```






## Plots

```{r}
plot(Phosphate ~ Distance, data=data2)
abline(coef(lm(Phosphate ~ Distance, data=data2)), col="red")

summary(lm(Phosphate ~ Distance, data=data2))
```


k