---
title: "Group 2: Bokashi/Compost Bioremediation"
author: "Marc Los Huertos"
date: "`r Sys.Date()`"
output: pdf_document
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load libraries

library(ggplot2)
library(dplyr)
library(broom)      # for tidy model summaries
#library(kableExtra) # for nice tables
set.seed(123)       # reproducibility


```

## Introduction

This document performs a Before-After-Control-Impact (BACI) analysis using simulated data.
The goal is to detect whether an environmental impact causes a change relative to a control site.

## Hypotheses

## Data Set

```{r}
# Load the data
fakedata.csv = "/home/mwl04747/RTricks/00_Project_Group_Demos/Group2_FakeData.csv"
group1data = read.csv(fakedata.csv, header = TRUE)
str(group1data)
names(group1data)
unique(group1data$Treatment)
unique(group1data$Before.After)
```
## Removing Water -- since that is a correction, not a factor
```{r}

# Remove water
group1data <- group1data %>%
  filter(Treatment != "Water (control)")
# Remove the "Water" treatment

# Check the data
unique(group1data$Treatment)
```

```{r}
# Make sure site and time are factors
group1data <- group1data %>%
  mutate(Treatment = factor(Treatment),
         Before.After = factor(Before.After),
         Concentraion..ppm. = factor(Concentraion..ppm.))

# Check the data
str(group1data)
group1data[sample(1:nrow(group1data), 8), ]
```


## Summary Stats

Fake data isn't really working -- since you don't have any variance with ppm. We are not going reporting absorbance values, but I'll do this now, since it looks like there is some variance there.

```{r}
# Summarize the data
summary_stats <- group1data %>%
  group_by(Treatment, Before.After, Concentraion..ppm.) %>%
  summarise(
    mean = mean(Absorption),
    sd = sd(Absorption),
    n = n()
  ) %>%
  ungroup()
summary_stats
```


## Hypothesis Tests

waiting to see hypotheses!

```{r}
# Fit a linear model
model <- lm(Absorption ~ Treatment + Concentraion..ppm. * Before.After, data = group1data)
# Summarize the model
anova(model)


model_summary <- tidy(model)
model_summary


# Key interaction term:
key_interaction <- coef(model)["Treatmentimpact:Before.Afterafter"]

```



## Plots
```{r}
ggplot(group1data, aes(x = Before.After, y = Absorption, color = Treatment, group = Treatment)) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun = mean, geom = "point") +
  labs(
    title = "BACI Design: Mean Measured Values",
    subtitle = "Before and After at Control vs Impact Sites",
    y = "Measured Value",
    x = "Time",
    color = "Treatment"
  ) +
  scale_color_manual(values = c("control" = "#1f77b4", "impact" = "#d62728")) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")
```




5. Conclusion
The analysis suggests:

There was a decrease in the measured value at the impact site after the event.

The significant site × time interaction supports a likely effect of the environmental disturbance.

6. Appendix (Optional)
You can improve the model by considering:

Mixed models (lmer) if you have random effects (e.g., multiple sites)

Repeated measures ANOVA

Adding covariates (e.g., weather, seasonality)