---
title: "Group 6: Land-use Impacts on Biodiversity along FW-Coastal Gradients"
author: "Marc Los Huertos"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Introduction

The \textbf{Shannon-Weaver Index}, also known as the \textbf{Shannon-Wiener Index} or simply the \textbf{Shannon Diversity Index}, is a widely used metric in ecology for measuring biodiversity. It incorporates both species richness and evenness into a single value.

\section*{Historical Background}

\begin{itemize}
    \item \textbf{1948}: Claude Shannon introduced the concept of \textit{entropy} in his paper \textit{``A Mathematical Theory of Communication''}, where he quantified the uncertainty in a message source using:
    \[
    H = -\sum p_i \log_2 p_i
    \]
    
    \item \textbf{Warren Weaver} later helped interpret and popularize this theory, leading to the combined attribution as the \textbf{Shannon-Weaver Index}.
    
    \item In the \textbf{1950s--60s}, ecologists adapted the entropy formula to describe biodiversity, treating species as symbols and their relative abundances as probabilities.
    
    \item Sometimes incorrectly referred to as the \textbf{Shannon-Wiener Index} due to confusion with Norbert Wiener.
\end{itemize}

## Implemented in R


```{r}
group6.csv = "/home/mwl04747/RTricks/00_Project_Group_Demos/Group6_FakeData.csv"

# Read in the data
group6 = read.csv(group6.csv, header = TRUE)

# check read data
head(group6)
names(group6)
```

## Calculating Shannon-diversity index

### Definition

The Shannon Diversity Index (also known as the Shannon-Wiener or Shannon-Weaver Index) is a measure of biodiversity that accounts for both species richness and species evenness. It is defined mathematically as:

\[
H' = -\sum_{i=1}^{S} p_i \ln p_i
\]

\noindent Where:
\begin{itemize}
    \item \( H' \) is the Shannon diversity index.
    \item \( S \) is the total number of species.
    \item \( p_i \) is the proportion of individuals belonging to the \( i \)-th species.
    \item \( \ln \) is the natural logarithm.
\end{itemize}

### Calculation Steps

To calculate the index:
\begin{enumerate}
    \item Count the total number of individuals across all species, denoted \( N \).
    \item For each species \( i \), calculate the proportion \( p_i = \frac{n_i}{N} \), where \( n_i \) is the number of individuals in species \( i \).
    \item Plug each \( p_i \) into the formula and compute the sum.
\end{enumerate}

\section*{Alternate Formula Using Counts}

Using species counts directly:

\[
H' = -\sum_{i=1}^{S} \left( \frac{n_i}{N} \ln \frac{n_i}{N} \right)
\]

### Example Calculation in R

```{r}

library(vegan)

# Example data: rows = sites, columns = species counts
species_data <- matrix(
  c(5, 10, 15,
    8, 12, 10),
  nrow = 2,
  byrow = TRUE
)

# Calculate Shannon index
shannon_index <- diversity(species_data, index = "shannon")

# Print result
print(shannon_index)
```


The \texttt{diversity()} function defaults to the Shannon index when \texttt{index = "shannon"}. It computes the index for each row (e.g., sample or site).

## Example Calculation with Fake Data!

  
```{r}
# Load the required library
library(vegan)
# Calculate the Shannon diversity index
shannon_diversity = function(data) {
  # Remove any rows with NA values
  data = na.omit(data)
  
  # Calculate the Shannon diversity index
  shannon_index = diversity(data, index = "shannon")
  
  return(shannon_index)
  
}

```

```{r}
# Calculate the Shannon diversity index for phytoplankton and fish from group6 data
shannon_diversity_phytoplankton <- shannon_diversity(group6[, c("Athecate.dinoflagellates", "Thecate.dinoflagellates", "Silicoflagellates", "Diatoms")]); shannon_diversity_phytoplankton

shannon_diversity_fish <- shannon_diversity(group6[, c("Coccolithophores", "Centrarchidae", "Salmonidae", "Cyprinidae", "Serranidae", "Sciaenidae")]); 

str(shannon_diversity_fish)

fish = as.data.frame(shannon_diversity_fish); fish
fish$Location = group6$X; fish

## Let's add the location to dataframe

shannon_diversity_phytoplankton$Location = group6$X; shannon_diversity_phytoplankton


```

## Combine with WQ data! super neat dataset -- great job!!



## Hypotheses

How do phytoplankton and fish diversity vary with sampling location?




## Plots