---
title: "EA30 Skill Tutorial: Water Quality Analysis with R"
author: "EA30 – Environmental Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(lubridate)
library(trend)
library(broom)
```

# Purpose (EA30 Skills Alignment)

This tutorial is a **skill‑based example** designed for **EA30 (Introduction to Environmental Analysis)**. The goal is to practice **core quantitative reasoning with R** using realistic, simulated **water quality data**, following **best practices used by environmental scientists**.

Students completing this tutorial will practice:

- Importing and inspecting environmental data
- Applying regulatory standards (EPA)
- Conducting statistical tests commonly used in water analysis
- Interpreting results in an environmental management context

This is **not about memorization**, but about **using tools to answer environmental questions**.

---

# Water Quality Context

We analyze three common water sources:

- **Groundwater** (wells)
- **Surface water** (streams/rivers)
- **Drinking water** (treated supply)

Key indicators:

- **Nitrate (NO3⁻)** – nutrient pollution
- **Lead (Pb)** – toxic metal
- **E. coli** – microbial contamination

---

# Regulatory Benchmarks (EPA)

```{r epa}
epa <- tibble(
  parameter = c("Nitrate", "Lead", "E_coli"),
  standard = c(10, 15, 1),
  units = c("mg/L", "µg/L", "MPN / 100 mL"),
  rule = c("MCL", "Action Level", "Presence/Absence")
)
epa
```

These benchmarks allow us to translate chemistry data into **risk and compliance questions**.

---

# Simulated EA30 Water Quality Dataset

The following data are **simulated but realistic**, reflecting seasonal variability and differences among water types.

```{r simulate-data}
set.seed(30)

n_months <- 10 * 12

data <- tibble(
  date = seq.Date(ymd("2015-01-01"), by = "month", length.out = n_months),
  month = month(date),
  type = rep(c("Groundwater", "Surface Water", "Drinking Water"), each = n_months),
  nitrate = c(
    rlnorm(n_months, log(7), 0.4),   # groundwater
    rlnorm(n_months, log(5), 0.5),   # surface water
    rlnorm(n_months, log(3), 0.3)    # drinking water
  ),
  lead = c(
    rlnorm(n_months, log(10), 0.7),
    rlnorm(n_months, log(6), 0.6),
    rlnorm(n_months, log(4), 0.5)
  ),
  ecoli = c(
    rpois(n_months, 0.2),
    rpois(n_months, 1.5),
    rpois(n_months, 0.05)
  )
)

head(data)
```

---

# Skill 1: Identify Exceedances

```{r exceedance}
data <- data %>%
  mutate(
    nitrate_exc = nitrate > 10,
    lead_exc = lead > 15,
    ecoli_exc = ecoli >= 1
  )

exceed_summary <- data %>%
  group_by(type) %>%
  summarise(across(ends_with("_exc"), mean))

exceed_summary
```

**Skill focus:** translating raw measurements into environmental meaning.

---

# Skill 2: Seasonal Kendall Trend Test (Best Practice)

The **Seasonal Kendall test** is widely used for long‑term environmental time series.

```{r sk}
sk_nitrate <- data %>%
  group_by(type) %>%
  do(tidy(seaken(.$nitrate, .$date)))

sk_nitrate
```

**Interpretation skill:** Is water quality improving, degrading, or stable?

---

# Skill 3: ANOVA – Comparing Water Sources

```{r anova}
anova_nitrate <- aov(nitrate ~ type, data = data)
summary(anova_nitrate)
```

**Skill focus:** detecting systematic differences among environmental systems.

---

# Skill 4: Toxicity & Risk (Logistic Regression)

We model the **probability of regulatory exceedance** as a measure of toxicity risk.

```{r logistic}
logit_model <- glm(lead_exc ~ type + nitrate,
                   family = binomial,
                   data = data)

tidy(logit_model, exponentiate = TRUE)
```

**Interpretation skill:** understanding predictors of environmental risk.

---

# Skill 5: Visualizing Data Relative to Standards

```{r plot}
ggplot(data, aes(date, nitrate, color = type)) +
  geom_line(alpha = 0.6) +
  geom_hline(yintercept = 10, linetype = "dashed") +
  labs(
    title = "Nitrate Concentrations Relative to EPA Standard",
    y = "Nitrate (mg/L)",
    x = "Year"
  )
```

---

# EA30 Takeaways

By completing this tutorial, you practiced:

✅ Using R to analyze real‑world environmental data
✅ Applying EPA standards to interpret results
✅ Using statistical tools common in water quality assessments
✅ Framing results as **environmental decisions**, not just numbers

These are foundational skills you will reuse throughout **environmental analysis, policy, and science courses**.

---

# Reflection (Optional)

- Which water source showed the highest risk?
- Which metric do you think poses the greatest human health concern?
- How would real data complicate this analysis?

