---
title: "Shifting Baselines and Changepoint Analysis"
author: "Marc Los Huertos"
date: "EA030 · Compiled `r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    self_contained: true
    theme: null
    highlight: null
    toc: false
knit: (function(inputFile, encoding) {
    rmarkdown::render(
      inputFile,
      encoding = encoding,
      output_file = file.path(dirname(inputFile),
                              "../docs/advanced_stats/05_Shifting_Baselines.html")
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  eval    = TRUE,
  fig.width  = 8,
  fig.height = 4
)
```

```{css, echo=FALSE}
@import url('https://fonts.googleapis.com/css2?family=Libre+Baskerville:ital,wght@0,400;0,700;1,400&family=Source+Code+Pro:wght@400;600&family=Lato:wght@300;400;700&display=swap');

*, *::before, *::after { box-sizing: border-box; margin: 0; padding: 0; }

:root {
  --forest:    #1e3a2f;
  --moss:      #2d5a3d;
  --fern:      #4a7c5e;
  --sage:      #8aad96;
  --mist:      #d4e6da;
  --cream:     #f7f3ed;
  --parchment: #ede8df;
  --ink:       #1a1a18;
  --charcoal:  #3a3a38;
  --dust:      #7a7a72;
  --amber:     #c8832a;
  --amber-lt:  #e8a84a;
  --white:     #ffffff;
  --card-shadow: 0 2px 10px rgba(30,58,47,.09);
}

html { scroll-behavior: smooth; }
body {
  font-family: 'Lato', sans-serif;
  background: var(--cream);
  color: var(--ink);
  font-size: 17px;
  line-height: 1.75;
  padding-left:  clamp(14px, 4vw, 48px);
  padding-right: clamp(14px, 4vw, 48px);
  padding-bottom: 80px;
}
.main-container { max-width: 920px !important; margin: 0 auto !important;
  padding: 0 clamp(10px, 2vw, 32px) 80px !important; }
.page-header {
  margin-left:  calc(-1 * clamp(14px, 4vw, 48px));
  margin-right: calc(-1 * clamp(14px, 4vw, 48px));
}
h1.title, h4.author, h4.date { display: none; }

.page-header {
  background-image:
    repeating-linear-gradient(120deg, transparent, transparent 60px,
      rgba(255,255,255,.018) 60px, rgba(255,255,255,.018) 61px),
    linear-gradient(160deg, #1e3a2f 0%, #2d5a3d 60%, #1a3d30 100%);
  color: #fff; padding: 44px 4vw 36px; margin: 0 -4vw 48px;
  border-bottom: 4px solid var(--amber); position: relative; overflow: hidden;
}
.page-header::after {
  content: ''; position: absolute; right: -40px; top: -40px;
  width: 260px; height: 260px; border-radius: 50%;
  background: radial-gradient(circle, rgba(138,173,150,.12) 0%, transparent 70%);
  pointer-events: none;
}
.page-header .module-tag {
  display: inline-block; font-family: 'Source Code Pro', monospace;
  font-size: .68rem; font-weight: 600; letter-spacing: .18em; text-transform: uppercase;
  color: var(--amber-lt); border: 1px solid rgba(200,131,42,.45);
  padding: 3px 10px; border-radius: 2px; margin-bottom: 12px;
}
.page-header h1 {
  font-family: 'Libre Baskerville', serif; font-size: clamp(1.6rem, 3vw, 2.3rem);
  font-weight: 700; color: #fff; line-height: 1.2; margin-bottom: 8px;
}
.page-header h1 span { color: var(--sage); }
.page-header .sub { font-size: .9rem; color: rgba(255,255,255,.65); font-weight: 300; }
.page-header .sub strong { color: rgba(255,255,255,.9); font-weight: 700; }
.back-link {
  display: inline-block; margin-top: 16px;
  font-family: 'Source Code Pro', monospace; font-size: .7rem; letter-spacing: .08em;
  color: var(--amber-lt); text-decoration: none;
  border: 1px solid rgba(200,131,42,.4); padding: 4px 11px; border-radius: 3px;
  transition: background .2s;
}
.back-link:hover { background: rgba(200,131,42,.18); }

.section-nav {
  display: flex; gap: 8px; flex-wrap: wrap; margin-bottom: 44px;
  padding-bottom: 20px; border-bottom: 1px solid var(--parchment);
}
.section-nav a {
  font-family: 'Source Code Pro', monospace; font-size: .7rem; font-weight: 600;
  letter-spacing: .08em; text-transform: uppercase;
  color: var(--fern); text-decoration: none;
  border: 1px solid var(--mist); padding: 4px 12px; border-radius: 3px;
  background: var(--white); transition: background .2s, color .2s;
}
.section-nav a:hover { background: var(--mist); color: var(--forest); }

h2 {
  font-family: 'Libre Baskerville', serif; font-size: 1.45rem; font-weight: 700;
  color: var(--forest); margin-top: 56px; margin-bottom: 6px;
  padding-bottom: 8px; border-bottom: 1px solid var(--parchment);
}
h3 {
  font-family: 'Libre Baskerville', serif; font-size: 1.1rem; font-weight: 700;
  color: var(--moss); margin-top: 28px; margin-bottom: 4px;
}
h4 {
  font-family: 'Lato', sans-serif; font-size: 1rem; font-weight: 700;
  color: var(--charcoal); margin-top: 20px; margin-bottom: 4px;
}

p { margin-bottom: 1em; }
strong { color: var(--forest); }
ol, ul { padding-left: 1.5em; margin-bottom: 1em; }
li { margin-bottom: .3em; }
blockquote {
  border-left: 4px solid var(--sage); margin: 1.2em 0; padding: .5em 1em;
  background: var(--white); border-radius: 0 5px 5px 0;
  font-size: .94em; color: var(--charcoal);
}

code {
  font-family: 'Source Code Pro', monospace; font-size: .84em;
  background: var(--mist); color: var(--moss); padding: 1px 5px; border-radius: 3px;
}
pre {
  background: var(--ink); color: #e8e6df; border-radius: 5px;
  padding: 14px 18px; font-size: .83rem; line-height: 1.55;
  overflow-x: auto; border-left: 4px solid var(--amber); margin: 1.2em 0;
}
pre code { background: transparent; color: inherit; padding: 0; font-size: 1em; }

.callout {
  border-left: 5px solid; padding: .65em 1.1em;
  margin: 1.4em 0; border-radius: 0 5px 5px 0; font-size: .92em; line-height: 1.6;
}
.callout > strong:first-child { display: block; margin-bottom: .3em; }
.callout-concept { border-color: var(--fern);  background: #f0f7f2; }
.callout-tip     { border-color: #2e7d32;      background: #f1f8f1; }
.callout-warn    { border-color: #e65100;      background: #fff8f4; }
.callout-formula { border-color: var(--amber); background: #fdf8f0;
                   font-family: 'Source Code Pro', monospace; font-size: .88em; }

.key-box {
  background: var(--forest); color: #fff; border-radius: 6px;
  padding: 20px 24px; margin: 2em 0; font-size: .9rem;
}
.key-box h4 {
  font-family: 'Libre Baskerville', serif; color: var(--sage);
  font-size: 1rem; margin-bottom: 10px;
}
.key-box ul { padding-left: 1.3em; }
.key-box li { margin-bottom: 5px; color: rgba(255,255,255,.85); }
.key-box strong { color: var(--amber-lt); }
.key-box code { background: rgba(255,255,255,.12); color: var(--amber-lt); }

.goals-box {
  background: var(--white); border: 1px solid var(--mist);
  border-left: 5px solid var(--fern); border-radius: 0 6px 6px 0;
  padding: 18px 22px 18px 20px; margin: 0 0 44px;
  box-shadow: var(--card-shadow);
}
.goals-box .goals-label {
  font-family: 'Source Code Pro', monospace;
  font-size: .68rem; font-weight: 700; letter-spacing: .18em; text-transform: uppercase;
  color: var(--fern); margin-bottom: 10px;
}
.goals-box ol { padding-left: 1.4em; margin: 0; }
.goals-box li { font-size: .92rem; color: var(--charcoal); margin-bottom: 5px; }

.exercises {
  border: 1px solid var(--parchment); border-radius: 6px;
  margin: 2.5em 0; overflow: hidden;
}
.exercises-header {
  background: var(--amber); color: #fff; padding: 10px 18px;
  font-family: 'Source Code Pro', monospace; font-size: .75rem;
  font-weight: 700; letter-spacing: .12em; text-transform: uppercase;
}
.exercise { padding: 16px 20px; border-bottom: 1px solid var(--parchment); }
.exercise:last-child { border-bottom: none; }
.exercise-num {
  font-family: 'Source Code Pro', monospace; font-size: .7rem; font-weight: 700;
  letter-spacing: .12em; text-transform: uppercase; color: var(--amber);
  margin-bottom: 5px;
}
.exercise p { font-size: .93rem; color: var(--charcoal); margin-bottom: 0; }
details.solution { margin-top: 10px; }
details.solution summary {
  font-family: 'Source Code Pro', monospace; font-size: .72rem;
  font-weight: 600; letter-spacing: .08em; text-transform: uppercase;
  color: var(--fern); cursor: pointer; padding: 5px 0; list-style: none;
}
details.solution summary::before { content: '▶  '; font-size: .65rem; }
details.solution[open] summary::before { content: '▼  '; }
details.solution summary::-webkit-details-marker { display: none; }
details.solution .solution-body {
  background: var(--cream); border-left: 3px solid var(--fern);
  border-radius: 0 4px 4px 0; padding: 12px 16px; margin-top: 8px;
  font-size: .88rem; color: var(--charcoal);
}

table { font-size: .9rem; border-collapse: collapse; width: 100%; margin: 1.2em 0; }
thead th {
  background: var(--forest); color: #fff;
  font-family: 'Source Code Pro', monospace; font-size: .76rem;
  letter-spacing: .06em; text-transform: uppercase; padding: 8px 12px; text-align: left;
}
tbody tr:nth-child(even) { background: var(--mist); }
tbody td { padding: 7px 12px; border-bottom: 1px solid var(--parchment); }

.btn-pdf {
  display: inline-flex; align-items: center; gap: 6px;
  font-family: 'Source Code Pro', monospace; font-size: .72rem;
  font-weight: 600; letter-spacing: .08em; text-transform: uppercase;
  background: var(--amber); color: #fff; border: none;
  padding: 6px 16px; border-radius: 4px; cursor: pointer;
  margin-top: 16px; transition: background .2s;
}
.btn-pdf:hover { background: var(--amber-lt); }

.page-footer {
  background: var(--ink); color: rgba(255,255,255,.42); text-align: center;
  padding: 24px 4vw; margin: 60px -4vw 0;
  font-size: .76rem; font-family: 'Source Code Pro', monospace; letter-spacing: .04em;
}
.page-footer a { color: var(--sage); text-decoration: none; }
.page-footer .divider { margin: 0 8px; opacity: .3; }

@media print {
  .section-nav, .back-link, .btn-pdf, .page-footer { display: none !important; }
  .page-header { margin: 0 !important; }
  body { padding: 0 !important; font-size: 13px !important; }
  pre { border-left: 2px solid #ccc !important;
        background: #f5f5f5 !important; color: #000 !important; }
  .callout { border-left: 3px solid #999 !important; background: #fafafa !important; }
  .key-box { background: #1e3a2f !important;
             -webkit-print-color-adjust: exact; print-color-adjust: exact; }
}
```

<div class="page-header">
<div class="module-tag">EA030 &mdash; Field &amp; Lab Activities &mdash; A12</div>
<h1>Shifting Baselines &amp; <span>Changepoint Analysis</span></h1>
<p class="sub"><strong>Prof. Marc Los Huertos</strong> &nbsp;&bull;&nbsp; Pomona College</p>
<a class="back-link" href="../index.html">&#8592; Back to RTricks</a>
<button class="btn-pdf" onclick="window.print()">&#8659; Save as PDF</button>
</div>

<div class="section-nav">
<a href="#background">Background</a>
<a href="#setup">Setup</a>
<a href="#offline">Offline Detection</a>
<a href="#variance">Variance Shifts</a>
<a href="#sequential">Sequential Detection</a>
<a href="#realdata">Real Data</a>
<a href="#exercises">Exercises</a>
</div>

<div class="goals-box">
<div class="goals-label">By the end of this module you will be able to</div>
<ol>
<li>Explain the shifting baseline problem and why it matters for long-term environmental monitoring.</li>
<li>Detect mean shifts in historical data using AMOC and PELT via the <code>changepoint</code> package.</li>
<li>Detect changes in variance using <code>cpt.var()</code> and interpret their ecological significance.</li>
<li>Apply sequential CUSUM monitoring and the <code>strucchange</code> package to test for structural breaks.</li>
<li>Choose the right penalty (AIC, BIC, MBIC) for a detection task and justify that choice.</li>
</ol>
</div>

> **What this module covers.** Environmental systems rarely stay put. Stream
> temperatures creep upward over decades, salmon populations collapse to a new
> lower equilibrium, droughts push groundwater down and keep it there. Standard
> statistics assume a stable reference state. This module teaches you to detect
> the moments when that reference permanently shifts, using both offline
> retrospective analysis and sequential monitoring tools.

---

## 1. Background {#background}

### 1.1 What is a baseline?

A **baseline** is the expected background state of a system against which
current conditions are compared. Baselines are not fixed facts of nature —
they are constructed reference periods, and they shift.

<div class="callout callout-concept">
<strong>📌 Shifting baseline syndrome</strong>
Fisheries biologist Daniel Pauly coined this term in 1995. Each generation of
scientists calibrates "normal" to the ecosystem state they first encountered.
Scientists starting in the 1950s defined healthy cod stocks by 1950s abundance;
those starting in the 1980s accepted the already-depleted state as normal.
The cumulative loss becomes invisible to any single generation.
</div>

### 1.2 What is a changepoint?

A **changepoint** is a time index $\tau$ at which a statistical property
— mean, variance, or both — undergoes an abrupt shift:

$$X_t \sim \begin{cases}
\mathcal{N}(\mu_1,\, \sigma_1^2) & t \leq \tau \\
\mathcal{N}(\mu_2,\, \sigma_2^2) & t > \tau
\end{cases}$$

This is distinct from a gradual trend. Mann-Kendall detects trends;
changepoint analysis detects **discrete jumps**.

<div class="callout callout-warn">
<strong>⚠️ Changepoints vs. trends</strong>
A gradual upward trend and a sudden step change can look similar in a raw
time plot. Choose your tool based on which kind of shift is ecologically
plausible for your system.
</div>

---

## 2. Setup {#setup}

```{r packages}
library(changepoint)   # AMOC, PELT (cpt.mean, cpt.var, cpt.meanvar)
library(strucchange)   # CUSUM, breakpoints with confidence intervals
library(ggplot2)
library(dplyr)
library(knitr)
```

```{r palette, include=FALSE}
col_forest <- "#1e3a2f"
col_fern   <- "#4a7c5e"
col_amber  <- "#c8832a"
col_dust   <- "#7a7a72"

theme_set(
  theme_minimal(base_size = 11) +
  theme(
    plot.title    = element_text(family = "serif", face = "bold",
                                 colour = col_forest, size = 12),
    plot.subtitle = element_text(colour = col_dust, size = 9.5),
    plot.caption  = element_text(colour = col_dust, size = 8),
    axis.title    = element_text(colour = "#3a3a38", size = 9.5),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "#ede8df", linewidth = 0.4)
  )
)
```

### 2.1 Simulated stream temperature data

We use a synthetic dataset mimicking Sierra Nevada foothill stream
temperatures during and after the 2012--2017 California drought:

- **Phase 1** (months 1--120): Cool pre-drought baseline, mean 14 deg C, SD 0.8
- **Phase 2** (months 121--204): Drought, mean rises to 18.2 deg C, SD 0.9
- **Phase 3** (months 205--288): Partial recovery to 16 deg C, but SD = 2.1 (flashier hydrology)

```{r simulate}
set.seed(2024)
n1 <- 120; n2 <- 84; n3 <- 84
N  <- n1 + n2 + n3

temp  <- c(
  rnorm(n1, mean = 14.0, sd = 0.8),
  rnorm(n2, mean = 18.2, sd = 0.9),
  rnorm(n3, mean = 16.1, sd = 2.1)
)
dates     <- seq(as.Date("2000-01-01"), by = "month", length.out = N)
true_cpts <- c(n1, n1 + n2)
stream_df <- data.frame(date = dates, temp = temp)
```

```{r raw-plot, fig.cap="Simulated monthly stream temperature. True changepoints are dashed amber lines; the shaded band marks the simulated drought window.", fig.alt="Line plot of 288 monthly temperatures showing a step up at month 120 and wider oscillations from month 204."}
ggplot(stream_df, aes(x = date, y = temp)) +
  annotate("rect",
    xmin = dates[n1 + 1], xmax = dates[n1 + n2],
    ymin = -Inf, ymax = Inf, fill = "#fff8f4", alpha = 0.9) +
  geom_line(colour = col_fern, linewidth = 0.45) +
  geom_vline(xintercept = dates[true_cpts],
             linetype = "dashed", colour = col_amber, linewidth = 0.8) +
  labs(title    = "Simulated Monthly Stream Temperature",
       subtitle = "Dashed lines = true changepoints  |  Shaded = drought window",
       x = NULL, y = "Temperature (deg C)")
```

---

## 3. Offline Changepoint Detection {#offline}

Offline methods work on the complete historical record. They find the
set of changepoints that minimises total within-segment cost plus a
penalty for model complexity.

### 3.1 AMOC — At Most One Change

AMOC evaluates every possible single split and picks the best one.
Useful when you expect exactly one regime shift.

```{r amoc, fig.cap="AMOC detects the single most significant mean shift — the large drought-onset step. It cannot detect a second changepoint by design.", fig.alt="Time series with one changepoint line and two horizontal mean lines."}
m_amoc <- cpt.mean(temp, method = "AMOC")

cat("AMOC changepoint at index:", cpts(m_amoc), "\n")
cat("  =>", format(dates[cpts(m_amoc)], "%b %Y"), "\n")
cat("Segment means:", round(param.est(m_amoc)$mean, 2), "deg C\n")

plot(m_amoc,
     main = "AMOC: Single Mean Changepoint",
     ylab = "Temperature (deg C)", xlab = "Month index",
     col  = col_fern, cpt.col = col_amber,
     cpt.width = 3, mean.col = col_forest)
```

### 3.2 PELT — Multiple Changepoints

PELT (Pruned Exact Linear Time) finds the globally optimal set of
changepoints efficiently. The key user decision is the **penalty**:

```{r penalty-table, echo=FALSE}
knitr::kable(
  data.frame(
    Penalty  = c("AIC", "BIC", "MBIC"),
    Behavior = c("Most permissive — more changepoints",
                 "Balanced — standard default",
                 "Conservative — fewer false positives"),
    Use      = c("Exploratory, dense sensor data",
                 "Most environmental records",
                 "High cost of false alarms")
  ),
  caption = "Penalty options for PELT mean changepoint detection."
)
```

```{r pelt-bic, fig.cap="PELT with BIC penalty recovers both true changepoints in the stream temperature record.", fig.alt="Time series with two changepoint lines at months 120 and 204 and three horizontal mean lines."}
m_pelt <- cpt.mean(temp, method = "PELT", penalty = "BIC")

cat("PELT (BIC) changepoints:", cpts(m_pelt), "\n")
cat("  =>", format(dates[cpts(m_pelt)], "%b %Y"), "\n")
cat("Segment means:", round(param.est(m_pelt)$mean, 2), "deg C\n")
cat("True changepoints at months:", true_cpts, "\n")

plot(m_pelt,
     main = "PELT (BIC): Multiple Mean Changepoints",
     ylab = "Temperature (deg C)", xlab = "Month index",
     col  = col_fern, cpt.col = col_amber,
     cpt.width = 3, mean.col = col_forest)
```

```{r pelt-compare, fig.cap="AIC vs BIC penalty comparison. AIC is more permissive and may over-segment the volatile post-drought phase.", fig.alt="Two stacked plots comparing PELT AIC and BIC results on the same data."}
m_aic <- cpt.mean(temp, method = "PELT", penalty = "AIC")

par(mfrow = c(2, 1), mar = c(3, 4, 2.5, 1))
plot(m_pelt, main = paste("BIC:", length(cpts(m_pelt)), "changepoints"),
     ylab = "deg C", col = col_fern, cpt.col = col_amber,
     cpt.width = 2, mean.col = col_forest, xlab = "")
plot(m_aic,  main = paste("AIC:", length(cpts(m_aic)),  "changepoints"),
     ylab = "deg C", col = col_fern, cpt.col = col_amber,
     cpt.width = 2, mean.col = col_forest, xlab = "Month index")
par(mfrow = c(1, 1))
```

<div class="callout callout-tip">
<strong>💡 Which penalty to use?</strong>
Start with BIC. Use AIC for dense exploratory work. Use MBIC when false
alarms carry regulatory costs. Always report and justify your choice.
</div>

---

## 4. Variance Shifts {#variance}

The post-drought phase has the same rough mean as recovery, but much
higher variance — the stream is flashier. This matters ecologically:

<div class="callout callout-formula">
<strong>Why variance shifts matter</strong>
Mean = 16 deg C, SD = 0.9: temperatures exceed 18 deg C less than 2% of days.
Mean = 16 deg C, SD = 2.1: temperatures exceed 18 deg C roughly 17% of days —
above the thermal tolerance limit of many cold-water fish.
</div>

```{r cpt-var, fig.cap="Variance changepoint detection with PELT. The detected shift near month 204 marks the onset of higher SD in the post-drought phase.", fig.alt="Time series with a variance changepoint showing wider oscillations in the right-hand segment."}
m_var <- cpt.var(temp, method = "PELT", penalty = "BIC")

cat("Variance changepoint(s):", cpts(m_var), "\n")
cat("  =>", format(dates[cpts(m_var)], "%b %Y"), "\n")
cat("Segment SDs:", round(sqrt(param.est(m_var)$variance), 2), "deg C\n")

plot(m_var,
     main = "Variance Changepoint Detection (PELT)",
     ylab = "Temperature (deg C)", xlab = "Month index",
     col  = col_fern, cpt.col = col_amber, cpt.width = 3)
```

```{r cpt-meanvar, fig.cap="Joint mean-and-variance changepoint detection. A shift in either the mean or the variance triggers a new changepoint.", fig.alt="Time series with three changepoints separating segments with different means and spreads."}
m_mv <- cpt.meanvar(temp, method = "PELT", penalty = "BIC")

cat("Mean+variance changepoints:", cpts(m_mv), "\n")
cat("Segment means:", round(param.est(m_mv)$mean, 2), "\n")
cat("Segment SDs:  ", round(sqrt(param.est(m_mv)$variance), 2), "\n")

plot(m_mv,
     main = "Mean and Variance Changepoints (PELT)",
     ylab = "Temperature (deg C)", xlab = "Month index",
     col  = col_fern, cpt.col = col_amber,
     cpt.width = 3, mean.col = col_forest)
```

---

## 5. Sequential Detection with strucchange {#sequential}

Offline methods need all the data at once. The `strucchange` package
provides two complementary tools for detecting and locating breaks:
a sequential fluctuation test and optimal breakpoints with timing CIs.

### 5.1 CUSUM fluctuation process

The OLS-CUSUM process accumulates deviations from a fitted intercept
model over time. Breaching the critical boundary signals structural change.

```{r cusum, fig.cap="OLS-CUSUM fluctuation process for the stream temperature series. Both boundary crossings correspond to the true changepoints.", fig.alt="CUSUM plot with a fluctuation line that crosses the upper significance boundary twice."}
efp_fit <- efp(temp ~ 1, type = "OLS-CUSUM")
plot(efp_fit,
     main = "OLS-CUSUM: Sequential Test for Mean Shift",
     ylab = "Empirical fluctuation process",
     col  = col_forest, lwd  = 1.2)
```

```{r sctest}
sc <- sctest(efp_fit)
cat(sprintf("OLS-CUSUM:  statistic = %.3f,  p-value = %.5f\n",
            sc$statistic, sc$p.value))
```

### 5.2 Optimal breakpoints with confidence intervals

`breakpoints()` finds the globally optimal breakpoints and — unlike
`changepoint` — reports **95% confidence intervals on timing**.
Knowing whether a shift occurred in 2009 vs 2013 often matters for
attributing cause to a specific event or policy.

```{r bps-fit}
# Step 1: fit the full breakpointsfull object
bp_full <- breakpoints(temp ~ 1)

# Step 2: select optimal number of breakpoints (minimises BIC)
bp_opt  <- breakpoints(bp_full)

# Step 3: confidence intervals come from the breakpointsfull object
ci      <- confint(bp_full)

cat("Optimal breakpoint indices:", bp_opt$breakpoints, "\n")
cat("  =>", format(dates[bp_opt$breakpoints], "%b %Y"), "\n")
cat("True changepoints at months:", true_cpts, "\n")
cat("\n95% CI on breakpoint timing (month indices):\n")
print(ci$confint)
```

```{r bps-plot, fig.cap="Stream temperature with breakpoints and 95% confidence intervals. Amber bands mark the timing uncertainty; horizontal lines show segment means.", fig.alt="Line plot with two amber CI bands and two horizontal segment mean lines."}
# Segment means via breakfactor
bf        <- breakfactor(bp_opt)
seg_means <- coef(lm(temp ~ bf - 1))

ci_mat  <- ci$confint
bp_idx  <- c(0, bp_opt$breakpoints, N)

mean_df <- data.frame(
  xmin = dates[bp_idx[-length(bp_idx)] + 1],
  xmax = dates[bp_idx[-1]],
  yval = seg_means
)
ci_df <- data.frame(
  xlo = dates[pmax(1, ci_mat[, 1])],
  xhi = dates[pmin(N, ci_mat[, 3])]
)

ggplot(stream_df, aes(x = date, y = temp)) +
  geom_rect(data = ci_df,
            aes(xmin = xlo, xmax = xhi, ymin = -Inf, ymax = Inf),
            inherit.aes = FALSE, fill = col_amber, alpha = 0.18) +
  geom_line(colour = col_fern, linewidth = 0.45, alpha = 0.8) +
  geom_segment(data = mean_df,
               aes(x = xmin, xend = xmax, y = yval, yend = yval),
               colour = col_forest, linewidth = 1.2) +
  geom_vline(xintercept = dates[ci_mat[, 2]],
             linetype = "dashed", colour = col_amber, linewidth = 0.8) +
  labs(title    = "Optimal Breakpoints with 95% Confidence Intervals",
       subtitle = "Amber bands = timing uncertainty on each changepoint",
       x = NULL, y = "Temperature (deg C)")
```

### 5.3 The shifting baseline problem in sequential monitoring

A CUSUM chart uses a fixed reference mean. Once the baseline shifts,
every observation in the new regime accumulates as a "deviation" from
the old reference — the statistic never resets and becomes meaningless.

```{r cusum-rolling, fig.cap="Rolling CUSUM from the pre-drought reference mean. The statistic keeps rising indefinitely after the first shift because the reference is no longer valid — this is the shifting baseline problem in action.", fig.alt="Line plot of cumulative standardised sum showing a steep upward rise after month 120 that never returns to baseline."}
ref_mean <- mean(temp[1:n1])
ref_sd   <- sd(temp[1:n1])
cusum_df <- data.frame(
  date  = dates,
  cusum = cumsum((temp - ref_mean) / ref_sd)
)

ggplot(cusum_df, aes(x = date, y = cusum)) +
  geom_hline(yintercept = c(-4, 4), linetype = "dashed",
             colour = "#c0392b", linewidth = 0.8) +
  geom_hline(yintercept = 0, colour = col_dust, linewidth = 0.3) +
  geom_line(colour = col_forest, linewidth = 0.5) +
  geom_vline(xintercept = dates[true_cpts],
             linetype = "dotted", colour = col_amber) +
  annotate("text", x = min(dates) + 30, y = 4.8,
           label = "Critical value (+/-4)", colour = "#c0392b", size = 2.8) +
  labs(title    = "Rolling CUSUM from Pre-Drought Reference",
       subtitle = "Statistic never resets after the first shift — the old reference is obsolete",
       x = NULL, y = "Cumulative standardised sum")
```

<div class="callout callout-warn">
<strong>⚠️ Operational fix</strong>
Reset the reference mean after each confirmed changepoint. Otherwise
your monitoring system perpetually alarms on the new normal.
</div>

---

## 6. Real Environmental Data: Nile River Flow {#realdata}

R's built-in `Nile` dataset records annual discharge at Aswan 1871--1970.
A documented structural break near 1899 reflects construction of the
Aswan Low Dam and upstream land-use change.

```{r nile-pelt, fig.cap="PELT detects one dominant mean shift in the Nile record, consistent with the Aswan Low Dam (completed 1902).", fig.alt="Nile discharge time series with a single changepoint near 1899 and two horizontal mean lines."}
data(Nile)
nile_df <- data.frame(year = as.numeric(time(Nile)),
                      discharge = as.numeric(Nile))

nile_pelt <- cpt.mean(as.numeric(Nile), method = "PELT", penalty = "BIC")
cp_year   <- 1871 + cpts(nile_pelt) - 1
means_n   <- param.est(nile_pelt)$mean

cat("PELT changepoint year:", cp_year, "\n")
cat("Segment means:", round(means_n), "x 10^8 m^3\n")
cat("Drop:", round(abs(diff(means_n))), "x 10^8 m^3",
    sprintf("(%.0f%%)\n", 100 * abs(diff(means_n)) / means_n[1]))

plot(nile_pelt,
     main = "Nile Discharge: PELT Mean Changepoint",
     ylab = "Discharge (10^8 m^3)", xlab = "Year index",
     col  = col_fern, cpt.col = col_amber,
     cpt.width = 3, mean.col = col_forest)
```

```{r nile-ci}
# strucchange: breakpoints with 95% CI on timing
# confint() must be called on the breakpointsfull object
nile_bp_full <- breakpoints(as.numeric(Nile) ~ 1)
nile_bp_opt  <- breakpoints(nile_bp_full)
nile_ci      <- confint(nile_bp_full)

bp_yr <- 1871 + nile_bp_opt$breakpoints - 1
ci_lo <- 1871 + nile_ci$confint[1, 1] - 1
ci_hi <- 1871 + nile_ci$confint[1, 3] - 1

cat(sprintf("strucchange breakpoint: %d  (95%% CI: %d - %d)\n",
            bp_yr, ci_lo, ci_hi))
cat("CI width:", ci_hi - ci_lo, "years\n")
```

```{r nile-final-plot, fig.cap="Nile discharge with segment means and 95% CI band on the changepoint year. The narrow CI supports attributing the break to the Aswan Low Dam construction.", fig.alt="Nile discharge time series with amber confidence band around 1899, two horizontal mean lines, and a dashed vertical changepoint line."}
nile_bf  <- breakfactor(nile_bp_opt)
nile_mu  <- coef(lm(as.numeric(Nile) ~ nile_bf - 1))
nile_idx <- c(0, nile_bp_opt$breakpoints, length(Nile))

nile_mean_df <- data.frame(
  xmin = 1871 + nile_idx[-length(nile_idx)],
  xmax = 1871 + nile_idx[-1] - 1,
  yval = nile_mu
)

ggplot(nile_df, aes(x = year, y = discharge)) +
  annotate("rect", xmin = ci_lo, xmax = ci_hi,
           ymin = -Inf, ymax = Inf, fill = col_amber, alpha = 0.2) +
  geom_line(colour = col_fern, linewidth = 0.5, alpha = 0.8) +
  geom_segment(data = nile_mean_df,
               aes(x = xmin, xend = xmax, y = yval, yend = yval),
               colour = col_forest, linewidth = 1.3) +
  geom_vline(xintercept = bp_yr, colour = col_amber,
             linetype = "dashed", linewidth = 1) +
  annotate("text", x = bp_yr + 1, y = 1370,
           label = sprintf("Break: %d\n(95%% CI: %d-%d)", bp_yr, ci_lo, ci_hi),
           hjust = 0, size = 2.8, colour = col_amber) +
  labs(title    = "Nile Discharge: PELT Means and strucchange 95% CI",
       x = "Year", y = "Discharge (10^8 m^3)")
```

---

<div class="key-box">
<h4>🔑 Key Takeaways</h4>
<ul>
<li><strong>Shifting baselines</strong> accumulate silently — statistical detection makes them explicit and quantifiable.</li>
<li><strong>AMOC</strong> finds one break; <strong>PELT</strong> finds the globally optimal set in O(n) time.</li>
<li><strong>Penalty choice</strong> (AIC / BIC / MBIC) controls sensitivity vs false-alarm rate — always report it.</li>
<li><strong>Variance shifts</strong> are ecologically important even when the mean is stable — use <code>cpt.var()</code>.</li>
<li><strong>strucchange</strong> adds 95% CIs on changepoint timing; call <code>confint()</code> on the <code>breakpointsfull</code> object, not on the result of <code>breakpoints(breakpointsfull)</code>.</li>
<li>Sequential CUSUM must be <strong>re-referenced</strong> after each confirmed shift or it becomes meaningless.</li>
</ul>
</div>

---

<div class="exercises">
<div class="exercises-header">✏️ Practice Exercises</div>

<div class="exercise">
<div class="exercise-num">Exercise 1 — Ozone changepoints</div>
<p>Load <code>airquality</code> and remove rows with missing <code>Ozone</code>.
Apply PELT with BIC penalty. (a) How many changepoints are detected?
(b) Convert indices to calendar dates. (c) Do the breaks correspond to
a meteorological explanation?</p>
<details class="solution">
<summary>Show solution</summary>
<div class="solution-body">
<pre><code>data(airquality)
ozone <- airquality[!is.na(airquality$Ozone), ]
m <- cpt.mean(ozone$Ozone, method = "PELT", penalty = "BIC")
cpts(m)
ozone$date <- as.Date(paste(1973, ozone$Month, ozone$Day, sep = "-"))
ozone$date[cpts(m)]
plot(m, ylab = "Ozone (ppb)", main = "Ozone changepoints 1973")</code></pre>
<p>A changepoint typically appears around late June, separating the cool
low-ozone spring from hot high-ozone midsummer. This marks the onset
of photochemical production driven by higher temperatures and longer days.</p>
</div>
</details>
</div>

<div class="exercise">
<div class="exercise-num">Exercise 2 — Variance shift in PM2.5</div>
<p>Simulate two years of daily PM2.5: Year 1 mean = 10, SD = 2;
Year 2 mean = 10, SD = 8 (wildfire smoke). Use <code>set.seed(42)</code>.
(a) Does <code>cpt.mean()</code> find a changepoint? (b) Does <code>cpt.var()</code>?
(c) What does this imply for monitoring standards based on annual means?</p>
<details class="solution">
<summary>Show solution</summary>
<div class="solution-body">
<pre><code>set.seed(42)
pm25 <- c(rnorm(365, 10, 2), rnorm(365, 10, 8))
m_mean <- cpt.mean(pm25, method = "PELT", penalty = "BIC")
m_var  <- cpt.var(pm25,  method = "PELT", penalty = "BIC")
cat("Mean cpts:", cpts(m_mean), "\n")
cat("Variance cpts:", cpts(m_var), "\n")</code></pre>
<p><code>cpt.mean()</code> finds nothing. <code>cpt.var()</code> detects the
shift near day 365. Even with the same annual mean, the high-variance year
has far more days above the 35 ug/m3 NAAQS threshold — health risk is driven
by peak events, not averages alone.</p>
</div>
</details>
</div>

<div class="exercise">
<div class="exercise-num">Exercise 3 — Penalty sensitivity on Nile data</div>
<p>Run PELT on the Nile dataset with "AIC", "BIC", and "MBIC" penalties.
(a) How many changepoints does each detect? (b) Is the primary break
year consistent? (c) Which penalty would you use in a regulatory report?</p>
<details class="solution">
<summary>Show solution</summary>
<div class="solution-body">
<pre><code>data(Nile)
for (pen in c("AIC", "BIC", "MBIC")) {
  m   <- cpt.mean(as.numeric(Nile), method = "PELT", penalty = pen)
  yrs <- 1871 + cpts(m) - 1
  cat(pen, ":", length(cpts(m)), "changepoint(s) at", yrs, "\n")
}</code></pre>
<p>AIC typically finds 2-3 breaks; BIC and MBIC agree on one near 1899.
For a regulatory report, BIC or MBIC is appropriate — false alarms could
trigger costly policy responses. The 1899 break is robust across all
penalties, lending credibility to the attribution.</p>
</div>
</details>
</div>

<div class="exercise">
<div class="exercise-num">Exercise 4 — Confidence intervals on timing</div>
<p>Using the Nile dataset and <code>strucchange</code>, obtain a 95% CI on the
breakpoint year. (a) How wide is the CI? (b) The Aswan Low Dam was
completed in 1902. Does your CI include this year?
(c) Why does CI width matter for causal attribution?</p>
<details class="solution">
<summary>Show solution</summary>
<div class="solution-body">
<pre><code>library(strucchange)
data(Nile)
bp  <- breakpoints(as.numeric(Nile) ~ 1)   # breakpointsfull
bpo <- breakpoints(bp)                       # optimal selection
ci  <- confint(bp)                           # CI from breakpointsfull
cat("Year:", 1871 + bpo$breakpoints - 1, "\n")
cat("95% CI:", 1871 + ci$confint[c(1, 3)] - 1, "\n")
cat("Width:", diff(ci$confint[c(1, 3)]), "years\n")</code></pre>
<p>The CI is typically 4-6 years and includes 1902. A narrow CI allows you to
say "the shift occurred during the dam construction window." A CI spanning
decades would be too wide to implicate any specific cause.</p>
</div>
</details>
</div>

</div>

```{r session-info, echo=FALSE}
# sessionInfo()
```

<footer class="page-footer">
EA030 Field &amp; Lab Activities &mdash; A12
<span class="divider">|</span> Prof. Marc Los Huertos
<span class="divider">|</span> Pomona College
<span class="divider">|</span> Last updated: `r format(Sys.Date(), "%B %Y")`
<span class="divider">|</span> <a href="https://github.com/marclos/RTricks">GitHub</a>
</footer>
