---
title: "Activity Title Here"
author: "Marc Los Huertos"
date: "EA030 · Compiled `r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    self_contained: true
    theme: null
    highlight: null
    toc: false
knit: (function(inputFile, encoding) {
    rmarkdown::render(
      inputFile,
      encoding = encoding,
      output_file = file.path(dirname(inputFile),
                              "../docs/Activities/AXX_ActivityName.html")
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  eval    = TRUE,
  fig.width  = 7,
  fig.height = 4
)
```

```{css, echo=FALSE}
@import url('https://fonts.googleapis.com/css2?family=Libre+Baskerville:ital,wght@0,400;0,700;1,400&family=Source+Code+Pro:wght@400;600&family=Lato:wght@300;400;700&display=swap');

*, *::before, *::after { box-sizing: border-box; margin: 0; padding: 0; }

:root {
  --forest:    #1e3a2f;
  --moss:      #2d5a3d;
  --fern:      #4a7c5e;
  --sage:      #8aad96;
  --mist:      #d4e6da;
  --cream:     #f7f3ed;
  --parchment: #ede8df;
  --ink:       #1a1a18;
  --charcoal:  #3a3a38;
  --dust:      #7a7a72;
  --amber:     #c8832a;
  --amber-lt:  #e8a84a;
  --white:     #ffffff;
  --card-shadow: 0 2px 10px rgba(30,58,47,.09);
}

html { scroll-behavior: smooth; }
body {
  font-family: 'Lato', sans-serif;
  background: var(--cream);
  color: var(--ink);
  font-size: 17px;
  line-height: 1.75;
  padding-left:  clamp(14px, 4vw, 48px);
  padding-right: clamp(14px, 4vw, 48px);
  padding-bottom: 80px;
}
.main-container { max-width: 920px !important; margin: 0 auto !important;
  padding: 0 clamp(10px, 2vw, 32px) 80px !important; }
.page-header {
  margin-left:  calc(-1 * clamp(14px, 4vw, 48px));
  margin-right: calc(-1 * clamp(14px, 4vw, 48px));
}
h1.title, h4.author, h4.date { display: none; }

/* ── Page header ─────────────────────────────────────────────────── */
.page-header {
  background-image:
    repeating-linear-gradient(120deg, transparent, transparent 60px,
      rgba(255,255,255,.018) 60px, rgba(255,255,255,.018) 61px),
    linear-gradient(160deg, #1e3a2f 0%, #2d5a3d 60%, #1a3d30 100%);
  color: #fff; padding: 44px 4vw 36px; margin: 0 -4vw 48px;
  border-bottom: 4px solid var(--amber); position: relative; overflow: hidden;
}
.page-header::after {
  content: ''; position: absolute; right: -40px; top: -40px;
  width: 260px; height: 260px; border-radius: 50%;
  background: radial-gradient(circle, rgba(138,173,150,.12) 0%, transparent 70%);
  pointer-events: none;
}
.page-header .module-tag {
  display: inline-block; font-family: 'Source Code Pro', monospace;
  font-size: .68rem; font-weight: 600; letter-spacing: .18em; text-transform: uppercase;
  color: var(--amber-lt); border: 1px solid rgba(200,131,42,.45);
  padding: 3px 10px; border-radius: 2px; margin-bottom: 12px;
}
.page-header h1 {
  font-family: 'Libre Baskerville', serif; font-size: clamp(1.6rem, 3vw, 2.3rem);
  font-weight: 700; color: #fff; line-height: 1.2; margin-bottom: 8px;
}
.page-header h1 span { color: var(--sage); }
.page-header .sub { font-size: .9rem; color: rgba(255,255,255,.65); font-weight: 300; }
.page-header .sub strong { color: rgba(255,255,255,.9); font-weight: 700; }
.back-link {
  display: inline-block; margin-top: 16px;
  font-family: 'Source Code Pro', monospace; font-size: .7rem; letter-spacing: .08em;
  color: var(--amber-lt); text-decoration: none;
  border: 1px solid rgba(200,131,42,.4); padding: 4px 11px; border-radius: 3px;
  transition: background .2s;
}
.back-link:hover { background: rgba(200,131,42,.18); }

/* ── Section nav ─────────────────────────────────────────────────── */
.section-nav {
  display: flex; gap: 8px; flex-wrap: wrap; margin-bottom: 44px;
  padding-bottom: 20px; border-bottom: 1px solid var(--parchment);
}
.section-nav a {
  font-family: 'Source Code Pro', monospace; font-size: .7rem; font-weight: 600;
  letter-spacing: .08em; text-transform: uppercase;
  color: var(--fern); text-decoration: none;
  border: 1px solid var(--mist); padding: 4px 12px; border-radius: 3px;
  background: var(--white); transition: background .2s, color .2s;
}
.section-nav a:hover { background: var(--mist); color: var(--forest); }

/* ── Headings ────────────────────────────────────────────────────── */
h2 {
  font-family: 'Libre Baskerville', serif; font-size: 1.45rem; font-weight: 700;
  color: var(--forest); margin-top: 56px; margin-bottom: 6px;
  padding-bottom: 8px; border-bottom: 1px solid var(--parchment);
}
h3 {
  font-family: 'Libre Baskerville', serif; font-size: 1.1rem; font-weight: 700;
  color: var(--moss); margin-top: 28px; margin-bottom: 4px;
}
h4 {
  font-family: 'Lato', sans-serif; font-size: 1rem; font-weight: 700;
  color: var(--charcoal); margin-top: 20px; margin-bottom: 4px;
}

/* ── Body text ───────────────────────────────────────────────────── */
p { margin-bottom: 1em; }
strong { color: var(--forest); }
ol, ul { padding-left: 1.5em; margin-bottom: 1em; }
li { margin-bottom: .3em; }
blockquote {
  border-left: 4px solid var(--sage); margin: 1.2em 0; padding: .5em 1em;
  background: var(--white); border-radius: 0 5px 5px 0;
  font-size: .94em; color: var(--charcoal);
}

/* ── Code ────────────────────────────────────────────────────────── */
code {
  font-family: 'Source Code Pro', monospace; font-size: .84em;
  background: var(--mist); color: var(--moss); padding: 1px 5px; border-radius: 3px;
}
pre {
  background: var(--ink); color: #e8e6df; border-radius: 5px;
  padding: 14px 18px; font-size: .83rem; line-height: 1.55;
  overflow-x: auto; border-left: 4px solid var(--amber); margin: 1.2em 0;
}
pre code { background: transparent; color: inherit; padding: 0; font-size: 1em; }

/* ── Callout boxes ───────────────────────────────────────────────── */
.callout {
  border-left: 5px solid; padding: .65em 1.1em;
  margin: 1.4em 0; border-radius: 0 5px 5px 0; font-size: .92em; line-height: 1.6;
}
.callout > strong:first-child { display: block; margin-bottom: .3em; }
.callout-concept { border-color: var(--fern);  background: #f0f7f2; }
.callout-tip     { border-color: #2e7d32;      background: #f1f8f1; }
.callout-warn    { border-color: #e65100;      background: #fff8f4; }
.callout-formula { border-color: var(--amber); background: #fdf8f0;
                   font-family: 'Source Code Pro', monospace; font-size: .88em; }

/* ── Key concept box ─────────────────────────────────────────────── */
.key-box {
  background: var(--forest); color: #fff; border-radius: 6px;
  padding: 20px 24px; margin: 2em 0; font-size: .9rem;
}
.key-box h4 {
  font-family: 'Libre Baskerville', serif; color: var(--sage);
  font-size: 1rem; margin-bottom: 10px;
}
.key-box ul { padding-left: 1.3em; }
.key-box li { margin-bottom: 5px; color: rgba(255,255,255,.85); }
.key-box code { background: rgba(255,255,255,.12); color: var(--amber-lt); }

/* ── Goals box ───────────────────────────────────────────────────── */
.goals-box {
  background: var(--white); border: 1px solid var(--mist);
  border-left: 5px solid var(--fern); border-radius: 0 6px 6px 0;
  padding: 18px 22px 18px 20px; margin: 0 0 44px;
  box-shadow: var(--card-shadow);
}
.goals-box .goals-label {
  font-family: 'Source Code Pro', monospace;
  font-size: .68rem; font-weight: 700; letter-spacing: .18em; text-transform: uppercase;
  color: var(--fern); margin-bottom: 10px;
}
.goals-box ol { padding-left: 1.4em; margin: 0; }
.goals-box li { font-size: .92rem; color: var(--charcoal); margin-bottom: 5px; }

/* ── Practice exercises ──────────────────────────────────────────── */
.exercises {
  border: 1px solid var(--parchment); border-radius: 6px;
  margin: 2.5em 0; overflow: hidden;
}
.exercises-header {
  background: var(--amber); color: #fff; padding: 10px 18px;
  font-family: 'Source Code Pro', monospace; font-size: .75rem;
  font-weight: 700; letter-spacing: .12em; text-transform: uppercase;
}
.exercise { padding: 16px 20px; border-bottom: 1px solid var(--parchment); }
.exercise:last-child { border-bottom: none; }
.exercise-num {
  font-family: 'Source Code Pro', monospace; font-size: .7rem; font-weight: 700;
  letter-spacing: .12em; text-transform: uppercase; color: var(--amber);
  margin-bottom: 5px;
}
.exercise p { font-size: .93rem; color: var(--charcoal); margin-bottom: 0; }
details.solution { margin-top: 10px; }
details.solution summary {
  font-family: 'Source Code Pro', monospace; font-size: .72rem;
  font-weight: 600; letter-spacing: .08em; text-transform: uppercase;
  color: var(--fern); cursor: pointer; padding: 5px 0; list-style: none;
}
details.solution summary::before { content: '▶  '; font-size: .65rem; }
details.solution[open] summary::before { content: '▼  '; }
details.solution summary::-webkit-details-marker { display: none; }
details.solution .solution-body {
  background: var(--cream); border-left: 3px solid var(--fern);
  border-radius: 0 4px 4px 0; padding: 12px 16px; margin-top: 8px;
  font-size: .88rem; color: var(--charcoal);
}

/* ── Tables ──────────────────────────────────────────────────────── */
table { font-size: .9rem; border-collapse: collapse; width: 100%; margin: 1.2em 0; }
thead th {
  background: var(--forest); color: #fff;
  font-family: 'Source Code Pro', monospace; font-size: .76rem;
  letter-spacing: .06em; text-transform: uppercase; padding: 8px 12px; text-align: left;
}
tbody tr:nth-child(even) { background: var(--mist); }
tbody td { padding: 7px 12px; border-bottom: 1px solid var(--parchment); }

/* ── PDF / Print button ──────────────────────────────────────────── */
.btn-pdf {
  display: inline-flex; align-items: center; gap: 6px;
  font-family: 'Source Code Pro', monospace; font-size: .72rem;
  font-weight: 600; letter-spacing: .08em; text-transform: uppercase;
  background: var(--amber); color: #fff; border: none;
  padding: 6px 16px; border-radius: 4px; cursor: pointer;
  margin-top: 16px; transition: background .2s;
}
.btn-pdf:hover { background: var(--amber-lt); }

/* ── Footer ──────────────────────────────────────────────────────── */
.page-footer {
  background: var(--ink); color: rgba(255,255,255,.42); text-align: center;
  padding: 24px 4vw; margin: 60px -4vw 0;
  font-size: .76rem; font-family: 'Source Code Pro', monospace; letter-spacing: .04em;
}
.page-footer a { color: var(--sage); text-decoration: none; }
.page-footer .divider { margin: 0 8px; opacity: .3; }

/* ── Print ───────────────────────────────────────────────────────── */
@media print {
  .section-nav, .back-link, .btn-pdf, .page-footer { display: none !important; }
  .page-header { margin: 0 !important; }
  body { padding: 0 !important; font-size: 13px !important; }
  pre { border-left: 2px solid #ccc !important;
        background: #f5f5f5 !important; color: #000 !important; }
  .callout { border-left: 3px solid #999 !important; background: #fafafa !important; }
  .key-box { background: #1e3a2f !important;
             -webkit-print-color-adjust: exact; print-color-adjust: exact; }
}
```

<!-- ============================================================ -->
<!--  PAGE HEADER — edit module-tag, h1, and back-link           -->
<!-- ============================================================ -->

<div class="page-header">
<div class="module-tag">EA030 &mdash; Field &amp; Lab Activities &mdash; AXX</div>
<h1>Activity <span>Title Here</span></h1>
<p class="sub"><strong>Prof. Marc Los Huertos</strong> &nbsp;&bull;&nbsp; Pomona College</p>
<a class="back-link" href="../index.html">&#8592; Back to RTricks</a>
<button class="btn-pdf" onclick="window.print()">&#8659; Save as PDF</button>
</div>

<!-- ============================================================ -->
<!--  SECTION NAV PILLS — add/remove anchors to match your h2s  -->
<!-- ============================================================ -->

<div class="section-nav">
<a href="#background">Background</a>
<a href="#setup">Setup</a>
<a href="#section1">Section 1</a>
<a href="#section2">Section 2</a>
<a href="#exercises">Exercises</a>
</div>

<!-- ============================================================ -->
<!--  GOALS BOX                                                   -->
<!-- ============================================================ -->

<div class="goals-box">
<div class="goals-label">By the end of this module you will be able to</div>
<ol>
<li>Goal one — a concrete, measurable skill.</li>
<li>Goal two.</li>
<li>Goal three.</li>
<li>Goal four.</li>
</ol>
</div>

> **What this module covers.** One or two sentences summarising the
> scientific and analytical problem this activity addresses. Write this
> for a student who is deciding whether to read further.

---

## 1. Background {#background}

Replace this text with your background section. Explain the
environmental context, why the analysis matters, and what the student
should already know before starting.

<div class="callout callout-concept">
<strong>📌 Key concept</strong>
Use callout-concept (green) for important ideas, callout-tip for
practical advice, and callout-warn for common mistakes.
</div>

---

## 2. Setup {#setup}

### 2.1 Packages

```{r packages}
# List packages required for this activity.
# Uncomment install lines and run once in the Console if needed:
# install.packages("packagename", repos = "https://cloud.r-project.org")

library(ggplot2)
library(dplyr)
```

### 2.2 Data

```{r load-data}
# Load or simulate data here.
# Example using a built-in dataset:
data(airquality)
head(airquality)
```

---

## 3. Section One {#section1}

### 3.1 Subsection

Explain the concept, then show the code.

```{r example-chunk, fig.cap="Caption describing what this figure shows.", fig.alt="Alt text describing the figure for screen readers."}
# R code here
ggplot(airquality, aes(x = Temp, y = Ozone)) +
  geom_point(colour = "#4a7c5e", alpha = 0.7) +
  labs(title = "Example plot", x = "Temperature (F)", y = "Ozone (ppb)") +
  theme_minimal()
```

---

## 4. Section Two {#section2}

### 4.1 Subsection

```{r another-chunk}
# More analysis here
summary(airquality)
```

<div class="callout callout-warn">
<strong>⚠️ Common mistake</strong>
Describe a typical error students make here.
</div>

---

<div class="key-box">
<h4>🔑 Key Takeaways</h4>
<ul>
<li>First takeaway — one sentence, bold the key term.</li>
<li>Second takeaway.</li>
<li>Third takeaway.</li>
</ul>
</div>

---

<!-- ============================================================ -->
<!--  EXERCISES                                                   -->
<!-- ============================================================ -->

<div class="exercises">
<div class="exercises-header">✏️ Practice Exercises</div>

<div class="exercise">
<div class="exercise-num">Exercise 1</div>
<p>Write the exercise question here. Be specific about what R code to write
and what output to interpret.</p>
<details class="solution">
<summary>Show solution</summary>
<div class="solution-body">
<pre><code># Solution code here
summary(airquality$Ozone)</code></pre>
<p>Written explanation of the solution and what students should observe.</p>
</div>
</details>
</div>

<div class="exercise">
<div class="exercise-num">Exercise 2</div>
<p>Second exercise question.</p>
<details class="solution">
<summary>Show solution</summary>
<div class="solution-body">
<pre><code># Solution code here</code></pre>
<p>Explanation.</p>
</div>
</details>
</div>

</div>

```{r session-info, echo=FALSE}
# sessionInfo()   # uncomment to include reproducibility info
```

<footer class="page-footer">
EA030 Field &amp; Lab Activities &mdash; AXX
<span class="divider">|</span> Prof. Marc Los Huertos
<span class="divider">|</span> Pomona College
<span class="divider">|</span> Last updated: `r format(Sys.Date(), "%B %Y")`
<span class="divider">|</span> <a href="https://github.com/marclos/RTricks">GitHub</a>
</footer>
