---
title: "Multiple Regression & Generalized Linear Models (GLMs)"
author: "EA30 / marc"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
    theme: readable
---

# Welcome

This tutorial introduces **multiple regression** and **generalized linear models (GLMs)** in R. These tools allow you to model relationships between a response variable and **multiple predictors**, test **interactions**, compare models using **AIC**, and analyze **count** and **binary** data.

By the end of this tutorial, you will be able to:

- Fit multiple linear regression models
- Interpret coefficients with multiple predictors
- Specify and interpret interaction terms
- Compare models using AIC
- Fit GLMs for **count** (Poisson) and **binary** (logistic) data

---

# Setup

```{r setup, message=FALSE}
# Load required packages
library(tidyverse)
```

---

# Part 1: Multiple Linear Regression

## Example Data

We will use the built-in `mtcars` dataset.

```{r}
data(mtcars)
head(mtcars)
```

Suppose we want to predict **fuel efficiency (`mpg`)** using:

- Car weight (`wt`)
- Horsepower (`hp`)

---

## Fitting a Multiple Regression Model

```{r}
model1 <- lm(mpg ~ wt + hp, data = mtcars)
summary(model1)
```

### Interpretation

- Each coefficient represents the effect of a predictor **holding the others constant**
- The intercept is the predicted `mpg` when all predictors are zero (often not meaningful)

---

## Visualizing Partial Relationships

```{r}
# Added-variable (partial regression) plot
library(car)
avPlots(model1)
```

---

# Part 2: Interaction Terms

Sometimes the effect of one predictor depends on another.

Example: Does the effect of **weight** on `mpg` depend on **horsepower**?

```{r}
model2 <- lm(mpg ~ wt * hp, data = mtcars)
summary(model2)
```

### Reading an Interaction

- `wt:hp` represents how the slope of `wt` changes with `hp`
- Interaction models should usually include the main effects

---

## Plotting an Interaction

```{r}
mtcars %>%
  mutate(hp_group = if_else(hp > median(hp), "High HP", "Low HP")) %>%
  ggplot(aes(wt, mpg, color = hp_group)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

---

# Part 3: Model Selection with AIC

AIC (Akaike Information Criterion) balances **model fit** and **complexity**.

Lower AIC is better.

```{r}
model_simple <- lm(mpg ~ wt, data = mtcars)
model_full   <- lm(mpg ~ wt + hp + cyl, data = mtcars)

AIC(model_simple, model_full, model1, model2)
```

### Rules of Thumb

- ΔAIC < 2 → models are similarly supported
- Prefer simpler models when AICs are close

---

# Part 4: Generalized Linear Models (GLMs)

Linear regression assumes:

- Normally distributed errors
- Constant variance

GLMs allow different **distributions** and **link functions**.

---

## GLM for Count Data (Poisson Regression)

### Example

Suppose we count events (e.g., number of species).

```{r}
# Simulated count data
set.seed(42)
df_count <- tibble(
  x = runif(50, 0, 10),
  count = rpois(50, lambda = exp(1 + 0.3 * x))
)
```

```{r}
poisson_model <- glm(count ~ x, family = poisson, data = df_count)
summary(poisson_model)
```

### Interpretation

- Coefficients are on the **log scale**
- `exp(coef)` gives multiplicative effects

```{r}
exp(coef(poisson_model))
```

---

## Checking Fit

```{r}
# Plot observed vs predicted
df_count$predicted <- predict(poisson_model, type = "response")

ggplot(df_count, aes(x, count)) +
  geom_point() +
  geom_line(aes(y = predicted), color = "blue")
```

---

# Part 5: GLM for Binary Data (Logistic Regression)

### Example

Binary outcomes (yes/no, presence/absence).

```{r}
# Simulated binary data
set.seed(123)
df_binary <- tibble(
  x = rnorm(100),
  y = rbinom(100, 1, plogis(-0.5 + 1.2 * x))
)
```

```{r}
logit_model <- glm(y ~ x, family = binomial, data = df_binary)
summary(logit_model)
```

### Interpretation

- Coefficients are in **log-odds**
- `exp(coef)` gives **odds ratios**

```{r}
exp(coef(logit_model))
```

---

## Predicted Probabilities

```{r}
df_binary$prob <- predict(logit_model, type = "response")

ggplot(df_binary, aes(x, prob)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess")
```

---

# Model Comparison for GLMs

AIC works for GLMs as well:

```{r}
logit_null <- glm(y ~ 1, family = binomial, data = df_binary)
AIC(logit_null, logit_model)
```

---

# Key Takeaways ✅

- Multiple regression allows **multiple predictors**
- Interaction terms change how predictors work together
- AIC helps compare non-nested models
- GLMs extend regression to **non-normal data**
- Interpretation depends on the **link function**

---

# Common Mistakes

- ❌ Ignoring interactions when theoretically important
- ❌ Using linear regression for count or binary data
- ❌ Interpreting GLM coefficients on the raw scale
- ❌ Overfitting complex models with little data

---

# Congratulations 🎉

You now have the tools to fit and interpret **multiple regression models** and **GLMs**—a major step toward real-world data analysis in environmental and social sciences.
