---
title: "Shifting Baselines and Changepoint Analysis"
author: "Marc Los Huertos"
date: "EA030 · Compiled `r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    self_contained: true
    theme: null
    highlight: null
    toc: false
    keep_md: true
knit: (function(inputFile, encoding) {
    rmarkdown::render(
      inputFile,
      encoding = encoding,
      output_file = file.path(dirname(inputFile),
                              "../docs/Activities/A12_Shifting_Baselines.html")
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 8,
  fig.height = 4
)
```

```{r shared-css, echo=FALSE, results='asis'}
.src_dir <- dirname(normalizePath(knitr::current_input()))
.css_candidates <- c(
  file.path(.src_dir, "_shared.css"),
  file.path(.src_dir, "..", "_shared.css"),
  file.path(.src_dir, "..", "..", "_shared.css")
)
.css_file <- Filter(file.exists, .css_candidates)
if (length(.css_file) > 0L) {
  cat("<style>\n",
      paste(readLines(.css_file[[1L]]), collapse = "\n"),
      "\n</style>\n", sep = "")
}
```

```{r module-css, echo=FALSE, results='asis'}
module_css <- "
/* Module-specific: A12 changepoint labels, workflow boxes */
.cpt-label {
  display: inline-block;
  font-family: 'Source Code Pro', monospace; font-size: .72rem;
  font-weight: 700; letter-spacing: .1em; text-transform: uppercase;
  padding: 3px 9px; border-radius: 3px; margin: 2px 4px 2px 0;
  vertical-align: middle;
}
.cpt-mean   { background: #e8f5e9; color: #2e7d32; border: 1px solid #a5d6a7; }
.cpt-var    { background: #fff8e1; color: #f57f17; border: 1px solid #ffe082; }
.cpt-online { background: #e3f2fd; color: #1565c0; border: 1px solid #90caf9; }
.penalty-table td:first-child {
  font-family: 'Source Code Pro', monospace; font-weight: 600; color: var(--forest);
}
.step-label {
  font-family: 'Source Code Pro', monospace; font-size: .7rem; font-weight: 700;
  letter-spacing: .12em; text-transform: uppercase; color: var(--amber);
  display: block; margin-bottom: 4px;
}
.workflow-box {
  display: grid; grid-template-columns: repeat(auto-fit, minmax(160px, 1fr));
  gap: 10px; margin: 1.4em 0;
}
.wf-step {
  background: var(--white); border: 1px solid var(--mist);
  border-top: 3px solid var(--fern); border-radius: 0 0 5px 5px;
  padding: 12px 14px; font-size: .86rem;
}
.wf-step strong { display: block; color: var(--forest); margin-bottom: 3px; }
"
cat(paste0('<style>', module_css, '</style>\n'))
```

<div class="page-header">
<div class="module-tag">EA030 &mdash; Field &amp; Lab Activities &mdash; A12</div>
<h1>Shifting Baselines &amp; <span>Changepoint Analysis</span></h1>
<p class="sub"><strong>Prof. Marc Los Huertos</strong> &nbsp;&bull;&nbsp; Pomona College</p>
<a class="back-link" href="../index.html">&#8592; Back to RTricks</a>
<button class="btn-pdf" onclick="window.print()">&#8659; Save as PDF</button>
</div>

<div class="section-nav">
<a href="#baselines">Shifting Baselines</a>
<a href="#setup">Setup</a>
<a href="#offline">Offline Detection</a>
<a href="#variance">Variance Shifts</a>
<a href="#online">Online Detection</a>
<a href="#application">Real Application</a>
<a href="#exercises">Exercises</a>
</div>

<div class="goals-box">
<div class="goals-label">By the end of this module you will be able to</div>
<ol>
<li>Explain the shifting baseline problem and why it matters for long-term environmental monitoring.</li>
<li>Apply AMOC and PELT algorithms to detect mean shifts in historical time-series data using the <code>changepoint</code> package.</li>
<li>Detect changes in variance (flashiness, volatility) as well as changes in mean.</li>
<li>Describe how Bayesian Online Changepoint Detection (BOCPD) and CUSUM sequential detection work, and implement sequential analysis with the <code>strucchange</code> package.</li>
<li>Choose the right penalty (AIC, BIC, MBIC) for a given monitoring context and justify that choice.</li>
</ol>
</div>

> **What this module covers.** Environmental systems rarely stay put. Stream temperatures
> creep upward over decades. Salmon populations collapse to a new, lower equilibrium.
> Drought pushes groundwater levels down and keeps them there long after rain returns.
> Standard statistics — t-tests, regression, even time-series decomposition — implicitly
> assume a stable reference state. This module teaches you to *detect* the moments when
> that reference state shifts, using two complementary approaches: offline analysis of
> historical records and online (real-time) monitoring for new breaks.

---

## 1. The Shifting Baseline Problem {#baselines}

### 1.1 What is a baseline?

In environmental monitoring, a **baseline** is the expected background state of a system
against which current conditions are compared. Baselines define "normal":

- Is today's stream temperature warm or cool? *Compared to the baseline.*
- Has this year's salmon return been poor? *Compared to the baseline.*
- Is current groundwater depth declining? *Compared to the baseline.*

Baselines are not fixed facts of nature — they are socially and scientifically constructed
reference periods, and they shift.

### 1.2 The problem: each generation inherits a different baseline

<div class="callout callout-concept">
<strong>📌 Shifting baseline syndrome</strong>
Fisheries biologist Daniel Pauly coined this term in 1995 to describe how each generation
of scientists calibrates "normal" to the ecosystem state they first encountered.
Scientists who began work in the 1950s defined healthy cod stocks by 1950s abundance;
those starting in the 1980s accepted the already-depleted 1980s state as normal.
The result: the baseline quietly shifts downward with each decade, and the cumulative
loss is invisible to any single generation.
</div>

The same phenomenon appears across environmental science:

| System | Historic baseline | Current state | Apparent magnitude of change |
|---|---|---|---|
| North Atlantic cod | ~200M tonnes biomass (pre-1900) | < 1% of historic | Enormous if measured from true baseline |
| CA Sierra snowpack | 1950–1999 April 1 SWE average | −30 to −60% in many basins | Severe |
| Chesapeake Bay oxygen | Pre-industrial dissolved O₂ | Seasonal hypoxia now "normal" | Underestimated |
| Global mean temperature | Pre-industrial 1850–1900 | +1.2°C | ~50% larger than if baseline = 1980 |

### 1.3 What is a changepoint?

A **changepoint** is a time index $\tau$ at which one or more statistical properties
of a series — mean, variance, autocorrelation structure — undergo an abrupt shift.

$$X_t \sim \begin{cases}
\mathcal{N}(\mu_1, \sigma_1^2) & t \leq \tau \\
\mathcal{N}(\mu_2, \sigma_2^2) & t > \tau
\end{cases}$$

The goal of changepoint analysis is to estimate $\tau$ (and its uncertainty) from the data.
This is fundamentally different from trend detection (Mann-Kendall, regression slope) —
we are looking for **discrete jumps**, not gradual slopes.

<div class="callout callout-warn">
<strong>⚠️ Changepoints ≠ trends</strong>
A long-term upward trend and a sudden step-change can look similar in a raw time plot.
Mann-Kendall will find both; changepoint analysis will only flag the step. Choose your
tool based on which kind of change you expect ecologically.
</div>

### 1.4 Two detection paradigms

<div class="workflow-box">
<div class="wf-step">
<strong>Offline (retrospective)</strong>
All data in hand. Find the best-fitting set of changepoints over the full record. Methods: AMOC, PELT. Use for: historical analysis, report writing, model building.
</div>
<div class="wf-step">
<strong>Online (sequential)</strong>
Data arrives one point at a time. Update the probability of a new changepoint with each observation. Methods: BOCPD, CUSUM. Use for: sensor networks, early-warning systems, adaptive management.
</div>
</div>

---

## 2. Setup {#setup}

We need two packages:

- **`changepoint`** — mature, CRAN-stable implementation of AMOC, PELT, and CROPS for
  offline analysis (Killick & Eckley 2014).
- **`strucchange`** — structural change tests and sequential detection (Zeileis et al. 2002).
  Provides CUSUM-based online monitoring, optimal breakpoints with confidence intervals,
  and the `breakpoints()` function used in econometrics and environmental science alike.
- **`ggplot2`** and **`dplyr`** for visualization and wrangling.

```{r packages}
if (!require(changepoint)) install.packages("changepoint", repos = "https://cloud.r-project.org")
if (!require(strucchange))  install.packages("strucchange",  repos = "https://cloud.r-project.org")

library(changepoint)
library(strucchange)
library(ggplot2)
library(dplyr)
library(tidyr)
```

```{r palette, include=FALSE}
# Shared colour palette (mirrors _shared.css tokens)
col_forest  <- "#1e3a2f"
col_fern    <- "#4a7c5e"
col_sage    <- "#8aad96"
col_amber   <- "#c8832a"
col_amberlt <- "#e8a84a"
col_ink     <- "#1a1a18"
col_dust    <- "#7a7a72"
col_mist    <- "#d4e6da"

theme_rtricks <- theme_minimal(base_size = 11) +
  theme(
    plot.title    = element_text(family = "serif", face = "bold",
                                 colour = col_forest, size = 12),
    plot.subtitle = element_text(colour = col_dust, size = 9.5),
    plot.caption  = element_text(colour = col_dust, size = 8),
    axis.title    = element_text(colour = "#3a3a38", size = 9.5),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "#ede8df", linewidth = 0.4)
  )
theme_set(theme_rtricks)
```

### 2.1 Simulated California stream temperature scenario

We will build a realistic synthetic dataset before moving to real data. The scenario:

- **2000–2009** (n = 120 months): Cool baseline, mean July temperature ≈ 14°C
- **2010–2016** (n = 84 months): Prolonged drought; irreversible step up to ~18°C
- **2017–2023** (n = 84 months): Post-drought partial recovery to ~16°C, but with
  **higher variance** — wetter winters followed by hotter summers (flashiness)

This mimics the documented pattern in several Sierra Nevada foothill streams during and
after the 2012–2017 California drought.

```{r simulate-data, fig.cap="Simulated monthly stream temperature series. Three distinct phases are visible: cool pre-drought baseline, a sharp step at the drought onset, and a volatile post-drought period. Grey shading marks the drought window.", fig.alt="Line plot of 288 monthly temperature observations showing a rise from ~14°C to ~18°C around month 120, then a slight decline with increased variability from month 204 onward. Two vertical dashed lines mark the true changepoints."}
set.seed(2024)

n1 <- 120; n2 <- 84; n3 <- 84
N  <- n1 + n2 + n3

temp <- c(
  rnorm(n1, mean = 14.0, sd = 0.8),   # pre-drought
  rnorm(n2, mean = 18.2, sd = 0.9),   # drought
  rnorm(n3, mean = 16.1, sd = 2.1)    # post-drought (higher SD)
)

# Monthly date sequence
dates <- seq(as.Date("2000-01-01"), by = "month", length.out = N)

stream_df <- data.frame(
  date  = dates,
  month = seq_along(dates),
  temp  = temp,
  phase = rep(c("Pre-drought", "Drought", "Post-drought"),
               times = c(n1, n2, n3))
)

# True changepoints for reference
true_cpts <- c(n1, n1 + n2)

ggplot(stream_df, aes(x = date, y = temp)) +
  annotate("rect",
    xmin = dates[n1 + 1], xmax = dates[n1 + n2],
    ymin = -Inf, ymax = Inf,
    fill = "#fff8f4", alpha = 0.8) +
  geom_line(colour = col_fern, linewidth = 0.45) +
  geom_vline(xintercept = dates[true_cpts], linetype = "dashed",
             colour = col_amber, linewidth = 0.8) +
  annotate("text", x = dates[true_cpts[1]] + 30, y = 20.5,
           label = "True CP 1", size = 2.8, colour = col_amber, hjust = 0) +
  annotate("text", x = dates[true_cpts[2]] + 30, y = 20.5,
           label = "True CP 2", size = 2.8, colour = col_amber, hjust = 0) +
  labs(title = "Simulated Monthly Stream Temperature — Sierra Foothill Station",
       subtitle = "Scenario: 2012–2016 California drought + post-drought volatility",
       x = NULL, y = "Temperature (°C)",
       caption = "Grey band = simulated drought window   |   Dashed lines = true changepoints")
```

---

## 3. Offline Changepoint Detection {#offline}

Offline methods work on the complete historical record. The core idea is to partition
the series into $m+1$ segments (separated by $m$ changepoints) to maximise the
fit of a within-segment model, penalised for complexity.

$$\text{Minimise:} \quad \sum_{i=1}^{m+1} \mathcal{C}(y_{\tau_{i-1}+1:\tau_i}) + \beta f(m)$$

where $\mathcal{C}$ is a cost function (negative log-likelihood for the within-segment
model) and $\beta f(m)$ is the penalty for adding changepoints.

### 3.1 AMOC — At Most One Change

AMOC asks a single question: *is there exactly one changepoint, and if so, where?*
It evaluates every possible split position and picks the one that most reduces
the total cost. Useful as a first pass, or when you have strong prior reason to expect
a single regime shift.

```{r amoc, fig.cap="AMOC detects the single most significant mean shift in the stream temperature series. The estimated segment means are shown as horizontal dashed lines.", fig.alt="Two-panel changepoint plot showing the time series with a single detected changepoint and two horizontal mean lines at approximately 14°C and 17°C."}
m_amoc <- cpt.mean(temp, method = "AMOC")

# Extract results
cpt_amoc    <- cpts(m_amoc)
means_amoc  <- param.est(m_amoc)$mean

cat("AMOC changepoint at index:", cpt_amoc, "\n")
cat("   → month", cpt_amoc, "=", format(dates[cpt_amoc], "%b %Y"), "\n")
cat("Segment means:", round(means_amoc, 2), "°C\n")

plot(m_amoc,
     main   = "AMOC: Single Mean Changepoint",
     ylab   = "Temperature (°C)", xlab = "Month index",
     col    = col_fern, cpt.col = col_amber,
     cpt.width = 3, mean.col = col_forest)
```

AMOC correctly identifies the large drought-onset shift but cannot, by definition,
find the second changepoint. This is why PELT is usually preferred for real
environmental records with complex histories.

### 3.2 PELT — Multiple Changepoints

The **Pruned Exact Linear Time** algorithm (Killick et al. 2012) finds the *globally
optimal* set of changepoints efficiently, in $O(n)$ time on average. It eliminates
candidate split positions using a pruning rule — once a partition is provably suboptimal
it is never revisited.

The key user decision is the **penalty**. Too small a penalty → too many changepoints
(overfitting noise); too large → important shifts are missed.

```{r penalty-table, echo=FALSE}
knitr::kable(
  data.frame(
    Penalty = c("AIC", "BIC", "MBIC", "Manual (numeric)"),
    Formula = c("2k", "k·log(n)", "3k·log(n)", "user-specified"),
    Behavior = c(
      "Most permissive; finds more changepoints",
      "Balanced; standard default for large n",
      "Conservative; reduces false positives",
      "Full control for domain-specific thresholds"
    ),
    `Best for` = c(
      "Exploratory analysis; dense monitoring",
      "Most environmental records",
      "Sparse data or high cost of false alarms",
      "Regulatory or management thresholds"
    )
  ),
  col.names = c("Penalty", "Formula (k = no. cpts)", "Behavior", "Best for"),
  caption   = "Penalty comparison for PELT changepoint detection."
)
```

```{r pelt-bic, fig.cap="PELT with BIC penalty detects two mean changepoints in the stream temperature record. Vertical lines mark detected changepoints; horizontal lines show estimated segment means.", fig.alt="Time series with two detected changepoints marked by vertical lines at approximately months 120 and 204, with three horizontal mean lines at 14, 18, and 16°C."}
m_pelt_bic <- cpt.mean(temp, method = "PELT", penalty = "BIC")

cpts_bic   <- cpts(m_pelt_bic)
means_bic  <- param.est(m_pelt_bic)$mean

cat("PELT (BIC) changepoints at indices:", cpts_bic, "\n")
cat("   →", format(dates[cpts_bic], "%b %Y"), "\n")
cat("Segment means:", round(means_bic, 2), "°C\n")
cat("True changepoints were at:", true_cpts, "\n")

plot(m_pelt_bic,
     main   = "PELT (BIC Penalty): Multiple Mean Changepoints",
     ylab   = "Temperature (°C)", xlab = "Month index",
     col    = col_fern, cpt.col = col_amber,
     cpt.width = 3, mean.col = col_forest)
```

```{r pelt-aic, fig.cap="Comparison of PELT with AIC vs BIC penalty. AIC is more permissive and may detect additional changepoints that represent noise rather than genuine regime shifts.", fig.alt="Two time-series plots stacked vertically comparing AIC and BIC penalty detections in the same dataset."}
m_pelt_aic <- cpt.mean(temp, method = "PELT", penalty = "AIC")

par(mfrow = c(2, 1), mar = c(3, 4, 2.5, 1))
plot(m_pelt_bic, main = "PELT — BIC penalty",
     ylab = "°C", col = col_fern, cpt.col = col_amber, cpt.width = 2,
     mean.col = col_forest, xlab = "")
plot(m_pelt_aic, main = "PELT — AIC penalty",
     ylab = "°C", col = col_fern, cpt.col = col_amber, cpt.width = 2,
     mean.col = col_forest, xlab = "Month index")
par(mfrow = c(1, 1))
```

<div class="callout callout-concept">
<strong>📌 Which penalty to use in practice?</strong>
Start with BIC. If you expect dense changepoints (e.g., weekly sensor data with frequent
regime shifts) try AIC. If false alarms carry regulatory costs — triggering a permit
review, declaring a fishery closure — use MBIC or a manually tuned penalty. Always
report which penalty you used and why; it is a modelling decision, not a black box.
</div>

### 3.3 Comparing estimated changepoints to ground truth

```{r accuracy-table, echo=FALSE}
knitr::kable(
  data.frame(
    Method   = c("AMOC", "PELT (BIC)", "PELT (AIC)"),
    Detected = c(
      paste(cpts(m_amoc), collapse = ", "),
      paste(cpts(m_pelt_bic), collapse = ", "),
      paste(cpts(m_pelt_aic), collapse = ", ")
    ),
    n_cpts   = c(
      length(cpts(m_amoc)),
      length(cpts(m_pelt_bic)),
      length(cpts(m_pelt_aic))
    ),
    True     = "120, 204",
    Notes    = c(
      "Finds only the dominant shift",
      "Matches true changepoints well",
      "May over-segment noisy post-drought phase"
    )
  ),
  col.names = c("Method", "Detected indices", "No. cpts", "True cpts", "Notes"),
  caption   = "Changepoint detection accuracy across methods. True changepoints at months 120 and 204."
)
```

---

## 4. Detecting Shifts in Variance {#variance}

So far we have looked for changes in the **mean**. But the post-drought scenario
specifically includes a variance increase — the stream became flashier (wider swings
between wet-year highs and dry-year lows). This is ecologically important: high
variance increases thermal stress events even if the mean temperature is tolerable.

<div class="callout callout-formula">
<strong>⚙️ Why variance shifts matter</strong>
If mean temp = 16°C and SD = 0.9°C, temperatures rarely exceed 18°C.
If mean temp = 16°C and SD = 2.1°C, temperatures exceed 18°C ~17% of the time —
above the thermal tolerance limit of many cold-water fish.
Detecting the variance shift is as ecologically important as detecting the mean shift.
</div>

### 4.1 Testing for a variance changepoint

```{r cpt-var, fig.cap="PELT applied to variance detects the onset of increased volatility in the post-drought phase. Note that the mean may appear similar while the spread of values is clearly wider.", fig.alt="Time series plot with a detected variance changepoint around month 204, showing noticeably wider oscillations in the right-hand segment."}
m_var <- cpt.var(temp, method = "PELT", penalty = "BIC")

cpts_var   <- cpts(m_var)
vars_est   <- param.est(m_var)$variance

cat("Variance changepoint(s) at index:", cpts_var, "\n")
cat("   →", format(dates[cpts_var], "%b %Y"), "\n")
cat("Estimated variances:", round(vars_est, 3), "(SD:",
    round(sqrt(vars_est), 2), ")\n")

plot(m_var,
     main      = "Variance Changepoint Detection (PELT)",
     ylab      = "Temperature (°C)", xlab = "Month index",
     col       = col_fern, cpt.col = col_amber,
     cpt.width = 3)
```

### 4.2 Simultaneous mean and variance changepoints

`cpt.meanvar()` searches for locations where *either* parameter changes — useful when
you do not want to pre-specify which parameter is shifting.

```{r cpt-meanvar, fig.cap="Joint mean-and-variance changepoint detection. Each segment has its own estimated mean and variance; shifts in either property trigger a new changepoint.", fig.alt="Time series with three detected changepoints showing segments of different means and spread widths."}
m_mv <- cpt.meanvar(temp, method = "PELT", penalty = "BIC")

cat("Mean+Variance changepoints:", cpts(m_mv), "\n")
cat("Segment means:",     round(param.est(m_mv)$mean, 2), "\n")
cat("Segment variances:", round(param.est(m_mv)$variance, 3), "\n")

plot(m_mv,
     main      = "Mean + Variance Changepoints (PELT)",
     ylab      = "Temperature (°C)", xlab = "Month index",
     col       = col_fern, cpt.col = col_amber,
     cpt.width = 3, mean.col = col_forest)
```

---

## 5. Sequential & Online Changepoint Detection {#online}

The offline methods above assume the full time series is available. But environmental
monitoring is often **sequential** — sensors stream data continuously and managers
need to know as soon as possible when a new regime has begun.

### 5.1 The conceptual basis: BOCPD and CUSUM

Two families of sequential methods dominate practice:

**Bayesian Online Changepoint Detection (BOCPD, Adams & MacKay 2007)**
maintains a probability distribution over the *run length* $r_t$ — the number of steps
since the last changepoint. At each new observation the distribution is updated via
Bayes' rule. The probability that $r_t = 0$ is the real-time alarm probability.

**CUSUM (Cumulative Sum Control Chart)**
accumulates signed deviations from a reference mean. When the cumulative sum exceeds
a threshold, an alarm fires. Unlike BOCPD, CUSUM is nonparametric and makes no
distributional assumptions — it simply asks whether the data have drifted from
the reference level.

<div class="callout callout-warn">
<strong>⚠️ The shifting baseline problem in sequential detection</strong>
Both BOCPD and CUSUM use a <em>fixed</em> reference state. If the true baseline has
drifted far from that reference over years (e.g., stream temperatures have risen 4°C
since the sensor was first calibrated), sensitivity degrades because ordinary
observations now look like outliers relative to the stale prior.<br><br>
Fixes include: (1) recalibrating the reference after each confirmed changepoint,
(2) using a moving window for the reference mean, or (3) applying a robust
adaptive hazard function in BOCPD.
</div>

### 5.2 CUSUM and structural change tests with `strucchange`

The `strucchange` package implements empirical fluctuation processes (EFP) —
running statistics that accumulate evidence of structural change over time.
The OLS-CUSUM process is the most widely used.

```{r cusum-stream, fig.cap="OLS-CUSUM fluctuation process for the simulated stream temperature series. The process accumulates deviations from the full-sample OLS fit. When it breaches the red boundary lines (95% significance envelope), structural change is detected. Both changepoints are clearly visible as boundary crossings.", fig.alt="CUSUM plot with a time axis and a fluctuation process line that crosses the significance boundary twice, around months 120 and 204."}
# Build a simple linear model framework: temp ~ 1 (intercept only)
# The CUSUM process tests whether that intercept is stable over time
efp_stream <- efp(temp ~ 1, type = "OLS-CUSUM")
plot(efp_stream,
     main = "OLS-CUSUM: Sequential Test for Mean Shift",
     ylab = "Empirical fluctuation process",
     col  = col_forest, lwd  = 1.2)
```

```{r sctest}
# Formal test: is there ANY structural break?
sc_test <- sctest(efp_stream)
cat(sprintf("OLS-CUSUM test: statistic = %.3f, p-value = %.4f\n",
            sc_test$statistic, sc_test$p.value))
cat("Conclusion:", ifelse(sc_test$p.value < 0.05,
    "Strong evidence of at least one structural break.",
    "No significant structural break detected."), "\n")
```

### 5.3 Optimal breakpoints with confidence intervals

`strucchange::breakpoints()` finds the globally optimal set of breakpoints under an
OLS framework — conceptually similar to PELT but reporting confidence intervals on
the *timing* of each break. This is particularly valuable for management: knowing
whether a shift occurred in 2009, 2011, or 2013 has very different implications
for attributing cause.

```{r strucchange-bps, fig.cap="Optimal breakpoints with 95% confidence intervals from strucchange. The vertical lines show the point estimates; the shaded bands show the 95% CI on each changepoint date. Narrower bands indicate a sharper, more precisely dated transition.", fig.alt="Time series of stream temperature with two vertical dashed lines at the detected changepoints and shaded confidence interval bands around each."}
bp_stream <- breakpoints(temp ~ 1)

# How many breakpoints minimise BIC?
summary(bp_stream)
plot(bp_stream, main = "BIC by number of breakpoints")
```

```{r bp-ci, fig.cap="Stream temperature with detected breakpoints and 95% confidence intervals. Segment means are shown as solid horizontal lines. The confidence bands illustrate the uncertainty in when each transition occurred.", fig.alt="Line plot of temperature over 288 months with two breakpoint confidence bands shaded in amber and two horizontal segment mean lines."}
# Select optimal number of breakpoints (BIC minimum)
bp_opt <- breakpoints(bp_stream)      # optimal selection
ci_bp  <- confint(bp_stream)          # CI requires breakpointsfull

cat("Optimal breakpoint indices:", bp_opt$breakpoints, "\n")
cat("   →", format(dates[bp_opt$breakpoints], "%b %Y"), "\n")
cat("True changepoints:         ", true_cpts, "\n")
cat("95% CI:", ci_bp$confint, "\n")

# Segment means: one intercept per segment via breakfactor()
bf        <- breakfactor(bp_opt)
seg_means <- coef(lm(temp ~ bf - 1))
cat("Segment means:", round(seg_means, 2), "°C\n")

# Plot with ggplot
n_bp   <- length(bp_opt$breakpoints)
bp_idx <- c(0, bp_opt$breakpoints, N)
mean_df <- data.frame(
  xmin = dates[bp_idx[-length(bp_idx)] + 1],
  xmax = dates[bp_idx[-1]],
  mean = seg_means
)
# confint row = breakpoint, cols = [lower, estimate, upper]
ci_mat <- ci_bp$confint
ci_df  <- data.frame(
  x   = dates[ci_mat[, 2]],
  xlo = dates[pmax(1, ci_mat[, 1])],
  xhi = dates[pmin(N, ci_mat[, 3])]
)

ggplot(stream_df, aes(x = date, y = temp)) +
  geom_rect(data = ci_df,
              aes(xmin = xlo, xmax = xhi, ymin = -Inf, ymax = Inf),
              inherit.aes = FALSE,
              fill = col_amber, alpha = 0.18) +
  geom_line(colour = col_fern, linewidth = 0.45, alpha = 0.8) +
  geom_segment(data = mean_df,
               aes(x = xmin, xend = xmax, y = mean, yend = mean),
               colour = col_forest, linewidth = 1.2) +
  geom_vline(xintercept = dates[ci_mat[, 2]],
             linetype = "dashed", colour = col_amber, linewidth = 0.8) +
  labs(title    = "Stream Temperature: Optimal Breakpoints with 95% CIs",
       subtitle = "Amber bands = 95% confidence interval on changepoint timing",
       x = NULL, y = "Temperature (°C)")
```

<div class="callout callout-concept">
<strong>📌 Why confidence intervals on timing matter</strong>
A point estimate says "the shift occurred at month 120." A confidence interval says
"we're 95% confident it occurred between months 115 and 125." In regulatory contexts
(e.g., attributing a water-quality decline to a specific upstream land-use change
permitted in month 118), the width of this interval is often the key evidence.
</div>

### 5.4 Simulating online detection with a rolling CUSUM

To mimic real-time monitoring, we compute the CUSUM statistic on expanding windows —
adding one observation at a time and checking whether the statistic exceeds the
critical value derived from the pre-change baseline.

```{r rolling-cusum, fig.cap="Rolling CUSUM statistic computed on expanding windows. The statistic accumulates evidence of a shift from the pre-drought baseline (months 1–120). The critical value (dashed red) is set at the 95th percentile of the null distribution. Exceedances flag a detected changepoint.", fig.alt="Line plot of rolling CUSUM statistic over time showing a sharp rise above the critical threshold around month 130, followed by another exceedance around month 210."}
# Baseline: use first 120 months (pre-change)
baseline_mean <- mean(temp[1:n1])
baseline_sd   <- sd(temp[1:n1])

# Normalised CUSUM: accumulate (x - mu_0) / sigma_0
cusum <- cumsum((temp - baseline_mean) / baseline_sd)

# Critical value: for a standard CUSUM, ~±1.96*sqrt(n) at 95%
# Use 4.0 as a conservative threshold (standard quality-control value)
crit <- 4.0

cusum_df <- data.frame(
  date  = dates,
  cusum = cusum,
  alarm = abs(cusum) > crit
)

ggplot(cusum_df, aes(x = date, y = cusum)) +
  geom_hline(yintercept = c(-crit, crit), linetype = "dashed",
             colour = "#c0392b", linewidth = 0.8) +
  geom_hline(yintercept = 0, colour = col_dust, linewidth = 0.4) +
  geom_line(colour = col_forest, linewidth = 0.5) +
  geom_point(data = filter(cusum_df, alarm),
             aes(x = date, y = cusum),
             colour = col_amber, size = 1.2, alpha = 0.6) +
  annotate("text", x = min(dates) + 30, y = crit + 1.5,
           label = "Critical value (±4.0)", size = 2.8,
           colour = "#c0392b") +
  labs(title    = "Rolling CUSUM — Online Detection from Pre-Drought Baseline",
       subtitle = "Reference = pre-drought mean; amber points = alarm threshold exceeded",
       x = NULL, y = "Cumulative standardised sum",
       caption  = "Note: CUSUM sensitivity degrades as the new baseline drifts from the reference mean")
```

Note the important pattern: once the drought-era mean is established, the CUSUM
statistic continues to rise because every new observation is above the pre-drought
reference. This illustrates the **shifting baseline problem** directly — the old
reference is no longer relevant but the statistic keeps accumulating. Resetting the
reference after the first confirmed changepoint is essential for detecting subsequent
transitions.

---
## 6. Real Environmental Application {#application}

### 6.1 Built-in data: Nile river flow

R includes the `Nile` dataset: annual flow of the Nile at Aswan, 1871–1970.
A famous changepoint corresponds to the construction of the Aswan Low Dam (~1898)
and later hydrological changes. This is a canonical environmental changepoint example.

```{r nile-data, fig.cap="Annual Nile river discharge at Aswan, 1871–1970 (millions of m³). A well-documented structural break occurs around 1898 following dam construction and upstream land use change.", fig.alt="Time series plot of Nile discharge showing a clear downward step around 1898, with lower and more stable flow in the second half of the record."}
data(Nile)
nile_df <- data.frame(
  year      = as.numeric(time(Nile)),
  discharge = as.numeric(Nile)
)

ggplot(nile_df, aes(x = year, y = discharge)) +
  geom_line(colour = col_fern, linewidth = 0.6) +
  geom_smooth(method = "loess", span = 0.3,
              colour = col_amber, se = FALSE, linewidth = 0.9) +
  labs(title    = "Annual Nile Discharge — Aswan, 1871–1970",
       subtitle = "Loess smoother (orange) highlights the structural break",
       x = "Year", y = "Discharge (10⁸ m³)",
       caption  = "Source: R built-in Nile dataset (Cobb 1978)")
```

### 6.2 Applying the full workflow

```{r nile-workflow}
# Step 1: PELT mean changepoint
nile_pelt <- cpt.mean(as.numeric(Nile), method = "PELT", penalty = "BIC")
cpt_year  <- 1871 + cpts(nile_pelt) - 1  # convert index to year
means_nile <- param.est(nile_pelt)$mean

cat("Detected changepoint year:", cpt_year, "\n")
cat("Segment means:", round(means_nile), "× 10⁸ m³\n")
cat("Change in mean:", round(diff(means_nile)), "× 10⁸ m³",
    sprintf("(%.0f%%)\n", 100 * diff(means_nile) / means_nile[1]))
```

```{r nile-pelt-plot, fig.cap="PELT changepoint detection on the Nile discharge record. The detected shift corresponds closely to the 1898 construction of the Aswan Low Dam and associated upstream hydrological changes.", fig.alt="Nile discharge time series with a single detected changepoint at approximately 1899, and two horizontal segment mean lines."}
# Step 2: Visualise with ggplot (manual, for clean styling)
cp_idx  <- cpts(nile_pelt)
cp_yr   <- cpt_year

anno_df <- data.frame(
  xmin   = c(1871, cp_yr + 1),
  xmax   = c(cp_yr, 1970),
  ymean  = means_nile
)

ggplot(nile_df, aes(x = year, y = discharge)) +
  geom_line(colour = col_fern, linewidth = 0.5, alpha = 0.8) +
  geom_segment(data = anno_df,
               aes(x = xmin, xend = xmax, y = ymean, yend = ymean),
               colour = col_forest, linewidth = 1.3, linetype = "solid") +
  geom_vline(xintercept = cp_yr, colour = col_amber,
             linewidth = 1, linetype = "dashed") +
  annotate("text", x = cp_yr + 1, y = 1370,
           label = paste0("Changepoint: ", cp_yr),
           hjust = 0, size = 3, colour = col_amber) +
  annotate("text", x = 1873, y = means_nile[1] + 35,
           label = paste0("Mean = ", round(means_nile[1])),
           size = 2.9, colour = col_forest) +
  annotate("text", x = cp_yr + 2, y = means_nile[2] + 35,
           label = paste0("Mean = ", round(means_nile[2])),
           size = 2.9, colour = col_forest) +
  labs(title    = "Nile Discharge: PELT Mean Changepoint",
       subtitle = paste0("Detected shift at ", cp_yr,
                         " — consistent with Aswan Low Dam construction (1898)"),
       x = "Year", y = "Discharge (10⁸ m³)")
```

```{r nile-struc, fig.cap="strucchange breakpoints with 95% confidence interval for the Nile discharge series. The detected break near 1899 is precisely dated; the narrow CI confirms the transition was sharp.", fig.alt="Nile discharge time series with a vertical dashed line at approximately 1899 and a shaded confidence interval band, plus two horizontal segment mean lines."}
# Step 3: Breakpoints with confidence intervals
nile_bp  <- breakpoints(as.numeric(Nile) ~ 1)
nile_opt <- breakpoints(nile_bp)                  # optimal selection
nile_ci  <- confint(nile_bp)                      # CI from breakpointsfull

bp_yr    <- 1871 + nile_opt$breakpoints - 1
ci_lo    <- 1871 + nile_ci$confint[1, 1] - 1
ci_hi    <- 1871 + nile_ci$confint[1, 3] - 1
bf_n   <- breakfactor(nile_opt)
seg_mu <- coef(lm(as.numeric(Nile) ~ bf_n - 1))

cat(sprintf("Breakpoint: %d  (95%% CI: %d – %d)\n", bp_yr, ci_lo, ci_hi))
cat("Segment means:", round(seg_mu), "× 10⁸ m³\n")

ggplot(nile_df, aes(x = year, y = discharge)) +
  annotate("rect", xmin = ci_lo, xmax = ci_hi,
           ymin = -Inf, ymax = Inf, fill = col_amber, alpha = 0.18) +
  geom_line(colour = col_fern, linewidth = 0.5, alpha = 0.8) +
  geom_segment(data = data.frame(
                 xmin = c(1871, bp_yr + 1), xmax = c(bp_yr, 1970),
                 mu   = seg_mu),
               aes(x = xmin, xend = xmax, y = mu, yend = mu),
               colour = col_forest, linewidth = 1.3) +
  geom_vline(xintercept = bp_yr, colour = col_amber,
             linetype = "dashed", linewidth = 1) +
  annotate("text", x = bp_yr + 1, y = 1370,
           label = sprintf("Break: %d\n(95%% CI: %d–%d)", bp_yr, ci_lo, ci_hi),
           hjust = 0, size = 2.8, colour = col_amber) +
  labs(title    = "Nile Discharge: strucchange Breakpoint + 95% CI",
       subtitle = "Amber band = confidence interval on timing of the structural break",
       x = "Year", y = "Discharge (10⁸ m³)")
```

### 6.3 Interpreting changepoints in context

Statistical detection is only the first step. The analysis above tells us *when*
a shift occurred and *how large* it was. Environmental interpretation requires
connecting these findings to mechanism:

```{r interpretation-table, echo=FALSE}
knitr::kable(
  data.frame(
    Question = c(
      "When did the shift occur?",
      "How large was it?",
      "Is it permanent?",
      "What caused it?",
      "What is the new baseline?"
    ),
    Tool = c(
      "cpt.mean(), breakpoints() + confint() — point estimate + CI",
      "Difference in segment means; effect size = Δμ / pooled SD",
      "PELT + strucchange CUSUM on rolling windows",
      "Domain knowledge, causal inference — statistics cannot answer this",
      "Mean of the most recent confirmed segment"
    ),
    `Nile example` = c(
      paste0(cp_yr, " (± ~2 years)"),
      paste0(round(abs(diff(means_nile))), " × 10⁸ m³ (", round(100*abs(diff(means_nile))/means_nile[1]), "% reduction)"),
      "Appears permanent over the 1871–1970 record",
      "Aswan Low Dam (1898); upstream irrigation expansion",
      paste0(round(means_nile[2]), " × 10⁸ m³ (post-1899 mean)")
    )
  ),
  col.names = c("Question", "Statistical tool", "Nile example"),
  caption   = "Connecting statistical output to environmental interpretation."
)
```

<div class="callout callout-concept">
<strong>📌 The new baseline is a management decision, not just a number</strong>
Once a changepoint is confirmed, the pre-change period is no longer a valid reference for
current conditions. Using pre-dam Nile flow as the "normal" baseline for contemporary
water allocation would be as misleading as using 1950 cod abundances to set current
fishing quotas. Changepoint analysis forces a reckoning with which period actually
represents the current system state.
</div>

---

## 7. Reporting Changepoint Analyses {#reporting}

A complete changepoint analysis report should include:

1. **Method and package**: `cpt.mean(..., method = "PELT", penalty = "BIC")` via `changepoint` v`r packageVersion("changepoint")`
2. **Detected changepoints**: indices or dates, with uncertainty where available
3. **Segment statistics**: mean and SD per segment
4. **Penalty justification**: why you chose AIC vs BIC vs MBIC
5. **Ecological interpretation**: what mechanism might explain the detected shift
6. **Caveats**: minimum detectable effect size; sensitivity to penalty choice

```{r session-report, echo=FALSE}
cat("changepoint version:", as.character(packageVersion("changepoint")), "\n")
cat("strucchange version:",  as.character(packageVersion("strucchange")), "\n")
cat("R version:",
    paste(R.version$major, R.version$minor, sep = "."), "\n")
```

---

<div class="key-box">
<h4>🔑 Key Takeaways</h4>
<ul>
<li><strong>Shifting baselines</strong> mean that the "normal" reference state changes over time — detecting these shifts is essential before any comparison to historical conditions.</li>
<li><strong>AMOC</strong> finds the single most significant break; <strong>PELT</strong> finds the globally optimal set of multiple breaks in O(n) time.</li>
<li><strong>Penalty choice</strong> (AIC / BIC / MBIC) controls the trade-off between sensitivity and false alarms — always report and justify it.</li>
<li><strong>Variance shifts</strong> can be ecologically significant even when the mean is unchanged — use <code>cpt.var()</code> or <code>cpt.meanvar()</code>.</li>
<li><strong>Sequential methods</strong> (<code>strucchange</code>) provide CUSUM monitoring and confidence intervals on changepoint <em>timing</em> — essential for regulatory attribution.</li>
<li>Statistical detection answers <em>when</em> and <em>how much</em> — ecological interpretation of <em>why</em> requires domain knowledge.</li>
</ul>
</div>

---

<div class="exercises">
<div class="exercises-header">✏️ Practice Exercises</div>

<div class="exercise">
<div class="exercise-num">Exercise 1 — Apply PELT to air quality data</div>
<p>The built-in <code>airquality</code> dataset contains 153 daily ozone readings (May–September 1973).
(a) Apply <code>cpt.mean()</code> with PELT and BIC penalty to the <code>Ozone</code> column (remove NAs first).
(b) Report detected changepoints as calendar dates (hint: <code>airquality$Month</code> and <code>Day</code>).
(c) Plot the result. Do the detected breaks correspond to any meteorological explanation?</p>
<details class="solution">
<summary>Show solution</summary>
<div class="solution-body">
<pre><code>library(changepoint)
data(airquality)
ozone_clean <- airquality[!is.na(airquality$Ozone), ]
m <- cpt.mean(ozone_clean$Ozone, method = "PELT", penalty = "BIC")
cpts(m)

# Convert index to date
ozone_clean$date <- as.Date(paste(1973, ozone_clean$Month, ozone_clean$Day, sep="-"))
ozone_clean$date[cpts(m)]
plot(m, ylab = "Ozone (ppb)", xlab = "Observation index",
     main = "Ozone changepoints — New York 1973")</code></pre>
<p>A changepoint typically appears around late June / early July, separating the
cool, low-ozone spring period from the hot, high-ozone midsummer. This corresponds
to the photochemical production peak driven by higher temperatures and longer days.</p>
</div>
</details>
</div>

<div class="exercise">
<div class="exercise-num">Exercise 2 — Variance shift in PM2.5 data</div>
<p>Simulate two years of daily PM2.5: Year 1 has mean = 10 µg/m³, SD = 2; Year 2 has
mean = 10 µg/m³, SD = 8 (more wildfire smoke events). Use <code>set.seed(42)</code>.
(a) Does <code>cpt.mean()</code> detect a changepoint? (b) Does <code>cpt.var()</code>?
(c) What does this tell you about the ecological importance of monitoring variance,
not just the mean?</p>
<details class="solution">
<summary>Show solution</summary>
<div class="solution-body">
<pre><code>set.seed(42)
pm25 <- c(rnorm(365, 10, 2), rnorm(365, 10, 8))

m_mean <- cpt.mean(pm25, method = "PELT", penalty = "BIC")
m_var  <- cpt.var(pm25,  method = "PELT", penalty = "BIC")

cat("Mean changepoints:", cpts(m_mean), "\n")
cat("Variance changepoints:", cpts(m_var), "\n")
par(mfrow = c(2,1))
plot(m_mean, main = "Mean changepoints (none expected)")
plot(m_var,  main = "Variance changepoint (expected at day 365)")</code></pre>
<p><code>cpt.mean()</code> finds no changepoint (both years have the same mean).
<code>cpt.var()</code> correctly identifies the onset of higher variance at day ~365.
This matters because even with the same annual mean PM2.5, the high-variance year
has far more days exceeding the 35 µg/m³ NAAQS threshold — health risk is driven
by peak events, not averages.</p>
</div>
</details>
</div>

<div class="exercise">
<div class="exercise-num">Exercise 3 — Penalty sensitivity analysis</div>
<p>Using the Nile dataset, run <code>cpt.mean()</code> with PELT and penalties "AIC", "BIC",
and "MBIC". (a) How many changepoints does each penalty detect?
(b) Are the detected years consistent across penalties?
(c) Which penalty would you recommend for a regulatory report to the Egyptian Ministry
of Water, and why?</p>
<details class="solution">
<summary>Show solution</summary>
<div class="solution-body">
<pre><code>data(Nile)
for (pen in c("AIC", "BIC", "MBIC")) {
  m <- cpt.mean(as.numeric(Nile), method = "PELT", penalty = pen)
  yrs <- 1871 + cpts(m) - 1
  cat(pen, ":", length(cpts(m)), "changepoints at", yrs, "\n")
}</code></pre>
<p>AIC typically finds 2–3 changepoints; BIC and MBIC usually agree on one dominant
changepoint (~1899). For a regulatory report, MBIC or BIC is appropriate — false
alarms (claiming a shift that isn't real) could trigger costly policy responses, so
a conservative penalty is justified. The primary changepoint at 1899 is robust across
all penalties, lending it scientific credibility.</p>
</div>
</details>
</div>

<div class="exercise">
<div class="exercise-num">Exercise 4 — Confidence intervals on changepoint timing</div>
<p>Using the Nile dataset, run <code>breakpoints(as.numeric(Nile) ~ 1)</code>
and extract the 95% confidence interval with <code>confint()</code>.
(a) How wide is the CI in years? (b) Now add <code>AR(1)</code> autocorrelation
correction: <code>breakpoints(as.numeric(Nile) ~ 1, h = 0.15)</code>. Does the
CI change? (c) Why does autocorrelation in the data affect the precision of
changepoint timing estimates?</p>
<details class="solution">
<summary>Show solution</summary>
<div class="solution-body">
<pre><code>library(strucchange)
data(Nile)

# Standard breakpoints
bp  <- breakpoints(as.numeric(Nile) ~ 1)
bpo <- breakpoints(bp)
  ci  <- confint(bp)
cat("Break year:", 1871 + bpo$breakpoints - 1, "
")
cat("95% CI:", 1871 + ci$confint[c(1,3)] - 1, "
")
cat("CI width:", diff(ci$confint[c(1,3)]), "years
")

# With smaller bandwidth (more sensitive to local change)
bp2 <- breakpoints(as.numeric(Nile) ~ 1, h = 0.15)
ci2 <- confint(bp2)
cat("Narrower h CI:", 1871 + ci2$confint[c(1,3)] - 1, "
")</code></pre>
<p>A wider CI results when the transition is gradual or when the series is
autocorrelated — adjacent observations carry redundant information, effectively
reducing the sample size available to pin down the exact transition point.
The dam construction at Aswan was fairly sudden, so the CI is typically narrow
(~4–6 years), lending credibility to the 1899 attribution.</p>
</div>
</details>
</div>
</div>

<div class="module-nav">
<span></span>
<a class="nav-index" href="../index.html">↑ Module Index</a>
<span></span>
</div>

<footer class="page-footer">
EA030 Field &amp; Lab Activities &mdash; A12
<span class="divider">|</span> Prof. Marc Los Huertos
<span class="divider">|</span> Pomona College
<span class="divider">|</span> Last updated: `r format(Sys.Date(), "%B %Y")`
<span class="divider">|</span> <a href="https://github.com/marclos/RTricks">GitHub</a>
</footer>
