---
title: "Course Time"
author: "Marc Los Huertos"
date: "`r Sys.Date()`"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## General Expections

Based on the "accepted" expectations, we should be a spending a minimum of 2 hours outside of class for every hour in class (2:1 ratio). Counting only the lecture and 1/2 of lab time that would be about atleast 6 hours outside of class. 

In general, I get worried if a class requires more than 8 hours outside of class for every hour in class on a consistent basis.

## Time Spent Outside of Class

```{r, echo=TRUE}

week2 <- c(1.5, 1, 15, 40, 3, 2, 10, 1.5, 
           0, 20, 2, 45, 7, 10, 2, 0, 2, 2, 2.5, 5, 0)

```

This is an R Markdown presentation. But code can be added just like the PDFs we have been knitting.


## Skewed Distribution

- Outliers
- Data Entry Mistakes, maybe?
- Excessive Readings, likely!
```{r}
hist(week2, breaks=20, col="lightblue", border="black", xlim=c(0,50), xlab="Time (hours)", main="Time spent outside in a week", ylab="Frequency")
```

## t-test a hypothesis: Different than 6 hours

```{r test, echo = TRUE}
t.test(week2, mu=6)
```


## t-test: Greater than 6 hours

```{r testgreater, echo = TRUE}
t.test(week2, mu=6, alternative="greater" )
```

## Marc's Conclusions

- Make sure the quiz directions are clear.
- Reduced Readings starting next week.
- Offer additional support: If you find the readings to take more than 6-8 hours, I can help you decipher the readings more efficiently.





