---
title: "USGS Stream Discharge Report"
author: "Marc Los Huertos"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
    highlight: tango
geometry: margin=1in
fontsize: 11pt
---

```{r setup, include=FALSE}
# Global chunk options applied to every code chunk in this document.
# echo=TRUE   : show code in the PDF
# message/warning=FALSE : suppress package startup noise
# fig settings: consistent figure sizing throughout
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 7,
  fig.height = 4,
  fig.align  = "center",
  out.width  = "100%"
)
```

---

# Overview

This document provides a fully reproducible workflow for downloading,
cleaning, summarising, and visualising **daily mean stream discharge**
from the USGS National Water Information System (NWIS).

All data are fetched directly from the USGS API using the
`dataRetrieval` package --- no manual downloads are needed. An
alternative path using a pre-downloaded CSV (e.g. from
<https://waterdata.usgs.gov/nwis>) is also provided.

> **To reuse for a different site:** change only `site_no`, `start_dt`,
> and `end_dt` in **Section 3 – Settings**, then click
> **Knit → Knit to PDF** (or press `Ctrl+Shift+K`) in RStudio.

---

# Required Packages

The four packages below cover all data access, manipulation, and
plotting needs. `dataRetrieval` is the official USGS R package and
handles authentication and API formatting automatically. Uncomment the
`install.packages` line the first time you run this document.

```{r packages}
# install.packages(c("dataRetrieval", "tidyverse", "lubridate",
#                    "scales", "knitr"))

library(dataRetrieval)  # USGS NWIS data retrieval
library(tidyverse)      # data wrangling (dplyr, tidyr) + ggplot2
library(lubridate)      # date/time parsing and arithmetic
library(scales)         # comma/log axis formatting in ggplot2
library(knitr)          # kable() for formatted tables
```

---

# Settings

Centralising all user-defined parameters in one chunk makes the
document easy to adapt. Change these four values and re-knit to
generate a complete report for any USGS gauging station.

| Variable    | Meaning                                        |
|-------------|------------------------------------------------|
| `site_no`   | 8-digit USGS site number                       |
| `param_cd`  | Parameter code (`00060` = discharge in cfs)    |
| `stat_cd`   | Statistic code (`00003` = daily mean)          |
| `start/end` | Period of record (YYYY-MM-DD)                  |

```{r settings}
site_no  <- "01646500"   # Potomac River at Little Falls, MD
param_cd <- "00060"      # 00060 = discharge (cfs); 00065 = gage height (ft)
stat_cd  <- "00003"      # 00003 = daily mean
start_dt <- "2022-01-01"
end_dt   <- "2023-12-31"
```

---

# Data Acquisition

## Option A: Download Directly in R *(recommended)*

`readNWISdv()` queries the NWIS daily-values service and returns a
tidy data frame. `renameNWISColumns()` converts the cryptic
auto-generated column name (e.g. `X_00060_00003`) to the human-readable
`Flow`. `readNWISsite()` fetches station metadata (name, coordinates,
drainage area) used for axis labels later.

```{r download}
raw <- readNWISdv(
  siteNumbers = site_no,
  parameterCd = param_cd,
  startDate   = start_dt,
  endDate     = end_dt
)

raw  <- renameNWISColumns(raw)   # -> adds column "Flow"
meta <- readNWISsite(site_no)    # station metadata

glimpse(raw)
```

## Option B: Load a Pre-Downloaded CSV

If you exported a CSV from <https://waterdata.usgs.gov/nwis>, use this
block instead. The expected column headers from an NWIS download are:

```
x, y, id, time_series_id, monitoring_location_id, parameter_code,
statistic_id, time, value, unit_of_measure, approval_status,
qualifier, last_modified
```

The block is set to `eval=FALSE` so it does not run by default.
To activate it: set `eval=TRUE` here and set `eval=FALSE` on the
Option A chunk above.

```{r load_csv, eval=FALSE}
# raw_csv <- read_csv("your_file.csv") %>%
#   mutate(time = ymd_hms(time)) %>%
#   filter(parameter_code == "00060") %>%   # discharge only
#   rename(Date = time, Flow = value) %>%
#   filter(approval_status == "A")          # Approved records only
#
# raw <- raw_csv %>% select(Date, Flow)
```

---

# Data Cleaning

This chunk selects only the columns needed downstream (`Date`, `Flow`),
removes any `NA` discharge values (caused by equipment malfunctions or
data gaps), and derives three helper columns used in the plots:

- `year`  — calendar year for annual grouping  
- `month` — abbreviated month label (factor, ordered Jan–Dec)  
- `doy`   — day of year (1–366), useful for seasonal analysis  

Two character labels (`site_name`, `units_lab`) are also created here
so they propagate consistently to every plot title and axis.

```{r clean}
df <- raw %>%
  select(Date, Flow) %>%
  filter(!is.na(Flow)) %>%
  mutate(
    year  = year(Date),
    month = month(Date, label = TRUE),
    doy   = yday(Date)
  )

site_name <- meta$station_nm
units_lab  <- "Discharge (cfs)"

cat("Records loaded :", nrow(df), "\n")
cat("Date range     :", as.character(min(df$Date)),
    "to", as.character(max(df$Date)), "\n")
cat("Site           :", site_name, "\n")
```

---

# Summary Statistics

Annual summary statistics are computed with `dplyr::summarise()` and
formatted as a publication-ready table with `knitr::kable()`. The
`booktabs = TRUE` option uses clean horizontal rules instead of the
default grid lines, which is standard in academic papers and reports.

```{r summary_table}
stats <- df %>%
  group_by(year) %>%
  summarise(
    N      = n(),
    Mean   = round(mean(Flow),   1),
    Median = round(median(Flow), 1),
    SD     = round(sd(Flow),     1),
    Min    = round(min(Flow),    1),
    Max    = round(max(Flow),    1),
    .groups = "drop"
  )

kable(
  stats,
  booktabs  = TRUE,
  caption   = paste("Annual discharge summary \u2014", site_name),
  col.names = c("Year", "N", "Mean", "Median", "SD", "Min", "Max"),
  format.args = list(big.mark = ",")
)
```

---

# Plots

## Daily Time Series

A linear y-axis time series is the most intuitive first look at the
record. The `date_breaks` and `date_labels` arguments in
`scale_x_date()` control the x-axis tick spacing and format.
`scales::comma` prevents R from using scientific notation on the y-axis
for large discharge values.

```{r plot_ts, fig.cap="Daily mean discharge over the full period of record."}
ggplot(df, aes(x = Date, y = Flow)) +
  geom_line(color = "steelblue", linewidth = 0.5) +
  scale_x_date(date_labels = "%b %Y", date_breaks = "3 months") +
  scale_y_continuous(labels = comma) +
  labs(
    title    = paste("Daily Mean Discharge \u2014", site_name),
    subtitle = paste("USGS site", site_no),
    x        = NULL,
    y        = units_lab
  ) +
  theme_bw(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Log-Scale Time Series

Discharge records are highly right-skewed: a few flood peaks can be
orders of magnitude larger than typical low flows. A log~10~ y-axis
compresses the peaks and expands the low-flow period, making both
visible simultaneously. `annotation_logticks()` adds the small tick
marks characteristic of log axes.

```{r plot_log, fig.cap="Daily mean discharge on a log-10 y-axis."}
ggplot(df, aes(x = Date, y = Flow)) +
  geom_line(color = "steelblue", linewidth = 0.5) +
  scale_x_date(date_labels = "%b %Y", date_breaks = "3 months") +
  scale_y_log10(labels = comma) +
  annotation_logticks(sides = "l") +
  labs(
    title    = paste("Daily Mean Discharge (log scale) \u2014", site_name),
    subtitle = paste("USGS site", site_no),
    x        = NULL,
    y        = paste0("log\u2081\u2080 ", units_lab)
  ) +
  theme_bw(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Monthly Box Plots

Box plots grouped by calendar month reveal the seasonal pattern of
discharge. Pooling all years together shows the typical low-flow and
high-flow seasons regardless of year-to-year variability. The log y-axis
is retained here because discharge distributions within months are also
skewed. `scale_fill_brewer(palette = "Paired")` assigns a distinct but
harmonious colour to each month.

```{r plot_box, fig.cap="Distribution of daily discharge by calendar month (all years combined)."}
ggplot(df, aes(x = month, y = Flow, fill = month)) +
  geom_boxplot(outlier.size  = 0.8,
               outlier.alpha = 0.4,
               show.legend   = FALSE) +
  scale_y_log10(labels = comma) +
  scale_fill_brewer(palette = "Paired") +
  annotation_logticks(sides = "l") +
  labs(
    title = paste("Monthly Discharge Distribution \u2014", site_name),
    x     = NULL,
    y     = paste0("log\u2081\u2080 ", units_lab)
  ) +
  theme_bw(base_size = 11)
```

## Flow Duration Curve

The **flow duration curve (FDC)** is a fundamental tool in hydrology.
It shows the percentage of time that streamflow equalled or exceeded a
given value over the period of record. The x-axis represents exceedance
probability (100 % = always exceeded; 0 % = never exceeded).
Low-gradient FDCs indicate flashy, rainfall-driven streams; steep FDCs
suggest groundwater-sustained baseflow regimes.

```{r plot_fdc, fig.cap="Flow duration curve for the full period of record."}
fdc <- df %>%
  arrange(desc(Flow)) %>%
  mutate(exceed_pct = 100 * row_number() / n())

ggplot(fdc, aes(x = exceed_pct, y = Flow)) +
  geom_line(color = "steelblue", linewidth = 0.7) +
  scale_y_log10(labels = comma) +
  scale_x_continuous(
    breaks = seq(0, 100, 10),
    labels = paste0(seq(0, 100, 10), "%")
  ) +
  annotation_logticks(sides = "l") +
  labs(
    title = paste("Flow Duration Curve \u2014", site_name),
    x     = "Percent of time flow equalled or exceeded",
    y     = paste0("log\u2081\u2080 ", units_lab)
  ) +
  theme_bw(base_size = 11)
```

---

# Session Info

Recording session info at the end of every report ensures
reproducibility: anyone re-running this document can see exactly which
package versions were used to generate the results.

```{r session}
sessionInfo()
```
