---
title: "Sentiment Analysis of Environmental News"
author: "Marc Los Huertos"
date: "EA030 · Compiled `r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    self_contained: true
    theme: null
    highlight: null
    toc: false
knit: (function(inputFile, encoding) {
    rmarkdown::render(
      inputFile,
      encoding = encoding,
      output_file = file.path(dirname(inputFile),
                              "../docs/Activities/A29_Sentiment_Analysis.html")
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 7,
  fig.height = 4.5,
  comment = "#>"
)

# ── Resolve namespace conflicts before loading packages ────────────────────
library(conflicted)
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)

# ── Core packages ──────────────────────────────────────────────────────────
library(tidyverse)
library(quanteda)
library(quanteda.textplots)
library(quanteda.textstats)
library(tidytext)
library(sentimentr)
library(knitr)
library(kableExtra)
library(lubridate)
library(scales)
```

```{css, echo=FALSE}
@import url('https://fonts.googleapis.com/css2?family=Libre+Baskerville:ital,wght@0,400;0,700;1,400&family=Source+Code+Pro:wght@400;600&family=Lato:wght@300;400;700&display=swap');

*, *::before, *::after { box-sizing: border-box; margin: 0; padding: 0; }

:root {
  --forest:    #1e3a2f;
  --moss:      #2d5a3d;
  --fern:      #4a7c5e;
  --sage:      #8aad96;
  --mist:      #d4e6da;
  --cream:     #f7f3ed;
  --parchment: #ede8df;
  --ink:       #1a1a18;
  --charcoal:  #3a3a38;
  --dust:      #7a7a72;
  --amber:     #c8832a;
  --amber-lt:  #e8a84a;
  --white:     #ffffff;
  --red-dark:  #7a1e1e;
  --card-shadow: 0 2px 10px rgba(30,58,47,.09);
}

html { scroll-behavior: smooth; }

body {
  font-family: 'Lato', sans-serif;
  background: var(--cream);
  color: var(--ink);
  font-size: 17px;
  line-height: 1.75;
  padding-left:  clamp(14px, 4vw, 48px);
  padding-right: clamp(14px, 4vw, 48px);
  padding-bottom: 80px;
}

.main-container {
  max-width: 920px !important;
  margin: 0 auto !important;
  padding: 0 clamp(10px, 2vw, 32px) 80px !important;
}

.page-header {
  margin-left:  calc(-1 * clamp(14px, 4vw, 48px));
  margin-right: calc(-1 * clamp(14px, 4vw, 48px));
}

h1.title, h4.author, h4.date { display: none; }

/* ── Page header ─────────────────────────────────────────────────────────── */
.page-header {
  background-image:
    repeating-linear-gradient(120deg, transparent, transparent 60px, rgba(255,255,255,.018) 60px, rgba(255,255,255,.018) 61px),
    linear-gradient(160deg, #1e3a2f 0%, #2d5a3d 60%, #1a3d30 100%);
  color: #fff; padding: 44px 4vw 36px; margin: 0 -4vw 48px;
  border-bottom: 4px solid var(--amber); position: relative; overflow: hidden;
}
.page-header::after {
  content: ''; position: absolute; right: -40px; top: -40px;
  width: 260px; height: 260px; border-radius: 50%;
  background: radial-gradient(circle, rgba(138,173,150,.12) 0%, transparent 70%);
  pointer-events: none;
}
.page-header .module-tag {
  display: inline-block; font-family: 'Source Code Pro', monospace;
  font-size: .68rem; font-weight: 600; letter-spacing: .18em; text-transform: uppercase;
  color: var(--amber-lt); border: 1px solid rgba(200,131,42,.45);
  padding: 3px 10px; border-radius: 2px; margin-bottom: 12px;
}
.page-header h1 {
  font-family: 'Libre Baskerville', serif;
  font-size: clamp(1.6rem, 3vw, 2.3rem);
  font-weight: 700; color: #fff; line-height: 1.2; margin-bottom: 8px;
}
.page-header h1 span { color: var(--sage); }
.page-header .sub { font-size: .9rem; color: rgba(255,255,255,.65); font-weight: 300; }
.page-header .sub strong { color: rgba(255,255,255,.9); font-weight: 700; }
.back-link {
  display: inline-block; margin-top: 16px;
  font-family: 'Source Code Pro', monospace; font-size: .7rem; letter-spacing: .08em;
  color: var(--amber-lt); text-decoration: none;
  border: 1px solid rgba(200,131,42,.4); padding: 4px 11px; border-radius: 3px;
  transition: background .2s;
}
.back-link:hover { background: rgba(200,131,42,.18); }
.header-links {
  display: flex; gap: 10px; flex-wrap: wrap; margin-top: 16px;
}
.pdf-btn {
  font-family: 'Source Code Pro', monospace; font-size: .7rem; letter-spacing: .08em;
  color: var(--amber-lt); background: transparent; border: 1px solid rgba(200,131,42,.4);
  padding: 4px 11px; border-radius: 3px; cursor: pointer; transition: background .2s;
}
.pdf-btn:hover { background: rgba(200,131,42,.18); }

/* ── Section nav ─────────────────────────────────────────────────────────── */
.section-nav {
  display: flex; gap: 8px; flex-wrap: wrap; margin-bottom: 44px;
  padding-bottom: 20px; border-bottom: 1px solid var(--parchment);
}
.section-nav a {
  font-family: 'Source Code Pro', monospace; font-size: .7rem; font-weight: 600;
  letter-spacing: .08em; text-transform: uppercase;
  color: var(--fern); text-decoration: none;
  border: 1px solid var(--mist); padding: 4px 12px; border-radius: 3px;
  background: var(--white); transition: background .2s, color .2s;
}
.section-nav a:hover { background: var(--mist); color: var(--forest); }

/* ── Headings ────────────────────────────────────────────────────────────── */
h2 {
  font-family: 'Libre Baskerville', serif; font-size: 1.45rem; font-weight: 700;
  color: var(--forest); margin-top: 56px; margin-bottom: 6px;
  padding-bottom: 8px; border-bottom: 1px solid var(--parchment);
}
h3 {
  font-family: 'Libre Baskerville', serif; font-size: 1.1rem; font-weight: 700;
  color: var(--moss); margin-top: 28px; margin-bottom: 4px;
}
h4 {
  font-family: 'Lato', sans-serif; font-size: 1rem; font-weight: 700;
  color: var(--charcoal); margin-top: 20px; margin-bottom: 4px;
}

p { margin-bottom: 1em; }
strong { color: var(--forest); }

/* ── Code ────────────────────────────────────────────────────────────────── */
code {
  font-family: 'Source Code Pro', monospace; font-size: .84em;
  background: var(--mist); color: var(--moss); padding: 1px 5px; border-radius: 3px;
}
pre {
  background: var(--ink); color: #e8e6df; border-radius: 5px;
  padding: 14px 18px; font-size: .83rem; line-height: 1.55;
  overflow-x: auto; border-left: 4px solid var(--amber); margin: 1.2em 0;
}
pre code { background: transparent; color: inherit; padding: 0; font-size: 1em; }

/* ── Callouts ────────────────────────────────────────────────────────────── */
.callout {
  border-left: 5px solid; padding: .65em 1.1em;
  margin: 1.4em 0; border-radius: 0 5px 5px 0; font-size: .92em; line-height: 1.6;
}
.callout > strong:first-child { display: block; margin-bottom: .3em; }
.callout-concept { border-color: var(--fern);  background: #f0f7f2; }
.callout-tip     { border-color: #2e7d32;      background: #f1f8f1; }
.callout-warn    { border-color: #e65100;      background: #fff8f4; }
.callout-myth    { border-color: var(--red-dark); background: #fdf2f2; }
.callout-formula { border-color: var(--amber); background: #fdf8f0;
                   font-family: 'Source Code Pro', monospace; font-size: .88em; }
.callout-try     { border-color: var(--amber); background: #fffaf0; }

/* ── Learning outcomes ───────────────────────────────────────────────────── */
.outcomes-box {
  background: var(--forest); color: rgba(255,255,255,.88); border-radius: 6px;
  padding: 20px 24px; margin: 0 0 2.2em; font-size: .9rem;
}
.outcomes-box h4 {
  font-family: 'Libre Baskerville', serif; color: var(--sage);
  font-size: 1rem; margin-bottom: 10px;
}
.outcomes-box ol { padding-left: 1.4em; }
.outcomes-box li { margin-bottom: .35em; }
.outcomes-box code { background: rgba(255,255,255,.12); color: var(--amber-lt); }

/* ── Key concept summary boxes ───────────────────────────────────────────── */
.key-box {
  background: var(--forest); color: #fff; border-radius: 6px;
  padding: 20px 24px; margin: 2em 0; font-size: .9rem;
}
.key-box h4 {
  font-family: 'Libre Baskerville', serif; color: var(--sage);
  font-size: 1rem; margin-bottom: 10px;
}
.key-box ul { padding-left: 1.3em; }
.key-box li { margin-bottom: 5px; color: rgba(255,255,255,.85); }
.key-box code { background: rgba(255,255,255,.12); color: var(--amber-lt); }

/* ── Comparison / method table ───────────────────────────────────────────── */
table { font-size: .9rem; border-collapse: collapse; width: 100%; margin: 1.2em 0; }
thead th {
  background: var(--forest); color: #fff;
  font-family: 'Source Code Pro', monospace; font-size: .76rem;
  letter-spacing: .06em; text-transform: uppercase; padding: 8px 12px; text-align: left;
}
tbody tr:nth-child(even) { background: var(--mist); }
tbody td { padding: 7px 12px; border-bottom: 1px solid var(--parchment); }

/* ── Pipeline graphic ────────────────────────────────────────────────────── */
.pipeline-row {
  display: flex; gap: 10px; flex-wrap: wrap; margin: 1.6em 0;
  align-items: stretch;
}
.pipeline-step {
  flex: 1; min-width: 130px; padding: 14px 16px; border-radius: 6px;
  box-shadow: var(--card-shadow); font-size: .85rem; text-align: center;
  position: relative;
}
.pipeline-step .step-num {
  font-family: 'Source Code Pro', monospace; font-size: .66rem;
  font-weight: 700; letter-spacing: .14em; text-transform: uppercase;
  margin-bottom: 5px;
}
.pipeline-step .step-label {
  font-family: 'Libre Baskerville', serif; font-size: .9rem;
  font-weight: 700; line-height: 1.3;
}
.ps-ingest  { background: var(--parchment); }
.ps-ingest .step-num  { color: var(--dust); }
.ps-ingest .step-label { color: var(--charcoal); }
.ps-lexicon { background: #f0f7f2; }
.ps-lexicon .step-num  { color: var(--fern); }
.ps-lexicon .step-label { color: var(--forest); }
.ps-transform { background: var(--forest); }
.ps-transform .step-num  { color: var(--sage); }
.ps-transform .step-label { color: #fff; }
.ps-validate { background: #fdf8f0; }
.ps-validate .step-num  { color: var(--amber); }
.ps-validate .step-label { color: var(--charcoal); }
.ps-timeseries { background: var(--ink); }
.ps-timeseries .step-num  { color: var(--amber-lt); }
.ps-timeseries .step-label { color: #fff; }

/* ── Footer ──────────────────────────────────────────────────────────────── */
.page-footer {
  background: var(--ink); color: rgba(255,255,255,.42); text-align: center;
  padding: 24px 4vw; margin: 60px -4vw 0;
  font-size: .76rem; font-family: 'Source Code Pro', monospace; letter-spacing: .04em;
}
.page-footer a { color: var(--sage); text-decoration: none; }
.page-footer .divider { margin: 0 8px; opacity: .3; }
```

<div class="page-header">
<div class="module-tag">EA030 &mdash; Activity A29</div>
<h1>Sentiment Analysis of <span>Environmental News</span></h1>
<p class="sub"><strong>Prof. Marc Los Huertos</strong> &nbsp;&bull;&nbsp; Pomona College &nbsp;&bull;&nbsp; EA030 Environmental Analysis</p>
<div class="header-links">
<a class="back-link" href="../activities_index.html">&#8592; Back to Activities</a>
<button class="pdf-btn" onclick="window.print()">&#8595; Save as PDF</button>
</div>
</div>

<div class="section-nav">
<a href="#why">Why Sentiment?</a>
<a href="#methods">Methods Landscape</a>
<a href="#corpus">Building a Corpus</a>
<a href="#lexicon">Lexicon Scoring</a>
<a href="#aspect">Aspect-Based</a>
<a href="#transformers">Transformers</a>
<a href="#timeseries">Time Series</a>
<a href="#validation">Validation</a>
<a href="#pipeline">Full Pipeline</a>
<a href="#extensions">Extensions</a>
</div>

<div class="outcomes-box">
<h4>&#127919; Learning Outcomes</h4>
<ol>
<li>Explain what sentiment analysis measures and why it matters for environmental research.</li>
<li>Build a text corpus from news articles using <code>quanteda</code> and <code>readtext</code>.</li>
<li>Apply dictionary-based sentiment scoring with LSD2015, AFINN, and NRC lexicons.</li>
<li>Handle valence shifters (negation, amplifiers) with <code>sentimentr</code>.</li>
<li>Perform aspect-based sentiment analysis toward specific entities or topics.</li>
<li>Call pre-trained transformer models (ClimateBERT) from R using the <code>text</code> package.</li>
<li>Aggregate document-level sentiment into time-series indices with <code>sentometrics</code>.</li>
<li>Validate sentiment models against hand-coded labels using Cohen's kappa and macro-F1.</li>
<li>Recognize the limitations of each approach and choose the right tool for a given research question.</li>
</ol>
</div>

> **What this module covers.** How do news outlets portray environmental issues 
> over time? Is coverage of climate policy becoming more or less alarmist? Do 
> articles about renewable energy carry a different tone than articles about 
> fossil fuels? These are **sentiment analysis** questions, and answering them 
> requires converting unstructured text into structured quantitative measures of 
> tone, emotion, or polarity. This tutorial walks the complete R workflow, from 
> raw text through validated time-series indices, using methods appropriate for 
> environmental and climate-related news.

<hr>

## 1. Why Sentiment Analysis for Environmental News? {#why}

Environmental scientists increasingly study not just physical systems but the 
**human dimensions** of environmental change: public perception, media framing, 
policy discourse, and stakeholder attitudes. News coverage shapes public opinion 
and policy priorities, so measuring its tone systematically is a legitimate 
research question.

Sentiment analysis converts text into numbers that represent **polarity** 
(positive/negative/neutral), **emotion** (anger, fear, joy, trust), or 
**stance** (opportunity vs. risk). Applied to a corpus of news articles, it 
lets you ask quantitative questions about qualitative data.

### Three motivating questions

These examples illustrate the kinds of research questions sentiment analysis 
can address in environmental science:

**1. Tone over time.** Has US newspaper coverage of wildfire become more 
negative over the past two decades? If so, does the shift correlate with 
increases in burned acreage, or does media framing lead or lag the physical 
trend?

**2. Entity-level framing.** Within a single article about the Inflation 
Reduction Act, is the tone toward the EPA positive while the tone toward the 
fossil-fuel industry is negative? A single "article-level" sentiment score 
would miss this entirely.

**3. Cross-outlet comparison.** Do articles from a regional paper in a 
coal-dependent community frame energy transition differently than articles 
from a national outlet? Comparing sentiment distributions across sources 
reveals structural differences in coverage.

<div class="callout callout-concept">
<strong>&#128161; Sentiment is a noisy signal, not ground truth</strong>
A sentiment score is a simplified, often imperfect measure of a complex 
property. News articles are written in a deliberately neutral register. 
Reporters attribute opinions to sources ("Critics called it a failure") 
without endorsing them. Lexicon-based methods cannot distinguish the 
reporter's voice from a quoted critic's voice. Keep this in mind throughout 
the tutorial: every sentiment number you produce is a proxy, and the quality 
of that proxy must be validated.
</div>

### What sentiment analysis is not

Sentiment analysis does **not** tell you what a reader thinks, feels, or 
believes. It tells you something about the language in the text itself, 
as measured by a particular dictionary or model. Conflating text-level 
polarity with audience reception is a common analytical error.

It also does not handle **framing** well. Two articles can have identical 
sentiment scores while framing the same issue in completely different ways 
(economic growth vs. ecological collapse). Frame analysis requires different 
tools, though sentiment can serve as one input.

<hr>

## 2. The Methods Landscape {#methods}

Sentiment analysis methods fall into three broad families. Each trades off 
transparency, accuracy, and computational cost differently.

```{r methods-table, echo=FALSE}
methods_df <- data.frame(
  Family = c("Dictionary / Lexicon", "Traditional ML", "Transformers & LLMs"),
  Approach = c(
    "Score words against a pre-built list of polarity values. Sum or average.",
    "Train a classifier (logistic regression, SVM, random forest) on TF-IDF features with labeled data.",
    "Fine-tune a pre-trained language model (BERT, RoBERTa, ClimateBERT) on labeled examples."
  ),
  Accuracy = c(
    "Low to moderate (27-65% on multi-class news)",
    "Moderate (58-71% on multi-class news)",
    "High (73-84% on multi-class news)"
  ),
  GPU = c("No", "No", "Recommended"),
  Labels = c("None required", "1,000-10,000", "500-5,000"),
  R_Package = c(
    "quanteda, tidytext, sentimentr, sentometrics",
    "tidymodels, caret, text2vec",
    "text (HuggingFace bridge)"
  )
)

kable(methods_df, col.names = c("Family", "Approach", "Typical Accuracy", 
                                 "GPU?", "Labeled Data", "R Packages"),
      caption = "Three families of sentiment analysis methods, compared on attributes relevant to environmental news projects.") %>%
  kable_styling(full_width = TRUE, font_size = 14) %>%
  column_spec(1, bold = TRUE, width = "12em")
```

The accuracy numbers in the table come from a 2024 study (Wang et al., 
*Soft Computing*) that compared all three families on 9,382 news articles 
about transboundary rivers, scored on a 5-class scale. AFINN scored 27.4% 
accuracy, the best traditional ML models clustered around 66-71%, and 
BERT reached 72.7%. On a simpler 3-class task, domain-adapted transformers 
like ClimateBERT report weighted F1 scores around 0.84.

<div class="callout callout-warn">
<strong>&#9888;&#65039; No public benchmark exists for environmental news specifically</strong>
ClimateBERT's 0.84 F1 was measured on paragraphs from corporate climate 
disclosures, not news articles. The Wang et al. numbers are the closest 
rigorous head-to-head on actual news. When you report results from your 
own analysis, be explicit about what corpus and what labels your accuracy 
is measured against.
</div>

### When to use which method

The decision depends on your sample size, research question, and 
computational resources:

| If... | Use this |
|---|---|
| Corpus < 1,000 articles, exploratory | `tidytext` + AFINN / Bing / NRC |
| Need defensible news-specific baseline | `quanteda` + LSD2015 + `sentimentr` |
| Need time-series indices for regression | `sentometrics` |
| Need entity-level sentiment, no GPU | `quanteda` targeted dictionary with +/- 10-word window |
| Need accuracy > ~0.70 macro-F1 | `text` package calling ClimateBERT or cardiffnlp RoBERTa |
| Ambiguous edge cases, sarcasm | LLM (Claude, GPT-4o) zero-shot or few-shot via API |

<div class="pipeline-row">
<div class="pipeline-step ps-ingest">
<div class="step-num">Stage 0</div>
<div class="step-label">Corpus Assembly</div>
</div>
<div class="pipeline-step ps-lexicon">
<div class="step-num">Stage 1</div>
<div class="step-label">Lexicon Baseline</div>
</div>
<div class="pipeline-step ps-transform">
<div class="step-num">Stage 2</div>
<div class="step-label">Transformer Layer</div>
</div>
<div class="pipeline-step ps-validate">
<div class="step-num">Stage 3</div>
<div class="step-label">Validation</div>
</div>
<div class="pipeline-step ps-timeseries">
<div class="step-num">Stage 4</div>
<div class="step-label">Time Series</div>
</div>
</div>

This tutorial follows the pipeline above. We start with the simplest methods 
(lexicons) and add complexity only when needed.

<hr>

## 3. Building a Corpus {#corpus}

Before scoring sentiment, you need a structured collection of texts with 
metadata. In `quanteda`, this is called a **corpus**.

### Simulated environmental news corpus

For this tutorial, we create a small simulated corpus of environmental news 
headlines and lead paragraphs. In a real project, you would import text from 
files, a database, or an API.

```{r corpus-create}
# ── Simulated environmental news articles ──────────────────────────────
articles <- tibble(
  doc_id = paste0("art_", sprintf("%03d", 1:20)),
  source = rep(c("National Tribune", "Valley Gazette",
                 "Coastal Herald", "Mountain Post"), each = 5),
  date   = seq(as.Date("2024-01-15"), by = "month", length.out = 20),
  topic  = rep(c("wildfire", "water", "energy", "biodiversity", "climate"),
               times = 4),
  text   = c(
    # National Tribune
    "Devastating wildfire season scorches 2.3 million acres across the West, forcing thousands of evacuations and overwhelming firefighting resources.",
    "New EPA water quality standards will impose stricter limits on agricultural runoff, drawing praise from conservation groups and criticism from farm lobbies.",
    "Solar installations surged 42% in 2023, marking a record year for clean energy deployment and creating thousands of new jobs in underserved communities.",
    "Alarming decline in migratory bird populations raises concerns about habitat loss and pesticide exposure across major flyways.",
    "Global temperatures shattered records for the twelfth consecutive month, intensifying calls for accelerated emissions reductions at the upcoming climate summit.",
    # Valley Gazette
    "Local ranchers struggle to balance wildfire risk with grazing traditions as prescribed burn programs expand into private lands.",
    "Community groups celebrate the restoration of riparian habitat along Bear Creek after years of advocacy and volunteer effort.",
    "The proposed wind farm faces opposition from residents concerned about noise, visual impact, and declining property values in the rural township.",
    "Native plant restoration project transforms abandoned lot into thriving pollinator garden, attracting dozens of butterfly species.",
    "Farmers report shorter growing seasons and unpredictable frost dates, attributing the changes to shifting climate patterns.",
    # Coastal Herald
    "After years of devastating fire seasons, a coalition of agencies unveils a bold new strategy for community wildfire preparedness and resilient landscapes.",
    "Toxic algal bloom shuts down water intake for coastal city, exposing vulnerabilities in aging infrastructure and nutrient management.",
    "Offshore wind project receives final permits after environmental review finds minimal impact on marine mammals and seabirds.",
    "Kelp forest recovery surprises scientists as urchin removal program shows rapid regrowth of canopy and return of marine biodiversity.",
    "Sea level projections released today paint a stark picture for low-lying neighborhoods, with managed retreat now part of the planning conversation.",
    # Mountain Post
    "Controlled burns reduce fuel loads but spark air quality warnings in mountain communities already burdened by summer smoke.",
    "Drought emergency declared as reservoir levels drop below 30% capacity, triggering mandatory conservation measures for the first time in a decade.",
    "Utility announces early retirement of its last coal plant, pledging to replace capacity with battery storage and community solar.",
    "Endangered species listing for mountain caribou prompts heated debate between conservation advocates and logging industry representatives.",
    "Glaciers in the range have lost 40% of their volume since 1980, according to a university survey released this week."
  )
)

# Build a quanteda corpus with document-level variables
corp <- corpus(articles, text_field = "text",
               docid_field = "doc_id")
```

```{r corpus-inspect}
# Quick look at the corpus
summary(corp, n = 5)
```

<div class="callout callout-concept">
<strong>&#128161; Document variables carry your metadata through the pipeline</strong>
The <code>source</code>, <code>date</code>, and <code>topic</code> columns you 
include at corpus creation travel with every token, dfm, and sentiment score 
downstream. This is what allows you to compare sentiment by newspaper, plot it 
over time, or break it out by topic later. Always attach metadata at the corpus 
stage, not after scoring.
</div>

### Tokenization and document-feature matrix

Most `quanteda` operations require tokenizing the corpus (splitting text into 
words) and converting it to a **document-feature matrix** (dfm), where rows 
are documents and columns are unique word types.

```{r tokenize}
# Tokenize: lowercase, remove punctuation and stopwords
toks <- tokens(corp,
               remove_punct  = TRUE,
               remove_numbers = TRUE) %>%
  tokens_tolower() %>%
  tokens_remove(stopwords("en"))

# Document-feature matrix
dfm_articles <- dfm(toks)
dfm_articles
```

```{r wordcloud, fig.cap="Word cloud of the simulated corpus. Larger words appear more frequently across all 20 articles. Environmental terms (wildfire, water, community, climate) dominate.", fig.height=5}
set.seed(42)
textplot_wordcloud(dfm_articles, max_words = 80,
                   color = c("#1e3a2f", "#4a7c5e", "#8aad96", 
                             "#c8832a", "#7a3a2f"))
```

<hr>

## 4. Dictionary-Based Sentiment Scoring {#lexicon}

The simplest approach to sentiment analysis is to look up each word in a 
**sentiment dictionary** (also called a lexicon) that assigns a polarity or 
emotion score to known words. The document's sentiment is the sum or average 
of its word scores.

### Available dictionaries in R

Several dictionaries are available through different R packages. Each was 
built for a different purpose.

```{r lexicon-table, echo=FALSE}
lex_df <- data.frame(
  Dictionary = c("LSD2015", "AFINN", "Bing", "NRC", "Loughran-McDonald", "VADER"),
  Words = c("4,567 + negation patterns", "2,477", "6,786", "~14,000", "~4,200", "~7,500"),
  Scale = c("Positive / negative", "-5 to +5", "Binary +/-", "8 emotions + +/-", "6 categories", "Compound [-1, +1]"),
  `Best For` = c("News and political text", "Quick exploratory scoring", "Simple positive/negative counts", "Emotion decomposition", "Financial / regulatory text", "Social media and headlines"),
  Package = c("quanteda (built-in)", "tidytext", "tidytext", "tidytext, lexicon", "lexicon, quanteda.sentiment", "sentimentr (VADER-like)")
)

kable(lex_df, col.names = c("Dictionary", "Vocabulary", "Scale", "Best For", "R Package"),
      caption = "Major sentiment dictionaries available in R.") %>%
  kable_styling(full_width = TRUE, font_size = 14) %>%
  column_spec(1, bold = TRUE)
```

<div class="callout callout-warn">
<strong>&#9888;&#65039; Different lexicons can give opposite answers</strong>
A tutorial analysis applying four standard lexicons (AFINN, Bing, Jockers, NRC) 
to the same 4,508 sentences found Fleiss' kappa of only 0.22 across the four 
methods. That is barely above chance agreement. Always run at least two 
lexicons and compare results. If they disagree, that tells you something 
important about the robustness of your findings.
</div>

### Scoring with LSD2015 (recommended for news)

The **Lexicoder Sentiment Dictionary** (LSD2015) was designed specifically for 
political and news text. It includes negation-aware word patterns 
(e.g., "not good" is captured as negative). It is built into `quanteda` as 
`data_dictionary_LSD2015`.

```{r lsd-score}
# Apply LSD2015 dictionary to the dfm
dfm_sent <- dfm_lookup(dfm_articles,
                        dictionary = data_dictionary_LSD2015)

# Convert to a tidy data frame
lsd_scores <- convert(dfm_sent, to = "data.frame") %>%
  rename(doc_id = doc_id) %>%
  mutate(
    net_sentiment = positive - negative,
    total_words   = ntoken(corp),
    sent_ratio    = net_sentiment / total_words
  ) %>%
  bind_cols(docvars(corp))

head(lsd_scores)
```

```{r lsd-plot, fig.cap="Net LSD2015 sentiment (positive minus negative word counts) by news source. Each point is one article; the dashed line at zero separates net-positive from net-negative.", fig.height=4.5}
ggplot(lsd_scores, aes(x = source, y = net_sentiment, fill = source)) +
  geom_boxplot(width = 0.45, outlier.shape = NA,
               colour = "#1e3a2f", alpha = 0.75, linewidth = 0.6) +
  geom_jitter(aes(colour = source), width = 0.15,
              size = 2, alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "#7a7a72") +
  scale_fill_manual(values = c(
    "Coastal Herald"     = "#4a7c5e",
    "Mountain Post"      = "#7a3a2f",
    "National Tribune"   = "#1e3a2f",
    "Valley Gazette"     = "#c8832a"
  )) +
  scale_colour_manual(values = c(
    "Coastal Herald"     = "#4a7c5e",
    "Mountain Post"      = "#7a3a2f",
    "National Tribune"   = "#1e3a2f",
    "Valley Gazette"     = "#c8832a"
  )) +
  labs(x = NULL, y = "Net sentiment (positive - negative words)",
       title = "LSD2015 Sentiment by News Source") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title         = element_text(family = "serif", face = "bold",
                                      colour = "#1e3a2f"),
    legend.position    = "none",
    panel.grid.major.x = element_blank()
  )
```

### Scoring with tidytext and multiple lexicons

The `tidytext` package provides access to AFINN, Bing, and NRC lexicons through 
a tidy one-word-per-row workflow. Running multiple lexicons in parallel is the 
best way to check whether your findings are robust to dictionary choice.

```{r tidytext-score}
# Tidy tokenization
tidy_articles <- articles %>%
  unnest_tokens(word, text)

# Score with AFINN
afinn_scores <- tidy_articles %>%
  inner_join(get_sentiments("afinn"), by = "word") %>%
  group_by(doc_id, source, date, topic) %>%
  summarize(
    afinn_mean = mean(value),
    afinn_sum  = sum(value),
    n_scored   = n(),
    .groups = "drop"
  )

# Score with Bing
bing_scores <- tidy_articles %>%
  inner_join(get_sentiments("bing"), by = "word") %>%
  count(doc_id, source, date, topic, sentiment) %>%
  pivot_wider(names_from = sentiment, values_from = n, values_fill = 0) %>%
  mutate(bing_net = positive - negative)

# Score with NRC (emotion categories)
nrc_emotions <- tidy_articles %>%
  inner_join(get_sentiments("nrc"), by = "word") %>%
  count(doc_id, source, topic, sentiment) %>%
  pivot_wider(names_from = sentiment, values_from = n, values_fill = 0)
```

```{r nrc-plot, fig.cap="NRC emotion profile aggregated by topic across all articles. Each bar represents the total word count matching that emotion category. Fear and negative words dominate wildfire and climate articles; trust and positive words are more common in energy and biodiversity articles.", fig.height=5}
nrc_topic <- tidy_articles %>%
  inner_join(get_sentiments("nrc"), by = "word") %>%
  filter(!sentiment %in% c("positive", "negative")) %>%
  count(topic, sentiment)

ggplot(nrc_topic, aes(x = reorder(sentiment, n), y = n, fill = topic)) +
  geom_col(position = "dodge", alpha = 0.85) +
  coord_flip() +
  scale_fill_manual(values = c(
    "biodiversity" = "#4a7c5e",
    "climate"      = "#7a3a2f",
    "energy"       = "#c8832a",
    "water"        = "#2d5a3d",
    "wildfire"     = "#8b3a2a"
  )) +
  labs(x = NULL, y = "Word count",
       title = "NRC Emotion Profiles by Topic",
       fill = "Topic") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(family = "serif", face = "bold",
                               colour = "#1e3a2f"),
    legend.position = "bottom"
  )
```

### Handling valence shifters with sentimentr

A critical weakness of bag-of-words lexicons is that they score "not good" as 
positive (the word "good" matches the dictionary). The `sentimentr` package 
handles **valence shifters** -- negators ("not"), amplifiers ("very"), 
de-amplifiers ("somewhat"), and adversative conjunctions ("but") -- at the 
sentence level.

```{r sentimentr-score}
# sentimentr works on raw text, not tokenized input
sr_scores <- sentiment_by(articles$text, by = articles$doc_id)

# Merge back with metadata
sr_df <- articles %>%
  left_join(sr_scores, by = c("doc_id" = "doc_id"))

head(sr_df %>% select(doc_id, source, topic, ave_sentiment))
```

```{r sentimentr-detail, fig.cap="Sentence-level sentiment from sentimentr for two example articles, one about wildfire (article 1) and one about solar energy (article 3). The valence-shifted scoring correctly identifies the overall negative tone of wildfire coverage and positive tone of the solar story.", fig.height=4}
# Sentence-level detail for two articles
detail <- sentiment(articles$text[c(1, 3)])
detail$article <- rep(c("art_001 (Wildfire)", "art_003 (Solar)"),
                      times = c(sum(detail$element_id == 1),
                                sum(detail$element_id == 2)))

ggplot(detail, aes(x = sentence_id, y = sentiment, fill = sentiment > 0)) +
  geom_col(alpha = 0.8, show.legend = FALSE) +
  facet_wrap(~ article, scales = "free_x") +
  scale_fill_manual(values = c("TRUE" = "#4a7c5e", "FALSE" = "#7a3a2f")) +
  labs(x = "Sentence", y = "Sentiment score",
       title = "Sentence-Level Sentiment (sentimentr)") +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(family = "serif", face = "bold",
                               colour = "#1e3a2f"),
    strip.text = element_text(face = "bold")
  )
```

<div class="callout callout-try">
<strong>&#128221; Try it: compare three lexicons</strong>
Pick one article from the corpus (e.g., article 6 about ranchers and 
prescribed burns). Score it with LSD2015, AFINN, and <code>sentimentr</code>. 
Do all three methods agree on the direction (positive or negative)? What 
about the magnitude? If they disagree, examine which words each method 
matches and identify the source of divergence. This exercise builds 
intuition for why multi-lexicon comparison matters.
</div>

<hr>

## 5. Aspect-Based Sentiment Analysis {#aspect}

A single article often discusses multiple entities, policies, or topics. 
Scoring the entire document with one number collapses this structure. 
**Aspect-based sentiment analysis (ABSA)** decomposes the text into 
(aspect, sentiment) pairs.

### Targeted dictionary lookup in quanteda

The simplest ABSA approach uses `quanteda`'s windowed token selection. 
You define your target entity, extract a window of +/- 10 tokens around 
every mention of that entity, and then apply a sentiment dictionary only 
to the text within that window.

```{r absa-quanteda}
# Define target entities
targets <- list(
  wildfire  = c("wildfire", "fire", "burn", "blaze"),
  renewable = c("solar", "wind", "clean energy", "renewable"),
  water     = c("water", "reservoir", "drought", "runoff")
)

# Function: extract sentiment around a target
target_sentiment <- function(toks, target_words, window = 10) {
  # Select tokens within +/- window of target
  toks_target <- tokens_keep(toks, pattern = target_words,
                              window = window, padding = FALSE)
  
  # Apply LSD2015
  dfm_target <- dfm(toks_target) %>%
    dfm_lookup(dictionary = data_dictionary_LSD2015)
  
  result <- convert(dfm_target, to = "data.frame") %>%
    mutate(
      net = positive - negative,
      target = paste(target_words, collapse = "/")
    )
  
  result
}

# Score each target
absa_results <- map_dfr(targets, ~ target_sentiment(toks, .x)) %>%
  bind_cols(
    docvars(corp) %>% slice(rep(1:n(), times = length(targets)))
  )
```

```{r absa-summary}
# Summarize by target
absa_summary <- absa_results %>%
  group_by(target) %>%
  summarize(
    mean_net   = mean(net, na.rm = TRUE),
    median_net = median(net, na.rm = TRUE),
    sd_net     = sd(net, na.rm = TRUE),
    n_docs     = n(),
    .groups = "drop"
  )

kable(absa_summary, digits = 2,
      col.names = c("Target Entity", "Mean Net", "Median Net", "SD", "N"),
      caption = "Aspect-level sentiment: average net LSD2015 sentiment in a +/- 10 word window around each target entity.") %>%
  kable_styling(full_width = FALSE, font_size = 14)
```

<div class="callout callout-concept">
<strong>&#128161; Why +/- 10 tokens?</strong>
The 10-token window is a pragmatic compromise. Smaller windows (5 tokens) 
risk missing relevant modifiers; larger windows (20+ tokens) bleed into 
adjacent topics. The <code>quanteda</code> documentation and tutorials 
use 10 as the default for targeted dictionary lookup on news articles. 
You can and should test sensitivity to this choice.
</div>

<hr>

## 6. Transformer Models from R {#transformers}

When dictionary methods are not accurate enough, **transformer models** 
offer a substantial improvement. These are deep-learning models pre-trained 
on large text corpora and then fine-tuned for sentiment classification.

### ClimateBERT: a domain-adapted model

**ClimateBERT** is a DistilRoBERTa model further pre-trained on over 2 million 
climate-related paragraphs from news, research articles, and corporate reports. 
The sentiment variant (`climatebert/distilroberta-base-climate-sentiment`) 
classifies text into three categories: **opportunity**, **neutral**, and 
**risk**. On its benchmark dataset, it reports weighted F1 of approximately 0.84.

### Calling transformers from R with the text package

The `text` package provides an R bridge to HuggingFace models. It requires 
a one-time Python environment setup (handled by `textrpp_install()`).

```{r text-setup, eval=FALSE}
# ── One-time setup (run once, not every session) ───────────────────────
# install.packages("text")
library(text)
textrpp_install()        # sets up a conda environment with PyTorch
textrpp_initialize()     # connects R to the Python backend
```

```{r text-classify, eval=FALSE}
# ── Classify articles with ClimateBERT ─────────────────────────────────
library(text)

# Use the climate-sentiment model
climate_sent <- textClassify(
  articles$text,
  model = "climatebert/distilroberta-base-climate-sentiment",
  device = "cpu",             # use "cuda" if GPU available
  return_incorrect_results = FALSE
)

# Result: a data frame with columns for each class probability
# and a predicted label
head(climate_sent)
```

<div class="callout callout-warn">
<strong>&#9888;&#65039; Transformers truncate long text</strong>
Most transformer models have a maximum input length of 512 tokens 
(roughly 400 words). The average news article is 600 to 1,000 words, 
so the model silently ignores the second half of many articles. The 
standard workaround is to split each article into paragraphs, classify 
each paragraph separately, and then aggregate (e.g., majority vote or 
mean probability). Always document whether you are scoring full articles 
or chunks, and how you aggregate.
</div>

### Other models worth knowing

| Model | Domain | Classes | Notes |
|---|---|---|---|
| `cardiffnlp/twitter-roberta-base-sentiment-latest` | General (trained on tweets) | 3: neg/neu/pos | Strong out-of-the-box for English headlines |
| `ProsusAI/finbert` | Financial news | 3: neg/neu/pos | 97% on Financial PhraseBank (full agreement subset) |
| `climatebert/distilroberta-base-climate-detector` | Climate | 2: climate/not | Useful as a filter before sentiment scoring |
| `yangheng/deberta-v3-base-absa-v1.1` | Aspect-based | Variable | State-of-the-art open ABSA model |

### When transformers are not worth the cost

If your hand-coded validation shows that your lexicon approach achieves 
macro-F1 above 0.65 and human inter-annotator agreement (Cohen's kappa) 
is below 0.55, adding a transformer layer will not produce meaningful 
gains. The annotation noise itself becomes the ceiling. Invest the time 
in better annotation guidelines instead.

<hr>

## 7. Time-Series Aggregation {#timeseries}

Individual document scores are rarely the final product. Most research 
questions require aggregating sentiment into **time-series indices** that 
can be plotted, correlated with external variables, or modeled.

### The three aggregation dimensions

Building a sentiment time series involves three sequential choices:

**1. Within-document:** How do you combine sentence-level scores into one 
document score? Options include the mean, median, proportional weighting 
(longer sentences count more), or max-absolute (capture peak intensity).

**2. Across-documents per period:** How do you combine document scores 
within a week or month? Equal weighting treats every article the same. 
Article-count proportional weighting lets outlets with more articles 
have more influence. Source-balanced weighting averages within each 
outlet first, then across outlets.

**3. Across-time smoothing:** Raw weekly scores are noisy. Smoothing 
options include moving averages, exponential decay, and Beta-weighted 
curves that emphasize recent observations.

### Building a simple monthly index

```{r monthly-index}
# Monthly sentiment index from LSD2015 scores
monthly <- lsd_scores %>%
  mutate(month = floor_date(date, "month")) %>%
  group_by(month) %>%
  summarize(
    mean_sentiment = mean(sent_ratio, na.rm = TRUE),
    sd_sentiment   = sd(sent_ratio, na.rm = TRUE),
    n_articles     = n(),
    se             = sd_sentiment / sqrt(n_articles),
    lower          = mean_sentiment - 1.96 * se,
    upper          = mean_sentiment + 1.96 * se,
    .groups = "drop"
  )
```

```{r monthly-plot, fig.cap="Monthly sentiment index from LSD2015 dictionary scoring. Ribbon shows 95% confidence intervals. With only 20 simulated articles, the intervals are wide; a real project would have hundreds or thousands of articles per month.", fig.height=4.5}
ggplot(monthly, aes(x = month, y = mean_sentiment)) +
  geom_ribbon(aes(ymin = lower, ymax = upper),
              fill = "#8aad96", alpha = 0.3) +
  geom_line(colour = "#1e3a2f", linewidth = 0.9) +
  geom_point(colour = "#1e3a2f", size = 2.5) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "#7a7a72") +
  scale_x_date(date_labels = "%b %Y") +
  labs(x = NULL, y = "Mean sentiment ratio",
       title = "Monthly Environmental News Sentiment Index",
       subtitle = "LSD2015 dictionary, 95% CI ribbon") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(family = "serif", face = "bold",
                                 colour = "#1e3a2f"),
    plot.subtitle = element_text(colour = "#7a7a72", size = 10)
  )
```

### The sentometrics package

For a more principled approach, the `sentometrics` package (Ardia et al., 
*Journal of Statistical Software* 2021) handles all three aggregation 
dimensions in a single workflow. It also provides elastic-net regression 
tools for relating sentiment indices to external variables.

```{r sentometrics-example, eval=FALSE}
# ── sentometrics workflow (conceptual) ─────────────────────────────────
library(sentometrics)

# 1. Create a sentocorpus (requires 'id', 'date', 'texts' columns,
#    plus optional feature columns for topic or source)
sc <- sento_corpus(articles,
                   corpusdf = TRUE,
                   do.clean = TRUE)

# 2. Define lexicons
lexicons <- sento_lexicons(list(
  LSD2015 = data_dictionary_LSD2015,   # from quanteda
  AFINN   = get_sentiments("afinn")     # from tidytext
))

# 3. Define aggregation controls
ctr <- ctr_agg(
  howWithin  = "proportional",     # weight by sentence length
  howDocs    = "equal_weight",     # equal across documents
  howTime    = "equal_weight",     # equal across time
  by         = "month",
  lag        = 6                   # 6-month rolling window
)

# 4. Compute sentiment measures
sento_measures <- sento_measures(sc, lexicons, ctr)

# 5. Plot
plot(sento_measures)
```

<div class="callout callout-concept">
<strong>&#128161; Volume matters for uncertainty</strong>
Environmental news coverage spikes around major events (COP summits, 
oil spills, IPCC report releases, wildfire seasons) and drops during 
quiet periods. A monthly index based on 3 articles in January and 
150 articles in August treats both months equally, even though the 
January estimate is far more uncertain. Always plot confidence intervals 
or credible intervals alongside your sentiment time series, and consider 
inverse-variance weighting so that data-sparse periods do not produce 
spurious extremes.
</div>

<hr>

## 8. Validation {#validation}

No sentiment analysis result should be published without validation against 
human judgment. This section covers the standard validation protocol.

### Step 1: Hand-code a sample

Draw a stratified random sample of 200 to 500 articles (stratified by 
source, time period, and topic). Have **two independent coders** classify 
each article on the same scale your model uses (e.g., positive / neutral / 
negative, or opportunity / neutral / risk).

### Step 2: Measure human agreement

Before comparing the model to humans, measure how well the humans agree 
with each other. If the coders disagree substantially, the task itself is 
too subjective for any model to solve reliably.

```{r kappa-example}
# Simulated coding data: two coders, 50 articles, 3-class scale
set.seed(123)
labels <- c("positive", "neutral", "negative")
coder1 <- sample(labels, 50, replace = TRUE, prob = c(0.25, 0.45, 0.30))
coder2 <- coder1   # start with agreement, then add noise
noise_idx <- sample(50, 12)
coder2[noise_idx] <- sample(labels, 12, replace = TRUE)

# Cohen's kappa
confusion <- table(Coder1 = coder1, Coder2 = coder2)
confusion
```

```{r kappa-calc}
# Manual Cohen's kappa calculation
n <- sum(confusion)
p_o <- sum(diag(confusion)) / n               # observed agreement
p_e <- sum(rowSums(confusion) * colSums(confusion)) / n^2  # expected by chance
kappa <- (p_o - p_e) / (1 - p_e)

cat("Observed agreement (p_o):", round(p_o, 3), "\n")
cat("Expected agreement (p_e):", round(p_e, 3), "\n")
cat("Cohen's kappa:           ", round(kappa, 3), "\n")
```

<div class="callout callout-concept">
<strong>&#128161; Interpreting kappa</strong>
Kappa values below 0.40 indicate poor agreement, and a model trained on 
such labels cannot be trusted. Values between 0.40 and 0.60 are 
"moderate" and typical for news sentiment. Values above 0.60 are 
"substantial." Achieving kappa above 0.80 on news text is unusual 
and may signal that the coding scheme is too coarse (e.g., only 2 
categories). For news sentiment specifically, expect kappa in the 
0.40 to 0.60 range. If your coders are below 0.40, revise the coding 
guidelines before proceeding.
</div>

### Step 3: Compare model to human consensus

After the two coders resolve disagreements (creating a consensus label), 
compute the same metrics between the model's predictions and the 
consensus labels.

```{r validation-metrics}
# Simulated model predictions vs consensus labels
consensus <- coder1  # in practice, resolved disagreements
model_pred <- consensus
# Introduce some model errors
model_err <- sample(50, 8)
model_pred[model_err] <- sample(labels, 8, replace = TRUE)

# Confusion matrix
conf_model <- table(Model = model_pred, Human = consensus)

# Per-class precision, recall, F1
calc_metrics <- function(conf_mat) {
  classes <- rownames(conf_mat)
  metrics <- map_dfr(classes, function(cls) {
    tp <- conf_mat[cls, cls]
    fp <- sum(conf_mat[cls, ]) - tp
    fn <- sum(conf_mat[, cls]) - tp
    precision <- ifelse(tp + fp > 0, tp / (tp + fp), 0)
    recall    <- ifelse(tp + fn > 0, tp / (tp + fn), 0)
    f1        <- ifelse(precision + recall > 0,
                        2 * precision * recall / (precision + recall), 0)
    tibble(Class = cls, Precision = precision,
           Recall = recall, F1 = f1)
  })
  metrics
}

per_class <- calc_metrics(conf_model)
macro_f1 <- mean(per_class$F1)

kable(per_class, digits = 3,
      caption = paste0("Per-class metrics (model vs. human consensus). ",
                       "Macro-F1 = ", round(macro_f1, 3), ".")) %>%
  kable_styling(full_width = FALSE, font_size = 14)
```

### What metrics to report

| Metric | When to use | Caveat |
|---|---|---|
| Macro-F1 | Primary headline metric | Treats all classes equally |
| Weighted F1 | When class imbalance is severe | Inflated by majority class |
| Cohen's kappa (model vs. human) | Always | Chance-corrected agreement |
| Confusion matrix | Always | Shows where errors concentrate |
| Accuracy | Only if classes are balanced | Misleading with skewed neutral class |

<hr>

## 9. Putting It Together: A Full Pipeline {#pipeline}

This section assembles the complete workflow from corpus to validated 
time-series index. In a real project, each step would involve more 
articles and more iterations. The structure, however, remains the same.

```{r pipeline-summary, echo=FALSE}
pipeline_df <- data.frame(
  Stage = c("0. Corpus assembly", "1. Lexicon baseline", "2. Aspect scoring",
            "3. Transformer layer", "4. Validation", "5. Time-series aggregation",
            "6. Reporting"),
  Action = c(
    "Import articles with readtext or API. Attach source, date, topic metadata to corpus.",
    "Score with LSD2015 + sentimentr in parallel. Compare two lexicons for robustness.",
    "Define target entities. Use quanteda tokens_keep(window=10) + LSD2015.",
    "Call ClimateBERT via the text package. Chunk long articles into paragraphs first.",
    "Hand-code 300 articles (2 coders). Compute kappa and macro-F1.",
    "Aggregate to monthly index. Plot with confidence intervals. Overlay event markers.",
    "Report lexicon AND transformer results side by side. Include kappa, per-class F1, and sensitivity to lexicon choice."
  )
)

kable(pipeline_df, col.names = c("Stage", "Action"),
      caption = "The complete sentiment analysis pipeline for environmental news.") %>%
  kable_styling(full_width = TRUE, font_size = 14) %>%
  column_spec(1, bold = TRUE, width = "14em")
```

### Anchoring sentiment to environmental events

One of the most informative analyses is to overlay your sentiment time 
series with known environmental events. This reveals whether sentiment 
shifts lead, lag, or coincide with real-world triggers.

```{r events-plot, fig.cap="Sentiment time series with event annotations. In a real analysis with hundreds of articles, you would mark COP summits, IPCC report releases, major disasters, and policy announcements.", fig.height=4.5}
# Simulated event markers
events <- tibble(
  date  = as.Date(c("2024-03-15", "2024-07-15", "2024-11-15")),
  event = c("IPCC Report\nRelease", "Record Wildfire\nSeason Begins", 
            "COP29\nSummit")
)

ggplot(monthly, aes(x = month, y = mean_sentiment)) +
  geom_ribbon(aes(ymin = lower, ymax = upper),
              fill = "#8aad96", alpha = 0.25) +
  geom_line(colour = "#1e3a2f", linewidth = 0.9) +
  geom_point(colour = "#1e3a2f", size = 2.5) +
  geom_vline(data = events, aes(xintercept = date),
             linetype = "dotted", colour = "#c8832a", linewidth = 0.7) +
  geom_label(data = events, aes(x = date, y = max(monthly$upper, na.rm = TRUE),
                                 label = event),
             size = 2.5, fill = "#fdf8f0", colour = "#c8832a",
             label.padding = unit(0.3, "lines"), family = "mono") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "#7a7a72") +
  scale_x_date(date_labels = "%b %Y") +
  labs(x = NULL, y = "Mean sentiment ratio",
       title = "Environmental News Sentiment with Event Markers") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(family = "serif", face = "bold",
                               colour = "#1e3a2f")
  )
```

<hr>

## 10. Extensions and Next Steps {#extensions}

### Challenges specific to environmental news

Environmental news has several properties that make sentiment analysis 
harder than in other domains:

**Neutral-tone reporting.** Wire services (AP, Reuters) follow style 
guides that enforce emotional flatness. This pushes lexicon scores toward 
zero and inflates the neutral class. Models trained on opinion text (tweets, 
reviews) systematically over-predict polarity on news wire text.

**Attributed quotes.** "Critics called the policy a disaster" reports 
someone else's negative sentiment without the article endorsing it. 
Lexicon methods cannot distinguish the reporter's voice from a quoted 
source. Quote detection (using dependency parsing via `spacyr`) can help.

**Domain vocabulary.** General-purpose lexicons mishandle domain-specific 
terms. "Stranded asset" is not a negative phrase about being abandoned; it 
is a precise financial term. "Loss and damage" is a policy concept, not an 
expression of grief. Domain-adapted models like ClimateBERT exist precisely 
to handle this vocabulary.

**Negativity bias in coverage.** Environmental news is structurally negative. 
Disasters, extinctions, and pollution events are news; stable ecosystems 
are not. Raw aggregate sentiment is therefore meaningless without a 
topic-conditional or publication-conditional baseline.

### Multiple lexicons as a robustness check

Always compute sentiment under at least two independent lexicons and compare. 
If LSD2015 and AFINN produce time series that move in opposite directions on 
the same corpus, that tells you the signal is driven by lexicon composition, 
not by genuine shifts in news tone.

```{r robustness, eval=FALSE}
# Rank correlation between two lexicon time series
cor.test(
  monthly_lsd$mean_sentiment,
  monthly_afinn$mean_sentiment,
  method = "spearman"
)
```

### Topic-conditional sentiment

Running an unsupervised topic model (LDA or BERTopic) on your corpus, 
then computing sentiment within each topic, reveals structure that 
aggregate indices hide. An overall "neutral" trend might mask simultaneous 
increases in negative wildfire sentiment and positive renewable-energy 
sentiment.

### Data sources for environmental news

| Source | Access | Notes |
|---|---|---|
| LexisNexis Academic | University subscription | Full text, structured metadata |
| ProQuest Newsstand | University subscription | Historical newspapers |
| MediaCloud | Free API | Open-source media analysis platform |
| GDELT | Free API | Pre-computed tone scores for global news |
| EPA AQS + newsapi.org | Mixed | Combine air quality data with news coverage |
| Google News RSS | Free | Headlines only, limited metadata |

<div class="callout callout-tip">
<strong>&#9989; Checklist before publishing sentiment results</strong>
Before reporting sentiment analysis findings in a paper or presentation, 
verify that you have: (1) run at least two lexicons and compared agreement, 
(2) hand-coded a validation sample with two coders and reported kappa, 
(3) reported macro-F1 for each method against the hand-coded consensus, 
(4) documented how you handled article truncation in transformer models, 
(5) plotted confidence intervals on any time-series index, and 
(6) tested sensitivity to your aggregation choices (weighting, window, 
smoothing).
</div>

---

<div class="key-box">
<h4>&#128273; Sentiment analysis of environmental news -- key takeaways</h4>
<ul>
<li><strong>Start with lexicons, escalate to transformers only when needed.</strong> Dictionary methods (LSD2015 + <code>sentimentr</code>) are transparent, fast, and require no training data. They are the right starting point for any project.</li>
<li><strong>Lexicon choice changes results.</strong> Different dictionaries produce different time series on the same corpus. Run at least two and check rank correlation. Disagreement is information.</li>
<li><strong>Aspect-based scoring beats document-level scoring for news.</strong> Articles discuss multiple entities and policies. A single polarity score per article collapses important structure. Use <code>tokens_keep(window = 10)</code> to score sentiment around specific targets.</li>
<li><strong>Domain adaptation matters.</strong> ClimateBERT outperforms generic transformers on climate text because environmental vocabulary confuses general models. Use domain-specific models when they exist.</li>
<li><strong>Validate, validate, validate.</strong> Report Cohen's kappa between human coders and macro-F1 between model and consensus. Without these numbers, your results are not credible.</li>
<li><strong>Confidence intervals are not optional.</strong> Environmental news volume fluctuates wildly. Months with few articles produce unreliable estimates. Always visualize uncertainty alongside your index.</li>
<li><strong>Sentiment is a noisy proxy for framing.</strong> Text polarity does not tell you how readers interpret the article, which political frame it employs, or whether its factual claims are accurate. Treat sentiment as one signal among many, not as a standalone finding.</li>
</ul>
</div>

---

### What's next?

- **[Time-Series Analysis](14_Time_Series.html)** -- decomposition, ARIMA, and forecasting for the temporal indices you build here
- **[Multiple Regression & GLMs](11_Multiple_Regression_GLMs.html)** -- regress sentiment indices against external predictors (carbon prices, policy events)
- **[Data Visualization](../basic_statistics/02_Visualization.html)** -- polished ggplot2 figures for publication-quality sentiment plots
- **[Tidyverse Data Wrangling](../RGuides/R09_Tidyverse_Wrangling.html)** -- the pipe-based data manipulation that underlies the entire workflow

```{r session-info, echo=FALSE}
# sessionInfo()
```

<footer class="page-footer">
EA030 Activity A29 &mdash; Sentiment Analysis
<span class="divider">|</span> Prof. Marc Los Huertos
<span class="divider">|</span> Pomona College
<span class="divider">|</span> Last updated: `r format(Sys.Date(), "%B %Y")`
<span class="divider">|</span> <a href="https://github.com/marclos/RTricks">GitHub</a>
</footer>
