---
title: "Urban Air Quality Monitoring with Low-Cost Sensors"
author: "Marc Los Huertos"
date: "EA030 · Compiled `r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    self_contained: true
    theme: null
    highlight: null
    toc: false
knit: (function(inputFile, encoding) {
    rmarkdown::render(
      inputFile,
      encoding = encoding,
      output_file = file.path(dirname(inputFile),
                              "../docs/Unit_Air/P09_Raspberry_Pi_Tutorial.html")
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.width = 9,
  fig.height = 5
)
```

```{css, echo=FALSE}
@import url('https://fonts.googleapis.com/css2?family=Libre+Baskerville:ital,wght@0,400;0,700;1,400&family=Source+Code+Pro:wght@400;600&family=Lato:wght@300;400;700&display=swap');

*, *::before, *::after { box-sizing: border-box; margin: 0; padding: 0; }

:root {
  --forest:    #1e3a2f;
  --moss:      #2d5a3d;
  --fern:      #4a7c5e;
  --sage:      #8aad96;
  --mist:      #d4e6da;
  --cream:     #f7f3ed;
  --parchment: #ede8df;
  --ink:       #1a1a18;
  --charcoal:  #3a3a38;
  --dust:      #7a7a72;
  --amber:     #c8832a;
  --amber-lt:  #e8a84a;
  --white:     #ffffff;
  --card-shadow: 0 2px 10px rgba(30,58,47,.09);
}

html { scroll-behavior: smooth; }
body {
  font-family: 'Lato', sans-serif;
  background: var(--cream);
  color: var(--ink);
  font-size: 17px;
  line-height: 1.75;
  padding-left:  clamp(14px, 4vw, 48px);
  padding-right: clamp(14px, 4vw, 48px);
  padding-bottom: 80px;
}
.main-container { max-width: 920px !important; margin: 0 auto !important; padding: 0 clamp(10px, 2vw, 32px) 80px !important; }
.page-header {
  margin-left:  calc(-1 * clamp(14px, 4vw, 48px));
  margin-right: calc(-1 * clamp(14px, 4vw, 48px));
}
h1.title, h4.author, h4.date { display: none; }

/* Page header */
.page-header {
  background-image:
    repeating-linear-gradient(120deg, transparent, transparent 60px, rgba(255,255,255,.018) 60px, rgba(255,255,255,.018) 61px),
    linear-gradient(160deg, #1e3a2f 0%, #2d5a3d 60%, #1a3d30 100%);
  color: #fff; padding: 44px 4vw 36px; margin: 0 0 48px;
  border-bottom: 4px solid var(--amber); position: relative; overflow: hidden;
}
.page-header::after {
  content: ''; position: absolute; right: -40px; top: -40px;
  width: 260px; height: 260px; border-radius: 50%;
  background: radial-gradient(circle, rgba(138,173,150,.12) 0%, transparent 70%);
  pointer-events: none;
}
.page-header .module-tag {
  display: inline-block; font-family: 'Source Code Pro', monospace;
  font-size: .68rem; font-weight: 600; letter-spacing: .18em; text-transform: uppercase;
  color: var(--amber-lt); border: 1px solid rgba(200,131,42,.45);
  padding: 3px 10px; border-radius: 2px; margin-bottom: 12px;
}
.page-header h1 {
  font-family: 'Libre Baskerville', serif; font-size: clamp(1.6rem, 3vw, 2.3rem);
  font-weight: 700; color: #fff; line-height: 1.2; margin-bottom: 8px;
}
.page-header h1 span { color: var(--sage); }
.page-header .sub { font-size: .9rem; color: rgba(255,255,255,.65); font-weight: 300; }
.page-header .sub strong { color: rgba(255,255,255,.9); font-weight: 700; }
.back-link {
  display: inline-block; margin-top: 16px;
  font-family: 'Source Code Pro', monospace; font-size: .7rem; letter-spacing: .08em;
  color: var(--amber-lt); text-decoration: none;
  border: 1px solid rgba(200,131,42,.4); padding: 4px 11px; border-radius: 3px;
  transition: background .2s;
}
.back-link:hover { background: rgba(200,131,42,.18); }

/* Section nav pills */
.section-nav {
  display: flex; gap: 8px; flex-wrap: wrap; margin-bottom: 44px;
  padding-bottom: 20px; border-bottom: 1px solid var(--parchment);
}
.section-nav a {
  font-family: 'Source Code Pro', monospace; font-size: .7rem; font-weight: 600;
  letter-spacing: .08em; text-transform: uppercase;
  color: var(--fern); text-decoration: none;
  border: 1px solid var(--mist); padding: 4px 12px; border-radius: 3px;
  background: var(--white); transition: background .2s, color .2s;
}
.section-nav a:hover { background: var(--mist); color: var(--forest); }

/* Section headings */
h1 {
  font-family: 'Libre Baskerville', serif; font-size: 1.65rem; font-weight: 700;
  color: var(--forest); margin-top: 60px; margin-bottom: 6px;
  padding-bottom: 8px; border-bottom: 2px solid var(--parchment);
}
h2 {
  font-family: 'Libre Baskerville', serif; font-size: 1.3rem; font-weight: 700;
  color: var(--moss); margin-top: 40px; margin-bottom: 4px;
  padding-bottom: 6px; border-bottom: 1px solid var(--parchment);
}
h3 {
  font-family: 'Libre Baskerville', serif; font-size: 1.05rem; font-weight: 700;
  color: var(--fern); margin-top: 24px; margin-bottom: 4px;
}
h4 {
  font-family: 'Source Code Pro', monospace; font-size: .9rem; font-weight: 600;
  color: var(--charcoal); margin-top: 20px; margin-bottom: 4px;
  letter-spacing: .04em; text-transform: uppercase;
}

/* Body text */
p { margin-bottom: 1em; }
strong { color: var(--forest); }
blockquote {
  border-left: 4px solid var(--sage); margin: 1.2em 0;
  padding: .5em 1em; background: var(--white); border-radius: 0 5px 5px 0;
  font-size: .94em; color: var(--charcoal);
}

/* Code */
code {
  font-family: 'Source Code Pro', monospace; font-size: .84em;
  background: var(--mist); color: var(--moss); padding: 1px 5px; border-radius: 3px;
}
pre {
  background: var(--ink); color: #e8e6df; border-radius: 5px;
  padding: 14px 18px; font-size: .83rem; line-height: 1.55;
  overflow-x: auto; border-left: 4px solid var(--amber); margin: 1.2em 0;
}
pre code { background: transparent; color: inherit; padding: 0; font-size: 1em; }

/* Callout boxes */
.callout {
  border-left: 5px solid; padding: .65em 1.1em;
  margin: 1.4em 0; border-radius: 0 5px 5px 0; font-size: .92em; line-height: 1.6;
}
.callout > strong:first-child { display: block; margin-bottom: .3em; }
.callout-concept { border-color: var(--fern);  background: #f0f7f2; }
.callout-tip     { border-color: #2e7d32;      background: #f1f8f1; }
.callout-warn    { border-color: #e65100;      background: #fff8f4; }
.callout-formula { border-color: var(--amber); background: #fdf8f0;
                   font-family: 'Source Code Pro', monospace; font-size: .88em; }

/* Tables — base styles; kableExtra injects its own CSS on top */
table { font-size: .88rem; border-collapse: collapse; width: 100%; margin: 1.2em 0; }
thead th {
  background: var(--forest); color: #fff;
  font-family: 'Source Code Pro', monospace; font-size: .74rem;
  letter-spacing: .06em; text-transform: uppercase; padding: 7px 11px; text-align: left;
}
tbody tr:nth-child(even) { background: var(--mist); }
tbody td { padding: 6px 11px; border-bottom: 1px solid var(--parchment); }

/* Key concept summary boxes */
.key-box {
  background: var(--forest); color: #fff; border-radius: 6px;
  padding: 20px 24px; margin: 2em 0; font-size: .9rem;
}
.key-box h4 {
  font-family: 'Libre Baskerville', serif; color: var(--sage);
  font-size: 1rem; margin-bottom: 10px; text-transform: none; letter-spacing: 0;
}
.key-box ul { padding-left: 1.3em; }
.key-box li { margin-bottom: 5px; color: rgba(255,255,255,.85); }
.key-box code { background: rgba(255,255,255,.12); color: var(--amber-lt); }

/* Footer */
.page-footer {
  background: var(--ink); color: rgba(255,255,255,.42); text-align: center;
  padding: 24px 4vw; margin: 60px calc(-1 * clamp(14px, 4vw, 48px)) 0;
  font-size: .76rem; font-family: 'Source Code Pro', monospace; letter-spacing: .04em;
}
.page-footer a { color: var(--sage); text-decoration: none; }
.page-footer .divider { margin: 0 8px; opacity: .3; }
```

<div class="page-header">
<div class="module-tag">EA030 &mdash; Field &amp; Lab Projects &mdash; P09</div>
<h1>Urban Air Quality: <span>Low-Cost Sensor Networks</span></h1>
<p class="sub"><strong>Prof. Marc Los Huertos</strong> &nbsp;&bull;&nbsp; Pomona College &nbsp;&bull;&nbsp; Spring 2026</p>
<a class="back-link" href="../index.html">&#8592; Back to RTricks</a>
</div>

<div class="section-nav">
<a href="#background">Background</a>
<a href="#setup">Setup</a>
<a href="#clock">Clock Correction</a>
<a href="#loading">Data Loading</a>
<a href="#descriptive">Descriptive Stats</a>
<a href="#visualization">Visualization</a>
<a href="#inference">Inference</a>
<a href="#advanced">Advanced</a>
<a href="#conclusions">Conclusions</a>
</div>

---

# 1 · Background and Research Goals {#background}

## 1.1 What Are We Measuring?

Particulate matter (PM) is a mixture of solid particles and liquid droplets suspended
in air. Regulatory agencies classify particles by their aerodynamic diameter:

| Fraction | Diameter      | Health relevance |
|----------|---------------|-----------------|
| **PM1**  | ≤ 1 µm        | Penetrates deepest into lung tissue; linked to cardiovascular effects |
| **PM2.5**| ≤ 2.5 µm      | EPA primary NAAQS standard; major focus of air quality policy |
| **PM10** | ≤ 10 µm       | Larger particles; irritate upper airways |

The EPA Air Quality Index (AQI) categories for PM2.5 (24-h average) are:

| Category              | PM2.5 (µg/m³) |
|-----------------------|--------------|
| Good                  | 0 – 12.0     |
| Moderate              | 12.1 – 35.4  |
| Unhealthy for Sensitive Groups (USG) | 35.5 – 55.4 |
| Unhealthy             | 55.5 – 150.4 |

## 1.2 Sensor Network

Students deployed **Raspberry Pi Zero** units equipped with Enviro+ boards
(PMS5003 optical particle sensor + BME280 meteorological sensor + MICS6814 gas sensor)
at multiple campus locations. Each unit logs:

- `pm1_ug_m3`, `pm25_ug_m3`, `pm10_ug_m3` — particulate concentrations (µg/m³)
- `temperature_C`, `humidity_pct`, `pressure_hPa` — meteorological variables
- `light_lux` — ambient light
- `oxidised_kO`, `reduced_kO`, `nh3_kO` — gas-phase proxies (oxidising gases,
  reducing gases, ammonia)

## 1.3 Research Questions

<div class="callout callout-concept">
**Primary questions:**

1. Do PM2.5 and PM1.0 concentrations differ significantly across campus locations?
2. Is there a consistent diurnal (time-of-day) pattern in PM concentrations?
3. Which co-measured variables (temperature, humidity, gas sensors) best predict PM2.5?
</div>

## 1.4 Hypotheses

<div class="callout callout-tip">
**H1 (Spatial):** Sensors near higher-traffic areas or denser vegetation will show
elevated PM2.5 relative to open-courtyard sensors, due to proximity to combustion
sources or local resuspension.

**H2 (Diurnal):** PM2.5 will peak during morning and evening hours, consistent with
traffic-related emissions patterns and reduced atmospheric mixing during stable
boundary-layer conditions.

**H3 (Meteorological):** PM2.5 will be positively correlated with humidity
(hygroscopic growth of particles) and negatively correlated with pressure
(higher pressure → stronger inversions → accumulation) and temperature
(higher temperature → stronger convective mixing → dilution).
</div>

---

# 2 · Setup {#setup}

## 2.1 Packages

Run the install block **once in the RStudio Console** before knitting — not inside the Rmd.

<div class="callout callout-warn">
<strong>⚠️ Install once in the Console, never inside a knitting chunk</strong>
The block below only installs packages you are missing. Run it interactively once,
then the <code>eval=FALSE</code> tag keeps it from re-running on every knit.
</div>

```{r install-packages, eval=FALSE}
pkgs <- c("tidyverse", "lubridate", "patchwork", "scales",
          "zoo",       "ggridges",  "GGally",   "moments",
          "broom",     "knitr",     "kableExtra")
to_install <- pkgs[!pkgs %in% rownames(installed.packages())]
if (length(to_install)) install.packages(to_install, quiet = TRUE)
```

```{r packages}
library(tidyverse); library(lubridate); library(patchwork); library(scales)
library(zoo);       library(ggridges);  library(GGally);    library(moments)
library(broom);     library(knitr);     library(kableExtra)

theme_set(theme_minimal(base_size = 13))

# Color palette — scales automatically to however many sensors are present
sensor_pal <- c(
  "#0072B2", "#D55E00", "#009E73", "#CC79A7",
  "#E69F00", "#56B4E9", "#F0E442", "#999999",
  "#882255", "#44AA99"
)
```

---

# 3 · The Clock-Correction Problem {#clock}

<div class="callout callout-warn">
**Why does this matter?**  Raspberry Pi Zeros have **no real-time clock (RTC) battery**.
Every time a unit loses power it resets its internal clock to an arbitrary default epoch.
If we analyzed the raw timestamps, each sensor's time series would be offset from real
time by a different amount — making cross-sensor comparisons meaningless.
</div>

## 3.1 Correction Strategy

At deployment, each team member recorded:

1. The **actual wall-clock time** (PST) when they powered the unit on.
2. The **time displayed on the Pi's LCD** at that same moment.

The offset is simply:

$$\text{offset} = t_{\text{real}} - t_{\text{Pi}}$$

This constant is then **added to every raw timestamp** in that sensor's CSV.

## 3.2 Calibration Table

```{r calibration}
# sensor_calibration.R must live in the same directory as this Rmd.
# It creates the `calibration` tibble with one row per sensor (two for PiZ15).
source("sensor_calibration.R")

clock_offsets <- calibration |>
  filter(reboot == 1) |>
  select(sensor, offset_sec)

calibration |>
  filter(reboot == 1) |>
  select(sensor, team_member, location, offset_label, notes) |>
  kable(caption = "Deployment metadata and clock offsets (reboot 1 only)") |>
  kable_styling(full_width = FALSE) |>
  column_spec(3, width = "22em") |>
  row_spec(which(is.na(calibration$offset_label[calibration$reboot == 1])),
           color = "#aaa")
```

<div class="callout callout-concept">
**Note on PiZ2W1:** The sheet entry reads "2/9 4:47 PM". Because the raw data
begins 2026-02-27 *and* every other Pi clock was behind real time (never ahead),
this is interpreted as **3/9** (March 9). Verify with the student if needed.
</div>

---

# 4 · Data Loading and Cleaning {#loading}

## 4.1 Reading the CSVs

We auto-discover every `*_ea30_sp26_v05.csv` file in `data_dir` so no manual
list is needed — and no trailing-comma errors.

```{r load-data}
# ── Set to the folder containing your CSV files ──────────────────────────────
data_dir <- "/home/mwl04747/RTricks/28_Raspberry_Pi_Analysis/Data_S2026"   # <-- change this path if your files are elsewhere

# ── Helper: read one CSV and apply the clock correction ──────────────────────
read_sensor <- function(path, sensor_id, offsets_tbl) {

  offset_s <- offsets_tbl |>
    filter(sensor == sensor_id) |>
    pull(offset_sec)

  if (length(offset_s) == 0 || is.na(offset_s)) {
    warning(sprintf("No clock offset for %s — timestamps uncorrected.", sensor_id))
    offset_s <- 0
  }

  # Force timestamp to character so readr auto-parsing cannot interfere,
  # then rebuild datetime via plain numeric arithmetic (avoids Period/vctrs issues)
  read_csv(path, col_types = cols(timestamp = col_character()),
           show_col_types = FALSE) |>
    mutate(
      sensor      = sensor_id,
      datetime_pi = ymd_hms(timestamp, tz = "America/Los_Angeles", quiet = TRUE),
      datetime    = as.POSIXct(
                      as.numeric(datetime_pi) + as.numeric(offset_s),
                      tz     = "America/Los_Angeles",
                      origin = "1970-01-01"
                    ),
      date        = as_date(datetime),
      hour        = hour(datetime),
      wday        = wday(datetime, label = TRUE, abbr = TRUE),
      tod         = cut(hour,
                        breaks = c(-1, 5, 11, 17, 20, 23),
                        labels = c("Night (0–5)", "Morning (6–11)",
                                   "Afternoon (12–17)", "Evening (18–20)",
                                   "Night (21–23)"))
    ) |>
    select(-timestamp, -datetime_pi)
}

# ── Auto-discover CSVs ────────────────────────────────────────────────────────
csv_paths  <- list.files(data_dir, pattern = "_ea30_sp26_v05\\.csv$",
                         full.names = TRUE)
sensor_ids <- sub("_ea30_sp26_v05\\.csv$", "", basename(csv_paths))

if (length(csv_paths) == 0)
  stop("No *_ea30_sp26_v05.csv files found. Set data_dir correctly.")

files <- setNames(as.list(csv_paths), sensor_ids)
message(sprintf("Discovered %d file(s): %s",
                length(files), paste(sensor_ids, collapse = ", ")))

raw <- imap_dfr(files, ~ read_sensor(.x, .y, clock_offsets)) |>
  mutate(datetime = as.POSIXct(datetime, tz = "America/Los_Angeles",
                               origin = "1970-01-01"))

glimpse(raw)
```

<div class="callout callout-warn">
<strong>⚠️ Check datetime class before continuing</strong>
The line <code>glimpse(raw)</code> above should show <code>datetime &lt;dttm&gt;</code> — not
<code>&lt;chr&gt;</code> or <code>&lt;dbl&gt;</code>. If it shows the wrong type, your
<code>timestamp</code> column format may not match <code>ymd_hms()</code>. Run
<code>class(raw$datetime)</code> and <code>head(raw$timestamp)</code> in the Console
to diagnose. A format like <code>"02/27/2026 14:23:45"</code> needs
<code>mdy_hms()</code> instead.
</div>

## 4.2 Quality Control

We apply three QC filters:

1. **Timestamp present** — rows with no parseable datetime are removed.
2. **PM2.5 present** — rows with `NA` in the target variable are removed.
3. **Minimum observations** — sensors with fewer than 10 valid PM2.5 readings
   are excluded from analysis (e.g. a unit that only logged one row).

```{r qc}
# Sensors with enough data
valid_sensors <- raw |>
  filter(!is.na(pm25_ug_m3)) |>
  count(sensor) |>
  filter(n >= 10) |>
  pull(sensor)

dropped <- setdiff(sensor_ids, valid_sensors)
if (length(dropped) > 0)
  message("Dropped (< 10 obs): ", paste(dropped, collapse = ", "))

df <- raw |>
  filter(sensor %in% valid_sensors,
         !is.na(pm25_ug_m3),
         !is.na(datetime))

# Attach location label for plot facets / legends
df <- df |>
  left_join(
    calibration |>
      filter(reboot == 1) |>
      select(sensor, location),
    by = "sensor"
  ) |>
  mutate(
    sensor      = factor(sensor),
    loc_label   = str_trunc(coalesce(location, as.character(sensor)), 35),
    # Re-assert types after join — left_join can silently coerce POSIXct to dbl
    datetime    = as.POSIXct(datetime, tz = "America/Los_Angeles", origin = "1970-01-01"),
    pm25_ug_m3  = as.numeric(pm25_ug_m3),
    pm1_ug_m3   = as.numeric(pm1_ug_m3),
    pm10_ug_m3  = as.numeric(pm10_ug_m3)
  )

# Hard stop if datetime still isn't POSIXct — prevents cryptic ggplot errors downstream
stopifnot(
  "datetime must be POSIXct — check glimpse(raw) output above" =
    inherits(df$datetime, "POSIXct")
)
```

Final dataset: **`r nrow(df)` rows** across **`r n_distinct(df$sensor)` sensor(s)**.

## 4.3 Coverage Summary

```{r coverage-table}
df |>
  group_by(sensor, loc_label) |>
  summarise(
    n_obs      = n(),
    start      = min(date),
    end        = max(date),
    days_span  = as.integer(end - start) + 1L,
    pct_missing_pm25 = mean(is.na(pm25_ug_m3)) * 100,
    .groups = "drop"
  ) |>
  kable(digits = 1,
        caption = "Data coverage per sensor after QC",
        col.names = c("Sensor","Location","N obs","Start","End",
                      "Days spanned","% PM2.5 missing")) |>
  kable_styling(full_width = FALSE)
```

---

# 5 · Descriptive Statistics {#descriptive}

## 5.1 Summary Table

```{r desc-stats}
desc <- df |>
  group_by(sensor) |>
  summarise(
    N        = n(),
    Mean     = mean(pm25_ug_m3),
    SD       = sd(pm25_ug_m3),
    Median   = median(pm25_ug_m3),
    IQR      = IQR(pm25_ug_m3),
    Skewness = moments::skewness(pm25_ug_m3),
    P5       = quantile(pm25_ug_m3, .05),
    P95      = quantile(pm25_ug_m3, .95),
    Max      = max(pm25_ug_m3),
    `% Good (≤12)`       = mean(pm25_ug_m3 <= 12)  * 100,
    `% Moderate (12–35)` = mean(pm25_ug_m3 > 12 & pm25_ug_m3 <= 35.4) * 100,
    `% USG (>35)`        = mean(pm25_ug_m3 > 35.4) * 100,
    .groups = "drop"
  )

kable(desc, digits = 2,
      caption = "PM2.5 descriptive statistics (µg/m³) with EPA AQI breakdown") |>
  kable_styling(full_width = TRUE) |>
  add_header_above(c(" " = 10, "EPA AQI Category (%)" = 3))
```

<div class="callout callout-concept">
**Reading the table:** High skewness (> 1) indicates a right-skewed distribution —
a small number of high-concentration events pull the mean above the median.
This is typical for PM data and motivates the use of non-parametric tests later.
</div>

## 5.2 PM1.0 vs PM2.5 — Are They Telling the Same Story?

```{r pm1-vs-pm25}
df |>
  select(sensor, pm1_ug_m3, pm25_ug_m3) |>
  pivot_longer(-sensor, names_to = "fraction", values_to = "conc") |>
  mutate(fraction = recode(fraction,
                           pm1_ug_m3  = "PM1.0",
                           pm25_ug_m3 = "PM2.5")) |>
  group_by(sensor, fraction) |>
  summarise(Mean = mean(conc), Median = median(conc),
            SD = sd(conc), .groups = "drop") |>
  pivot_wider(names_from = fraction,
              values_from = c(Mean, Median, SD)) |>
  mutate(`PM1/PM2.5 ratio (mean)` = `Mean_PM1.0` / `Mean_PM2.5`) |>
  kable(digits = 2,
        caption = "PM1.0 vs PM2.5 summary — a ratio near 1 means nearly all fine
                   particles are in the sub-1µm range (combustion dominated)") |>
  kable_styling(full_width = FALSE)
```

---

# 6 · Visualization {#visualization}

## 6.1 Time Series with Rolling Mean

The raw signal is noisy (sensor noise, brief spikes). A centred **11-point rolling
mean** (~55 minutes at 5-min sampling) reveals the underlying trend.

```{r timeseries, fig.height=5}
n_sensors <- n_distinct(df$sensor)
pal       <- setNames(sensor_pal[seq_len(n_sensors)], levels(df$sensor))

# zoo::rollmean: 11-point centred moving average; fill = NA keeps vector length
df |>
  arrange(sensor, datetime) |>
  group_by(sensor) |>
  mutate(pm25_roll = as.numeric(
           zoo::rollmean(pm25_ug_m3, k = 11, fill = NA, align = "center")
         )) |>
  ungroup() |>
  ggplot(aes(x = datetime, color = sensor)) +
  geom_line(aes(y = pm25_ug_m3), alpha = 0.18, linewidth = 0.3) +
  geom_line(aes(y = pm25_roll),  linewidth = 0.8) +
  geom_hline(yintercept = 12,   linetype = "dashed",
             color = "forestgreen", linewidth = 0.6) +
  geom_hline(yintercept = 35.4, linetype = "dashed",
             color = "darkorange",  linewidth = 0.6) +
  annotate("text", x = min(df$datetime, na.rm = TRUE), y = 13.2,
           label = "EPA Good/Moderate (12 µg/m³)",
           hjust = 0, size = 3.2, color = "forestgreen") +
  annotate("text", x = min(df$datetime, na.rm = TRUE), y = 36.6,
           label = "EPA Moderate/USG (35.4 µg/m³)",
           hjust = 0, size = 3.2, color = "darkorange") +
  scale_color_manual(values = pal) +
  scale_x_datetime(date_breaks = "3 days", date_labels = "%b %d") +
  labs(title    = "PM2.5 Time Series — All Sensors",
       subtitle = "Faint lines = raw 5-min readings; bold = 11-pt centred rolling mean",
       x = NULL, y = "PM2.5 (µg/m³)", color = "Sensor") +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 30, hjust = 1))
```

> **Interpretation:** Look for (a) periods where sensors diverge — suggests local
> sources near individual sensors; (b) synchronised spikes across all sensors —
> suggests a regional pollution event; (c) whether any sensor consistently exceeds
> EPA thresholds.

## 6.2 PM1.0 and PM2.5 Together

```{r pm-fractions, fig.height=6}
df |>
  select(sensor, datetime, pm1_ug_m3, pm25_ug_m3) |>
  pivot_longer(c(pm1_ug_m3, pm25_ug_m3),
               names_to = "fraction", values_to = "conc") |>
  mutate(fraction = recode(fraction,
                           pm1_ug_m3  = "PM1.0",
                           pm25_ug_m3 = "PM2.5")) |>
  arrange(sensor, fraction, datetime) |>
  group_by(sensor, fraction) |>
  mutate(conc_roll = as.numeric(
           zoo::rollmean(conc, k = 11, fill = NA, align = "center")
         )) |>
  ungroup() |>
  ggplot(aes(x = datetime, y = conc_roll,
             color = fraction, linetype = fraction)) +
  geom_line(linewidth = 0.8, alpha = 0.85) +
  facet_wrap(~ sensor, ncol = 2, scales = "free_y") +
  scale_color_manual(values = c("PM1.0" = "#7B2D8B", "PM2.5" = "#D55E00")) +
  scale_x_datetime(date_breaks = "4 days", date_labels = "%b %d") +
  labs(title    = "Rolling-Mean PM1.0 vs PM2.5 by Sensor",
       subtitle = "11-pt centred rolling mean; free y-scale per sensor",
       x = NULL, y = "Concentration (µg/m³)",
       color = "Fraction", linetype = "Fraction") +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 30, hjust = 1),
        strip.text = element_text(face = "bold"))
```

> **Interpretation:** When PM1.0 tracks PM2.5 almost exactly (ratio ≈ 1), the
> aerosol is dominated by fine combustion particles. A large gap between the two
> lines indicates a contribution from coarser particles (dust, pollen, sea spray).

## 6.3 Distributions

```{r distributions, fig.height=5}
p_dens <- df |>
  ggplot(aes(x = pm25_ug_m3, fill = sensor, color = sensor)) +
  geom_density(alpha = 0.30, linewidth = 0.7) +
  geom_vline(xintercept = c(12, 35.4),
             linetype = "dashed",
             color    = c("forestgreen", "darkorange"),
             linewidth = 0.7) +
  scale_fill_manual(values  = pal) +
  scale_color_manual(values = pal) +
  labs(title = "PM2.5 Density", x = "PM2.5 (µg/m³)",
       y = "Density", fill = "Sensor", color = "Sensor") +
  theme(legend.position = "top")

p_box <- df |>
  ggplot(aes(x = sensor, y = pm25_ug_m3, fill = sensor)) +
  geom_boxplot(outlier.alpha = 0.25, outlier.size = 0.8, width = 0.5) +
  geom_jitter(width = 0.12, alpha = 0.04, size = 0.5) +
  stat_summary(fun = mean, geom = "point",
               shape = 23, fill = "white", size = 2.5) +
  scale_fill_manual(values = pal) +
  labs(title    = "PM2.5 by Sensor",
       subtitle = "◇ = mean; box = IQR",
       x = "Sensor", y = "PM2.5 (µg/m³)") +
  theme(legend.position = "none")

p_dens + p_box
```

<div class="callout callout-concept">
**Right skew is normal for PM data.** Most readings are low, but episodic events
produce a long right tail. The mean will always exceed the median. Use median and IQR
for central tendency, and non-parametric tests for inference.
</div>

## 6.4 Diurnal Pattern (Time-of-Day)

```{r diurnal, fig.height=4.5}
df |>
  group_by(sensor, hour) |>
  summarise(
    mean_pm25 = mean(pm25_ug_m3),
    se_pm25   = sd(pm25_ug_m3) / sqrt(n()),
    mean_pm1  = mean(pm1_ug_m3),
    se_pm1    = sd(pm1_ug_m3)  / sqrt(n()),
    .groups   = "drop"
  ) |>
  pivot_longer(c(mean_pm25, mean_pm1),
               names_to = "fraction", values_to = "mean") |>
  mutate(
    se       = if_else(fraction == "mean_pm25", se_pm25, se_pm1),
    fraction = recode(fraction, mean_pm25 = "PM2.5", mean_pm1 = "PM1.0")
  ) |>
  ggplot(aes(x = hour, y = mean, color = sensor, linetype = fraction)) +
  geom_ribbon(aes(ymin = mean - se, ymax = mean + se,
                  fill = sensor, group = interaction(sensor, fraction)),
              alpha = 0.10, color = NA) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~ fraction) +
  scale_color_manual(values = pal) +
  scale_fill_manual(values  = pal) +
  scale_x_continuous(breaks = seq(0, 23, 3),
                     labels = sprintf("%02d:00", seq(0, 23, 3))) +
  labs(title    = "Diurnal Profile of PM1.0 and PM2.5",
       subtitle = "Mean ± 1 SE by hour of day",
       x = "Hour of Day", y = "Concentration (µg/m³)",
       color = "Sensor", fill = "Sensor", linetype = NULL) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "top")
```

> **What to look for:** Morning peaks (~7–9 AM) or evening peaks (~5–8 PM) would
> support H2. A midday dip is consistent with stronger atmospheric mixing (higher
> planetary boundary layer) diluting surface-level emissions.

## 6.5 Date × Hour Heatmap

```{r heatmap, fig.height=max(4, n_distinct(df$sensor) * 2)}
df |>
  group_by(sensor, date, hour) |>
  summarise(pm25 = mean(pm25_ug_m3), .groups = "drop") |>
  ggplot(aes(x = hour, y = date, fill = pm25)) +
  geom_tile(color = "white", linewidth = 0.08) +
  facet_wrap(~ sensor, ncol = 2) +
  scale_fill_viridis_c(option = "magma", name = "PM2.5\n(µg/m³)") +
  scale_x_continuous(breaks = seq(0, 23, 6)) +
  scale_y_date(date_breaks = "3 days", date_labels = "%b %d") +
  labs(title = "PM2.5 Heatmap: Date × Hour of Day",
       x = "Hour", y = NULL) +
  theme(strip.text = element_text(face = "bold"))
```

> **Reading the heatmap:** Bright columns at consistent hours → reliable diurnal
> signal. Bright rows → day-wide pollution events (e.g. smoke, dust). Patchy
> brightness → intermittent local sources.

## 6.6 Correlation Matrix

These heatmaps show pairwise Pearson correlations between PM2.5 and all co-measured
variables. They help identify which predictors to include in the regression model.

```{r corr-matrix, fig.height=8, results='asis'}
cor_vars <- c("pm25_ug_m3", "pm1_ug_m3", "pm10_ug_m3",
              "temperature_C", "humidity_pct", "pressure_hPa",
              "light_lux", "oxidised_kO", "reduced_kO", "nh3_kO")

for (s in levels(df$sensor)) {
  cat(sprintf("\n\n#### Sensor: %s\n\n", s))
  p <- df |>
    filter(sensor == s) |>
    select(all_of(cor_vars)) |>
    drop_na() |>
    GGally::ggcorr(
      method      = c("pairwise.complete.obs", "pearson"),
      label       = TRUE, label_round = 2, label_size = 3,
      hjust       = 0.75,
      low         = "#D55E00", mid = "white", high = "#0072B2",
      legend.size = 10
    ) +
    labs(title = paste("Correlation Matrix —", s))
  print(p)
}
```

---

# 7 · Basic Statistical Analysis {#inference}

## 7.1 Are the Data Normally Distributed?

```{r normality}
set.seed(42)
norm_res <- df |>
  group_by(sensor) |>
  summarise(
    n       = n(),
    W_pm25  = shapiro.test(sample(pm25_ug_m3, min(n(), 5000)))$statistic,
    p_pm25  = shapiro.test(sample(pm25_ug_m3, min(n(), 5000)))$p.value,
    W_pm1   = shapiro.test(sample(pm1_ug_m3,  min(n(), 5000)))$statistic,
    p_pm1   = shapiro.test(sample(pm1_ug_m3,  min(n(), 5000)))$p.value,
    .groups = "drop"
  ) |>
  mutate(
    normal_pm25 = if_else(p_pm25 > 0.05, "Yes", "No"),
    normal_pm1  = if_else(p_pm1  > 0.05, "Yes", "No")
  )

kable(norm_res, digits = 4,
      caption = "Shapiro-Wilk normality test (sampled to ≤ 5000 obs)") |>
  kable_styling(full_width = FALSE)
```

<div class="callout callout-concept">
**Interpretation:** A p-value < 0.05 rejects the null hypothesis of normality.
PM data are almost always non-normal (right-skewed, discrete integer values at low
concentrations). This means we should use **non-parametric tests** for inference,
not t-tests or ANOVA.
</div>

## 7.2 Do Sensors Differ? — Wilcoxon Rank-Sum Test

The Wilcoxon rank-sum test (Mann-Whitney U) is the non-parametric equivalent of
the two-sample t-test. It tests whether the distributions of two groups have
the same location (median), without assuming normality.

**Hypotheses:** H₀: PM2.5 at sensor A = PM2.5 at sensor B (same distribution);
H₁: they differ.

```{r wilcoxon}
sensors_v <- levels(df$sensor)
combs     <- combn(sensors_v, 2, simplify = FALSE)

wilcox_res <- map_dfr(combs, function(pair) {
  x  <- df |> filter(sensor == pair[1]) |> pull(pm25_ug_m3)
  y  <- df |> filter(sensor == pair[2]) |> pull(pm25_ug_m3)
  wt <- wilcox.test(x, y, conf.int = TRUE, exact = FALSE)
  n1 <- length(x); n2 <- length(y)
  r  <- 1 - 2 * wt$statistic / (n1 * n2)   # rank-biserial effect size
  tibble(
    Comparison      = paste(pair[1], "vs", pair[2]),
    W               = wt$statistic,
    p_value         = wt$p.value,
    Location_shift  = wt$estimate,
    CI_lower        = wt$conf.int[1],
    CI_upper        = wt$conf.int[2],
    Effect_r        = r,
    Significant     = if_else(p_value < 0.05, "✓", "—")
  )
})

wilcox_res |>
  mutate(Significant = cell_spec(
    Significant, bold = TRUE,
    color = if_else(Significant == "✓", "forestgreen", "grey60")
  )) |>
  kable(digits = 4, escape = FALSE,
        caption = "Pairwise Wilcoxon tests for PM2.5; effect size r = rank-biserial correlation") |>
  kable_styling(full_width = FALSE)
```

<div class="callout callout-concept">
**Effect size (rank-biserial r):** |r| < 0.1 = negligible; 0.1–0.3 = small;
0.3–0.5 = medium; > 0.5 = large. A statistically significant result with a
negligible effect size is not practically meaningful.
</div>

## 7.3 Does PM2.5 Vary by Time of Day? — Kruskal-Wallis Test

```{r kruskal}
kw_res <- df |>
  group_by(sensor) |>
  summarise(
    chi_sq  = kruskal.test(pm25_ug_m3 ~ tod)$statistic,
    df_kw   = kruskal.test(pm25_ug_m3 ~ tod)$parameter,
    p_value = kruskal.test(pm25_ug_m3 ~ tod)$p.value,
    .groups = "drop"
  ) |>
  mutate(significant = if_else(p_value < 0.05, "✓", "—"),
         # Epsilon-squared effect size for KW
         eta_sq = (chi_sq - df_kw + 1) / (n_distinct(df$sensor) - df_kw))

kable(kw_res, digits = 4,
      caption = "Kruskal-Wallis test: does PM2.5 differ across time-of-day groups?") |>
  kable_styling(full_width = FALSE)
```

```{r tod-boxplot, fig.height=4.5}
df |>
  ggplot(aes(x = tod, y = pm25_ug_m3, fill = tod)) +
  geom_boxplot(outlier.alpha = 0.15, outlier.size = 0.6, width = 0.6) +
  stat_summary(fun = mean, geom = "point",
               shape = 23, fill = "white", size = 2) +
  facet_wrap(~ sensor, nrow = 1) +
  scale_fill_brewer(palette = "RdYlGn", direction = -1) +
  labs(title    = "PM2.5 by Time of Day",
       subtitle = "Kruskal-Wallis test p-values shown above",
       x = NULL, y = "PM2.5 (µg/m³)") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 35, hjust = 1))
```

---

# 8 · Advanced Analysis {#advanced}

## 8.1 What Predicts PM2.5? — Multiple Linear Regression

Even though PM2.5 is not normally distributed, OLS regression can still be used
to identify *predictors*, because the assumptions concern the **residuals**, not the
raw response. We will check residuals below.

**Model:**  
$$\text{PM}_{2.5} = \beta_0 + \beta_1 T + \beta_2 \text{RH} + \beta_3 P + \beta_4 \text{light} + \beta_5 \text{ox} + \beta_6 \text{red} + \beta_7 \text{NH}_3 + \varepsilon$$

```{r regression}
lm_list <- df |>
  group_by(sensor) |>
  group_map(function(d, key) {
    mod <- lm(pm25_ug_m3 ~ temperature_C + humidity_pct + pressure_hPa +
                light_lux + oxidised_kO + reduced_kO + nh3_kO,
              data = d)
    list(sensor = key$sensor, model = mod,
         tidy   = tidy(mod)   |> mutate(sensor = key$sensor, .before=1),
         glance = glance(mod) |> mutate(sensor = key$sensor, .before=1))
  })

# Coefficient table
map_dfr(lm_list, "tidy") |>
  mutate(sig = case_when(p.value < 0.001 ~ "***",
                         p.value < 0.01  ~ "**",
                         p.value < 0.05  ~ "*",
                         TRUE            ~ "")) |>
  kable(digits = 4,
        caption = "OLS regression coefficients (*** p<.001, ** p<.01, * p<.05)") |>
  kable_styling(full_width = TRUE)
```

```{r r-squared}
map_dfr(lm_list, "glance") |>
  select(sensor, r.squared, adj.r.squared,
         statistic, p.value, nobs) |>
  kable(digits = 4, caption = "Model fit statistics") |>
  kable_styling(full_width = FALSE)
```

<div class="callout callout-concept">
**Interpreting R²:** An R² of 0.35 means the seven predictors together explain 35%
of the variance in PM2.5. The remaining 65% is unexplained — likely due to
unmeasured factors (wind speed, source proximity, indoor infiltration).
</div>

## 8.2 Residual Diagnostics

Check the four key OLS assumptions visually:

```{r residuals, fig.height=6, results='asis'}
for (item in lm_list) {
  cat(sprintf("\n\n#### Sensor: %s\n\n", item$sensor))
  par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
  plot(item$model, main = item$sensor)
  par(mfrow = c(1, 1))
}
```

<div class="callout callout-warn">
**What to look for:**

- **Residuals vs Fitted** — should show no pattern (random scatter). A funnel shape
  means heteroscedasticity; a curve means a non-linear relationship.
- **Q-Q Plot** — points should follow the diagonal. Heavy tails suggest non-normality
  of residuals (log-transforming PM2.5 often helps).
- **Scale-Location** — similar to plot 1 but easier to see variance changes.
- **Cook's Distance** — points > 0.5 are influential and worth investigating.

If assumptions are badly violated, consider: (a) log-transform the response,
(b) use a Gamma GLM (natural for positive continuous data), or (c) a robust
regression.
</div>

## 8.3 Log-Transformed Response

PM concentrations are strictly positive and right-skewed — a natural candidate for
log-transformation, which often improves residual normality.

```{r log-regression}
lm_log_list <- df |>
  filter(pm25_ug_m3 > 0) |>    # log(0) = -Inf
  group_by(sensor) |>
  group_map(function(d, key) {
    mod <- lm(log(pm25_ug_m3) ~ temperature_C + humidity_pct + pressure_hPa +
                light_lux + oxidised_kO + reduced_kO + nh3_kO,
              data = d)
    list(sensor = key$sensor, model = mod,
         glance = glance(mod) |> mutate(sensor = key$sensor, .before=1))
  })

map_dfr(lm_log_list, "glance") |>
  select(sensor, r.squared, adj.r.squared, nobs) |>
  kable(digits = 4, caption = "Log-model R² per sensor") |>
  kable_styling(full_width = FALSE)
```

## 8.4 Temporal Autocorrelation — ACF / PACF

Air quality measurements taken minutes apart are correlated — a high reading now
predicts a high reading in 5 minutes. This **autocorrelation** violates the
independence assumption of OLS and inflates false-positive rates in standard tests.
The ACF (autocorrelation function) and PACF (partial ACF) quantify this.

```{r acf-pacf, fig.height=3.5, results='asis'}
for (s in levels(df$sensor)) {
  cat(sprintf("\n\n#### Sensor: %s\n\n", s))
  pm_vec <- df |> filter(sensor == s) |>
    arrange(datetime) |> pull(pm25_ug_m3)
  par(mfrow = c(1, 2))
  acf(pm_vec,  lag.max = 48,
      main = paste("ACF — PM2.5 |", s),
      col  = "#0072B2", lwd = 1.5)
  pacf(pm_vec, lag.max = 48,
       main = paste("PACF — PM2.5 |", s),
       col  = "#D55E00", lwd = 1.5)
  par(mfrow = c(1, 1))
}
```

<div class="callout callout-concept">
**Reading the ACF/PACF:**

- **ACF** shows total correlation at each lag. A slowly decaying ACF indicates
  strong persistence (common in PM data).
- **PACF** shows *direct* correlation at each lag, controlling for shorter lags.
  If the PACF cuts off sharply after lag *p*, that suggests an AR(*p*) process
  would be an appropriate time-series model.
- Significant bars at lags 12 and 24 would suggest diurnal periodicity (if
  sampling is hourly) or 1-hour / 2-hour cycles (if sampling is every 5 min).
</div>

## 8.5 Strongest Relationships: PM2.5 vs Top Predictors

Using the regression results, we plot the two strongest predictors (by |t-statistic|)
for each sensor.

```{r top-predictors, fig.height=5}
top_preds <- map_dfr(lm_list, "tidy") |>
  filter(term != "(Intercept)") |>
  group_by(sensor) |>
  slice_max(abs(statistic), n = 2) |>
  ungroup()

# Build scatter plots for each sensor × top predictor pair
plots <- top_preds |>
  group_by(sensor) |>
  group_map(function(d_top, key) {
    s    <- key$sensor
    preds <- d_top$term
    map(preds, function(pred) {
      df |>
        filter(sensor == s) |>
        ggplot(aes(x = .data[[pred]], y = pm25_ug_m3)) +
        geom_point(alpha = 0.06, size = 0.8, color = pal[[s]]) +
        geom_smooth(method = "lm", color = "black",
                    linewidth = 0.8, se = TRUE) +
        labs(title    = paste(s, "—", pred),
             subtitle = sprintf("t = %.2f, p = %.4f",
                                d_top$statistic[d_top$term == pred],
                                d_top$p.value[d_top$term == pred]),
             x = pred, y = "PM2.5 (µg/m³)") +
        theme(plot.title    = element_text(size = 10),
              plot.subtitle = element_text(size = 8, color = "grey40"))
    })
  }) |>
  unlist(recursive = FALSE)

wrap_plots(plots, ncol = min(4, length(plots)))
```

---

# 9 · Summary and Conclusions {#conclusions}

```{r summary-table}
# Combine key statistics into one final summary
df |>
  group_by(sensor) |>
  summarise(
    Location      = first(loc_label),
    `N obs`       = n(),
    `Median PM2.5`= median(pm25_ug_m3),
    `Mean PM2.5`  = mean(pm25_ug_m3),
    `Median PM1`  = median(pm1_ug_m3),
    `% EPA Good`  = mean(pm25_ug_m3 <= 12) * 100,
    .groups = "drop"
  ) |>
  left_join(
    map_dfr(lm_list, "glance") |> select(sensor, adj.r.squared),
    by = "sensor"
  ) |>
  rename(`Regression adj-R²` = adj.r.squared) |>
  kable(digits = 2, caption = "Final cross-sensor summary") |>
  kable_styling(full_width = TRUE)
```

<div class="callout callout-tip">
**Return to hypotheses:**

- **H1 (Spatial):** Were there significant differences between sensors?
  Check the Wilcoxon results in Section 7.2.
- **H2 (Diurnal):** Did PM2.5 vary significantly by time of day?
  Check the Kruskal-Wallis results in Section 7.3 and the diurnal plot in 6.4.
- **H3 (Meteorological):** Which variables were the strongest predictors?
  Check the top-predictor plots in Section 8.5 and coefficients in 8.1.
</div>

---

# 10 · Session Info {#session}

```{r session-info}
sessionInfo()
```

<footer class="page-footer">
EA030 Field &amp; Lab Projects — P09
<span class="divider">|</span> Prof. Marc Los Huertos
<span class="divider">|</span> Pomona College
<span class="divider">|</span> Last updated: `r format(Sys.Date(), "%B %Y")`
<span class="divider">|</span> <a href="https://github.com/marclos/RTricks">GitHub</a>
</footer>
