---
title: "Climate Change Project"
author: "Marc Los Huertos"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Template Updated: 2025-08-30

This template cannot be run without making substantion changes. However, it does provide the skeleton and some of the code needed to complete the project, in particular the sourcing of the functions and the order of operations.

# Guide 1

## Chunk 1 Source Guide1functions.R

Load Guide Functions

```{r, Guide1_1}
source("/home/mwl04747/RTricks/05_Regional_Climate_Trends/Guides/Guide1functions.R")
```

## Chunk 2 Read 
  
Select State and Read Inventory of Stations. In my example, I have selected Nevada (NV).

```{r, chunk1_2}
my.state = "NV"
filename.csv <- "/home/mwl04747/RTricks/05_Regional_Climate_Trends/stations.active.oldest.csv"
my.inventory <- readInventory.fun(filename.csv, my.state)
```

## Chunk 3 Download Stations
```{r, chunk1_3, results='hide'}
datapath = "/home/mwl04747/RTricks/05_Regional_Climate_Trends/Data/FA25/"
downloadStations.fun(datapath, my.inventory)
```

# Guide 2

## Chunk 1 Source Guide2functions

```{r chunk2_1}
source("/home/mwl04747/RTricks/05_Regional_Climate_Trends/Guides/Guide2functions.R")
```

## Chunk 2 Read Stations
```{r, chunk2_2}
datafolder = "/home/mwl04747/RTricks/05_Regional_Climate_Trends/Data/SP25/"
ReadStations2.fun(datafolder)
```

## Chunk 3 Sort Stations --> Skip

```{r}
# sortstations.fun() Nor working yet. 
```

## Chunk 4 (Fix Date)
```{r, chunk2_4}
USW00024128a <- fixDates.fun(USW00024128)
#USW00014898a <- fixDates.fun(USW00014898)

```

## Chunk 5 Check Coverage
```{r, chunk2_5}
coverage.fun(USC00042294a)
```

## Chunk 6 Fix Values
```{r, chunk2_6}
USW00024128b <- fixValues.fun(USW00024128a)
# USW00014898b <- fixValues.fun(USW00014898a)
```

## Chunk 7 QAQC
```{r, chunk2_7}
QAQC.fun(USW00024128b)
```

## Chunk 8 # Monthly Means and Normals
```{r, chunk2_8}
USW00024128.monthly <- MonthlyValues.fun(USW00024128b)
USW00024128.normals <- MonthlyNormals.fun(USW00024128b)

# USW00014898.monthly <- MonthlyValues.fun(USW00014898b)
# USW00014898.normals <- MonthlyNormals.fun(USW00014898b)

```


## Chunk 9 Anomalies
```{r, chunk2_9}
USW00024128.anomalies <- MonthlyAnomalies.fun(USW00024128.monthly, USW00024128.normals)

#USC00042294.anomalies <- MonthlyAnomalies.fun(USC00042294.monthly, USC00042294.normals)
```


# Guide 3

## Chunk 1 Source Guide3fuctions

You might see a note about file not found, this is for Marc, you can ignore.
```{r, chunk3_1}
source("/home/mwl04747/RTricks/05_Regional_Climate_Trends/Guides/Guide3functions.R")
```

## Chunk 2 Analyze for Trends by Month
```{r chunk3_2}
USW00024128.trends <- monthlyTrend.fun(USW00024128.anomalies)
```

## Chunk 3 TBD
```{r}

```

# Guide 4

## Chunk 1 Source Guide4functions

You might see a note about file note found, this is for Marc, you can ignore.
```{r chunk4_1}
source("/home/mwl04747/RTricks/05_Regional_Climate_Trends/Guides/Guide4functions.R")
```

## Chunk 2 Plot Trend
```{r}
plotTrend2.fun(USW00024128.anomalies, "TMIN", 7)
```

almost done!


# Zip Files for Class

Marc's has created a function to zip all the costum functions and active stations file into a zip file for easy download/upload: Guides_Guidefunctions.zip.
```{r, echo=FALSE}
zip(zipfile = "/home/mwl04747/RTricks/05_Regional_Climate_Trends/Guides/Guides_Guidefunctions.zip", 
    files = c("/home/mwl04747/RTricks/05_Regional_Climate_Trends/Guides/Guide1functions.R",
      "/home/mwl04747/RTricks/05_Regional_Climate_Trends/Guides/Guide2functions.R",
      "/home/mwl04747/RTricks/05_Regional_Climate_Trends/Guides/Guide3functions.R",
      "/home/mwl04747/RTricks/05_Regional_Climate_Trends/Guides/0_Project_Overview_and_Set_Up.pdf",
      "/home/mwl04747/RTricks/05_Regional_Climate_Trends/Guides/1_Obtaining_State_Weather_Data.pdf",
      "/home/mwl04747/RTricks/05_Regional_Climate_Trends/Guides/2_Cleaning_and_Preprocessing_Station_Data.pdf",
      "/home/mwl04747/RTricks/05_Regional_Climate_Trends/Guides/3_Analyzing_and_Visualizaton_of_Climate_Trends.pdf",
      "/home/mwl04747/RTricks/05_Regional_Climate_Trends/Guides/MarcsTemplate.Rmd",      
      "/home/mwl04747/RTricks/05_Regional_Climate_Trends/stations.active.oldest.csv"),
    
    flags= "-j")
```


# Putting it together


```{r, eval=FALSE}
# Guide 1
source("/home/mwl04747/RTricks/05_Regional_Climate_Trends/Guides/Guide1functions.R") # 1_1

my.state = "MT"
filename.csv <- "/home/mwl04747/RTricks/05_Regional_Climate_Trends/stations.active.oldest.csv" # 1_1
my.inventory <- readInventory.fun(filename.csv, my.state) # 1_2
datapath = "/home/mwl04747/RTricks/04_Regional_Climate_Trends/Data/SP24/"
downloadStations.fun(datapath, my.inventory) # 1_3

# Guide 2
source("/home/mwl04747/RTricks/05_Regional_Climate_Trends/Guides/Guide2functions.R") # 2_1

datafolder = "/home/mwl04747/RTricks/05_Regional_Climate_Trends/Data/SP24/"
ReadStations2.fun(datafolder) # 2_2

#station = 	USC00246918
#stationID = "USW00054797"

	
USC00246918a <- fixDates.fun(USC00246918) # 2_4
coverage.fun(USC00246918a) # 2_5
USC00246918b <- fixValues.fun(USC00246918a) # 2_6
QAQC.fun(USC00246918b) # 2_7
USC00246918.monthly <- MonthlyValues.fun(USC00246918b) # 2_8
USC00246918.normals <- MonthlyNormals.fun(USC00246918b) # 2_8

USC00246918.anomalies <- MonthlyAnomalies.fun(USC00246918.monthly, USC00246918.normals) # 2_9

# Guide 3
source("/home/mwl04747/RTricks/04_Regional_Climate_Trends/Guides/Guide3functions.R") # 3_1
USC00246918.trends <- monthlyTrend.fun(USC00246918.anomalies) # 3_2
plotTrend2.fun(USC00246918.anomalies, "TMIN", 7) # 4_2


#Guide 4
#source("/home/mwl04747/RTricks/04_Regional_Climate_Trends/Guides/Guide4functions.R") # 4_1

```

















