---
title: "Data Visualization"
author: "Marc Los Huertos"
date: "EA030 · Compiled `r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    self_contained: true
    theme: null
    highlight: null
    toc: false
knit: (function(inputFile, encoding) {
    rmarkdown::render(
      inputFile,
      encoding = encoding,
      output_file = file.path(dirname(inputFile),
                              "../docs/basic_statistics/02_Visualization.html")
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 7,
  fig.height = 4
)
library(ggplot2)
library(dplyr)
library(tidyr)
```

```{css, echo=FALSE}
@import url('https://fonts.googleapis.com/css2?family=Libre+Baskerville:ital,wght@0,400;0,700;1,400&family=Source+Code+Pro:wght@400;600&family=Lato:wght@300;400;700&display=swap');

*, *::before, *::after { box-sizing: border-box; margin: 0; padding: 0; }

:root {
  --forest:    #1e3a2f;
  --moss:      #2d5a3d;
  --fern:      #4a7c5e;
  --sage:      #8aad96;
  --mist:      #d4e6da;
  --cream:     #f7f3ed;
  --parchment: #ede8df;
  --ink:       #1a1a18;
  --charcoal:  #3a3a38;
  --dust:      #7a7a72;
  --amber:     #c8832a;
  --amber-lt:  #e8a84a;
  --white:     #ffffff;
  --card-shadow: 0 2px 10px rgba(30,58,47,.09);
}

html { scroll-behavior: smooth; }
body {
  font-family: 'Lato', sans-serif;
  background: var(--cream);
  color: var(--ink);
  font-size: 17px;
  line-height: 1.75;
  padding-left:  clamp(14px, 4vw, 48px);
  padding-right: clamp(14px, 4vw, 48px);
  padding-bottom: 80px;
}
.main-container { max-width: 920px !important; margin: 0 auto !important; padding: 0 clamp(10px, 2vw, 32px) 80px !important; }
.page-header {
  margin-left:  calc(-1 * clamp(14px, 4vw, 48px));
  margin-right: calc(-1 * clamp(14px, 4vw, 48px));
}
h1.title, h4.author, h4.date { display: none; }

/* Page header */
.page-header {
  background-image:
    repeating-linear-gradient(120deg, transparent, transparent 60px, rgba(255,255,255,.018) 60px, rgba(255,255,255,.018) 61px),
    linear-gradient(160deg, #1e3a2f 0%, #2d5a3d 60%, #1a3d30 100%);
  color: #fff; padding: 44px 4vw 36px; margin: 0 -4vw 48px;
  border-bottom: 4px solid var(--amber); position: relative; overflow: hidden;
}
.page-header::after {
  content: ''; position: absolute; right: -40px; top: -40px;
  width: 260px; height: 260px; border-radius: 50%;
  background: radial-gradient(circle, rgba(138,173,150,.12) 0%, transparent 70%);
  pointer-events: none;
}
.page-header .module-tag {
  display: inline-block; font-family: 'Source Code Pro', monospace;
  font-size: .68rem; font-weight: 600; letter-spacing: .18em; text-transform: uppercase;
  color: var(--amber-lt); border: 1px solid rgba(200,131,42,.45);
  padding: 3px 10px; border-radius: 2px; margin-bottom: 12px;
}
.page-header h1 {
  font-family: 'Libre Baskerville', serif; font-size: clamp(1.6rem, 3vw, 2.3rem);
  font-weight: 700; color: #fff; line-height: 1.2; margin-bottom: 8px;
}
.page-header h1 span { color: var(--sage); }
.page-header .sub { font-size: .9rem; color: rgba(255,255,255,.65); font-weight: 300; }
.page-header .sub strong { color: rgba(255,255,255,.9); font-weight: 700; }
.back-link {
  display: inline-block; margin-top: 16px;
  font-family: 'Source Code Pro', monospace; font-size: .7rem; letter-spacing: .08em;
  color: var(--amber-lt); text-decoration: none;
  border: 1px solid rgba(200,131,42,.4); padding: 4px 11px; border-radius: 3px;
  transition: background .2s;
}
.back-link:hover { background: rgba(200,131,42,.18); }

/* Section nav pills */
.section-nav {
  display: flex; gap: 8px; flex-wrap: wrap; margin-bottom: 44px;
  padding-bottom: 20px; border-bottom: 1px solid var(--parchment);
}
.section-nav a {
  font-family: 'Source Code Pro', monospace; font-size: .7rem; font-weight: 600;
  letter-spacing: .08em; text-transform: uppercase;
  color: var(--fern); text-decoration: none;
  border: 1px solid var(--mist); padding: 4px 12px; border-radius: 3px;
  background: var(--white); transition: background .2s, color .2s;
}
.section-nav a:hover { background: var(--mist); color: var(--forest); }

/* Section headings */
h2 {
  font-family: 'Libre Baskerville', serif; font-size: 1.45rem; font-weight: 700;
  color: var(--forest); margin-top: 56px; margin-bottom: 6px;
  padding-bottom: 8px; border-bottom: 1px solid var(--parchment);
}
h3 {
  font-family: 'Libre Baskerville', serif; font-size: 1.1rem; font-weight: 700;
  color: var(--moss); margin-top: 28px; margin-bottom: 4px;
}

/* Body text */
p { margin-bottom: 1em; }
strong { color: var(--forest); }

/* Code */
code {
  font-family: 'Source Code Pro', monospace; font-size: .84em;
  background: var(--mist); color: var(--moss); padding: 1px 5px; border-radius: 3px;
}
pre {
  background: var(--ink); color: #e8e6df; border-radius: 5px;
  padding: 14px 18px; font-size: .83rem; line-height: 1.55;
  overflow-x: auto; border-left: 4px solid var(--amber); margin: 1.2em 0;
}
pre code { background: transparent; color: inherit; padding: 0; font-size: 1em; }

/* Callout boxes */
.callout {
  border-left: 5px solid; padding: .65em 1.1em;
  margin: 1.4em 0; border-radius: 0 5px 5px 0; font-size: .92em; line-height: 1.6;
}
.callout > strong:first-child { display: block; margin-bottom: .3em; }
.callout-concept { border-color: var(--fern);  background: #f0f7f2; }
.callout-tip     { border-color: #2e7d32;      background: #f1f8f1; }
.callout-warn    { border-color: #e65100;      background: #fff8f4; }
.callout-formula { border-color: var(--amber); background: #fdf8f0;
                   font-family: 'Source Code Pro', monospace; font-size: .88em; }

/* Plot type selector grid */
.plot-grid {
  display: grid;
  grid-template-columns: repeat(3, 1fr);
  gap: 10px; margin: 2em 0 2.5em;
}
.plot-card {
  background: var(--white); border: 1px solid var(--parchment);
  border-radius: 6px; padding: 14px 16px;
  box-shadow: var(--card-shadow);
}
.plot-card .pc-type {
  font-family: 'Source Code Pro', monospace; font-size: .72rem;
  font-weight: 600; letter-spacing: .1em; text-transform: uppercase;
  color: var(--amber); margin-bottom: 4px;
}
.plot-card .pc-name {
  font-family: 'Libre Baskerville', serif; font-size: .95rem;
  font-weight: 700; color: var(--forest); margin-bottom: 3px;
}
.plot-card .pc-when { font-size: .82rem; color: var(--charcoal); line-height: 1.45; }
.plot-card a { color: var(--fern); font-size: .78rem; text-decoration: none; }

/* Key concept summary boxes */
.key-box {
  background: var(--forest); color: #fff; border-radius: 6px;
  padding: 20px 24px; margin: 2em 0; font-size: .9rem;
}
.key-box h4 {
  font-family: 'Libre Baskerville', serif; color: var(--sage);
  font-size: 1rem; margin-bottom: 10px;
}
.key-box ul { padding-left: 1.3em; }
.key-box li { margin-bottom: 5px; color: rgba(255,255,255,.85); }
.key-box code { background: rgba(255,255,255,.12); color: var(--amber-lt); }

/* Do / Don't side-by-side */
.do-dont { display: grid; grid-template-columns: 1fr 1fr; gap: 12px; margin: 1.4em 0; }
.do-box, .dont-box {
  padding: .6em 1em; border-radius: 5px; font-size: .88em; line-height: 1.55;
}
.do-box   { background: #f1f8f1; border-left: 5px solid #2e7d32; }
.dont-box { background: #fff8f4; border-left: 5px solid #e65100; }
.do-box strong   { color: #1b5e20; display: block; margin-bottom: .25em; }
.dont-box strong { color: #bf360c; display: block; margin-bottom: .25em; }

/* Tables */
table { font-size: .9rem; border-collapse: collapse; width: 100%; margin: 1.2em 0; }
thead th {
  background: var(--forest); color: #fff;
  font-family: 'Source Code Pro', monospace; font-size: .76rem;
  letter-spacing: .06em; text-transform: uppercase; padding: 8px 12px; text-align: left;
}
tbody tr:nth-child(even) { background: var(--mist); }
tbody td { padding: 7px 12px; border-bottom: 1px solid var(--parchment); }

/* Footer */
.page-footer {
  background: var(--ink); color: rgba(255,255,255,.42); text-align: center;
  padding: 24px 4vw; margin: 60px -4vw 0;
  font-size: .76rem; font-family: 'Source Code Pro', monospace; letter-spacing: .04em;
}
.page-footer a { color: var(--sage); text-decoration: none; }
.page-footer .divider { margin: 0 8px; opacity: .3; }
```

<div class="page-header">
<div class="module-tag">EA030 &mdash; Basic Statistics &mdash; Module 03</div>
<h1>Data <span>Visualization</span></h1>
<p class="sub"><strong>Prof. Marc Los Huertos</strong> &nbsp;&bull;&nbsp; Pomona College</p>
<a class="back-link" href="index.html">&#8592; Back to Basic Statistics</a>
</div>

<div class="section-nav">
<a href="#why">Why Visualize?</a>
<a href="#grammar">Grammar of Graphics</a>
<a href="#one-continuous">One Continuous</a>
<a href="#one-categorical">One Categorical</a>
<a href="#cont-by-group">Continuous × Group</a>
<a href="#two-continuous">Two Continuous</a>
<a href="#time-series">Time Series</a>
<a href="#design">Design Principles</a>
<a href="#choosing">Choosing a Plot</a>
</div>

> **What this module covers.** A well-chosen graph can reveal structure that
> no summary statistic captures — and an ill-chosen one can hide it. This module
> builds a systematic toolkit for data visualization in R using `ggplot2`, from
> the underlying *grammar of graphics* to the practical decisions that make a
> plot honest, clear, and publication-ready.

---

## 1. Why Visualize Before You Model? {#why}

Anscombe's Quartet (1973) is the classic demonstration that four datasets with
*identical* summary statistics can look completely different. The same principle
applies constantly in environmental data: means and standard deviations are not
enough. 

[]

```{r anscombe-data}
# Anscombe's Quartet is built into base R
data(anscombe)

# Reshape to long format for ggplot
ansc_long <- data.frame(
  set = rep(1:4, each = 11),
  x   = c(anscombe$x1, anscombe$x2, anscombe$x3, anscombe$x4),
  y   = c(anscombe$y1, anscombe$y2, anscombe$y3, anscombe$y4)
)
```

```{r anscombe-stats}
# All four datasets have (nearly) identical summary statistics
ansc_long |>
  group_by(set) |>
  summarise(
    mean_x = round(mean(x), 2),
    mean_y = round(mean(y), 2),
    sd_x   = round(sd(x),   2),
    sd_y   = round(sd(y),   2),
    r      = round(cor(x, y), 2)
  ) |>
  knitr::kable(
    col.names = c("Dataset", "Mean X", "Mean Y", "SD X", "SD Y", "Correlation r"),
    caption   = "Anscombe's Quartet: four structurally different datasets with identical summary statistics."
  )
```

```{r anscombe-plot, echo=FALSE, fig.cap="Anscombe's Quartet. Identical means, SDs, and correlations — but completely different structures. Dataset 1 alone suits linear regression; the others reveal a curve, an outlier, and a leverage point. Without plotting, you would never know.", fig.height=5}
ggplot(ansc_long, aes(x, y)) +
  geom_point(colour = "#4a7c5e", size = 2.2, alpha = 0.85) +
  geom_smooth(method = "lm", se = FALSE,
              colour = "#c8832a", linewidth = 0.9, linetype = "dashed") +
  facet_wrap(~ paste("Dataset", set), nrow = 2) +
  labs(x = "X", y = "Y",
       title = "Anscombe's Quartet: Always Plot Your Data") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title   = element_text(family = "serif", face = "bold",
                                colour = "#1e3a2f", size = 13),
    strip.text   = element_text(family = "mono", size = 9,
                                colour = "#2d5a3d", face = "bold"),
    panel.border = element_rect(colour = "#d4e6da", fill = NA)
  )
```

<div class="callout callout-concept">
<strong>📌 Visualize → Summarize → Model</strong>

This is the correct order of operations. Plotting first lets you:
detect outliers and data entry errors before they corrupt an analysis;
check whether a linear model is even appropriate; identify subgroups or
bimodal distributions that a single mean obscures; and communicate results
to an audience that cannot read a regression table.
</div>

---

## 2. The Grammar of Graphics {#grammar}

`ggplot2` is built on a systematic *grammar* — a set of independent components
that can be composed to describe any statistical graphic. Understanding the
grammar means you are never guessing syntax; you are assembling known parts.

### The seven layers

Every `ggplot2` plot is built from up to seven layers stacked on top of each other:

```{r grammar-table, echo=FALSE}
knitr::kable(
  data.frame(
    Layer       = c("**Data**", "**Aesthetics** `aes()`", "**Geometry** `geom_*()`",
                    "**Statistics** `stat_*()`", "**Scales** `scale_*()`",
                    "**Facets** `facet_*()`", "**Theme** `theme()`"),
    Purpose     = c("The data frame being plotted",
                    "Map variables to visual properties: x, y, colour, fill, size, shape",
                    "The geometric object: point, line, bar, boxplot, ribbon…",
                    "Statistical transformation before drawing: bin, smooth, summarise",
                    "Control axis limits, colour palettes, axis labels",
                    "Split the plot into small multiples by a grouping variable",
                    "All non-data ink: fonts, grid lines, background, margins"),
    Example     = c("`data = river_df`",
                    "`aes(x = temp, y = DO, colour = site)`",
                    "`geom_point()`, `geom_line()`, `geom_boxplot()`",
                    "`stat_smooth()`, `stat_bin()`",
                    "`scale_colour_manual()`, `scale_x_log10()`",
                    "`facet_wrap(~ year)`, `facet_grid(site ~ season)`",
                    "`theme_minimal()`, `theme(legend.position = 'bottom')`")
  ),
  col.names = c("Layer", "Purpose", "Example"),
  escape    = FALSE
)
```

### Minimal working example

```{r grammar-example, fig.cap="A minimal ggplot2 scatterplot using only three required layers: data, aesthetics, and geometry. Every additional layer is optional refinement.", fig.height=3.5}
set.seed(1)
river <- data.frame(
  temp_C = runif(50, 12, 24),
  DO_mgl = 14 - 0.35 * runif(50, 12, 24) + rnorm(50, 0, 0.5),
  site   = sample(c("Upstream", "Downstream"), 50, replace = TRUE)
)

# Layer 1: data.  Layer 2: aesthetics.  Layer 3: geometry.
ggplot(data = river,                        # Layer 1 — data
       aes(x = temp_C, y = DO_mgl,          # Layer 2 — aesthetics
           colour = site)) +
  geom_point(size = 2.5, alpha = 0.8)       # Layer 3 — geometry
```

Now add refinement layers one at a time:

```{r grammar-full, fig.cap="The same plot with scales, a smoother, and a theme added. Each layer is independent — remove any one without breaking the others.", fig.height=4}
ggplot(river, aes(x = temp_C, y = DO_mgl, colour = site)) +
  geom_point(size = 2.2, alpha = 0.75) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8) +   # stat layer
  scale_colour_manual(                                          # scale layer
    values = c("Upstream" = "#4a7c5e", "Downstream" = "#c8832a"),
    name   = "Site"
  ) +
  labs(                                                         # labels
    x     = "Water temperature (°C)",
    y     = "Dissolved oxygen (mg/L)",
    title = "DO vs. Temperature — Two River Sites"
  ) +
  theme_minimal(base_size = 12) +                              # theme layer
  theme(
    plot.title  = element_text(family = "serif", face = "bold",
                               colour = "#1e3a2f"),
    legend.position = "top"
  )
```

<div class="callout callout-formula">
<strong>⚙️ The ggplot2 template — memorise this</strong><br>
<code>ggplot(data, aes(...)) + geom_*() + scale_*() + labs() + theme_*()</code><br><br>
Only <code>ggplot()</code> and one <code>geom_*()</code> are required.
Everything else is optional. Add layers with <code>+</code>, never <code>|></code>.
</div>

---

## 3. One Continuous Variable {#one-continuous}

When you have a single continuous measurement, your first goal is to understand
its **distribution** — shape, centre, spread, and any unusual features.

### Histogram

The histogram bins the data and plots counts (or density) per bin. The choice
of bin width is consequential: too wide hides structure, too narrow produces noise.

```{r hist-data}
set.seed(5)
# Daily maximum ozone (ppb) at a monitoring station — one year
ozone <- data.frame(
  ppb  = c(rnorm(200, mean = 38, sd = 9),    # typical summer days
           rnorm(165, mean = 22, sd = 6)),    # winter / clean days
  season = c(rep("Summer", 200), rep("Winter", 165))
)
```

```{r histogram, fig.cap="Distribution of daily maximum ozone. Using `bins = 30` (the default) vs. a carefully chosen `binwidth` — the narrower bins reveal the bimodal structure that the coarser histogram obscures.", fig.height=4, fig.width=7}
p1 <- ggplot(ozone, aes(ppb)) +
  geom_histogram(bins = 10, fill = "#4a7c5e", colour = "white", alpha = 0.9) +
  labs(title = "bins = 10 (too coarse)", x = "Ozone (ppb)", y = "Count") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(family = "serif", colour = "#1e3a2f"))

p2 <- ggplot(ozone, aes(ppb)) +
  geom_histogram(binwidth = 2, fill = "#4a7c5e", colour = "white", alpha = 0.9) +
  labs(title = "binwidth = 2 (reveals bimodality)", x = "Ozone (ppb)", y = "Count") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(family = "serif", colour = "#1e3a2f"))

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

### Density plot

A kernel density estimate (KDE) is a smoothed histogram — useful when you want
to compare distributions without the bin-width problem. Overlay multiple groups
using transparency (`alpha`).

```{r density, fig.cap="Kernel density estimates of ozone by season. Overlapping semi-transparent fills allow direct comparison without stacking. The bimodal overall distribution is clearly driven by the two seasons.", fig.height=4}
ggplot(ozone, aes(ppb, fill = season, colour = season)) +
  geom_density(alpha = 0.35, linewidth = 0.9) +
  scale_fill_manual(values  = c("Summer" = "#c8832a", "Winter" = "#4a7c5e")) +
  scale_colour_manual(values = c("Summer" = "#c8832a", "Winter" = "#4a7c5e")) +
  labs(x = "Ozone (ppb)", y = "Density",
       title  = "Ozone Distribution by Season",
       fill   = NULL, colour = NULL) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title      = element_text(family = "serif", face = "bold", colour = "#1e3a2f"),
    legend.position = "top"
  )
```

<div class="callout callout-tip">
<strong>💡 Histogram vs. density plot</strong>

Use a **histogram** when showing actual counts matters — for instance, to
communicate sample size visually alongside distribution shape. Use a
**density plot** when overlaying two or more groups for comparison; raw
count histograms of different-sized groups are visually misleading.
To get the best of both: `geom_histogram(aes(y = after_stat(density)))` puts
histograms on the density scale so an overlaid `geom_density()` aligns correctly.
</div>

---

## 4. One Categorical Variable {#one-categorical}

When you need to show counts or proportions across categories, the bar chart
is the natural choice — provided you follow a few conventions that prevent
distortion.

```{r bar-data}
set.seed(8)
# Land-cover classification from remote sensing (n = 300 parcels)
landcover <- data.frame(
  class = sample(
    c("Forest", "Shrubland", "Grassland", "Wetland", "Developed"),
    size = 300, replace = TRUE,
    prob = c(0.34, 0.22, 0.18, 0.12, 0.14)
  )
)
```

```{r barplot, fig.cap="Land-cover class frequencies. Categories are sorted by count (most to least common) using `fct_infreq()`. A sorted bar chart lets the reader rank categories at a glance; alphabetical ordering forces the eye to search.", fig.height=4}
landcover |>
  count(class) |>
  mutate(class = forcats::fct_reorder(class, n, .desc = TRUE)) |>
  ggplot(aes(class, n, fill = class)) +
  geom_col(width = 0.6, colour = "#1e3a2f", alpha = 0.88) +
  geom_text(aes(label = n), vjust = -0.4,
            family = "mono", size = 3.5, colour = "#1e3a2f") +
  scale_fill_manual(values = c(
    "Forest"    = "#1e3a2f",
    "Shrubland" = "#4a7c5e",
    "Grassland" = "#8aad96",
    "Wetland"   = "#2e7d94",
    "Developed" = "#c8832a"
  )) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(x = NULL, y = "Number of parcels",
       title = "Land-Cover Classification — 300 Parcels") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title      = element_text(family = "serif", face = "bold", colour = "#1e3a2f"),
    legend.position = "none",
    panel.grid.major.x = element_blank()
  )
```

<div class="callout callout-warn">
<strong>⚠️ The "bar chart with error bars" problem</strong>

A bar chart of means with error bars — sometimes called a "dynamite plot" —
is one of the most overused and misleading plot types in ecology. The bar
implies that data begin at zero, which focuses attention on the wrong feature
(the mean height, not the spread). It also hides whether the distribution is
symmetric, skewed, bimodal, or contains outliers. See Section 5 for better
alternatives when the goal is to compare group means.
</div>

---

## 5. Continuous Variable by Group {#cont-by-group}

This is the most common plotting task in environmental science: comparing a
measurement across two or more groups. A family of plot types exists, each
with different strengths.

```{r group-data}
set.seed(14)
# Macroinvertebrate EPT richness (sensitive taxa) at four stream sites
ept <- data.frame(
  site     = rep(c("Reference", "Urban", "Agricultural", "Mining"), each = 30),
  richness = c(
    round(rnorm(30, 12, 2.1)),   # Reference  — highest quality
    round(rnorm(30,  7, 2.8)),   # Urban
    round(rnorm(30,  9, 2.4)),   # Agricultural
    round(rnorm(30,  4, 1.9))    # Mining — most impaired
  )
) |>
  mutate(site = factor(site,
    levels = c("Reference", "Agricultural", "Urban", "Mining")))
```

### Boxplot — the workhorse

Shows median, IQR, whiskers (1.5×IQR), and outliers. Always add the raw data
as jittered points so the reader sees both distribution and sample size.

```{r boxplot, fig.cap="EPT richness across four stream sites. Boxes show the IQR; horizontal bars the median; jittered points show all individual observations. Colour is redundant with the x-axis position — it aids quick group identification.", fig.height=4.5}
ggplot(ept, aes(site, richness, fill = site)) +
  geom_boxplot(width = 0.45, outlier.shape = NA,     # hide outlier points —
               colour = "#1e3a2f", alpha = 0.75,     # we show all raw data below
               linewidth = 0.6) +
  geom_jitter(aes(colour = site), width = 0.18,
              size = 1.6, alpha = 0.45) +
  scale_fill_manual(values = c(
    "Reference"    = "#1e3a2f",
    "Agricultural" = "#8aad96",
    "Urban"        = "#c8832a",
    "Mining"       = "#7a3a2f"
  )) +
  scale_colour_manual(values = c(
    "Reference"    = "#1e3a2f",
    "Agricultural" = "#4a7c5e",
    "Urban"        = "#c8832a",
    "Mining"       = "#7a3a2f"
  )) +
  labs(x = NULL, y = "EPT richness (no. of taxa)",
       title = "Macroinvertebrate EPT Richness by Stream Site") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title         = element_text(family = "serif", face = "bold", colour = "#1e3a2f"),
    legend.position    = "none",
    panel.grid.major.x = element_blank()
  )
```

### Violin plot — shows the full distribution shape

When sample size is large enough (n ≥ 25 per group), a violin plot is superior
to a boxplot because it reveals bimodality and skewness that the box hides.

```{r violin, fig.cap="Violin plots of EPT richness. The width of each violin at any given y-value is proportional to the density of observations at that value. The embedded boxplot preserves the median and IQR summary.", fig.height=4.5}
ggplot(ept, aes(site, richness, fill = site)) +
  geom_violin(trim = TRUE, colour = "#1e3a2f",
              alpha = 0.7, linewidth = 0.5) +
  geom_boxplot(width = 0.08, fill = "white",
               colour = "#1e3a2f", outlier.shape = NA, linewidth = 0.7) +
  scale_fill_manual(values = c(
    "Reference"    = "#1e3a2f",
    "Agricultural" = "#8aad96",
    "Urban"        = "#c8832a",
    "Mining"       = "#7a3a2f"
  )) +
  labs(x = NULL, y = "EPT richness (no. of taxa)",
       title = "Violin + Boxplot: Full Distribution Shape") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title         = element_text(family = "serif", face = "bold", colour = "#1e3a2f"),
    legend.position    = "none",
    panel.grid.major.x = element_blank()
  )
```

### Point + CI plot — for inference

When your goal is *comparison of means* (i.e., you are about to run a t-test or
ANOVA), show the mean and 95% confidence interval explicitly. This is more honest
than a "dynamite plot" and more focused than a boxplot.

```{r point-ci, fig.cap="Mean EPT richness ± 95% CI by site. Non-overlapping intervals between Reference and Mining provide strong visual evidence of a meaningful difference. This plot directly mirrors what a t-test or ANOVA tests.", fig.height=4}
ept_summary <- ept |>
  group_by(site) |>
  summarise(
    n    = n(),
    mean = mean(richness),
    se   = sd(richness) / sqrt(n),
    t_crit = qt(0.975, df = n - 1),
    lower  = mean - t_crit * se,
    upper  = mean + t_crit * se
  )

ggplot(ept_summary, aes(site, mean, colour = site)) +
  geom_hline(yintercept = ept_summary$mean[ept_summary$site == "Reference"],
             linetype = "dashed", colour = "#aaaaaa", linewidth = 0.6) +
  geom_errorbar(aes(ymin = lower, ymax = upper),
                width = 0.15, linewidth = 1.1) +
  geom_point(size = 4) +
  scale_colour_manual(values = c(
    "Reference"    = "#1e3a2f",
    "Agricultural" = "#4a7c5e",
    "Urban"        = "#c8832a",
    "Mining"       = "#7a3a2f"
  )) +
  labs(x = NULL, y = "Mean EPT richness ± 95% CI",
       title = "Mean EPT Richness with 95% Confidence Intervals") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title      = element_text(family = "serif", face = "bold", colour = "#1e3a2f"),
    legend.position = "none"
  )
```

<div class="callout callout-concept">
<strong>📌 When to use each group-comparison plot</strong>

<strong>Boxplot:</strong> default choice for most ecological data; works well for n ≥ 10 per group; always layer raw data on top.<br>
<strong>Violin:</strong> upgrade when n ≥ 25 and distribution shape matters — detecting bimodality in species abundance, for instance.<br>
<strong>Point + CI:</strong> use when your primary message is inferential — "these groups are or are not significantly different". Pairs naturally with ANOVA results.
</div>

---

## 6. Two Continuous Variables {#two-continuous}

### Scatterplot with smoother

Use `geom_smooth()` to overlay a trend line. The default method `"loess"` is
non-parametric and reveals non-linear patterns. Use `method = "lm"` when you
are specifically asserting a linear relationship.

```{r scatter-data}
set.seed(21)
n <- 80
soil <- data.frame(
  organic_pct = runif(n, 1, 8),
  moisture    = NA,
  bulk_density = NA
)
soil$moisture     <- 12 + 4.5 * soil$organic_pct + rnorm(n, 0, 3)
soil$bulk_density <- 1.85 - 0.08 * soil$organic_pct + rnorm(n, 0, 0.08)
soil$land_use     <- sample(c("Forest", "Cropland"), n, replace = TRUE,
                             prob = c(0.5, 0.5))
```

```{r scatter-loess, fig.cap="Soil moisture increases with organic matter content. The loess smoother (blue) reveals a slightly non-linear relationship — the slope flattens at high organic matter. The linear fit (dashed) would miss this. Point colour encodes land use.", fig.height=4.5}
ggplot(soil, aes(organic_pct, moisture, colour = land_use)) +
  geom_point(size = 2.2, alpha = 0.7) +
  geom_smooth(aes(group = 1), method  = "loess", span = 0.8,
              colour = "#1e3a2f", fill = "#d4e6da",
              linewidth = 1, se = TRUE) +
  geom_smooth(aes(group = 1), method = "lm", se = FALSE,
              colour = "#c8832a", linetype = "dashed",
              linewidth = 0.8) +
  scale_colour_manual(
    values = c("Forest" = "#4a7c5e", "Cropland" = "#c8832a"),
    name   = "Land use"
  ) +
  labs(
    x     = "Soil organic matter (%)",
    y     = "Volumetric moisture (%)",
    title = "Soil Moisture vs. Organic Matter"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title      = element_text(family = "serif", face = "bold", colour = "#1e3a2f"),
    legend.position = "top"
  )
```

### Small multiples (facets)

When you have a third grouping variable, `facet_wrap()` produces **small
multiples** — the same axes and scales repeated across groups, which makes
cross-group comparison easier than overlaying everything on one plot.

```{r facets, fig.cap="The organic matter–moisture relationship faceted by land use. Keeping the axes identical across panels allows direct comparison of both slope and intercept. Cropland soils retain less moisture at a given organic matter level.", fig.height=4}
ggplot(soil, aes(organic_pct, moisture)) +
  geom_point(aes(colour = land_use), size = 1.9, alpha = 0.65) +
  geom_smooth(method = "lm", colour = "#1e3a2f",
              fill = "#d4e6da", linewidth = 1) +
  facet_wrap(~ land_use) +
  scale_colour_manual(values = c("Forest" = "#4a7c5e", "Cropland" = "#c8832a")) +
  labs(x = "Soil organic matter (%)", y = "Volumetric moisture (%)") +
  theme_minimal(base_size = 12) +
  theme(
    strip.text      = element_text(family = "mono", face = "bold",
                                   colour = "#1e3a2f", size = 10),
    legend.position = "none"
  )
```

### Correlation matrix for exploratory analysis

When you have several continuous variables and want a quick overview of their
pairwise relationships, a pairs plot is invaluable.

```{r pairs, fig.cap="Lower triangle: scatterplots with loess smoothers. Diagonal: variable name. Upper triangle would show correlation coefficients (see `GGally::ggpairs()` for a polished version). Bulk density is inversely related to both moisture and organic matter — classic compaction signal.", fig.height=5, fig.width=6}
pairs(
  soil[, c("organic_pct", "moisture", "bulk_density")],
  pch    = 20, cex = 0.7,
  col    = ifelse(soil$land_use == "Forest", "#4a7c5e", "#c8832a"),
  labels = c("Organic\nmatter (%)", "Moisture\n(%)", "Bulk\ndensity\n(g/cm³)"),
  panel  = function(x, y, ...) {
    points(x, y, ...)
    lines(lowess(x, y), col = "#1e3a2f", lwd = 1.5)
  },
  main = "Soil Variable Pairs"
)
```

---

## 7. Time Series {#time-series}

Environmental data are often collected over time. Connecting observations with
a line communicates temporal continuity; adding points marks the actual
measurement occasions.

```{r ts-data}
set.seed(33)
# Monthly mean streamflow (m³/s) — 5 years of a hypothetical CA watershed
months <- seq(as.Date("2019-01-01"), as.Date("2023-12-01"), by = "month")
n_mo   <- length(months)

# Seasonal signal + long-term decline + noise
seasonal   <- 12 * sin((seq_len(n_mo) - 2) * 2 * pi / 12) + 8
trend      <- seq(0, -4, length.out = n_mo)
streamflow <- data.frame(
  date  = months,
  flow  = pmax(0.2, seasonal + trend + rnorm(n_mo, 0, 2)),
  year  = format(months, "%Y")
)
```

```{r ts-annual, fig.cap="Monthly mean streamflow over five years. The line reveals the annual wet-season pulse; points mark individual monthly measurements. The declining trend is subtle but visible by comparing peak flows across years.", fig.height=4}
ggplot(streamflow, aes(date, flow)) +
  geom_area(fill = "#d4e6da", colour = NA, alpha = 0.6) +
  geom_line(colour = "#2d5a3d", linewidth = 0.85) +
  geom_point(colour = "#1e3a2f", size = 1.2, alpha = 0.6) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(
    x     = NULL,
    y     = "Mean monthly streamflow (m³/s)",
    title = "Streamflow — Hypothetical CA Watershed, 2019–2023"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title      = element_text(family = "serif", face = "bold", colour = "#1e3a2f"),
    panel.grid.minor = element_blank()
  )
```

### Seasonal decomposition plot

Faceting by year allows direct visual comparison of the seasonal cycle
across years — much clearer than trying to read inter-annual differences
from the single time series above.

```{r ts-faceted, fig.cap="Streamflow by month, faceted by year. Each panel shares the same x-axis (month of year) and y-axis scale, making it easy to compare peak-flow timing and magnitude across years. The declining late-season flow is especially clear.", fig.height=4}
streamflow |>
  mutate(month_num = as.integer(format(date, "%m")),
         month_lbl = format(date, "%b")) |>
  mutate(month_lbl = factor(month_lbl, levels = month.abb)) |>
  ggplot(aes(month_lbl, flow, group = year, colour = year)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.8) +
  scale_colour_manual(
    values = c("2019" = "#1e3a2f", "2020" = "#2d5a3d",
               "2021" = "#4a7c5e", "2022" = "#8aad96", "2023" = "#c8832a"),
    name = "Year"
  ) +
  labs(x = NULL, y = "Streamflow (m³/s)",
       title = "Monthly Streamflow by Year") +
  theme_minimal(base_size = 11) +
  theme(
    plot.title      = element_text(family = "serif", face = "bold", colour = "#1e3a2f"),
    axis.text.x     = element_text(size = 8),
    legend.position = "top"
  )
```

---

## 8. Design Principles {#design}

Technical correctness is necessary but not sufficient. A plot also needs to
communicate clearly. These principles have empirical support from perceptual
research.

### Colour

Colour is the most misused visual channel. Three rules cover most situations:

<div class="do-dont">
<div class="do-box">
<strong>✅ Do</strong>
Use a <strong>qualitative</strong> palette (e.g., ColorBrewer Set2) for nominal categories
where no ordering exists. Use a <strong>sequential</strong> palette (dark→light) for
ordered or quantitative data. Use <strong>diverging</strong> palettes (two hues meeting
at a neutral midpoint) for data centred on a meaningful zero, like anomalies
from a long-term average. Test your palette with a colour-blindness simulator.
</div>
<div class="dont-box">
<strong>❌ Don't</strong>
Use the default rainbow palette — it has no perceptual order and is
uninterpretable to red-green colourblind readers (~8% of men).
Never use colour alone to encode a data dimension; also vary shape or line
type so the plot is legible in black-and-white print and to colourblind readers.
Do not use more than ~6–7 distinct colours in one plot.
</div>
</div>

```{r colour-demo, fig.cap="Left: rainbow palette — perceptually non-monotone, inaccessible. Right: ColorBrewer 'Dark2' qualitative palette — balanced, colourblind-friendly. Both show the same EPT richness data.", fig.height=3.5, fig.width=7}
p_bad <- ggplot(ept, aes(site, richness, fill = site)) +
  geom_boxplot(outlier.shape = NA, linewidth = 0.5) +
  scale_fill_manual(values = c("red", "orange", "green", "blue")) +
  labs(title = "❌ Rainbow colours", x = NULL, y = "EPT richness") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none",
        plot.title = element_text(colour = "#bf360c", face = "bold"))

p_good <- ggplot(ept, aes(site, richness, fill = site)) +
  geom_boxplot(outlier.shape = NA, linewidth = 0.5) +
  scale_fill_brewer(palette = "Dark2") +
  labs(title = "✅ ColorBrewer 'Dark2'", x = NULL, y = "EPT richness") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none",
        plot.title = element_text(colour = "#1b5e20", face = "bold"))

gridExtra::grid.arrange(p_bad, p_good, ncol = 2)
```

### Axis scales and the zero baseline

```{r axes-demo, fig.cap="Left: y-axis truncated at 3 — the Urban–Mining difference looks enormous. Right: y-axis starting at 0 shows the true proportion of the range occupied by the difference. Truncating axes exaggerates effects and can mislead readers.", fig.height=3.5, fig.width=7}
p_trunc <- ggplot(ept_summary, aes(site, mean, colour = site)) +
  geom_errorbar(aes(ymin = lower, ymax = upper),
                width = 0.15, linewidth = 1) +
  geom_point(size = 3.5) +
  coord_cartesian(ylim = c(3, 14)) +
  scale_colour_manual(values = c("#1e3a2f","#4a7c5e","#c8832a","#7a3a2f")) +
  labs(title = "❌ Truncated y-axis", x = NULL, y = "Mean EPT richness") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none",
        plot.title = element_text(colour = "#bf360c", face = "bold"))

p_zero <- ggplot(ept_summary, aes(site, mean, colour = site)) +
  geom_errorbar(aes(ymin = lower, ymax = upper),
                width = 0.15, linewidth = 1) +
  geom_point(size = 3.5) +
  expand_limits(y = 0) +
  scale_colour_manual(values = c("#1e3a2f","#4a7c5e","#c8832a","#7a3a2f")) +
  labs(title = "✅ Axis from zero", x = NULL, y = "Mean EPT richness") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none",
        plot.title = element_text(colour = "#1b5e20", face = "bold"))

gridExtra::grid.arrange(p_trunc, p_zero, ncol = 2)
```

<div class="callout callout-warn">
<strong>⚠️ The zero-baseline exception</strong>

The "start at zero" rule applies to <strong>bar charts and area charts</strong>,
where the encoded quantity is the bar's height or the filled area.
For <strong>point-and-line plots and scatterplots</strong>, forcing a zero baseline
often wastes most of the plot area and obscures variation that is scientifically
meaningful. Judgment is required; the key is that the axis range should not
exaggerate or minimise the effect relative to what is natural for the variable.
</div>

### The data-ink ratio

Edward Tufte's principle: every drop of ink should encode data. Remove
gridlines, borders, and tick marks that do not help the reader.

```{r theme-demo, fig.cap="Left: default ggplot2 grey theme — heavy grid, bold background, double-encoded axis. Right: `theme_minimal()` with further refinement — the same data, more signal, less noise.", fig.height=3.5, fig.width=7}
base_plot <- ggplot(soil, aes(organic_pct, moisture)) +
  geom_point(colour = "#4a7c5e", alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, colour = "#1e3a2f") +
  labs(x = "Organic matter (%)", y = "Moisture (%)")

p_default <- base_plot +
  labs(title = "❌ Default grey theme") +
  theme_grey(base_size = 11) +
  theme(plot.title = element_text(colour = "#bf360c", face = "bold"))

p_clean <- base_plot +
  labs(title = "✅ Minimal theme, refined") +
  theme_minimal(base_size = 11) +
  theme(
    plot.title       = element_text(family = "serif", face = "bold",
                                    colour = "#1b5e20"),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "#ede8df", linewidth = 0.5),
    axis.line        = element_line(colour = "#3a3a38", linewidth = 0.4)
  )

gridExtra::grid.arrange(p_default, p_clean, ncol = 2)
```

### Labelling and annotation

Axis labels must include **units**. Titles should state the finding, not just
the variable names. Annotations (`annotate()`, `geom_text()`) can replace
or supplement a legend, reducing eye travel.

```{r annotation, fig.cap="Direct annotation replaces the legend entirely — the reader's eye never has to leave the plot. The title states the finding ('Moisture increases with organic matter') rather than just naming the axes.", fig.height=4}
soil_means <- soil |>
  group_by(land_use) |>
  summarise(x = mean(organic_pct), y = mean(moisture))

ggplot(soil, aes(organic_pct, moisture, colour = land_use)) +
  geom_point(size = 2, alpha = 0.55) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  geom_text(data = soil_means,
            aes(x = x + 0.3, y = y + 2, label = land_use, colour = land_use),
            fontface = "bold", family = "mono", size = 3.8) +
  scale_colour_manual(values = c("Forest" = "#1e3a2f", "Cropland" = "#c8832a")) +
  labs(
    x     = "Soil organic matter (%)",
    y     = "Volumetric moisture (%)",
    title = "Moisture increases with organic matter — more strongly in forest soils"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title      = element_text(family = "serif", face = "bold",
                                   colour = "#1e3a2f", size = 11),
    legend.position = "none"
  )
```

---

## 9. Choosing the Right Plot {#choosing}

```{r choice-table, echo=FALSE}
knitr::kable(
  data.frame(
    Goal = c(
      "Understand distribution of one variable",
      "Understand distribution of one variable",
      "Count observations per category",
      "Compare distributions across groups",
      "Compare distributions across groups (large n)",
      "Compare means across groups (inference)",
      "Show relationship between two continuous variables",
      "Show three variables simultaneously",
      "Show change over time",
      "Compare many variables pairwise"
    ),
    `Best choice` = c(
      "Histogram (`geom_histogram()`)",
      "Density plot (`geom_density()`)",
      "Bar chart, sorted (`geom_col()` + `fct_reorder()`)",
      "Boxplot + jitter (`geom_boxplot()` + `geom_jitter()`)",
      "Violin + embedded boxplot (`geom_violin()` + `geom_boxplot()`)",
      "Point + CI (`geom_point()` + `geom_errorbar()`)",
      "Scatterplot + smoother (`geom_point()` + `geom_smooth()`)",
      "Scatterplot, colour/size encodes 3rd variable",
      "Line plot (`geom_line()` + `scale_x_date()`)",
      "Pairs plot (`pairs()` or `GGally::ggpairs()`)"
    ),
    `Avoid` = c(
      "Pie chart; bars of means",
      "Histogram with too few or too many bins",
      "Pie chart; 3-D bar chart",
      "Bar chart of means only ('dynamite plot')",
      "Single mean ± SD bar",
      "Bar chart of means without CI",
      "Line chart (implies continuity between discrete points)",
      "3-D scatterplot (almost never interpretable)",
      "Bar chart; connecting non-temporal measurements",
      "Correlation heatmap without scatterplot check"
    )
  ),
  col.names = c("Goal", "Best choice", "Avoid"),
  caption   = "Plot selection guide. The best choice depends on your data structure and the question you want to answer visually."
)
```

<div class="key-box">
<h4>🔑 Key visualization takeaways</h4>
<ul>
<li><strong>Plot first, always.</strong> Summary statistics can be identical across structurally different datasets (Anscombe's Quartet). A histogram or scatterplot takes 2 lines of code and catches errors that hours of modelling will not.</li>
<li><strong>The grammar of graphics</strong> is data + aesthetics + geometry + optional refinements. Build plots layer by layer; each component is independent.</li>
<li><strong>Show the raw data</strong> whenever n is small enough. Overlay jittered points on boxplots; show individual dots on bar charts. Summary statistics that hide data are dishonest.</li>
<li><strong>Use 95% CI bars, not SD bars</strong>, when making inferential comparisons. Label error bars explicitly — readers cannot tell SD from SE from CI by looking.</li>
<li><strong>Colour for communication, not decoration.</strong> Use colourblind-safe palettes, never more than ~6 categories, and always a non-colour backup encoding (shape, linetype).</li>
<li><strong>Titles state findings</strong> ("Moisture increases with organic matter"), not variable names ("Organic matter vs. moisture"). Your reader should understand the main message without reading the caption.</li>
</ul>
</div>

---

### What's next?

- **[Describing Data](01_Describing_Data.html)** — means, SD, SE, and CIs
- **[Four Frameworks](02_Four_Frameworks.html)** — choosing the right statistical test
- **[t-test in depth](TimeOutSideClass.html)** — worked example with course data

```{r session-info, echo=FALSE}
# sessionInfo()   # uncomment for reproducibility reporting
```

<footer class="page-footer">
EA030 Basic Statistics &mdash; Module 03
<span class="divider">|</span> Prof. Marc Los Huertos
<span class="divider">|</span> Pomona College
<span class="divider">|</span> Last updated: `r format(Sys.Date(), "%B %Y")`
<span class="divider">|</span> <a href="https://github.com/marclos/RTricks">GitHub</a>
</footer>
