---
title: "Four Frameworks of Statistical Analyses"
author: "Marc Los Huertos"
date: "EA030 · Compiled `r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    self_contained: true
    theme: null
    highlight: null
    toc: false
knit: (function(inputFile, encoding) {
    rmarkdown::render(
      inputFile,
      encoding = encoding,
      output_file = file.path(dirname(inputFile),
                              "../docs/basic_statistics/03_Four_Frameworks.html")
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 7,
  fig.height = 4
)
library(ggplot2)
library(dplyr)
```

```{css, echo=FALSE}
@import url('https://fonts.googleapis.com/css2?family=Libre+Baskerville:ital,wght@0,400;0,700;1,400&family=Source+Code+Pro:wght@400;600&family=Lato:wght@300;400;700&display=swap');

*, *::before, *::after { box-sizing: border-box; margin: 0; padding: 0; }

:root {
  --forest:    #1e3a2f;
  --moss:      #2d5a3d;
  --fern:      #4a7c5e;
  --sage:      #8aad96;
  --mist:      #d4e6da;
  --cream:     #f7f3ed;
  --parchment: #ede8df;
  --ink:       #1a1a18;
  --charcoal:  #3a3a38;
  --dust:      #7a7a72;
  --amber:     #c8832a;
  --amber-lt:  #e8a84a;
  --white:     #ffffff;
  --card-shadow: 0 2px 10px rgba(30,58,47,.09);
}

html { scroll-behavior: smooth; }
body { font-family: 'Lato', sans-serif; background: var(--cream); color: var(--ink); font-size: 17px; line-height: 1.75; }
.main-container { max-width: 920px !important; margin: 0 auto !important; padding: 0 4vw 80px !important; }
h1.title, h4.author, h4.date { display: none; }

/* Page header */
.page-header {
  background-image:
    repeating-linear-gradient(120deg, transparent, transparent 60px, rgba(255,255,255,.018) 60px, rgba(255,255,255,.018) 61px),
    linear-gradient(160deg, #1e3a2f 0%, #2d5a3d 60%, #1a3d30 100%);
  color: #fff; padding: 44px 4vw 36px; margin: 0 -4vw 48px;
  border-bottom: 4px solid var(--amber); position: relative; overflow: hidden;
}
.page-header::after {
  content: ''; position: absolute; right: -40px; top: -40px;
  width: 260px; height: 260px; border-radius: 50%;
  background: radial-gradient(circle, rgba(138,173,150,.12) 0%, transparent 70%);
  pointer-events: none;
}
.page-header .module-tag {
  display: inline-block; font-family: 'Source Code Pro', monospace;
  font-size: .68rem; font-weight: 600; letter-spacing: .18em; text-transform: uppercase;
  color: var(--amber-lt); border: 1px solid rgba(200,131,42,.45);
  padding: 3px 10px; border-radius: 2px; margin-bottom: 12px;
}
.page-header h1 {
  font-family: 'Libre Baskerville', serif; font-size: clamp(1.6rem, 3vw, 2.3rem);
  font-weight: 700; color: #fff; line-height: 1.2; margin-bottom: 8px;
}
.page-header h1 span { color: var(--sage); }
.page-header .sub { font-size: .9rem; color: rgba(255,255,255,.65); font-weight: 300; }
.page-header .sub strong { color: rgba(255,255,255,.9); font-weight: 700; }
.back-link {
  display: inline-block; margin-top: 16px;
  font-family: 'Source Code Pro', monospace; font-size: .7rem; letter-spacing: .08em;
  color: var(--amber-lt); text-decoration: none;
  border: 1px solid rgba(200,131,42,.4); padding: 4px 11px; border-radius: 3px;
  transition: background .2s;
}
.back-link:hover { background: rgba(200,131,42,.18); }

/* Section nav pills */
.section-nav {
  display: flex; gap: 8px; flex-wrap: wrap; margin-bottom: 44px;
  padding-bottom: 20px; border-bottom: 1px solid var(--parchment);
}
.section-nav a {
  font-family: 'Source Code Pro', monospace; font-size: .7rem; font-weight: 600;
  letter-spacing: .08em; text-transform: uppercase;
  color: var(--fern); text-decoration: none;
  border: 1px solid var(--mist); padding: 4px 12px; border-radius: 3px;
  background: var(--white); transition: background .2s, color .2s;
}
.section-nav a:hover { background: var(--mist); color: var(--forest); }

/* Section headings */
h2 {
  font-family: 'Libre Baskerville', serif; font-size: 1.45rem; font-weight: 700;
  color: var(--forest); margin-top: 56px; margin-bottom: 6px;
  padding-bottom: 8px; border-bottom: 1px solid var(--parchment);
}
h3 {
  font-family: 'Libre Baskerville', serif; font-size: 1.1rem; font-weight: 700;
  color: var(--moss); margin-top: 28px; margin-bottom: 4px;
}

/* Body text */
p { margin-bottom: 1em; }
strong { color: var(--forest); }

/* Code */
code {
  font-family: 'Source Code Pro', monospace; font-size: .84em;
  background: var(--mist); color: var(--moss); padding: 1px 5px; border-radius: 3px;
}
pre {
  background: var(--ink); color: #e8e6df; border-radius: 5px;
  padding: 14px 18px; font-size: .83rem; line-height: 1.55;
  overflow-x: auto; border-left: 4px solid var(--amber); margin: 1.2em 0;
}
pre code { background: transparent; color: inherit; padding: 0; font-size: 1em; }

/* Callout boxes */
.callout {
  border-left: 5px solid; padding: .65em 1.1em;
  margin: 1.4em 0; border-radius: 0 5px 5px 0; font-size: .92em; line-height: 1.6;
}
.callout > strong:first-child { display: block; margin-bottom: .3em; }
.callout-concept { border-color: var(--fern);  background: #f0f7f2; }
.callout-tip     { border-color: #2e7d32;      background: #f1f8f1; }
.callout-warn    { border-color: #e65100;      background: #fff8f4; }
.callout-formula { border-color: var(--amber); background: #fdf8f0;
                   font-family: 'Source Code Pro', monospace; font-size: .88em; }

/* Framework overview matrix */
.framework-matrix {
  width: 100%; border-collapse: separate; border-spacing: 6px;
  margin: 2em 0 2.5em;
}
.framework-matrix th {
  font-family: 'Source Code Pro', monospace; font-size: .75rem;
  letter-spacing: .09em; text-transform: uppercase; padding: 10px 14px;
  background: var(--forest); color: #fff; border-radius: 4px;
}
.framework-matrix th.row-header {
  background: var(--moss); writing-mode: horizontal-tb;
  min-width: 120px; text-align: center;
}
.fm-cell {
  background: var(--white); border: 1px solid var(--parchment);
  border-radius: 6px; padding: 14px 16px; vertical-align: top;
  box-shadow: var(--card-shadow);
}
.fm-cell .fm-label {
  font-family: 'Source Code Pro', monospace; font-size: .72rem;
  font-weight: 600; letter-spacing: .1em; text-transform: uppercase;
  color: var(--amber); margin-bottom: 4px;
}
.fm-cell .fm-name {
  font-family: 'Libre Baskerville', serif; font-size: 1rem;
  font-weight: 700; color: var(--forest); margin-bottom: 4px;
}
.fm-cell .fm-desc { font-size: .85rem; color: var(--charcoal); line-height: 1.5; }
.fm-cell a { color: var(--fern); text-decoration: none; font-size: .8rem; }
.fm-cell a:hover { text-decoration: underline; }

/* Key concept summary boxes */
.key-box {
  background: var(--forest); color: #fff; border-radius: 6px;
  padding: 20px 24px; margin: 2em 0; font-size: .9rem;
}
.key-box h4 {
  font-family: 'Libre Baskerville', serif; color: var(--sage);
  font-size: 1rem; margin-bottom: 10px;
}
.key-box ul { padding-left: 1.3em; }
.key-box li { margin-bottom: 5px; color: rgba(255,255,255,.85); }
.key-box code { background: rgba(255,255,255,.12); color: var(--amber-lt); }

/* assumption list */
.assumption-list { padding-left: 1.4em; margin: .6em 0 1em; }
.assumption-list li { margin-bottom: .3em; font-size: .91em; }

/* Tables */
table { font-size: .9rem; border-collapse: collapse; width: 100%; margin: 1.2em 0; }
thead th {
  background: var(--forest); color: #fff;
  font-family: 'Source Code Pro', monospace; font-size: .76rem;
  letter-spacing: .06em; text-transform: uppercase; padding: 8px 12px; text-align: left;
}
tbody tr:nth-child(even) { background: var(--mist); }
tbody td { padding: 7px 12px; border-bottom: 1px solid var(--parchment); }

/* Footer */
.page-footer {
  background: var(--ink); color: rgba(255,255,255,.42); text-align: center;
  padding: 24px 4vw; margin: 60px -4vw 0;
  font-size: .76rem; font-family: 'Source Code Pro', monospace; letter-spacing: .04em;
}
.page-footer a { color: var(--sage); text-decoration: none; }
.page-footer .divider { margin: 0 8px; opacity: .3; }
```

<div class="page-header">
<div class="module-tag">EA030 &mdash; Basic Statistics &mdash; Module 02</div>
<h1>Four <span>Frameworks of Statistical Analysis</span></h1>
<p class="sub"><strong>Prof. Marc Los Huertos</strong> &nbsp;&bull;&nbsp; Pomona College</p>
<a class="back-link" href="index.html">&#8592; Back to Basic Statistics</a>
</div>

<div class="section-nav">
<a href="#overview">Overview</a>
<a href="#anova">ANOVA / t-test</a>
<a href="#linear">Linear Regression</a>
<a href="#chisq">Chi-square</a>
<a href="#logistic">Logistic Regression</a>
<a href="#choosing">Choosing a Test</a>
</div>

> **What this module covers.** Every statistical test you will encounter in
> environmental science fits into one of four frameworks determined by a single
> question: *what type of variable is your outcome (dependent), and what type of
> variable is your predictor (independent)?* This module maps the landscape, then
> walks through each framework in depth with environmental examples, diagnostic
> plots, and worked R code.

---

## The Organising Principle {#overview}

Before running any analysis, ask two questions:

1. **What kind of variable is my dependent (response) variable?**
   Is it a measurement on a continuous scale, or does it represent a category or
   binary outcome?

2. **What kind of variable is my independent (predictor) variable?**
   Is it a group label (categorical), or a quantity that varies smoothly (continuous)?

The answers place you in exactly one of four cells:

<table class="framework-matrix">
<thead>
<tr>
  <th style="background:transparent; border:none;"></th>
  <th>Independent: <strong>Categorical</strong><br><span style="font-weight:300;font-size:.8em;">groups, treatments, species, sites</span></th>
  <th>Independent: <strong>Continuous</strong><br><span style="font-weight:300;font-size:.8em;">temperature, concentration, time</span></th>
</tr>
</thead>
<tbody>
<tr>
  <th class="row-header">Dependent:<br><strong>Continuous</strong><br><span style="font-weight:300;font-size:.8em;">mass, pH, height, PM2.5</span></th>
  <td class="fm-cell">
    <div class="fm-label">Framework I</div>
    <div class="fm-name">t-test &amp; ANOVA</div>
    <div class="fm-desc">Compare means across two or more groups.<br>
    <em>Do burned and unburned plots differ in soil nitrogen?</em></div>
    <a href="#anova">Jump to section ↓</a>
  </td>
  <td class="fm-cell">
    <div class="fm-label">Framework II</div>
    <div class="fm-name">Linear Regression</div>
    <div class="fm-desc">Model a straight-line relationship between two quantities.<br>
    <em>Does temperature predict dissolved oxygen?</em></div>
    <a href="#linear">Jump to section ↓</a>
  </td>
</tr>
<tr>
  <th class="row-header">Dependent:<br><strong>Categorical</strong><br><span style="font-weight:300;font-size:.8em;">present/absent, survived/died</span></th>
  <td class="fm-cell">
    <div class="fm-label">Framework III</div>
    <div class="fm-name">Chi-square Test</div>
    <div class="fm-desc">Test whether category frequencies differ between groups.<br>
    <em>Is species presence associated with habitat type?</em></div>
    <a href="#chisq">Jump to section ↓</a>
  </td>
  <td class="fm-cell">
    <div class="fm-label">Framework IV</div>
    <div class="fm-name">Logistic Regression</div>
    <div class="fm-desc">Model the probability of a binary outcome from a continuous predictor.<br>
    <em>Does pollution concentration predict fish mortality?</em></div>
    <a href="#logistic">Jump to section ↓</a>
  </td>
</tr>
</tbody>
</table>

<div class="callout callout-concept">
<strong>📌 The core logic is the same in every framework</strong>

Every test below asks the same fundamental question: **is the pattern in my
data larger than what chance alone would produce?** The answer is formalised
as a *test statistic* and a *p-value*. What differs is how the test statistic
is computed — because the geometry of the data changes depending on variable types.
</div>

---

## Framework I — t-test and ANOVA {#anova}

**When to use it:** Your response variable is continuous (a measurement), and
your predictor is a categorical group variable.

- **Two groups** → use a **t-test**
- **Three or more groups** → use **one-way ANOVA** (or its non-parametric
  equivalent, the Kruskal-Wallis test, when assumptions are violated)

### The core question

Is the difference in group means large enough — relative to the within-group
variability — that it is unlikely to have arisen by chance?

$$t = \frac{\bar{x}_1 - \bar{x}_2}{\sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}}}$$

For ANOVA, the same logic is generalised using the **F-ratio**: variance
*between* groups divided by variance *within* groups.

$$F = \frac{MS_{\text{between}}}{MS_{\text{within}}}$$

A large F means the group means are more spread out than individual
observations within groups — evidence against the null hypothesis that all
group means are equal.

### Environmental example: soil nitrogen across land-use types

```{r anova-data}
set.seed(42)
# Soil total nitrogen (g/kg) at three land-use types
nitrogen <- data.frame(
  landuse = rep(c("Forest", "Grassland", "Cropland"), each = 25),
  N_gkg   = c(
    rnorm(25, mean = 4.8, sd = 0.7),   # Forest — highest
    rnorm(25, mean = 3.2, sd = 0.9),   # Grassland
    rnorm(25, mean = 1.9, sd = 0.6)    # Cropland — lowest
  )
)
```

#### Step 1 — Visualise the groups

Always plot your data before running any test. Boxplots show the median,
interquartile range, and any outliers simultaneously.

```{r anova-plot, fig.cap="Soil nitrogen (g/kg) across three land-use types. The horizontal bar is the median; the box spans the IQR; whiskers extend to 1.5×IQR. Cropland shows consistently lower nitrogen than forest.", fig.height=4}
ggplot(nitrogen, aes(x = landuse, y = N_gkg, fill = landuse)) +
  geom_boxplot(width = 0.5, outlier.shape = 21, outlier.size = 1.8,
               colour = "#1e3a2f", alpha = 0.85) +
  geom_jitter(width = 0.15, size = 0.9,
              colour = "#1e3a2f", alpha = 0.35) +
  scale_fill_manual(values = c("#4a7c5e", "#8aad96", "#c8832a")) +
  labs(x = NULL, y = "Soil nitrogen (g/kg)",
       title = "Soil Nitrogen by Land-Use Type") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        panel.grid.major.x = element_blank(),
        plot.title = element_text(family = "serif", face = "bold",
                                  colour = "#1e3a2f"))
```

#### Step 2 — Run the ANOVA

```{r anova-run}
model_aov <- aov(N_gkg ~ landuse, data = nitrogen)
summary(model_aov)
```

The F-statistic and its p-value tell you whether *any* groups differ.
A significant ANOVA only tells you that the group means are not all equal —
it does not tell you *which pairs* differ.

#### Step 3 — Post-hoc comparisons

When ANOVA is significant, use **Tukey's Honest Significant Difference (HSD)**
test to identify which specific pairs of groups differ, while controlling
the overall Type I error rate.

```{r tukey}
TukeyHSD(model_aov)
```

```{r tukey-plot, echo=FALSE, fig.cap="Tukey HSD 95% confidence intervals for all pairwise differences. Intervals that do not cross zero indicate a statistically significant difference between those groups.", fig.height=3.5}
tukey_df <- as.data.frame(TukeyHSD(model_aov)$landuse)
tukey_df$comparison <- rownames(tukey_df)

ggplot(tukey_df, aes(x = diff, y = comparison)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "#c8832a", linewidth = 0.8) +
  geom_errorbarh(aes(xmin = lwr, xmax = upr),
                 height = 0.25, colour = "#1e3a2f", linewidth = 1) +
  geom_point(size = 3, colour = "#4a7c5e") +
  labs(x = "Difference in means (g/kg)", y = NULL,
       title = "Tukey HSD: Pairwise Comparisons") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(family = "serif", face = "bold",
                                  colour = "#1e3a2f"))
```

#### Checking assumptions

ANOVA rests on three assumptions. Always check them with diagnostic plots.

```{r anova-diagnostics, fig.cap="Left: residuals should scatter randomly around zero (no pattern = good). Right: residuals on a Q-Q plot should follow the diagonal line (normality check).", fig.height=3.5}
par(mfrow = c(1, 2), mar = c(4, 4, 3, 1), las = 1)
plot(model_aov, which = 1, main = "Residuals vs Fitted")
plot(model_aov, which = 2, main = "Normal Q-Q")
par(mfrow = c(1, 1))
```

<div class="callout callout-warn">
<strong>⚠️ When ANOVA assumptions are violated</strong>

If residuals are strongly non-normal or variances are very unequal across groups,
consider: **(1)** log-transforming the response variable (common for skewed
environmental data like concentrations), **(2)** using Welch's ANOVA
(`oneway.test()`) which does not assume equal variances, or **(3)** the
non-parametric **Kruskal-Wallis test** (`kruskal.test()`), which tests
whether group medians differ without assuming normality.
</div>

<div class="callout callout-formula">
<strong>⚙️ Two-group case: the t-test</strong><br>
When you have exactly two groups, `t.test(y ~ group, data = df)` is equivalent
to a two-group ANOVA (F = t²). Use `var.equal = TRUE` if Levene's test
confirms equal variances; otherwise the default Welch correction is safer.
</div>

---

## Framework II — Linear Regression {#linear}

**When to use it:** Both your response and predictor are continuous.
You want to describe *how much* the response changes per unit increase in
the predictor, and test whether that relationship is stronger than chance.

### The model

$$Y_i = \beta_0 + \beta_1 X_i + \varepsilon_i \qquad \varepsilon_i \sim \mathcal{N}(0,\, \sigma^2)$$

- $\beta_0$ — **intercept**: predicted value of Y when X = 0
- $\beta_1$ — **slope**: expected change in Y for a one-unit increase in X
- $\varepsilon_i$ — **residuals**: the part of Y not explained by X

The parameters are estimated by **Ordinary Least Squares (OLS)**: choosing the
line that minimises the sum of squared vertical distances between observed and
predicted values.

### Environmental example: water temperature and dissolved oxygen

Dissolved oxygen (DO) decreases as water warms — a critical relationship for
aquatic organisms.

```{r lm-data}
set.seed(7)
n_obs <- 60
temp_C <- runif(n_obs, min = 10, max = 28)          # water temperature (°C)
DO_mgl <- 14.6 - 0.38 * temp_C + rnorm(n_obs, 0, 0.6)  # DO (mg/L); ~real physics
stream <- data.frame(temp_C, DO_mgl)
```

#### Step 1 — Scatterplot with regression line

```{r lm-plot, fig.cap="Dissolved oxygen vs. water temperature in a stream dataset. The blue OLS line summarises the negative relationship; the grey ribbon is the 95% confidence interval on the mean prediction.", fig.height=4}
ggplot(stream, aes(temp_C, DO_mgl)) +
  geom_point(colour = "#4a7c5e", alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", colour = "#1e3a2f", fill = "#d4e6da",
              linewidth = 1, se = TRUE) +
  labs(x = "Water temperature (°C)", y = "Dissolved oxygen (mg/L)",
       title = "Dissolved Oxygen Declines with Temperature") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(family = "serif", face = "bold",
                                  colour = "#1e3a2f"))
```

#### Step 2 — Fit the model and read the output

```{r lm-fit}
model_lm <- lm(DO_mgl ~ temp_C, data = stream)
summary(model_lm)
```

**Reading the output:**

- `(Intercept)` — estimated DO at 0 °C (extrapolation here, but gives the line its anchor)
- `temp_C` — the slope: each 1 °C increase is associated with a −0.38 mg/L change in DO
- `Std. Error` — uncertainty in each coefficient estimate
- `t value` / `Pr(>|t|)` — is the coefficient significantly different from zero?
- `Multiple R-squared` — proportion of variance in DO explained by temperature
- `F-statistic` — overall test of the model (same as testing $\beta_1 = 0$ in simple regression)

```{r lm-coefs}
coef(model_lm)
confint(model_lm)   # 95% CIs on the coefficients
```

#### Step 3 — Partition the variance: what does R² mean?

```{r r-squared}
# How much variance does the model explain?
ss_total   <- sum((stream$DO_mgl - mean(stream$DO_mgl))^2)
ss_resid   <- sum(residuals(model_lm)^2)
ss_model   <- ss_total - ss_resid
r_sq       <- ss_model / ss_total

cat("SS Total :", round(ss_total, 2), "\n")
cat("SS Model :", round(ss_model, 2), "\n")
cat("SS Resid :", round(ss_resid, 2), "\n")
cat("R²       :", round(r_sq, 3), " (=", round(r_sq*100, 1), "% variance explained)\n")
```

#### Step 4 — Diagnostic plots

```{r lm-diagnostics, fig.cap="Regression diagnostics. Top-left: residuals vs fitted — look for random scatter with no funnel shape (homoscedasticity). Top-right: Q-Q plot — residuals should follow the diagonal. Bottom: Scale-Location confirms homoscedasticity; Cook's Distance flags high-influence observations.", fig.height=5.5, fig.width=7}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1), las = 1)
plot(model_lm)
par(mfrow = c(1, 1))
```

<div class="callout callout-concept">
<strong>📌 Correlation vs. regression — what is the difference?</strong>

**Correlation** (Pearson's *r*) quantifies the *strength and direction* of a
linear relationship. It is symmetric: cor(X, Y) = cor(Y, X).

**Regression** quantifies the *rate of change*: how many units does Y change
per unit increase in X? It is asymmetric — regressing Y on X gives a
different line than regressing X on Y.

Use correlation when you simply want to know "are these related?"
Use regression when you want to predict Y from X or estimate the effect size.
Note: $r^2 = R^2$ in simple linear regression — they are the same quantity.
</div>

<div class="callout callout-warn">
<strong>⚠️ Association ≠ causation</strong>
A significant slope does not prove that X causes Y. Temperature and dissolved
oxygen have a known physical mechanism. But in many environmental datasets,
confounding variables drive apparent correlations. Causal inference requires
careful study design, not just a significant p-value.
</div>

---

## Framework III — Chi-square Test {#chisq}

**When to use it:** Both your response and predictor are categorical.
You have counts of observations falling into the cells of a contingency table
and want to test whether the distribution of one variable depends on the other.

### The core question

If the two variables were independent, what counts would we *expect* in each cell?
The chi-square statistic measures how far observed counts deviate from expected:

$$\chi^2 = \sum_{\text{all cells}} \frac{(O - E)^2}{E}$$

where $O$ = observed count and $E = \frac{(\text{row total}) \times (\text{col total})}{\text{grand total}}$.

### Environmental example: species presence by habitat type

Is the presence of a focal riparian plant species associated with habitat type
(riparian buffer vs. degraded bank)?

```{r chisq-data}
set.seed(3)
n_sites <- 120

habitat  <- rep(c("Riparian buffer", "Degraded bank"), each = n_sites / 2)
# Species more likely in intact riparian buffer
present  <- c(rbinom(60, 1, prob = 0.72),   # buffer
              rbinom(60, 1, prob = 0.31))    # degraded

ecology <- data.frame(habitat, present = factor(present, labels = c("Absent","Present")))
```

#### Step 1 — Build the contingency table

```{r contingency-table}
(ct <- table(ecology$habitat, ecology$present))
```

```{r proportions}
# Row proportions: P(Present | habitat)
prop.table(ct, margin = 1) |> round(2)
```

#### Step 2 — Visualise the association

```{r chisq-plot, fig.cap="Proportion of sites where the focal species is present, by habitat type. Error bars are 95% Wilson confidence intervals. The difference in detection rates is visually clear.", fig.height=4}
prop_df <- ecology |>
  group_by(habitat) |>
  summarise(
    n       = n(),
    x       = sum(present == "Present"),
    prop    = x / n,
    # Wilson CI
    lower   = (prop + qnorm(0.975)^2 / (2*n) -
                 qnorm(0.975) * sqrt(prop*(1-prop)/n + qnorm(0.975)^2/(4*n^2))) /
              (1 + qnorm(0.975)^2 / n),
    upper   = (prop + qnorm(0.975)^2 / (2*n) +
                 qnorm(0.975) * sqrt(prop*(1-prop)/n + qnorm(0.975)^2/(4*n^2))) /
              (1 + qnorm(0.975)^2 / n)
  )

ggplot(prop_df, aes(habitat, prop, fill = habitat)) +
  geom_col(width = 0.5, colour = "#1e3a2f", alpha = 0.85) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.12,
                linewidth = 0.9, colour = "#1e3a2f") +
  scale_fill_manual(values = c("#4a7c5e", "#c8832a")) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  labs(x = NULL, y = "Proportion of sites with species present",
       title = "Species Presence by Habitat Type") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        panel.grid.major.x = element_blank(),
        plot.title = element_text(family = "serif", face = "bold",
                                  colour = "#1e3a2f"))
```

#### Step 3 — Run the chi-square test

```{r chisq-test}
chisq.test(ct)
```

```{r chisq-expected}
# Always check expected counts — all should be ≥ 5 for chi-square to be valid
chisq.test(ct)$expected
```

#### Step 4 — Effect size: Cramér's V

A significant chi-square only tells you there is an association; it does not
tell you how *strong* it is. Cramér's V standardises the chi-square statistic
to a 0–1 scale.

```{r cramers-v}
chi_val <- chisq.test(ct)$statistic
n_total <- sum(ct)
k       <- min(nrow(ct), ncol(ct))   # min(rows, cols)
cramers_v <- sqrt(chi_val / (n_total * (k - 1)))
cat("Cramér's V =", round(cramers_v, 3),
    " (0 = no association, 1 = perfect association)\n")
```

<div class="callout callout-concept">
<strong>📌 When to use Fisher's Exact Test instead</strong>

The chi-square test requires that all **expected** cell counts are ≥ 5.
When you have small samples or rare events (common in species detection studies),
use `fisher.test(ct)` instead. Fisher's exact test works with any sample size
and is preferred for 2×2 tables with small marginal totals.
</div>

<div class="callout callout-formula">
<strong>⚙️ Chi-square vs. logistic regression for 2×2 tables</strong><br>
Chi-square tests association. Logistic regression models the *probability* of
the outcome and produces an **odds ratio** — the most useful effect-size
measure for categorical outcomes. For a single binary predictor, both give the
same p-value; logistic regression adds the odds ratio and allows control
for confounders. See Framework IV.
</div>

---

## Framework IV — Logistic Regression {#logistic}

**When to use it:** Your response is binary (0/1, absent/present, survived/died),
and your predictor is continuous (or a mix of continuous and categorical).

### Why not just use linear regression on 0/1 data?

A linear model can predict probabilities below 0 or above 1, which is
nonsensical. Logistic regression solves this by modelling the **log-odds**
(logit) of the outcome as a linear function of the predictors:

$$\log\!\left(\frac{p}{1-p}\right) = \beta_0 + \beta_1 X$$

Solving for $p$ gives the **logistic (sigmoid) curve** — bounded between
0 and 1:

$$p = \frac{1}{1 + e^{-(\beta_0 + \beta_1 X)}}$$

### Environmental example: fish mortality as a function of pollutant concentration

At what pollutant concentration does survival probability drop below 50%?

```{r logistic-data}
set.seed(11)
n_fish <- 120
conc   <- runif(n_fish, min = 0, max = 10)   # pollutant (mg/L)
# True model: logit(p) = 2.5 - 0.8*conc  →  LC50 ≈ 3.1 mg/L
log_odds <- 2.5 - 0.8 * conc
prob     <- 1 / (1 + exp(-log_odds))
survived <- rbinom(n_fish, 1, prob)

fish <- data.frame(conc, survived)
```

#### Step 1 — Visualise the binary outcome

Binary data cannot be plotted as a scatterplot without modification.
Jitter the points vertically and overlay the model curve.

```{r logistic-plot, fig.cap="Fish survival (1 = survived, 0 = died) vs. pollutant concentration. Points are jittered to reduce overplotting. The curve is the fitted logistic regression model; the dashed line marks 50% survival probability and the estimated LC50.", fig.height=4.5}
model_glm <- glm(survived ~ conc, data = fish, family = binomial)

# Prediction curve
pred_df <- data.frame(conc = seq(0, 10, length.out = 300))
pred_df$prob <- predict(model_glm, newdata = pred_df, type = "response")

# LC50: concentration where p = 0.5  → conc = -b0/b1
lc50 <- -coef(model_glm)[1] / coef(model_glm)[2]

ggplot(fish, aes(conc, survived)) +
  geom_jitter(height = 0.04, width = 0, alpha = 0.4, size = 1.8,
              colour = "#4a7c5e") +
  geom_line(data = pred_df, aes(y = prob),
            colour = "#1e3a2f", linewidth = 1.4) +
  geom_hline(yintercept = 0.5, linetype = "dashed",
             colour = "#c8832a", linewidth = 0.8) +
  geom_vline(xintercept = lc50, linetype = "dashed",
             colour = "#c8832a", linewidth = 0.8) +
  annotate("text", x = lc50 + 0.4, y = 0.08,
           label = paste0("LC50 ≈ ", round(lc50, 1), " mg/L"),
           colour = "#c8832a", hjust = 0, size = 3.8, fontface = "italic") +
  scale_y_continuous(labels = scales::percent_format(),
                     breaks = c(0, 0.25, 0.5, 0.75, 1)) +
  labs(x = "Pollutant concentration (mg/L)",
       y = "P(survival)",
       title = "Logistic Regression: Fish Survival vs. Pollutant Concentration") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(family = "serif", face = "bold",
                                  colour = "#1e3a2f"))
```

#### Step 2 — Fit and interpret the model

```{r logistic-fit}
summary(model_glm)
```

**Reading the output:**

The coefficients are on the **log-odds scale**, which is hard to interpret
directly. Convert to **odds ratios** by exponentiating:

```{r odds-ratios}
exp(coef(model_glm))          # odds ratios
exp(confint(model_glm))       # 95% CIs on odds ratios
```

- `(Intercept)` exponentiated: the odds of survival when concentration = 0
- `conc` exponentiated: for each 1 mg/L increase in concentration, the
  **odds of survival** are multiplied by this factor (< 1 = decreasing odds)

```{r lc50}
lc50_val <- -coef(model_glm)["(Intercept)"] / coef(model_glm)["conc"]
cat("Estimated LC50:", round(lc50_val, 2), "mg/L\n")
```

#### Step 3 — Model fit statistics

Unlike linear regression, logistic regression has no single R² analogue.
Use **pseudo-R²** (McFadden's) and the **Akaike Information Criterion (AIC)**
to assess and compare models.

```{r logistic-fit-stats}
# Null model (intercept only) for comparison
model_null <- glm(survived ~ 1, data = fish, family = binomial)

# McFadden's pseudo-R²
mcfadden_r2 <- 1 - (logLik(model_glm) / logLik(model_null))
cat("McFadden R² =", round(mcfadden_r2, 3), "\n")

# AIC (lower = better fit, penalises complexity)
cat("AIC (full model) =", round(AIC(model_glm), 1), "\n")
cat("AIC (null model) =", round(AIC(model_null), 1), "\n")
```

#### Step 4 — Classification accuracy

```{r classification}
# Classify: predict survival if p > 0.5
fish$predicted <- ifelse(
  predict(model_glm, type = "response") > 0.5, 1, 0)

conf_mat <- table(Predicted = fish$predicted, Actual = fish$survived)
conf_mat

accuracy <- sum(diag(conf_mat)) / sum(conf_mat)
cat("Overall accuracy:", round(accuracy * 100, 1), "%\n")
```

<div class="callout callout-tip">
<strong>💡 Extending logistic regression</strong>

Logistic regression generalises naturally:

- **Multiple predictors** — add more terms: `glm(y ~ x1 + x2 + x3, family = binomial)`
- **Interaction terms** — `x1:x2` or `x1*x2` in the formula
- **Ordered outcomes (3+ levels)** — ordinal logistic regression (`MASS::polr()`)
- **Unordered outcomes (3+ levels)** — multinomial logistic regression (`nnet::multinom()`)

The underlying principle — modelling the log-odds as a linear function —
remains the same throughout.
</div>

---

## Choosing the Right Framework {#choosing}

Use this decision process before every analysis:

```{r decision-table, echo=FALSE}
knitr::kable(
  data.frame(
    `Dependent variable`  = c("Continuous", "Continuous",
                               "Binary (0/1)", "Binary (0/1)"),
    `Independent variable` = c("Categorical (2 groups)",
                                "Categorical (3+ groups)",
                                "Categorical",
                                "Continuous"),
    `Primary test`        = c("t-test  (`t.test()`)",
                               "One-way ANOVA  (`aov()`)",
                               "Chi-square  (`chisq.test()`)",
                               "Logistic regression  (`glm(..., family=binomial)`)"),
    `Effect size`         = c("Cohen's d",
                               "η² (eta-squared)",
                               "Cramér's V",
                               "Odds ratio"),
    `Non-parametric alt.` = c("Wilcoxon (`wilcox.test()`)",
                               "Kruskal-Wallis (`kruskal.test()`)",
                               "Fisher's exact (`fisher.test()`)",
                               "— (logistic is already non-normal)")
  ),
  col.names = c("Dependent variable", "Independent variable",
                "Primary test", "Effect size", "Non-parametric alt."),
  caption   = "Framework selection guide. When in doubt: plot first, then choose."
)
```

<div class="key-box">
<h4>🔑 The four frameworks — key takeaways</h4>
<ul>
<li><strong>Framework I (ANOVA / t-test):</strong> compares means across groups using the F- or t-statistic. Extend with Tukey's HSD for pairwise comparisons. Check residual normality and equal variances.</li>
<li><strong>Framework II (Linear regression):</strong> estimates a slope (effect size with units). R² tells you how much variance is explained. Four diagnostic plots catch violations of OLS assumptions.</li>
<li><strong>Framework III (Chi-square):</strong> tests independence in a contingency table. Use Cramér's V for effect size. Switch to Fisher's exact test when expected counts are small.</li>
<li><strong>Framework IV (Logistic regression):</strong> models P(outcome) as a sigmoid function of a continuous predictor. Coefficients are log-odds; exponentiate for odds ratios. McFadden's R² and AIC assess model fit.</li>
<li><strong>In every framework:</strong> visualise first, check assumptions, report effect sizes alongside p-values, and show uncertainty (CIs, not just stars).</li>
</ul>
</div>

---

### What's next?

- **[Describing Data](01_Describing_Data.html)** — means, SD, SE, and confidence intervals
- **[t-test in depth](TimeOutSideClass.html)** — worked example with course data
- **[ANOVA in depth](ANOVA.html)** — factorial designs and interaction effects
- **[Regression in depth](Regression.html)** — multiple predictors and model selection

```{r session-info, echo=FALSE}
# sessionInfo()   # uncomment for reproducibility reporting
```

<footer class="page-footer">
EA030 Basic Statistics &mdash; Module 02
<span class="divider">|</span> Prof. Marc Los Huertos
<span class="divider">|</span> Pomona College
<span class="divider">|</span> Last updated: `r format(Sys.Date(), "%B %Y")`
<span class="divider">|</span> <a href="https://github.com/marclos/RTricks">GitHub</a>
</footer>
