---
title: "Signal Over Noise"
author: "Marc Los Huertos"
date: "EA030 · Compiled `r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    self_contained: true
    theme: null
    highlight: null
    toc: false
knit: (function(inputFile, encoding) {
    rmarkdown::render(
      inputFile,
      encoding = encoding,
      output_file = file.path(dirname(inputFile),
                              "../docs/basic_statistics/03_Signal_Over_Noise.html")
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 7,
  fig.height = 4
)
library(ggplot2)
library(dplyr)
```

```{css, echo=FALSE}
@import url('https://fonts.googleapis.com/css2?family=Libre+Baskerville:ital,wght@0,400;0,700;1,400&family=Source+Code+Pro:wght@400;600&family=Lato:wght@300;400;700&display=swap');

*, *::before, *::after { box-sizing: border-box; margin: 0; padding: 0; }

:root {
  --forest:    #1e3a2f;
  --moss:      #2d5a3d;
  --fern:      #4a7c5e;
  --sage:      #8aad96;
  --mist:      #d4e6da;
  --cream:     #f7f3ed;
  --parchment: #ede8df;
  --ink:       #1a1a18;
  --charcoal:  #3a3a38;
  --dust:      #7a7a72;
  --amber:     #c8832a;
  --amber-lt:  #e8a84a;
  --white:     #ffffff;
  --card-shadow: 0 2px 10px rgba(30,58,47,.09);
}

html { scroll-behavior: smooth; }
body {
  font-family: 'Lato', sans-serif;
  background: var(--cream);
  color: var(--ink);
  font-size: 17px;
  line-height: 1.75;
  padding-left:  clamp(14px, 4vw, 48px);
  padding-right: clamp(14px, 4vw, 48px);
  padding-bottom: 80px;
}
.main-container { max-width: 920px !important; margin: 0 auto !important; padding: 0 clamp(10px, 2vw, 32px) 80px !important; }
.page-header {
  margin-left:  calc(-1 * clamp(14px, 4vw, 48px));
  margin-right: calc(-1 * clamp(14px, 4vw, 48px));
}
h1.title, h4.author, h4.date { display: none; }

.page-header {
  background-image:
    repeating-linear-gradient(120deg, transparent, transparent 60px, rgba(255,255,255,.018) 60px, rgba(255,255,255,.018) 61px),
    linear-gradient(160deg, #1e3a2f 0%, #2d5a3d 60%, #1a3d30 100%);
  color: #fff; padding: 44px 4vw 36px; margin: 0 -4vw 48px;
  border-bottom: 4px solid var(--amber); position: relative; overflow: hidden;
}
.page-header::after {
  content: ''; position: absolute; right: -40px; top: -40px;
  width: 260px; height: 260px; border-radius: 50%;
  background: radial-gradient(circle, rgba(138,173,150,.12) 0%, transparent 70%);
  pointer-events: none;
}
.page-header .module-tag {
  display: inline-block; font-family: 'Source Code Pro', monospace;
  font-size: .68rem; font-weight: 600; letter-spacing: .18em; text-transform: uppercase;
  color: var(--amber-lt); border: 1px solid rgba(200,131,42,.45);
  padding: 3px 10px; border-radius: 2px; margin-bottom: 12px;
}
.page-header h1 {
  font-family: 'Libre Baskerville', serif; font-size: clamp(1.6rem, 3vw, 2.3rem);
  font-weight: 700; color: #fff; line-height: 1.2; margin-bottom: 8px;
}
.page-header h1 span { color: var(--sage); }
.page-header .sub { font-size: .9rem; color: rgba(255,255,255,.65); font-weight: 300; }
.page-header .sub strong { color: rgba(255,255,255,.9); font-weight: 700; }
.back-link {
  display: inline-block; margin-top: 16px;
  font-family: 'Source Code Pro', monospace; font-size: .7rem; letter-spacing: .08em;
  color: var(--amber-lt); text-decoration: none;
  border: 1px solid rgba(200,131,42,.4); padding: 4px 11px; border-radius: 3px;
}
.back-link:hover { background: rgba(200,131,42,.18); }

.section-nav {
  display: flex; gap: 8px; flex-wrap: wrap; margin-bottom: 44px;
  padding-bottom: 20px; border-bottom: 1px solid var(--parchment);
}
.section-nav a {
  font-family: 'Source Code Pro', monospace; font-size: .7rem; font-weight: 600;
  letter-spacing: .08em; text-transform: uppercase;
  color: var(--fern); text-decoration: none;
  border: 1px solid var(--mist); padding: 4px 12px; border-radius: 3px;
  background: var(--white); transition: background .2s, color .2s;
}
.section-nav a:hover { background: var(--mist); color: var(--forest); }

h2 {
  font-family: 'Libre Baskerville', serif; font-size: 1.45rem; font-weight: 700;
  color: var(--forest); margin-top: 56px; margin-bottom: 6px;
  padding-bottom: 8px; border-bottom: 1px solid var(--parchment);
}
h3 {
  font-family: 'Libre Baskerville', serif; font-size: 1.1rem; font-weight: 700;
  color: var(--moss); margin-top: 28px; margin-bottom: 4px;
}

p { margin-bottom: 1em; }
strong { color: var(--forest); }

code {
  font-family: 'Source Code Pro', monospace; font-size: .84em;
  background: var(--mist); color: var(--moss); padding: 1px 5px; border-radius: 3px;
}
pre {
  background: var(--ink); color: #e8e6df; border-radius: 5px;
  padding: 14px 18px; font-size: .83rem; line-height: 1.55;
  overflow-x: auto; border-left: 4px solid var(--amber); margin: 1.2em 0;
}
pre code { background: transparent; color: inherit; padding: 0; font-size: 1em; }

.callout {
  border-left: 5px solid; padding: .65em 1.1em;
  margin: 1.4em 0; border-radius: 0 5px 5px 0; font-size: .92em; line-height: 1.6;
}
.callout > strong:first-child { display: block; margin-bottom: .3em; }
.callout-concept { border-color: var(--fern);  background: #f0f7f2; }
.callout-tip     { border-color: #2e7d32;      background: #f1f8f1; }
.callout-warn    { border-color: #e65100;      background: #fff8f4; }
.callout-formula { border-color: var(--amber); background: #fdf8f0;
                   font-family: 'Source Code Pro', monospace; font-size: .88em; }

/* Signal/noise ratio diagram */
.ratio-display {
  display: flex; align-items: center; justify-content: center;
  gap: 16px; margin: 2em 0; font-family: 'Libre Baskerville', serif;
}
.ratio-num, .ratio-den {
  text-align: center; padding: 16px 28px; border-radius: 6px;
  box-shadow: var(--card-shadow);
}
.ratio-num { background: var(--forest); color: #fff; }
.ratio-num .rn-label { font-size: .72rem; font-family: 'Source Code Pro', monospace;
  letter-spacing: .12em; text-transform: uppercase; color: var(--sage); margin-bottom: 4px; }
.ratio-num .rn-value { font-size: 1.5rem; font-weight: 700; color: #fff; }
.ratio-den { background: var(--parchment); color: var(--ink); }
.ratio-den .rn-label { font-size: .72rem; font-family: 'Source Code Pro', monospace;
  letter-spacing: .12em; text-transform: uppercase; color: var(--dust); margin-bottom: 4px; }
.ratio-den .rn-value { font-size: 1.5rem; font-weight: 700; color: var(--charcoal); }
.ratio-div { font-size: 2.5rem; color: var(--amber); font-weight: 700; line-height: 1; }
.ratio-eq  { font-size: 1.3rem; color: var(--dust); }
.ratio-result { font-size: 1.6rem; font-weight: 700;
  font-family: 'Source Code Pro', monospace; color: var(--amber); }

/* Step badges */
.step-badge {
  display: inline-block; background: var(--forest); color: #fff;
  font-family: 'Source Code Pro', monospace; font-size: .68rem; font-weight: 700;
  letter-spacing: .1em; text-transform: uppercase;
  padding: 3px 10px; border-radius: 3px; margin-right: 8px; vertical-align: middle;
}

.key-box {
  background: var(--forest); color: #fff; border-radius: 6px;
  padding: 20px 24px; margin: 2em 0; font-size: .9rem;
}
.key-box h4 {
  font-family: 'Libre Baskerville', serif; color: var(--sage);
  font-size: 1rem; margin-bottom: 10px;
}
.key-box ul { padding-left: 1.3em; }
.key-box li { margin-bottom: 5px; color: rgba(255,255,255,.85); }
.key-box code { background: rgba(255,255,255,.12); color: var(--amber-lt); }

table { font-size: .9rem; border-collapse: collapse; width: 100%; margin: 1.2em 0; }
thead th {
  background: var(--forest); color: #fff;
  font-family: 'Source Code Pro', monospace; font-size: .76rem;
  letter-spacing: .06em; text-transform: uppercase; padding: 8px 12px; text-align: left;
}
tbody tr:nth-child(even) { background: var(--mist); }
tbody td { padding: 7px 12px; border-bottom: 1px solid var(--parchment); }

.page-footer {
  background: var(--ink); color: rgba(255,255,255,.42); text-align: center;
  padding: 24px 4vw; margin: 60px -4vw 0;
  font-size: .76rem; font-family: 'Source Code Pro', monospace; letter-spacing: .04em;
}
.page-footer a { color: var(--sage); text-decoration: none; }
.page-footer .divider { margin: 0 8px; opacity: .3; }
```

<div class="page-header">
<div class="module-tag">EA030 &mdash; Basic Statistics &mdash; Module 04</div>
<h1>Signal <span>Over Noise</span></h1>
<p class="sub"><strong>Prof. Marc Los Huertos</strong> &nbsp;&bull;&nbsp; Pomona College</p>
<a class="back-link" href="index.html">&#8592; Back to Basic Statistics</a>
</div>

<div class="section-nav">
<a href="#core-idea">The Core Idea</a>
<a href="#null-world">The Null World</a>
<a href="#monte-carlo">Monte Carlo</a>
<a href="#permutation">Permutation Test</a>
<a href="#t-dist">The t-Distribution</a>
<a href="#f-dist">The F-Distribution</a>
<a href="#convergence">Two Roads, One Answer</a>
</div>

> **What this module covers.** Every classical statistical test — t-test, ANOVA,
> regression — is a formalisation of the same underlying question: *is the pattern
> I see in my data larger than what I would expect from chance variation alone?*
> This module builds that intuition from scratch using Monte Carlo simulation,
> then shows how the t and F distributions are just efficient shortcuts to the
> same answer.

---

## 1. The Core Idea: Signal Over Noise {#core-idea}

Every statistical test computes some version of this ratio:

<div class="ratio-display">
<div class="ratio-num">
  <div class="rn-label">Signal</div>
  <div class="rn-value">effect size</div>
</div>
<div class="ratio-div">÷</div>
<div class="ratio-den">
  <div class="rn-label">Noise</div>
  <div class="rn-value">chance variation</div>
</div>
<div class="ratio-eq">=</div>
<div class="ratio-result">test statistic</div>
</div>

**Signal** is the pattern you care about: the difference between group means,
the slope of a regression line, the change in frequency across categories.

**Noise** is the variation that exists even when nothing interesting is happening:
individual differences between organisms, measurement error, spatial and temporal
heterogeneity, the irreducible randomness of the natural world.

A test statistic close to **zero** means signal and noise are the same size —
the pattern is indistinguishable from background variation. A large test statistic
means the signal is much larger than expected from noise alone — the pattern is
*rare* under the assumption that nothing is happening.

```{r signal-noise-intuition, echo=FALSE, fig.cap="Three scenarios with different signal-to-noise ratios. Left: large signal, small noise — the groups are clearly different. Middle: moderate signal, moderate noise — uncertain. Right: small signal, large noise — the groups are indistinguishable. The *ratio* determines what a test can detect, not either quantity alone.", fig.height=3.8}
set.seed(42)
par(mfrow = c(1, 3), mar = c(3, 3, 3, 1), las = 1)

scenarios <- list(
  list(mu = c(5, 9), sd = 0.8, title = "Large signal\nSmall noise\nS/N = 5.0"),
  list(mu = c(5, 7), sd = 1.5, title = "Moderate signal\nModerate noise\nS/N = 1.3"),
  list(mu = c(5, 5.5), sd = 3.0, title = "Small signal\nLarge noise\nS/N = 0.2")
)

for (s in scenarios) {
  grp <- rep(c("A", "B"), each = 30)
  y   <- c(rnorm(30, s$mu[1], s$sd), rnorm(30, s$mu[2], s$sd))
  boxplot(y ~ grp,
          col    = c("#4a7c5e", "#c8832a"),
          border = "#1e3a2f",
          main   = s$title,
          xlab   = "", ylab   = "",
          outline = TRUE,
          cex.main = 0.88,
          frame.plot = FALSE)
  stripchart(y ~ grp, method = "jitter", jitter = 0.12,
             pch = 20, cex = 0.6, col = adjustcolor("#1e3a2f", 0.4),
             vertical = TRUE, add = TRUE)
}
par(mfrow = c(1, 1))
```

### Signal-to-noise in the t-statistic

For a two-group comparison, the signal-to-noise ratio is the t-statistic:

$$t = \frac{\underbrace{\bar{x}_A - \bar{x}_B}_{\text{signal: observed difference}}}
           {\underbrace{SE_{\text{diff}}}_{\text{noise: expected sampling variation}}}$$

For the F-statistic in ANOVA:

$$F = \frac{\underbrace{MS_{\text{between}}}_{\text{signal: variance explained by groups}}}
           {\underbrace{MS_{\text{within}}}_{\text{noise: variance within groups}}}$$

Both are signal-over-noise ratios. They differ only in what "signal" and "noise" mean
geometrically. The core question is always the same: **how often would we see a ratio
this large if there were actually no effect?**

---

## 2. The Null World — What If Nothing Were Happening? {#null-world}

To decide whether an observed signal-to-noise ratio is large, we need to know the
**distribution of that ratio under the null hypothesis** — i.e., in a world where the
groups have identical population means and any apparent difference is pure chance.

There are two ways to construct this null distribution:

1. **Simulation (Monte Carlo / permutation):** Shuffle the data to destroy any real
   group structure. Compute the test statistic many times. The histogram of those
   values *is* the empirical null distribution.

2. **Mathematics (t, F distributions):** Mathematicians derived the exact shape of
   these null distributions analytically. We compare our observed statistic to the
   theoretical curve rather than simulating.

Both approaches answer the same question. Simulation is slower but requires no
assumptions beyond exchangeability. Theory is instant but requires the data to satisfy
certain conditions (normality, equal variance, independence). We build the simulation
first because it makes the logic completely transparent.

---

## 3. Monte Carlo Simulation: Building Intuition {#monte-carlo}

Before looking at real data, let us build intuition by *watching* what happens when
we know for certain that the null hypothesis is true.

### Generating null data

We will create two groups drawn from **identical** populations. Any difference
between their means is pure sampling noise — there is no signal by construction.

```{r null-generation}
set.seed(7)

# Two groups from the SAME population — null is true by construction
n_per_group  <- 20
true_mean    <- 10
true_sd      <- 3

group_A_null <- rnorm(n_per_group, mean = true_mean, sd = true_sd)
group_B_null <- rnorm(n_per_group, mean = true_mean, sd = true_sd)

cat("Group A mean:", round(mean(group_A_null), 2), "\n")
cat("Group B mean:", round(mean(group_B_null), 2), "\n")
cat("Observed difference:", round(mean(group_A_null) - mean(group_B_null), 2), "\n")
```

Even though both groups were drawn from the same distribution, we observe a
non-zero difference. This is sampling variation — noise. The question is: *how
often does sampling produce a difference at least this large?*

### Repeating the experiment 5,000 times

We simulate 5,000 "experiments" under the null (both groups from the same population),
record the mean difference each time, and build the null distribution.

```{r monte-carlo-null, cache=FALSE}
set.seed(99)
n_sims       <- 5000
null_diffs   <- numeric(n_sims)

for (i in seq_len(n_sims)) {
  A <- rnorm(n_per_group, mean = true_mean, sd = true_sd)
  B <- rnorm(n_per_group, mean = true_mean, sd = true_sd)
  null_diffs[i] <- mean(A) - mean(B)
}

cat("Null distribution summary:\n")
round(quantile(null_diffs, c(0.025, 0.05, 0.25, 0.50, 0.75, 0.95, 0.975)), 2)
```

```{r mc-null-plot, echo=FALSE, fig.cap="The null distribution of mean differences when both groups are drawn from the same population (n = 20 each, 5,000 simulations). The distribution is centred on zero and roughly normal. The amber shaded tails contain 5% of simulations — our conventional Type I error threshold.", fig.height=4}
par(mar = c(4, 4, 3, 1), las = 1)

hist(null_diffs,
     breaks = 60,
     col    = "#d4e6da",
     border = "white",
     main   = "Null Distribution of Mean Differences (5,000 simulations)",
     xlab   = expression(bar(x)[A] - bar(x)[B]),
     ylab   = "Frequency",
     cex.main = 0.92)

# Shade the 5% tails
q025 <- quantile(null_diffs, 0.025)
q975 <- quantile(null_diffs, 0.975)

tail_breaks_lo <- hist(null_diffs[null_diffs <= q025], breaks = 60, plot = FALSE)
tail_breaks_hi <- hist(null_diffs[null_diffs >= q975], breaks = 60, plot = FALSE)

rect(tail_breaks_lo$breaks[-length(tail_breaks_lo$breaks)], 0,
     tail_breaks_lo$breaks[-1], tail_breaks_lo$counts,
     col = adjustcolor("#c8832a", 0.7), border = NA)
rect(tail_breaks_hi$breaks[-length(tail_breaks_hi$breaks)], 0,
     tail_breaks_hi$breaks[-1], tail_breaks_hi$counts,
     col = adjustcolor("#c8832a", 0.7), border = NA)

abline(v = 0,    col = "#1e3a2f",  lwd = 2, lty = 1)
abline(v = q025, col = "#c8832a",  lwd = 1.5, lty = 2)
abline(v = q975, col = "#c8832a",  lwd = 1.5, lty = 2)

legend("topright",
       legend = c("Null mean (0)", "2.5th / 97.5th percentile", "5% tail region"),
       col    = c("#1e3a2f", "#c8832a", adjustcolor("#c8832a", 0.7)),
       lty    = c(1, 2, NA), pch = c(NA, NA, 15),
       bty    = "n", cex = 0.82)

text(q025 - 0.05, max(table(cut(null_diffs, 60))) * 0.85,
     paste0("2.5%\n(", round(q025, 2), ")"),
     col = "#c8832a", cex = 0.75, adj = 1)
text(q975 + 0.05, max(table(cut(null_diffs, 60))) * 0.85,
     paste0("97.5%\n(", round(q975, 2), ")"),
     col = "#c8832a", cex = 0.75, adj = 0)
```

<div class="callout callout-concept">
<strong>📌 What the null distribution tells us</strong>

The amber tails contain exactly 5% of all simulated differences — values so
extreme they would arise by chance less than 1 in 20 times when the null is
true. This 5% is our conventional **α = 0.05** threshold. An observed difference
that falls in the tail is called "statistically significant" — not because it
is important, but because it is <em>rare under the null hypothesis</em>.
</div>

---

## 4. Permutation Test on Real Data {#permutation}

Now we apply the same logic to real data — but instead of generating new data from
a known distribution, we **shuffle** the existing observations to destroy the real
group structure. This is a **permutation test**, and it is the most assumption-free
way to construct a null distribution.

### The dataset: lead in drinking water by pipe material

```{r lead-data}
set.seed(3)
# Lead concentration (µg/L) sampled from buildings with different pipe material
# EPA action level: 15 µg/L
lead <- data.frame(
  pipe    = rep(c("Lead/Lead-solder", "Copper/PVC"), c(28, 32)),
  lead_ppb = c(
    round(rlnorm(28, meanlog = 2.6, sdlog = 0.55)),   # older lead pipes
    round(rlnorm(32, meanlog = 1.6, sdlog = 0.50))    # modern copper/PVC
  )
)
```

```{r lead-plot, fig.cap="Lead concentration by pipe material. Each point is one building sample. The EPA action level (15 µg/L) is shown as a dashed line. Lead/lead-solder pipes produce substantially higher concentrations.", fig.height=4}
ggplot(lead, aes(pipe, lead_ppb, fill = pipe)) +
  geom_boxplot(width = 0.45, outlier.shape = NA,
               colour = "#1e3a2f", alpha = 0.75, linewidth = 0.6) +
  geom_jitter(aes(colour = pipe), width = 0.18,
              size = 1.8, alpha = 0.5) +
  geom_hline(yintercept = 15, linetype = "dashed",
             colour = "#bf360c", linewidth = 0.9) +
  annotate("text", x = 2.48, y = 17,
           label = "EPA action level (15 µg/L)",
           colour = "#bf360c", size = 3.2, hjust = 1, fontface = "italic") +
  scale_fill_manual(values  = c("Lead/Lead-solder" = "#7a3a2f",
                                "Copper/PVC"       = "#4a7c5e")) +
  scale_colour_manual(values = c("Lead/Lead-solder" = "#7a3a2f",
                                 "Copper/PVC"       = "#4a7c5e")) +
  labs(x = NULL, y = "Lead concentration (µg/L)",
       title = "Tap Water Lead Concentration by Pipe Material") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title         = element_text(family = "serif", face = "bold", colour = "#1e3a2f"),
    legend.position    = "none",
    panel.grid.major.x = element_blank()
  )
```

```{r lead-observed}
# The observed difference in means — our "signal"
obs_diff <- mean(lead$lead_ppb[lead$pipe == "Lead/Lead-solder"]) -
            mean(lead$lead_ppb[lead$pipe == "Copper/PVC"])
cat("Observed mean difference:", round(obs_diff, 2), "µg/L\n")
```

### <span class="step-badge">Step 1</span> Understand what shuffling does

The key insight: if pipe material has **no effect** on lead concentration, then the
labels "Lead/Lead-solder" and "Copper/PVC" are arbitrary — we could randomly
reassign them and the difference in means would look just like our real difference.

Shuffling the labels destroys any real association while preserving the marginal
distribution of the data. The collection of mean differences from many shuffles
*is* the null distribution.

```{r shuffle-demo}
# One single shuffle — what does a "null" difference look like?
set.seed(42)
shuffled_labels <- sample(lead$pipe)    # randomly reassign pipe labels
one_null_diff   <- mean(lead$lead_ppb[shuffled_labels == "Lead/Lead-solder"]) -
                   mean(lead$lead_ppb[shuffled_labels == "Copper/PVC"])
cat("One null difference (shuffled):", round(one_null_diff, 2), "µg/L\n")
cat("Observed difference:           ", round(obs_diff, 2),      "µg/L\n")
```

### <span class="step-badge">Step 2</span> Repeat 9,999 times

```{r permutation-test, cache=FALSE}
set.seed(1234)
n_perms     <- 9999
perm_diffs  <- numeric(n_perms)

for (i in seq_len(n_perms)) {
  shuffled        <- sample(lead$pipe)
  perm_diffs[i]   <- mean(lead$lead_ppb[shuffled == "Lead/Lead-solder"]) -
                     mean(lead$lead_ppb[shuffled == "Copper/PVC"])
}

# p-value: proportion of null differences >= observed (one-tailed)
# Two-tailed: proportion of |null differences| >= |observed|
p_perm <- mean(abs(perm_diffs) >= abs(obs_diff))
cat("Permutation p-value (two-tailed):", p_perm, "\n")
cat("Observed difference falls in the",
    round(mean(perm_diffs < obs_diff) * 100, 1),
    "th percentile of the null distribution.\n")
```

### <span class="step-badge">Step 3</span> Visualise the observed statistic against the null

```{r permutation-plot, echo=FALSE, fig.cap="The permutation null distribution of mean differences (9,999 shuffles). The observed difference of ~9 µg/L (red line) sits far into the right tail — it would almost never arise by chance if pipe material had no effect. The permutation p-value is the proportion of null differences exceeding the red line.", fig.height=4.5}
par(mar = c(4.5, 4, 3.5, 1), las = 1)

hist(perm_diffs,
     breaks   = 80,
     col      = "#d4e6da",
     border   = "white",
     main     = "Permutation Null Distribution — Lead Concentration",
     xlab     = "Mean difference: Lead pipes − Copper/PVC (µg/L)",
     ylab     = "Frequency",
     cex.main = 0.92,
     xlim     = range(c(perm_diffs, obs_diff * 1.1)))

# Shade region more extreme than observed
extreme_lo <- perm_diffs[perm_diffs <=  -abs(obs_diff)]
extreme_hi <- perm_diffs[perm_diffs >=   abs(obs_diff)]

if (length(extreme_hi) > 0) {
  h_hi <- hist(extreme_hi, breaks = 80, plot = FALSE)
  rect(h_hi$breaks[-length(h_hi$breaks)], 0,
       h_hi$breaks[-1], h_hi$counts,
       col = adjustcolor("#c8832a", 0.65), border = NA)
}
if (length(extreme_lo) > 0) {
  h_lo <- hist(extreme_lo, breaks = 80, plot = FALSE)
  rect(h_lo$breaks[-length(h_lo$breaks)], 0,
       h_lo$breaks[-1], h_lo$counts,
       col = adjustcolor("#c8832a", 0.65), border = NA)
}

abline(v =  obs_diff, col = "#bf360c", lwd = 2.5, lty = 1)
abline(v = -obs_diff, col = "#bf360c", lwd = 2.5, lty = 2)
abline(v = 0,         col = "#1e3a2f", lwd = 1.5, lty = 3)

legend("topright",
       legend = c(
         paste0("Observed diff = +", round(obs_diff, 1), " µg/L"),
         paste0("Mirror (−", round(obs_diff, 1), " µg/L)"),
         paste0("Null mean (0)"),
         paste0("p = ", p_perm, "  (shaded area)")
       ),
       col  = c("#bf360c", "#bf360c", "#1e3a2f", adjustcolor("#c8832a", 0.65)),
       lty  = c(1, 2, 3, NA), pch = c(NA, NA, NA, 15),
       bty  = "n", cex = 0.80)
```

<div class="callout callout-formula">
<strong>⚙️ Computing the permutation p-value</strong><br>
<code>p_value &lt;- mean(abs(perm_diffs) &gt;= abs(obs_diff))</code><br><br>
This counts the fraction of shuffled experiments that produced a difference
at least as extreme as the one we observed. That fraction is the probability
of seeing our data — or something more extreme — if the null were true.
No formula required. The simulation <em>is</em> the calculation.
</div>

---

## 5. From Simulation to Theory: The t-Distribution {#t-dist}

The permutation test is the most honest approach, but it requires computing
thousands of shuffles. Mathematicians asked: *what shape does the null distribution
of a signal-to-noise ratio always take, if the data are normally distributed?*
The answer is the **t-distribution**.

### Watching the t-statistic accumulate under the null

We repeat the same logic as the Monte Carlo section, but now we standardise the
mean difference by its standard error — computing the t-statistic each time.

```{r t-null-build, cache=FALSE}
set.seed(77)
n_sims  <- 10000
t_nulls <- numeric(n_sims)

for (i in seq_len(n_sims)) {
  A <- rnorm(20, mean = 0, sd = 1)   # both from N(0,1) — null true
  B <- rnorm(20, mean = 0, sd = 1)
  t_nulls[i] <- t.test(A, B)$statistic
}
```

```{r t-dist-plot, echo=FALSE, fig.cap="Histogram of 10,000 simulated t-statistics when the null is true (both groups from N(0,1), n = 20 each). The theoretical t-distribution with df = 38 (amber curve) fits perfectly. The simulation and the formula describe the same null distribution — the formula is just much faster.", fig.height=4.5}
par(mar = c(4.5, 4, 3.5, 1), las = 1)

hist(t_nulls,
     breaks  = 70,
     freq    = FALSE,
     col     = "#d4e6da",
     border  = "white",
     main    = "Simulated t-statistics vs. Theoretical t-distribution (df = 38)",
     xlab    = "t-statistic",
     ylab    = "Density",
     xlim    = c(-5, 5),
     cex.main = 0.92)

# Overlay theoretical t-distribution
x_seq <- seq(-5, 5, length.out = 500)
lines(x_seq, dt(x_seq, df = 38),
      col = "#c8832a", lwd = 2.5)

# Mark critical values
t_crit <- qt(0.975, df = 38)
abline(v =  t_crit, col = "#1e3a2f", lwd = 1.5, lty = 2)
abline(v = -t_crit, col = "#1e3a2f", lwd = 1.5, lty = 2)

legend("topright",
       legend = c("Simulated t-statistics", "Theoretical t(df=38)",
                  paste0("±", round(t_crit, 2), "  (α = 0.05 cutoffs)")),
       col    = c(adjustcolor("#4a7c5e", 0.8), "#c8832a", "#1e3a2f"),
       lty    = c(NA, 1, 2), pch = c(15, NA, NA),
       bty    = "n", cex = 0.82)
```

### Applying the t-test to the lead data

```{r t-test-lead}
t_result <- t.test(lead_ppb ~ pipe, data = lead)
print(t_result)
```

```{r t-test-plot, echo=FALSE, fig.cap="The observed t-statistic from the lead data (red line) placed on the theoretical t-distribution under the null. The shaded amber region is the two-tailed rejection area (α = 0.05). Our observed t sits far into the tail — the same conclusion reached by the permutation test, now from a formula.", fig.height=4.5}
df_t     <- t_result$parameter          # degrees of freedom
t_obs    <- abs(t_result$statistic)     # observed |t|
t_crit_2 <- qt(0.975, df = df_t)

par(mar = c(4.5, 4, 3.5, 1), las = 1)
x_seq <- seq(-6, 6, length.out = 600)

plot(x_seq, dt(x_seq, df = df_t),
     type = "l", lwd = 2.5, col = "#1e3a2f",
     main = paste0("t-distribution (df = ", round(df_t, 0),
                   ") — Lead Pipe vs. Copper/PVC"),
     xlab = "t-statistic", ylab = "Density",
     ylim = c(0, 0.42), cex.main = 0.92)

# Shade rejection tails
x_lo <- x_seq[x_seq <= -t_crit_2]
x_hi <- x_seq[x_seq >=  t_crit_2]
polygon(c(x_lo[1], x_lo, x_lo[length(x_lo)]),
        c(0, dt(x_lo, df_t), 0),
        col = adjustcolor("#c8832a", 0.55), border = NA)
polygon(c(x_hi[1], x_hi, x_hi[length(x_hi)]),
        c(0, dt(x_hi, df_t), 0),
        col = adjustcolor("#c8832a", 0.55), border = NA)

# Shade area more extreme than observed
x_obs_hi <- x_seq[x_seq >=  t_obs]
x_obs_lo <- x_seq[x_seq <= -t_obs]
polygon(c(x_obs_hi[1], x_obs_hi, x_obs_hi[length(x_obs_hi)]),
        c(0, dt(x_obs_hi, df_t), 0),
        col = adjustcolor("#bf360c", 0.45), border = NA)
polygon(c(x_obs_lo[1], x_obs_lo, x_obs_lo[length(x_obs_lo)]),
        c(0, dt(x_obs_lo, df_t), 0),
        col = adjustcolor("#bf360c", 0.45), border = NA)

abline(v =  t_obs, col = "#bf360c", lwd = 2.5)
abline(v = -t_obs, col = "#bf360c", lwd = 2.5, lty = 2)
abline(v =  t_crit_2, col = "#c8832a", lwd = 1.4, lty = 3)
abline(v = -t_crit_2, col = "#c8832a", lwd = 1.4, lty = 3)

legend("topright",
       legend = c(
         paste0("Observed |t| = ", round(t_obs, 2)),
         paste0("Critical t = ±", round(t_crit_2, 2), "  (α = 0.05)"),
         paste0("p = ", formatC(t_result$p.value, format = "e", digits = 2))
       ),
       col  = c("#bf360c", "#c8832a", adjustcolor("#bf360c", 0.45)),
       lty  = c(1, 3, NA), pch = c(NA, NA, 15),
       bty  = "n", cex = 0.82)
```

### The t-distribution family: degrees of freedom

The t-distribution is actually a *family* of curves, indexed by **degrees of freedom**
(df). With small samples, the tails are heavier — reflecting greater uncertainty.
As df → ∞, the t-distribution converges to the standard normal.

```{r t-family, echo=FALSE, fig.cap="The t-distribution for different degrees of freedom. Small df produces heavy tails — small samples generate more extreme statistics by chance, so the critical value must be larger to maintain α = 0.05. With df ≥ 30, the t-distribution is almost indistinguishable from the normal.", fig.height=4}
par(mar = c(4, 4, 3, 1), las = 1)
x_seq <- seq(-5, 5, length.out = 500)

plot(x_seq, dnorm(x_seq), type = "l",
     col = "#1a1a18", lwd = 2.5, lty = 1,
     ylim = c(0, 0.42),
     main = "t-distribution Family by Degrees of Freedom",
     xlab = "t", ylab = "Density",
     cex.main = 0.92)

dfs   <- c(3, 5, 10, 30)
cols  <- c("#bf360c", "#c8832a", "#4a7c5e", "#8aad96")
lwidths <- c(2, 1.8, 1.5, 1.2)

for (k in seq_along(dfs)) {
  lines(x_seq, dt(x_seq, df = dfs[k]),
        col = cols[k], lwd = lwidths[k], lty = 1)
}

t_crits <- c(qt(0.975, 3), qt(0.975, 5), qt(0.975, 10),
             qt(0.975, 30), qnorm(0.975))
crit_labels <- paste0(c("df=3", "df=5", "df=10", "df=30", "Normal"),
                      "\nt*=", round(t_crits, 2))

legend("topright",
       legend = crit_labels,
       col    = c(cols, "#1a1a18"),
       lwd    = c(lwidths, 2.5), lty = 1,
       bty    = "n", cex = 0.72)
```

<div class="callout callout-concept">
<strong>📌 Why degrees of freedom shrink the distribution's tails</strong>

With only a few observations, your sample standard deviation is an unreliable
estimate of the population SD. Sometimes it underestimates by a lot — which
makes the t-statistic artificially large even when the null is true.
The t-distribution with small df compensates by having heavier tails:
it expects to see large values occasionally just from estimation error,
so it demands a larger observed t before calling a result "rare."
As sample size grows, the SD estimate becomes reliable and the t-distribution
narrows toward the normal.
</div>

---

## 6. The F-Distribution: Signal Over Noise for Multiple Groups {#f-dist}

When comparing **three or more** groups, the t-statistic generalises to the
**F-statistic**, which is the ratio of variance *between* groups to variance
*within* groups. The same signal-over-noise logic applies.

### Building the F null distribution by simulation

```{r f-null-build, cache=FALSE}
set.seed(55)
n_sims  <- 10000
f_nulls <- numeric(n_sims)

# Null: three groups from identical populations
for (i in seq_len(n_sims)) {
  grp <- factor(rep(c("A","B","C"), each = 15))
  y   <- rnorm(45, mean = 0, sd = 1)
  f_nulls[i] <- summary(aov(y ~ grp))[[1]]["grp", "F value"]
}
```

```{r f-null-plot, echo=FALSE, fig.cap="Simulated F-statistics when all three groups are drawn from the same population (n = 15 per group). The theoretical F-distribution F(2, 42) (amber curve) fits the simulation perfectly. The F-distribution is right-skewed and bounded below by zero — unlike the symmetric t-distribution.", fig.height=4.5}
par(mar = c(4.5, 4, 3.5, 1), las = 1)

hist(f_nulls,
     breaks  = 80,
     freq    = FALSE,
     col     = "#d4e6da",
     border  = "white",
     main    = "Simulated F-statistics vs. Theoretical F(2, 42) — Null True",
     xlab    = "F-statistic",
     ylab    = "Density",
     xlim    = c(0, 8),
     cex.main = 0.92)

x_f <- seq(0.001, 8, length.out = 500)
lines(x_f, df(x_f, df1 = 2, df2 = 42),
      col = "#c8832a", lwd = 2.5)

f_crit <- qf(0.95, df1 = 2, df2 = 42)
abline(v = f_crit, col = "#1e3a2f", lwd = 1.8, lty = 2)

# shade rejection region
x_rej <- x_f[x_f >= f_crit]
polygon(c(x_rej[1], x_rej, x_rej[length(x_rej)]),
        c(0, df(x_rej, 2, 42), 0),
        col = adjustcolor("#c8832a", 0.5), border = NA)

legend("topright",
       legend = c("Simulated F-statistics", "Theoretical F(2, 42)",
                  paste0("Critical F = ", round(f_crit, 2), "  (α = 0.05)"),
                  "Rejection region (5%)"),
       col    = c(adjustcolor("#4a7c5e", 0.8), "#c8832a", "#1e3a2f",
                  adjustcolor("#c8832a", 0.5)),
       lty    = c(NA, 1, 2, NA), pch = c(15, NA, NA, 15),
       bty    = "n", cex = 0.80)
```

### Applying the F-test: nitrogen across land-use types

```{r f-real-data}
set.seed(42)
nitrogen <- data.frame(
  landuse = rep(c("Forest", "Grassland", "Cropland"), each = 25),
  N_gkg   = c(rnorm(25, 4.8, 0.7),
              rnorm(25, 3.2, 0.9),
              rnorm(25, 1.9, 0.6))
)

model_aov <- aov(N_gkg ~ landuse, data = nitrogen)
aov_table <- summary(model_aov)[[1]]
F_obs     <- aov_table["landuse", "F value"]
df1       <- aov_table["landuse", "Df"]
df2       <- aov_table["Residuals", "Df"]

cat("Between-group MS (signal):", round(aov_table["landuse",   "Mean Sq"], 3), "\n")
cat("Within-group MS  (noise): ", round(aov_table["Residuals", "Mean Sq"], 3), "\n")
cat("F = signal / noise =",       round(F_obs, 2), "\n")
cat("p-value:                 ",  round(aov_table["landuse", "Pr(>F)"], 6), "\n")
```

```{r f-observed-plot, echo=FALSE, fig.cap="The observed F-statistic (red line) from the soil nitrogen ANOVA placed on the F(2, 72) null distribution. The observed F is so large it falls entirely beyond the plotted rejection region — correspondingly, p < 0.0001. The ratio of between-group to within-group variance is far larger than chance produces.", fig.height=4.5}
par(mar = c(4.5, 4, 3.5, 1), las = 1)

x_f2 <- seq(0.001, max(F_obs * 1.15, 6), length.out = 600)
f_crit2 <- qf(0.95, df1 = df1, df2 = df2)

plot(x_f2, df(x_f2, df1, df2),
     type = "l", lwd = 2.5, col = "#1e3a2f",
     main = paste0("F(", df1, ", ", df2, ") — Soil Nitrogen ANOVA"),
     xlab = "F-statistic", ylab = "Density",
     xlim = c(0, max(F_obs * 1.1, 6)),
     ylim = c(0, 0.80),
     cex.main = 0.92)

# Shade rejection region (theoretical)
x_rej2 <- x_f2[x_f2 >= f_crit2]
polygon(c(x_rej2[1], x_rej2, x_rej2[length(x_rej2)]),
        c(0, df(x_rej2, df1, df2), 0),
        col = adjustcolor("#c8832a", 0.5), border = NA)

# Shade area beyond observed F
x_obs_f <- x_f2[x_f2 >= F_obs]
if (length(x_obs_f) > 0) {
  polygon(c(x_obs_f[1], x_obs_f, x_obs_f[length(x_obs_f)]),
          c(0, df(x_obs_f, df1, df2), 0),
          col = adjustcolor("#bf360c", 0.5), border = NA)
}

abline(v = f_crit2, col = "#c8832a", lwd = 1.8, lty = 2)
abline(v = F_obs,   col = "#bf360c", lwd = 2.5)

legend("topright",
       legend = c(
         paste0("Observed F = ", round(F_obs, 1)),
         paste0("Critical F = ", round(f_crit2, 2), "  (α = 0.05)"),
         "5% rejection region",
         paste0("p(F ≥ ", round(F_obs, 1), ") ≈ 0")
       ),
       col  = c("#bf360c", "#c8832a",
                adjustcolor("#c8832a", 0.5), adjustcolor("#bf360c", 0.5)),
       lty  = c(1, 2, NA, NA), pch = c(NA, NA, 15, 15),
       bty  = "n", cex = 0.82)

# Decomposition annotation
legend("topleft",
       legend = c(
         paste0("Signal (MS_between) = ", round(aov_table["landuse",   "Mean Sq"], 2)),
         paste0("Noise  (MS_within)  = ", round(aov_table["Residuals", "Mean Sq"], 2)),
         paste0("F = ", round(F_obs, 2))
       ),
       bty = "n", cex = 0.80,
       text.col = c("#1e3a2f", "#4a7c5e", "#bf360c"))
```

### Decomposing the variance in a table

```{r anova-decomposition}
# Manual variance decomposition — showing the signal/noise arithmetic
grand_mean  <- mean(nitrogen$N_gkg)
group_means <- tapply(nitrogen$N_gkg, nitrogen$landuse, mean)
group_sizes <- tapply(nitrogen$N_gkg, nitrogen$landuse, length)

SS_between <- sum(group_sizes * (group_means - grand_mean)^2)
SS_within  <- sum(tapply(seq_along(nitrogen$N_gkg), nitrogen$landuse,
                  function(i) sum((nitrogen$N_gkg[i] - mean(nitrogen$N_gkg[i]))^2)))
SS_total   <- sum((nitrogen$N_gkg - grand_mean)^2)

df_between <- length(group_means) - 1
df_within  <- nrow(nitrogen) - length(group_means)

MS_between <- SS_between / df_between
MS_within  <- SS_within  / df_within

knitr::kable(
  data.frame(
    Source     = c("Between groups (signal)", "Within groups (noise)", "Total"),
    SS         = round(c(SS_between, SS_within, SS_total), 2),
    df         = c(df_between, df_within, df_between + df_within),
    MS         = round(c(MS_between, MS_within, NA), 3),
    F          = c(round(MS_between / MS_within, 2), NA, NA),
    `p-value`  = c(round(pf(MS_between/MS_within, df_between, df_within,
                            lower.tail = FALSE), 6), NA, NA)
  ),
  col.names = c("Source", "Sum of Squares", "df", "Mean Square", "F", "p-value"),
  caption = "ANOVA decomposition: the F-statistic is the ratio of between-group mean square (signal) to within-group mean square (noise).",
  na.encode = FALSE
)
```

---

## 7. Two Roads to the Same Answer {#convergence}

The permutation test and the t/F-test give the same conclusion because they are
asking the same question in different ways. Here we verify this side-by-side.

```{r two-roads, cache=FALSE}
set.seed(888)

# ── Observed F for the nitrogen data ──────────────────────────────────────────
F_obs_check <- summary(aov(N_gkg ~ landuse, data = nitrogen))[[1]][1, "F value"]

# ── Permutation F-distribution ────────────────────────────────────────────────
n_perm   <- 9999
perm_F   <- numeric(n_perm)

for (i in seq_len(n_perm)) {
  shuffled_groups <- sample(nitrogen$landuse)
  perm_F[i] <- summary(
    aov(nitrogen$N_gkg ~ shuffled_groups)
  )[[1]][1, "F value"]
}

p_permutation <- mean(perm_F >= F_obs_check)

# ── Parametric p-value ────────────────────────────────────────────────────────
p_parametric <- summary(model_aov)[[1]]["landuse", "Pr(>F)"]

cat("─── Permutation test ────────────────────────────────\n")
cat("Permutation p-value:", p_permutation, "\n\n")
cat("─── Parametric F-test ───────────────────────────────\n")
cat("Parametric  p-value:", round(p_parametric, 6), "\n")
```

```{r two-roads-plot, echo=FALSE, fig.cap="Left: the permutation null distribution of F (grey bars) with the observed F marked in red. Right: the theoretical F(2, 72) null distribution with the same observed statistic. Both methods place the observed F far into the tail. The tiny difference between the two p-values is pure Monte Carlo sampling noise.", fig.height=4.5, fig.width=8}
par(mfrow = c(1, 2), mar = c(4, 4, 3, 1), las = 1)

# --- Permutation ---
hist(perm_F,
     breaks = 60, freq = FALSE,
     col = "#d4e6da", border = "white",
     main = "Permutation Null\n(9,999 shuffles)",
     xlab = "F-statistic", ylab = "Density",
     xlim = c(0, max(perm_F, F_obs_check) * 1.05),
     cex.main = 0.9)
extreme_perm <- perm_F[perm_F >= F_obs_check]
if (length(extreme_perm) > 0) {
  h_ex <- hist(extreme_perm, breaks = 60, plot = FALSE)
  rect(h_ex$breaks[-length(h_ex$breaks)], 0,
       h_ex$breaks[-1], h_ex$counts / (length(perm_F) * diff(h_ex$breaks[1:2])),
       col = adjustcolor("#c8832a", 0.6), border = NA)
}
abline(v = F_obs_check, col = "#bf360c", lwd = 2.5)
text(F_obs_check * 0.82, par("usr")[4] * 0.92,
     paste0("p = ", p_permutation),
     col = "#bf360c", cex = 0.82, font = 2)

# --- Theoretical ---
x_f3 <- seq(0.001, max(F_obs_check * 1.1, 6), length.out = 500)
plot(x_f3, df(x_f3, df1 = 2, df2 = 72),
     type = "l", lwd = 2.5, col = "#1e3a2f",
     main = "Theoretical F(2, 72)\n(parametric test)",
     xlab = "F-statistic", ylab = "Density",
     xlim = c(0, max(F_obs_check * 1.1, 6)),
     cex.main = 0.9)
x_rej3 <- x_f3[x_f3 >= F_obs_check]
if (length(x_rej3) > 0) {
  polygon(c(x_rej3[1], x_rej3, x_rej3[length(x_rej3)]),
          c(0, df(x_rej3, 2, 72), 0),
          col = adjustcolor("#c8832a", 0.55), border = NA)
}
abline(v = F_obs_check, col = "#bf360c", lwd = 2.5)
text(F_obs_check * 0.82, par("usr")[4] * 0.92,
     paste0("p ≈ ", formatC(p_parametric, format = "e", digits = 1)),
     col = "#bf360c", cex = 0.82, font = 2)

par(mfrow = c(1, 1))
```

<div class="callout callout-tip">
<strong>💡 When the two approaches disagree</strong>

Permutation and parametric tests give almost identical p-values when the
parametric assumptions (normality, equal variances) are satisfied — as they
do here. They diverge when the data are strongly non-normal or when group
variances differ dramatically. In those cases, the permutation p-value is
more trustworthy because it makes no distributional assumptions. The practical
rule: if `shapiro.test()` or a Q-Q plot raises concerns, check your parametric
p-value against a permutation or bootstrap equivalent.
</div>

---

<div class="key-box">
<h4>🔑 Signal over noise — key takeaways</h4>
<ul>
<li><strong>Every test statistic is a signal-to-noise ratio.</strong> Signal is the pattern you care about (difference in means, slope, variance between groups). Noise is the variation that exists even when nothing is happening (within-group spread, estimation uncertainty).</li>
<li><strong>The p-value is a rarity statement:</strong> "If the null hypothesis were true, data as extreme as mine would arise <em>p</em> × 100% of the time." It is not the probability that the null is true, nor a measure of effect size.</li>
<li><strong>Monte Carlo simulation makes the logic transparent.</strong> Shuffle the labels, compute the statistic, repeat 10,000 times. The histogram you get <em>is</em> the null distribution. The p-value is just the fraction of bars to the right of your observed value.</li>
<li><strong>The t and F distributions are mathematical shortcuts</strong> to the same null distribution, valid when data are approximately normal with equal group variances. Use <code>t.test()</code> and <code>aov()</code> — they are fast and well-studied. Use permutation tests when assumptions are violated.</li>
<li><strong>Degrees of freedom control tail heaviness.</strong> Small samples produce larger test statistics by chance alone, so the critical value must be larger — which is why a t-test with n = 5 requires a bigger observed difference than one with n = 50 to reach the same α threshold.</li>
<li><strong>Statistical significance ≠ practical importance.</strong> A very large sample can make a tiny, ecologically irrelevant difference highly significant. Always report effect sizes (Cohen's d, η², odds ratio) alongside p-values.</li>
</ul>
</div>

---

### What's next?

- **[Four Frameworks](02_Four_Frameworks.html)** — which test to use for each data type
- **[Describing Data](01_Describing_Data.html)** — mean, SD, SE, and confidence intervals
- **[t-test in depth](TimeOutSideClass.html)** — full worked example with course data

```{r session-info, echo=FALSE}
# sessionInfo()
```

<footer class="page-footer">
EA030 Basic Statistics &mdash; Module 04
<span class="divider">|</span> Prof. Marc Los Huertos
<span class="divider">|</span> Pomona College
<span class="divider">|</span> Last updated: `r format(Sys.Date(), "%B %Y")`
<span class="divider">|</span> <a href="https://github.com/marclos/RTricks">GitHub</a>
</footer>
