---
title: "Hypothesis Testing"
author: "Marc Los Huertos"
date: "EA030 · Compiled `r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    self_contained: true
    theme: null
    highlight: null
    toc: false
knit: (function(inputFile, encoding) {
    rmarkdown::render(
      inputFile,
      encoding = encoding,
      output_file = file.path(dirname(inputFile),
                              "../docs/basic_statistics/05_Hypothesis_Testing.html")
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 7,
  fig.height = 4
)
library(ggplot2)
library(dplyr)
```

```{css, echo=FALSE}
@import url('https://fonts.googleapis.com/css2?family=Libre+Baskerville:ital,wght@0,400;0,700;1,400&family=Source+Code+Pro:wght@400;600&family=Lato:wght@300;400;700&display=swap');

*, *::before, *::after { box-sizing: border-box; margin: 0; padding: 0; }

:root {
  --forest:    #1e3a2f;
  --moss:      #2d5a3d;
  --fern:      #4a7c5e;
  --sage:      #8aad96;
  --mist:      #d4e6da;
  --cream:     #f7f3ed;
  --parchment: #ede8df;
  --ink:       #1a1a18;
  --charcoal:  #3a3a38;
  --dust:      #7a7a72;
  --amber:     #c8832a;
  --amber-lt:  #e8a84a;
  --white:     #ffffff;
  --red-dark:  #7a1e1e;
  --card-shadow: 0 2px 10px rgba(30,58,47,.09);
}

html { scroll-behavior: smooth; }

body {
  font-family: 'Lato', sans-serif;
  background: var(--cream);
  color: var(--ink);
  font-size: 17px;
  line-height: 1.75;
  padding-left:  clamp(14px, 4vw, 48px);
  padding-right: clamp(14px, 4vw, 48px);
  padding-bottom: 80px;
}

.main-container {
  max-width: 920px !important;
  margin: 0 auto !important;
  padding: 0 clamp(10px, 2vw, 32px) 80px !important;
}

.page-header {
  margin-left:  calc(-1 * clamp(14px, 4vw, 48px));
  margin-right: calc(-1 * clamp(14px, 4vw, 48px));
}

h1.title, h4.author, h4.date { display: none; }

/* ── Page header ─────────────────────────────────────────────────────────── */
.page-header {
  background-image:
    repeating-linear-gradient(120deg, transparent, transparent 60px,
      rgba(255,255,255,.018) 60px, rgba(255,255,255,.018) 61px),
    linear-gradient(160deg, #1e3a2f 0%, #2d5a3d 60%, #1a3d30 100%);
  color: #fff; padding: 44px 4vw 36px; margin: 0 -4vw 48px;
  border-bottom: 4px solid var(--amber); position: relative; overflow: hidden;
}
.page-header::after {
  content: ''; position: absolute; right: -40px; top: -40px;
  width: 260px; height: 260px; border-radius: 50%;
  background: radial-gradient(circle, rgba(138,173,150,.12) 0%, transparent 70%);
  pointer-events: none;
}
.page-header .module-tag {
  display: inline-block; font-family: 'Source Code Pro', monospace;
  font-size: .68rem; font-weight: 600; letter-spacing: .18em; text-transform: uppercase;
  color: var(--amber-lt); border: 1px solid rgba(200,131,42,.45);
  padding: 3px 10px; border-radius: 2px; margin-bottom: 12px;
}
.page-header h1 {
  font-family: 'Libre Baskerville', serif;
  font-size: clamp(1.6rem, 3vw, 2.3rem);
  font-weight: 700; color: #fff; line-height: 1.2; margin-bottom: 8px;
}
.page-header h1 span { color: var(--sage); }
.page-header .sub { font-size: .9rem; color: rgba(255,255,255,.65); font-weight: 300; }
.page-header .sub strong { color: rgba(255,255,255,.9); font-weight: 700; }
.back-link {
  display: inline-block; margin-top: 16px;
  font-family: 'Source Code Pro', monospace; font-size: .7rem; letter-spacing: .08em;
  color: var(--amber-lt); text-decoration: none;
  border: 1px solid rgba(200,131,42,.4); padding: 4px 11px; border-radius: 3px;
  transition: background .2s;
}
.back-link:hover { background: rgba(200,131,42,.18); }

/* ── Section nav ─────────────────────────────────────────────────────────── */
.section-nav {
  display: flex; gap: 8px; flex-wrap: wrap; margin-bottom: 44px;
  padding-bottom: 20px; border-bottom: 1px solid var(--parchment);
}
.section-nav a {
  font-family: 'Source Code Pro', monospace; font-size: .7rem; font-weight: 600;
  letter-spacing: .08em; text-transform: uppercase;
  color: var(--fern); text-decoration: none;
  border: 1px solid var(--mist); padding: 4px 12px; border-radius: 3px;
  background: var(--white); transition: background .2s, color .2s;
}
.section-nav a:hover { background: var(--mist); color: var(--forest); }

/* ── Headings ────────────────────────────────────────────────────────────── */
h2 {
  font-family: 'Libre Baskerville', serif; font-size: 1.45rem; font-weight: 700;
  color: var(--forest); margin-top: 56px; margin-bottom: 6px;
  padding-bottom: 8px; border-bottom: 1px solid var(--parchment);
}
h3 {
  font-family: 'Libre Baskerville', serif; font-size: 1.1rem; font-weight: 700;
  color: var(--moss); margin-top: 28px; margin-bottom: 4px;
}

p { margin-bottom: 1em; }
strong { color: var(--forest); }

/* ── Code ────────────────────────────────────────────────────────────────── */
code {
  font-family: 'Source Code Pro', monospace; font-size: .84em;
  background: var(--mist); color: var(--moss); padding: 1px 5px; border-radius: 3px;
}
pre {
  background: var(--ink); color: #e8e6df; border-radius: 5px;
  padding: 14px 18px; font-size: .83rem; line-height: 1.55;
  overflow-x: auto; border-left: 4px solid var(--amber); margin: 1.2em 0;
}
pre code { background: transparent; color: inherit; padding: 0; font-size: 1em; }

/* ── Callouts ────────────────────────────────────────────────────────────── */
.callout {
  border-left: 5px solid; padding: .65em 1.1em;
  margin: 1.4em 0; border-radius: 0 5px 5px 0; font-size: .92em; line-height: 1.6;
}
.callout > strong:first-child { display: block; margin-bottom: .3em; }
.callout-concept { border-color: var(--fern);  background: #f0f7f2; }
.callout-tip     { border-color: #2e7d32;      background: #f1f8f1; }
.callout-warn    { border-color: #e65100;      background: #fff8f4; }
.callout-myth    { border-color: var(--red-dark); background: #fdf2f2; }
.callout-formula { border-color: var(--amber); background: #fdf8f0;
                   font-family: 'Source Code Pro', monospace; font-size: .88em; }

/* ── Hypothesis framework display ───────────────────────────────────────── */
.hyp-row {
  display: flex; gap: 12px; margin: 1.6em 0;
}
.hyp-card {
  flex: 1; padding: 14px 18px; border-radius: 6px;
  box-shadow: var(--card-shadow); font-size: .9rem;
}
.hyp-card .hyp-label {
  font-family: 'Source Code Pro', monospace; font-size: .68rem;
  font-weight: 700; letter-spacing: .14em; text-transform: uppercase;
  margin-bottom: 5px;
}
.hyp-card .hyp-stmt {
  font-family: 'Libre Baskerville', serif; font-size: 1rem;
  line-height: 1.4;
}
.hyp-null { background: var(--parchment); }
.hyp-null .hyp-label { color: var(--dust); }
.hyp-null .hyp-stmt  { color: var(--charcoal); }
.hyp-alt  { background: var(--forest); color: #fff; }
.hyp-alt  .hyp-label { color: var(--sage); }
.hyp-alt  .hyp-stmt  { color: rgba(255,255,255,.9); }

/* ── Error type matrix ───────────────────────────────────────────────────── */
.error-matrix {
  width: 100%; border-collapse: separate; border-spacing: 6px; margin: 1.8em 0;
}
.error-matrix th {
  font-family: 'Source Code Pro', monospace; font-size: .72rem;
  letter-spacing: .09em; text-transform: uppercase;
  padding: 9px 14px; border-radius: 4px; text-align: center;
}
.em-header-top  { background: var(--forest); color: #fff; }
.em-header-side { background: var(--moss);   color: #fff; writing-mode: horizontal-tb; }
.em-correct { background: #f1f8f1; border: 1px solid #c8e6c9;
              padding: 12px 16px; border-radius: 5px; text-align: center;
              font-size: .88rem; color: #1b5e20; }
.em-error   { background: #fff8f4; border: 1px solid #ffccbc;
              padding: 12px 16px; border-radius: 5px; text-align: center;
              font-size: .88rem; color: #bf360c; }
.em-label   { font-weight: 700; display: block; margin-bottom: 3px;
              font-family: 'Source Code Pro', monospace; font-size: .78rem;
              letter-spacing: .06em; }

/* ── Effect size gauge ───────────────────────────────────────────────────── */
.effect-gauge {
  display: flex; height: 28px; border-radius: 5px; overflow: hidden;
  margin: .8em 0 .4em; font-family: 'Source Code Pro', monospace; font-size: .72rem;
}
.eg-neg-large { flex: 1.2; background: #7a1e1e; color: #fff;
                display: flex; align-items: center; justify-content: center; }
.eg-neg-med   { flex: 0.8; background: #c8832a; color: #fff;
                display: flex; align-items: center; justify-content: center; }
.eg-neg-small { flex: 0.5; background: #e8a84a; color: #fff;
                display: flex; align-items: center; justify-content: center; }
.eg-zero      { flex: 0.3; background: var(--dust); color: #fff;
                display: flex; align-items: center; justify-content: center; }
.eg-pos-small { flex: 0.5; background: #8aad96; color: var(--forest);
                display: flex; align-items: center; justify-content: center; }
.eg-pos-med   { flex: 0.8; background: #4a7c5e; color: #fff;
                display: flex; align-items: center; justify-content: center; }
.eg-pos-large { flex: 1.2; background: #1e3a2f; color: #fff;
                display: flex; align-items: center; justify-content: center; }

/* ── Misconception cards ─────────────────────────────────────────────────── */
.myth-grid {
  display: grid; grid-template-columns: 1fr 1fr; gap: 14px; margin: 1.8em 0;
}
.myth-card {
  border-radius: 6px; overflow: hidden; box-shadow: var(--card-shadow);
}
.myth-header {
  background: var(--red-dark); color: #fff; padding: 9px 14px;
  font-family: 'Source Code Pro', monospace; font-size: .72rem;
  font-weight: 700; letter-spacing: .12em; text-transform: uppercase;
}
.myth-body { background: #fdf2f2; padding: 12px 14px; }
.myth-body .myth-myth {
  font-size: .88rem; color: #7a1e1e; font-style: italic;
  margin-bottom: 6px; line-height: 1.4;
}
.myth-body .myth-truth {
  font-size: .87rem; color: var(--charcoal); line-height: 1.5;
}
.myth-body .myth-truth strong { color: var(--forest); }

/* ── Step badges ─────────────────────────────────────────────────────────── */
.step-badge {
  display: inline-block; background: var(--forest); color: #fff;
  font-family: 'Source Code Pro', monospace; font-size: .68rem; font-weight: 700;
  letter-spacing: .1em; text-transform: uppercase;
  padding: 3px 10px; border-radius: 3px; margin-right: 8px; vertical-align: middle;
}

/* ── Key box ─────────────────────────────────────────────────────────────── */
.key-box {
  background: var(--forest); color: #fff; border-radius: 6px;
  padding: 20px 24px; margin: 2em 0; font-size: .9rem;
}
.key-box h4 {
  font-family: 'Libre Baskerville', serif; color: var(--sage);
  font-size: 1rem; margin-bottom: 10px;
}
.key-box ul { padding-left: 1.3em; }
.key-box li { margin-bottom: 5px; color: rgba(255,255,255,.85); }
.key-box code { background: rgba(255,255,255,.12); color: var(--amber-lt); }

/* ── Tables ──────────────────────────────────────────────────────────────── */
table { font-size: .9rem; border-collapse: collapse; width: 100%; margin: 1.2em 0; }
thead th {
  background: var(--forest); color: #fff;
  font-family: 'Source Code Pro', monospace; font-size: .76rem;
  letter-spacing: .06em; text-transform: uppercase; padding: 8px 12px; text-align: left;
}
tbody tr:nth-child(even) { background: var(--mist); }
tbody td { padding: 7px 12px; border-bottom: 1px solid var(--parchment); }

/* ── Footer ──────────────────────────────────────────────────────────────── */
.page-footer {
  background: var(--ink); color: rgba(255,255,255,.42); text-align: center;
  padding: 24px 4vw; margin: 60px -4vw 0;
  font-size: .76rem; font-family: 'Source Code Pro', monospace; letter-spacing: .04em;
}
.page-footer a { color: var(--sage); text-decoration: none; }
.page-footer .divider { margin: 0 8px; opacity: .3; }
```

<div class="page-header">
<div class="module-tag">EA030 &mdash; Basic Statistics &mdash; Module 05</div>
<h1>Hypothesis <span>Testing</span></h1>
<p class="sub"><strong>Prof. Marc Los Huertos</strong> &nbsp;&bull;&nbsp; Pomona College</p>
<a class="back-link" href="../index.html">&#8592; Back to Index</a>
</div>

<div class="section-nav">
<a href="#framework">The Framework</a>
<a href="#one-sample">One-Sample t</a>
<a href="#two-sample">Two-Sample t</a>
<a href="#paired">Paired t</a>
<a href="#wilcoxon">Wilcoxon</a>
<a href="#proportions">Proportions</a>
<a href="#effect-size">Effect Sizes</a>
<a href="#pvalues">p-values</a>
<a href="#misconceptions">Misconceptions</a>
</div>

> **What this module covers.** Hypothesis testing is the formal machinery for
> deciding whether a pattern in your sample is strong enough to generalize to the
> population it came from. This module builds the complete workflow — from stating
> hypotheses through choosing a test, checking assumptions, interpreting output,
> and reporting results honestly — and then systematically dismantles the most
> common ways the framework is misunderstood.

---

## 1. The Hypothesis Testing Framework {#framework}

Every hypothesis test follows the same five-step logic regardless of which
specific test you use. Understanding the framework first makes every test
a variation on a familiar theme rather than a new procedure to memorise.

### The five steps

**Step 1 — State the hypotheses.**
Every test has exactly two complementary hypotheses. The **null hypothesis
(H₀)** is the default claim — usually "no effect", "no difference", or
"the parameter equals a specific value". The **alternative hypothesis (H₁)**
is what you would conclude if the null is implausible given the data.

**Step 2 — Choose a significance level (α).**
Before collecting data, decide how often you are willing to reject a true null
hypothesis by mistake. Convention is α = 0.05 (5%), but this is a social norm,
not a law of nature. For high-stakes environmental decisions (e.g., declaring a
site contaminated), you might use α = 0.01.

**Step 3 — Compute the test statistic.**
Reduce the data to a single signal-to-noise ratio (t, F, z, W…) that measures
how far the observed data are from what the null predicts.

**Step 4 — Find the p-value.**
Ask: if H₀ were true, how often would we observe a test statistic at least as
extreme as the one we computed? This probability is the p-value.

**Step 5 — Make a decision and report.**
If p ≤ α: reject H₀ ("the data are inconsistent with no effect").
If p > α: fail to reject H₀ ("the data do not provide sufficient evidence
against no effect"). Report the test statistic, degrees of freedom, p-value,
*and* an effect size. Never report only "p < 0.05".

### Type I and Type II errors

No matter what we decide, we can be wrong in two distinct ways:

<table class="error-matrix">
<thead>
<tr>
  <th style="background:transparent;border:none;"></th>
  <th class="em-header-top">H₀ is actually TRUE</th>
  <th class="em-header-top">H₀ is actually FALSE</th>
</tr>
</thead>
<tbody>
<tr>
  <th class="em-header-side">We <strong>reject</strong> H₀</th>
  <td class="em-error">
    <span class="em-label">Type I Error (α)</span>
    False positive. We declare an effect that does not exist.<br>
    Controlled by our choice of α.
  </td>
  <td class="em-correct">
    <span class="em-label">Correct Decision ✓</span>
    True positive. We detected a real effect.<br>
    Probability = 1 − β (statistical power).
  </td>
</tr>
<tr>
  <th class="em-header-side">We <strong>fail to reject</strong> H₀</th>
  <td class="em-correct">
    <span class="em-label">Correct Decision ✓</span>
    True negative. We correctly concluded no effect.
  </td>
  <td class="em-error">
    <span class="em-label">Type II Error (β)</span>
    False negative. A real effect exists but we missed it.<br>
    Reduced by larger sample size.
  </td>
</tr>
</tbody>
</table>

<div class="callout callout-warn">
<strong>⚠️ The α–β trade-off</strong>

Making α smaller (stricter) reduces Type I errors but increases Type II
errors — you miss more real effects. The only way to reduce *both simultaneously*
is to collect more data. Power analysis before data collection determines the
sample size needed to detect a specified effect at chosen α and β.
</div>

---

## 2. One-Sample t-Test {#one-sample}

**Question:** Is the population mean equal to a specific hypothesised value?

**Environmental context:** The EPA Clean Water Act standard for dissolved
oxygen (DO) in coldwater fisheries is 7 mg/L. We sample 22 sites to test
whether the regional mean meets this standard.

<div class="hyp-row">
<div class="hyp-card hyp-null">
  <div class="hyp-label">Null Hypothesis H₀</div>
  <div class="hyp-stmt">µ = 7 mg/L &nbsp;—&nbsp; the regional mean DO equals the standard</div>
</div>
<div class="hyp-card hyp-alt">
  <div class="hyp-label">Alternative Hypothesis H₁</div>
  <div class="hyp-stmt">µ ≠ 7 mg/L &nbsp;—&nbsp; the regional mean differs from the standard (two-tailed)</div>
</div>
</div>

```{r one-sample-data}
set.seed(12)
# DO measurements (mg/L) at 22 stream sites
do_sites <- c(6.1, 7.4, 5.8, 6.9, 6.2, 7.1, 5.5, 6.8, 6.3, 7.0,
              5.9, 6.4, 6.7, 5.6, 6.1, 7.2, 6.0, 5.7, 6.5, 6.9, 6.3, 5.8)
mu_0     <- 7    # hypothesised value (EPA standard)
n        <- length(do_sites)

cat("n =", n, "\n")
cat("Sample mean  =", round(mean(do_sites), 3), "mg/L\n")
cat("Sample SD    =", round(sd(do_sites),   3), "mg/L\n")
cat("SE           =", round(sd(do_sites)/sqrt(n), 3), "mg/L\n")
```

### Step 1 — Visualise against the null value

Always plot the data and mark the hypothesised mean before running any test.

```{r one-sample-plot, fig.cap="DO measurements at 22 stream sites. Each point is one site; the box shows median and IQR; the dashed red line marks the EPA standard (7 mg/L). Most sites fall below the standard.", fig.height=3.8}
do_df <- data.frame(site = seq_along(do_sites), DO = do_sites)

ggplot(do_df, aes(x = "", y = DO)) +
  geom_hline(yintercept = mu_0, linetype = "dashed",
             colour = "#bf360c", linewidth = 0.9) +
  geom_boxplot(width = 0.35, fill = "#d4e6da", colour = "#1e3a2f",
               outlier.shape = NA, linewidth = 0.7) +
  geom_jitter(width = 0.12, size = 2.2, colour = "#4a7c5e", alpha = 0.7) +
  annotate("text", x = 1.28, y = mu_0 + 0.08,
           label = "EPA standard (7 mg/L)", colour = "#bf360c",
           size = 3.4, hjust = 0, fontface = "italic") +
  stat_summary(fun = mean, geom = "point", shape = 18,
               size = 4, colour = "#c8832a") +
  labs(x = NULL, y = "Dissolved oxygen (mg/L)",
       title = "DO at 22 Stream Sites vs. EPA Standard") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title         = element_text(family = "serif", face = "bold", colour = "#1e3a2f"),
    axis.text.x        = element_blank(),
    panel.grid.major.x = element_blank()
  )
```

### Step 2 — Compute by hand, then verify

$$t = \frac{\bar{x} - \mu_0}{s / \sqrt{n}}$$

```{r one-sample-byhand}
t_hand <- (mean(do_sites) - mu_0) / (sd(do_sites) / sqrt(n))
p_hand <- 2 * pt(t_hand, df = n - 1)   # two-tailed

cat("t (by hand) =", round(t_hand, 4), "\n")
cat("p (by hand) =", round(p_hand, 5), "\n")
```

```{r one-sample-ttest}
t.test(do_sites, mu = mu_0)
```

### Step 3 — Visualise the t-statistic on the null distribution

```{r one-sample-tdist, echo=FALSE, fig.cap="The observed t-statistic (red line) on the t(21) null distribution. The amber shaded tails are the 5% rejection region. Our t falls well into the left tail — the sample mean is significantly below the EPA standard.", fig.height=4}
df_os  <- n - 1
t_obs  <- t_hand
t_crit <- qt(0.975, df_os)

# Widen x range to always cover the observed t, however extreme it is
x_lo  <- min(-6, t_obs - 0.5)
x_hi  <- max( 6, t_obs + 0.5)
x_seq <- seq(x_lo, x_hi, length.out = 600)

par(mar = c(4, 4, 3, 1), las = 1)
plot(x_seq, dt(x_seq, df_os), type = "l", lwd = 2.5, col = "#1e3a2f",
     main = paste0("One-Sample t-test: t(", df_os, ") null distribution"),
     xlab = "t-statistic", ylab = "Density", ylim = c(0, 0.42),
     cex.main = 0.92)

# Helper: shade a polygon only when at least 2 points exist
shade_poly <- function(xr, df, fill) {
  if (length(xr) >= 2)
    polygon(c(xr[1], xr, xr[length(xr)]),
            c(0, dt(xr, df), 0),
            col = fill, border = NA)
}

# Shade rejection tails (amber)
shade_poly(x_seq[x_seq <= -t_crit], df_os, adjustcolor("#c8832a", 0.45))
shade_poly(x_seq[x_seq >=  t_crit], df_os, adjustcolor("#c8832a", 0.45))

# Shade p-value region beyond the observed t (red) — left tail for negative t
if (t_obs < 0) {
  shade_poly(x_seq[x_seq <= t_obs], df_os, adjustcolor("#bf360c", 0.45))
} else {
  shade_poly(x_seq[x_seq >= t_obs], df_os, adjustcolor("#bf360c", 0.45))
}

abline(v = t_obs,   col = "#bf360c", lwd = 2.5)
abline(v =  t_crit, col = "#c8832a", lwd = 1.5, lty = 2)
abline(v = -t_crit, col = "#c8832a", lwd = 1.5, lty = 2)

legend("topright",
       legend = c(paste0("t = ", round(t_obs, 2)),
                  paste0("±", round(t_crit, 2), "  (α = 0.05)"),
                  paste0("p = ", round(p_hand, 4))),
       col  = c("#bf360c", "#c8832a", adjustcolor("#bf360c", 0.45)),
       lty  = c(1, 2, NA), pch = c(NA, NA, 15),
       bty  = "n", cex = 0.82)
```

### Step 4 — Report

> A one-sample t-test indicated that the regional mean dissolved oxygen
> (M = `r round(mean(do_sites), 2)` mg/L, SD = `r round(sd(do_sites), 2)`)
> was significantly below the EPA coldwater standard of 7 mg/L,
> t(`r df_os`) = `r round(t_hand, 2)`, p = `r round(p_hand, 4)`.

---

## 3. Two-Sample t-Test {#two-sample}

**Question:** Do two independent groups have different population means?

**Environmental context:** We compare soil carbon stocks (Mg C/ha) under
two restoration treatments — passive recovery vs. active replanting — after
10 years.

<div class="hyp-row">
<div class="hyp-card hyp-null">
  <div class="hyp-label">Null Hypothesis H₀</div>
  <div class="hyp-stmt">µ<sub>passive</sub> = µ<sub>active</sub></div>
</div>
<div class="hyp-card hyp-alt">
  <div class="hyp-label">Alternative Hypothesis H₁</div>
  <div class="hyp-stmt">µ<sub>passive</sub> ≠ µ<sub>active</sub> &nbsp;(two-tailed)</div>
</div>
</div>

```{r two-sample-data}
set.seed(7)
passive <- round(rnorm(18, mean = 34, sd = 6.5), 1)  # Mg C/ha
active  <- round(rnorm(22, mean = 41, sd = 7.2), 1)
carbon  <- data.frame(
  treatment = rep(c("Passive recovery", "Active replanting"),
                  c(length(passive), length(active))),
  C_MgCha   = c(passive, active)
)
```

```{r two-sample-plot, fig.cap="Soil carbon after 10 years by restoration treatment. The diamond marks the group mean. Active replanting appears to accumulate more carbon, but with considerable overlap between groups.", fig.height=4}
ggplot(carbon, aes(treatment, C_MgCha, fill = treatment)) +
  geom_boxplot(width = 0.45, outlier.shape = NA, colour = "#1e3a2f",
               alpha = 0.7, linewidth = 0.6) +
  geom_jitter(aes(colour = treatment), width = 0.17, size = 1.9, alpha = 0.5) +
  stat_summary(fun = mean, geom = "point", shape = 18,
               size = 4.5, colour = "#1e3a2f") +
  scale_fill_manual(values   = c("Passive recovery" = "#8aad96",
                                 "Active replanting" = "#1e3a2f")) +
  scale_colour_manual(values = c("Passive recovery" = "#4a7c5e",
                                 "Active replanting" = "#2d5a3d")) +
  labs(x = NULL, y = "Soil carbon (Mg C/ha)",
       title = "Soil Carbon by Restoration Treatment (10 years)") +
  theme_minimal(base_size = 12) +
  theme(plot.title         = element_text(family = "serif", face = "bold",
                                          colour = "#1e3a2f"),
        legend.position    = "none",
        panel.grid.major.x = element_blank())
```

### Welch's t-test (the safe default)

R's `t.test()` uses **Welch's correction** by default — it does not assume
equal variances between groups, making it robust to heteroscedasticity.
Only use `var.equal = TRUE` if you have strong evidence that population
variances are equal (Levene's test, or subject-matter knowledge).

```{r two-sample-test}
(result_2t <- t.test(C_MgCha ~ treatment, data = carbon))
```

```{r two-sample-levene}
# Levene's test for equal variances (car package, or manual approach)
var.test(C_MgCha ~ treatment, data = carbon)
```

<div class="callout callout-formula">
<strong>⚙️ One-tailed vs. two-tailed tests</strong><br>
A <strong>two-tailed</strong> test (default) rejects H₀ if the observed
difference is extreme in <em>either direction</em>. Use this unless you have a
strong, pre-registered directional hypothesis.<br><br>
A <strong>one-tailed</strong> test concentrates all of α in one tail — it has
more power to detect a specific direction but will miss effects in the other.
Specify with <code>alternative = "greater"</code> or <code>"less"</code>.
One-tailed tests chosen <em>after</em> looking at the data are not valid.
</div>

---

## 4. Paired t-Test {#paired}

**Question:** Did the same units change between two conditions?

The paired t-test is used when each observation in group A is matched to a
specific observation in group B — before/after measurements on the same site,
left vs. right bank of the same stream, paired field plots. By working with
*differences*, the paired test removes between-unit variability that would
otherwise inflate the noise denominator.

**Environmental context:** Turbidity (NTU) measured at 16 stream sites
*before* and *after* installation of riparian buffer strips (same sites,
two time points).

<div class="hyp-row">
<div class="hyp-card hyp-null">
  <div class="hyp-label">Null Hypothesis H₀</div>
  <div class="hyp-stmt">µ<sub>difference</sub> = 0 &nbsp;—&nbsp; buffers have no effect on turbidity</div>
</div>
<div class="hyp-card hyp-alt">
  <div class="hyp-label">Alternative Hypothesis H₁</div>
  <div class="hyp-stmt">µ<sub>difference</sub> &lt; 0 &nbsp;—&nbsp; turbidity is lower after installation (one-tailed)</div>
</div>
</div>

```{r paired-data}
set.seed(41)
n_sites  <- 16
before   <- round(rnorm(n_sites, mean = 28, sd = 9), 1)   # NTU
# Buffer reduces turbidity by ~8 NTU on average, with some site-level variation
after    <- round(pmax(2, before - rnorm(n_sites, mean = 8, sd = 4)), 1)
diff_ntu <- after - before

turbidity <- data.frame(site = 1:n_sites, before, after, diff = diff_ntu)
head(turbidity, 6)
```

```{r paired-plot, fig.cap="Left: turbidity before and after buffer installation at each site (connected lines show individual site trajectories). Right: the paired differences — negative values mean turbidity decreased. If the null were true, differences would scatter symmetrically around zero.", fig.height=4.5, fig.width=8}
par(mfrow = c(1, 2), mar = c(4, 4.5, 3, 1), las = 1)

# Before / after connected
plot(c(0.8, 2.2), range(c(before, after)) + c(-2, 2),
     type = "n", xaxt = "n",
     main = "Before vs. After (paired sites)",
     xlab = "", ylab = "Turbidity (NTU)", cex.main = 0.9)
axis(1, at = c(1, 2), labels = c("Before", "After"))
for (i in seq_len(n_sites)) {
  lines(c(1, 2), c(before[i], after[i]),
        col = adjustcolor(ifelse(diff_ntu[i] < 0, "#4a7c5e", "#c8832a"), 0.6),
        lwd = 1.3)
  points(c(1, 2), c(before[i], after[i]),
         pch = 20, cex = 0.9,
         col = adjustcolor(ifelse(diff_ntu[i] < 0, "#4a7c5e", "#c8832a"), 0.7))
}
points(c(1, 2), c(mean(before), mean(after)),
       pch = 18, cex = 2, col = "#1e3a2f")
lines(c(1, 2), c(mean(before), mean(after)),
      lwd = 2.5, col = "#1e3a2f")

# Differences
hist(diff_ntu, breaks = 12,
     col = "#d4e6da", border = "white",
     main = "Paired differences (After − Before)",
     xlab = "ΔNTU", ylab = "Count", cex.main = 0.9)
abline(v = 0,              col = "#c8832a", lwd = 1.5, lty = 2)
abline(v = mean(diff_ntu), col = "#1e3a2f", lwd = 2)
legend("topright",
       legend = c("Zero (null)", paste0("Mean diff = ", round(mean(diff_ntu), 1))),
       col    = c("#c8832a", "#1e3a2f"), lty = c(2, 1), lwd = 2,
       bty    = "n", cex = 0.8)
par(mfrow = c(1, 1))
```

```{r paired-test}
# Paired t-test: alternative = "less" because we predict After < Before
t.test(after, before, paired = TRUE, alternative = "less")
```

```{r paired-verify}
# Identical to a one-sample t-test on the differences
t.test(diff_ntu, mu = 0, alternative = "less")
```

<div class="callout callout-concept">
<strong>📌 Why pairing increases power</strong>

In the unpaired version, site-to-site variation in baseline turbidity becomes
noise — it makes it harder to see the treatment effect. In the paired version,
we subtract out that baseline, leaving only the treatment effect plus
measurement noise. The paired test always has equal or greater power than
the unpaired test when the within-unit correlation is positive. It is
<em>invalid</em> to pair data post-hoc to inflate significance.
</div>

---

## 5. Wilcoxon Tests: When Normality Fails {#wilcoxon}

The t-tests above assume the data (or their differences) are approximately
normally distributed. When this assumption is badly violated — typically
with small samples that are strongly skewed or contain outliers — the
**Wilcoxon tests** provide non-parametric alternatives.

Instead of working with raw values, Wilcoxon tests work with **ranks**.
This makes them robust to extreme observations and appropriate for ordinal
data, but they lose some efficiency when the data actually are normal.

### Wilcoxon signed-rank test (one-sample / paired)

The non-parametric equivalent of the one-sample or paired t-test. Tests
whether the *median* (not mean) equals a hypothesised value, or whether the
median difference equals zero.

**Environmental context:** Heavy metal concentration (mg/kg) in a
contaminated floodplain — small sample, right-skewed distribution.

```{r wilcoxon-data}
set.seed(19)
# Heavy metal (Pb) — right-skewed, small n
pb_mg   <- c(12, 18, 145, 23, 9, 31, 88, 14, 27, 19, 52, 11)
pb_limit <- 40   # remediation threshold (mg/kg)
n_pb    <- length(pb_mg)

cat("n =", n_pb, "\n")
cat("Mean =", round(mean(pb_mg), 1), "mg/kg\n")
cat("Median =", median(pb_mg), "mg/kg\n")
cat("Shapiro-Wilk p =", round(shapiro.test(pb_mg)$p.value, 4),
    "— normality", ifelse(shapiro.test(pb_mg)$p.value < 0.05,
                          "REJECTED", "not rejected"), "\n")
```

```{r wilcoxon-compare, fig.cap="Pb concentration in a small, skewed sample. The single extreme value at 145 mg/kg violates the normality assumption that the t-test requires. The Shapiro-Wilk test confirms non-normality (p < 0.05).", fig.height=3.8}
pb_df <- data.frame(pb = pb_mg)

ggplot(pb_df, aes(x = "", y = pb)) +
  geom_hline(yintercept = pb_limit, linetype = "dashed",
             colour = "#bf360c", linewidth = 0.9) +
  geom_boxplot(width = 0.35, fill = "#d4e6da", colour = "#1e3a2f",
               linewidth = 0.7) +
  geom_jitter(width = 0.1, size = 2.5, colour = "#4a7c5e", alpha = 0.75) +
  annotate("text", x = 1.28, y = pb_limit + 5,
           label = "Remediation threshold (40 mg/kg)",
           colour = "#bf360c", size = 3.2, hjust = 0, fontface = "italic") +
  labs(x = NULL, y = "Lead concentration (mg/kg)",
       title = "Pb in Floodplain Soil — Small Skewed Sample") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(family = "serif", face = "bold", colour = "#1e3a2f"),
        axis.text.x = element_blank(),
        panel.grid.major.x = element_blank())
```

```{r wilcoxon-vs-t}
# t-test: sensitive to the outlier, pulls the mean up
cat("── t-test (sensitive to outlier) ──────────────────\n")
print(t.test(pb_mg, mu = pb_limit))

cat("\n── Wilcoxon signed-rank (robust to outlier) ────────\n")
print(wilcox.test(pb_mg, mu = pb_limit, conf.int = TRUE))
```

### Wilcoxon rank-sum test (two independent groups)

The non-parametric equivalent of the two-sample t-test. Also known as the
**Mann-Whitney U test**. Tests whether one group tends to have larger values
than the other.

```{r ranksum-data}
set.seed(5)
# Species richness on degraded vs. restored plots — count data, right-skewed
restored  <- c(12, 18, 9, 24, 15, 11, 22, 8, 17, 20, 13, 27)
degraded  <- c(4,  7,  3, 11, 6,  2,  9,  5,  8,  4,  6,  3)

cat("Restored  median =", median(restored), "  mean =", round(mean(restored), 1), "\n")
cat("Degraded  median =", median(degraded), "  mean =", round(mean(degraded), 1), "\n")
```

```{r ranksum-test}
wilcox.test(restored, degraded, conf.int = TRUE)
```

```{r ranksum-plot, fig.cap="Species richness on restored vs. degraded plots. Count data are right-skewed with unequal variances — the Wilcoxon rank-sum test is preferred over the t-test here.", fig.height=3.8}
rich_df <- data.frame(
  status    = rep(c("Restored", "Degraded"), c(length(restored), length(degraded))),
  richness  = c(restored, degraded)
)
rich_df$status <- factor(rich_df$status, levels = c("Degraded", "Restored"))

ggplot(rich_df, aes(status, richness, fill = status)) +
  geom_boxplot(width = 0.45, outlier.shape = NA, colour = "#1e3a2f",
               alpha = 0.8, linewidth = 0.6) +
  geom_jitter(aes(colour = status), width = 0.16, size = 2, alpha = 0.6) +
  scale_fill_manual(values   = c("Degraded" = "#c8832a", "Restored" = "#4a7c5e")) +
  scale_colour_manual(values = c("Degraded" = "#c8832a", "Restored" = "#4a7c5e")) +
  labs(x = NULL, y = "Species richness",
       title = "Plant Species Richness: Restored vs. Degraded Plots") +
  theme_minimal(base_size = 12) +
  theme(plot.title         = element_text(family = "serif", face = "bold",
                                          colour = "#1e3a2f"),
        legend.position    = "none",
        panel.grid.major.x = element_blank())
```

```{r wilcoxon-choice-table, echo=FALSE}
knitr::kable(
  data.frame(
    Scenario    = c("One sample vs. known value", "Two independent groups",
                    "Paired / before–after"),
    Parametric  = c("One-sample t-test", "Welch two-sample t-test",
                    "Paired t-test"),
    `R function` = c("`t.test(x, mu = µ₀)`",
                     "`t.test(y ~ grp)`",
                     "`t.test(x, y, paired = TRUE)`"),
    `Non-parametric alt.` = c("Wilcoxon signed-rank",
                               "Wilcoxon rank-sum (Mann-Whitney)",
                               "Wilcoxon signed-rank on differences"),
    `R function NP` = c("`wilcox.test(x, mu = µ₀)`",
                        "`wilcox.test(x, y)` or `wilcox.test(y ~ grp)`",
                        "`wilcox.test(x, y, paired = TRUE)`")
  ),
  col.names = c("Scenario", "Parametric test", "R code",
                "Non-parametric alternative", "R code (NP)"),
  caption = "t-test and Wilcoxon test selection guide."
)
```

<div class="callout callout-tip">
<strong>💡 When to choose Wilcoxon over t</strong>

Use the Wilcoxon test when: (1) n &lt; 15 and the data are visibly skewed or
have outliers; (2) your variable is ordinal (ranks, damage scores); or (3)
Shapiro-Wilk is significant and transformation does not help. With n &gt; 30,
the t-test is robust to moderate non-normality by the Central Limit Theorem
and is usually the better choice because it provides a mean estimate and
confidence interval with direct physical interpretation.
</div>

---

## 6. Tests for Proportions {#proportions}

When your outcome is binary — a proportion of sites that exceed a threshold,
the fraction of individuals that survive a treatment — the t-test is not
appropriate. Use **one- and two-proportion z-tests** or the exact binomial test.

### One-proportion z-test

**Question:** Does the true proportion of exceedances equal a specified value?

**Environmental context:** In a regulatory sample of 85 wells, 23 exceed the
nitrate drinking-water limit (10 mg/L NO₃). Does this fraction differ from
the statewide exceedance rate of 15%?

<div class="hyp-row">
<div class="hyp-card hyp-null">
  <div class="hyp-label">Null Hypothesis H₀</div>
  <div class="hyp-stmt">p = 0.15 &nbsp;—&nbsp; exceedance rate equals the statewide rate</div>
</div>
<div class="hyp-card hyp-alt">
  <div class="hyp-label">Alternative Hypothesis H₁</div>
  <div class="hyp-stmt">p ≠ 0.15 &nbsp;—&nbsp; exceedance rate differs from statewide (two-tailed)</div>
</div>
</div>

```{r prop-one}
n_wells  <- 85
x_exceed <- 23
p_0      <- 0.15        # statewide rate (null)
p_hat    <- x_exceed / n_wells
cat("Observed proportion:", round(p_hat, 3), "\n")

# Exact binomial test (preferred when np < 10)
binom.test(x_exceed, n_wells, p = p_0)
```

```{r prop-z-test}
# Large-sample z-test (valid when np ≥ 10 and n(1-p) ≥ 10)
prop.test(x_exceed, n_wells, p = p_0, correct = FALSE)
```

### Two-proportion z-test

**Question:** Do two independent groups have the same proportion?

**Environmental context:** Compare the fraction of streams exceeding
copper criteria upstream vs. downstream of a mining district.

```{r prop-two}
# Upstream: 8 of 40 exceed criterion.  Downstream: 22 of 45.
prop.test(x = c(8, 22), n = c(40, 45), correct = FALSE)
```

```{r prop-plot, fig.cap="Proportion of streams exceeding the copper criterion, with 95% Wilson confidence intervals. The upstream rate (~20%) and downstream rate (~49%) are clearly separated — consistent with the significant p-value.", fig.height=3.8}
prop_df <- data.frame(
  location = c("Upstream", "Downstream"),
  x = c(8, 22), n = c(40, 45)
) |>
  mutate(
    p     = x / n,
    lower = mapply(function(xi, ni) binom.test(xi, ni)$conf.int[1], x, n),
    upper = mapply(function(xi, ni) binom.test(xi, ni)$conf.int[2], x, n)
  )
prop_df$location <- factor(prop_df$location, levels = c("Upstream", "Downstream"))

ggplot(prop_df, aes(location, p, fill = location)) +
  geom_col(width = 0.45, colour = "#1e3a2f", alpha = 0.85) +
  geom_errorbar(aes(ymin = lower, ymax = upper),
                width = 0.12, linewidth = 0.9, colour = "#1e3a2f") +
  scale_fill_manual(values = c("Upstream" = "#4a7c5e", "Downstream" = "#7a3a2f")) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 0.75)) +
  labs(x = NULL, y = "Proportion exceeding Cu criterion",
       title = "Stream Copper Exceedances Near Mining District") +
  theme_minimal(base_size = 12) +
  theme(plot.title         = element_text(family = "serif", face = "bold",
                                          colour = "#1e3a2f"),
        legend.position    = "none",
        panel.grid.major.x = element_blank())
```

---

## 7. Effect Sizes: How Big Is the Effect? {#effect-size}

A p-value tells you whether an effect is unlikely under the null. It does
**not** tell you whether the effect is large enough to matter. Effect sizes
are standardised measures of magnitude that are independent of sample size.

<div class="effect-gauge">
  <div class="eg-neg-large">Large−</div>
  <div class="eg-neg-med">Medium−</div>
  <div class="eg-neg-small">Small−</div>
  <div class="eg-zero">≈0</div>
  <div class="eg-pos-small">Small+</div>
  <div class="eg-pos-med">Medium+</div>
  <div class="eg-pos-large">Large+</div>
</div>

### Cohen's d (for t-tests)

Cohen's d expresses the difference in means as a multiple of the pooled
standard deviation. It is interpretable across studies and variables.

$$d = \frac{\bar{x}_1 - \bar{x}_2}{s_{\text{pooled}}}$$

Cohen's benchmarks: |d| = 0.2 small · 0.5 medium · 0.8 large.

```{r cohens-d}
cohens_d <- function(x, y) {
  nx <- length(x); ny <- length(y)
  sp <- sqrt(((nx - 1)*var(x) + (ny - 1)*var(y)) / (nx + ny - 2))
  (mean(x) - mean(y)) / sp
}

d_carbon <- cohens_d(active, passive)
d_rich   <- cohens_d(restored, degraded)

cat("Carbon (active vs. passive)  d =", round(d_carbon, 3), "\n")
cat("Richness (restored vs. degraded) d =", round(d_rich, 3), "\n")
```

```{r cohens-d-visual, echo=FALSE, fig.cap="Visualising Cohen's d. Each panel shows two normal distributions with d = 0.2, 0.5, 0.8, and 1.4. The overlap shrinks as d grows — d = 0.2 means the distributions are nearly identical; d = 1.4 means a typical value in one group exceeds ~92% of values in the other.", fig.height=4.5}
par(mfrow = c(2, 2), mar = c(3, 2, 2.5, 1), las = 1)
ds    <- c(0.2, 0.5, 0.8, 1.4)
labels <- c("Small (d = 0.2)", "Medium (d = 0.5)",
            "Large (d = 0.8)", "Very large (d = 1.4)")

for (k in seq_along(ds)) {
  d <- ds[k]
  x <- seq(-4, 4 + d, length.out = 400)
  y1 <- dnorm(x, 0, 1)
  y2 <- dnorm(x, d, 1)

  plot(x, y1, type = "l", lwd = 0, col = "white",
       ylim = c(0, 0.42), xlab = "", ylab = "",
       main = labels[k], cex.main = 0.88, frame.plot = FALSE)

  polygon(c(x, rev(x)), c(y1, rep(0, length(x))),
          col = adjustcolor("#4a7c5e", 0.45), border = NA)
  polygon(c(x, rev(x)), c(y2, rep(0, length(x))),
          col = adjustcolor("#c8832a", 0.45), border = NA)

  lines(x, y1, lwd = 2, col = "#4a7c5e")
  lines(x, y2, lwd = 2, col = "#c8832a")

  overlap <- round(2 * pnorm(-d/2) * 100)
  text(mean(range(x)), 0.38,
       paste0("Overlap ≈ ", overlap, "%"),
       cex = 0.78, col = "#1e3a2f")
}
par(mfrow = c(1, 1))
```

### r (for Wilcoxon tests)

For the Wilcoxon rank-sum, the effect size is:
$$r = \frac{Z}{\sqrt{N}}$$

where Z is the normal approximation of the W statistic and N is the total
sample size. Benchmarks: r = 0.1 small · 0.3 medium · 0.5 large.

```{r wilcoxon-effect}
w_result <- wilcox.test(restored, degraded, exact = FALSE)
Z_approx <- qnorm(w_result$p.value / 2)
r_effect <- abs(Z_approx) / sqrt(length(restored) + length(degraded))
cat("W =", w_result$statistic, "  Z ≈", round(Z_approx, 3),
    "  r =", round(r_effect, 3), "\n")
```

### h (for proportions tests)

Cohen's h standardises the difference between two proportions via the
arcsine transformation, which stabilises variance.

$$h = 2 \arcsin\!\left(\sqrt{p_1}\right) - 2 \arcsin\!\left(\sqrt{p_2}\right)$$

```{r cohens-h}
cohens_h <- function(p1, p2) 2*asin(sqrt(p1)) - 2*asin(sqrt(p2))
h_copper <- cohens_h(22/45, 8/40)
cat("Cohen's h =", round(h_copper, 3),
    " (|h| benchmarks: 0.2 small, 0.5 medium, 0.8 large)\n")
```

```{r effect-size-table, echo=FALSE}
knitr::kable(
  data.frame(
    Test        = c("t-test (two groups)", "t-test (one sample / paired)",
                    "Wilcoxon rank-sum", "Proportions test"),
    `Effect size` = c("Cohen's d", "Cohen's d", "r = Z/√N", "Cohen's h"),
    Small = c("0.2", "0.2", "0.1", "0.2"),
    Medium = c("0.5", "0.5", "0.3", "0.5"),
    Large  = c("0.8", "0.8", "0.5", "0.8"),
    `R calculation` = c("`(mean(x)-mean(y)) / sd_pooled`",
                        "`(mean(x)-mu0) / sd(x)`",
                        "`abs(qnorm(p/2)) / sqrt(N)`",
                        "`2*asin(sqrt(p1)) - 2*asin(sqrt(p2))`")
  ),
  col.names = c("Test", "Effect size", "Small", "Medium", "Large", "R calculation"),
  caption   = "Standardised effect sizes by test type, with Cohen's conventional benchmarks."
)
```

---

## 8. Understanding p-values {#pvalues}

The p-value is the most widely reported and most widely misunderstood quantity
in applied statistics. Before examining what it is *not*, here is a precise
statement of what it *is*.

> **The p-value is the probability of observing a test statistic at least as
> extreme as the one computed, *given that the null hypothesis is true*.**

That is it. Nothing more. The following simulation illustrates the sampling
behaviour of p-values — which behaves very differently from what most people expect.

### P-values are uniform under the null

When H₀ is true, p-values are **uniformly distributed** between 0 and 1.
There is nothing special about 0.05 except our convention to use it as a
threshold.

```{r pvalue-dist, cache=FALSE}
set.seed(321)
n_trials <- 10000
p_null   <- replicate(n_trials, {
  A <- rnorm(20, 0, 1)
  B <- rnorm(20, 0, 1)     # both from same distribution — null TRUE
  t.test(A, B)$p.value
})
cat("Proportion p < 0.05 when H0 true:", mean(p_null < 0.05), "\n")
cat("(Expected: exactly 0.05 — these are false positives)\n")
```

```{r pvalue-uniform-plot, echo=FALSE, fig.cap="Distribution of p-values from 10,000 t-tests when H₀ is exactly true. The distribution is flat (uniform) — every p-value between 0 and 1 is equally likely. The amber bar at the left (p < 0.05) contains exactly 5% of all tests — these are Type I errors. There is nothing special about values just below 0.05 vs. just above it.", fig.height=4}
par(mar = c(4, 4.5, 3, 1), las = 1)
hist(p_null, breaks = 40, col = "#d4e6da", border = "white",
     main = "P-value distribution when H₀ is TRUE (10,000 simulations)",
     xlab = "p-value", ylab = "Frequency", cex.main = 0.9)

# Shade p < 0.05
extreme_p <- p_null[p_null < 0.05]
hb <- hist(extreme_p, breaks = seq(0, 0.05, length.out = 4), plot = FALSE)
rect(hb$breaks[-length(hb$breaks)], 0, hb$breaks[-1], hb$counts,
     col = adjustcolor("#c8832a", 0.65), border = NA)

abline(v = 0.05, col = "#c8832a", lwd = 2, lty = 2)
text(0.05 + 0.01, max(table(cut(p_null, 40))) * 0.9,
     "α = 0.05", col = "#c8832a", cex = 0.85, adj = 0)
text(0.025, max(table(cut(p_null, 40))) * 0.6,
     paste0(round(mean(p_null < 0.05)*100, 1), "% false\npositives"),
     col = "#7a1e1e", cex = 0.8)
```

### How sample size and effect size jointly determine p

```{r pvalue-surface, cache=FALSE, echo=FALSE, fig.cap="P-values from two-sample t-tests across varying sample sizes (x-axis) and true effect sizes (line colour). A tiny effect (d = 0.1) stays non-significant even with n = 200 per group. A large effect (d = 1.0) reaches significance with n ≈ 10. The p-value is a function of *both* effect size and sample size — it is not a measure of effect importance.", fig.height=4.5}
set.seed(9)
n_range   <- c(5, 10, 15, 20, 30, 50, 75, 100, 150, 200)
d_vals    <- c(0.1, 0.2, 0.5, 0.8, 1.0)
reps_each <- 500

pval_matrix <- sapply(d_vals, function(d) {
  sapply(n_range, function(n) {
    ps <- replicate(reps_each, {
      t.test(rnorm(n, 0, 1), rnorm(n, d, 1))$p.value
    })
    mean(ps)   # average p-value at this (n, d) combination
  })
})

cols_d <- c("#d4e6da", "#8aad96", "#4a7c5e", "#c8832a", "#7a1e1e")
par(mar = c(4.5, 4, 3, 6), las = 1, xpd = FALSE)
plot(NULL, xlim = range(n_range), ylim = c(0, 0.7),
     xlab = "n per group", ylab = "Mean p-value",
     main = "Mean p-value vs. sample size, by effect size (d)",
     cex.main = 0.9)
abline(h = 0.05, col = "#1e3a2f", lty = 2, lwd = 1.2)
text(max(n_range), 0.07, "α = 0.05", col = "#1e3a2f", cex = 0.78, adj = 1)

for (k in seq_along(d_vals)) {
  lines(n_range, pval_matrix[, k], col = cols_d[k], lwd = 2.2)
  points(n_range, pval_matrix[, k], col = cols_d[k], pch = 20, cex = 1)
}
par(xpd = TRUE)
legend(par("usr")[2] + 3, mean(par("usr")[3:4]) + 0.2,
       legend = paste0("d = ", d_vals),
       col = cols_d, lwd = 2, pch = 20, bty = "n", cex = 0.8, title = "Effect size")
par(xpd = FALSE)
```

---

## 9. Common Misconceptions {#misconceptions}

Statistical hypothesis testing is routinely misinterpreted — not just by
students, but in published papers and policy documents. The following
misconceptions are documented in the statistics literature and directly
relevant to environmental science.

<div class="myth-grid">

<div class="myth-card">
<div class="myth-header">Myth 1</div>
<div class="myth-body">
<div class="myth-myth">"p = 0.03 means there is a 3% probability that the null hypothesis is true."</div>
<div class="myth-truth"><strong>Reality:</strong> The p-value is a probability about
<em>data</em>, not about hypotheses. It says: if H₀ were true, data at least this
extreme would arise 3% of the time. It says nothing about the probability that H₀
is or is not true — that would require Bayesian reasoning with a prior. The null is
either true or not; it has no probability in the frequentist framework.</div>
</div>
</div>

<div class="myth-card">
<div class="myth-header">Myth 2</div>
<div class="myth-body">
<div class="myth-myth">"p > 0.05 means the groups are the same / the null is true."</div>
<div class="myth-truth"><strong>Reality:</strong> Failing to reject H₀ is not the
same as accepting H₀. It means the data are insufficient to rule out chance.
With small samples, we have low power — we may miss a real effect entirely.
Always compute power or a confidence interval to understand what you can rule out.
The phrase "no significant difference" should never be reported without also
stating the sample size and effect size confidence interval.</div>
</div>
</div>

<div class="myth-card">
<div class="myth-header">Myth 3</div>
<div class="myth-body">
<div class="myth-myth">"A smaller p-value means a larger or more important effect."</div>
<div class="myth-truth"><strong>Reality:</strong> P-values shrink as sample size
grows, regardless of effect size. With n = 10,000, a biologically trivial
difference (d = 0.01) will be statistically significant at p &lt; 0.001. With
n = 5, a large effect (d = 1.0) may fail to reach significance. Always report
an effect size alongside the p-value. Significance is about rarity under H₀;
importance is about the size of the effect relative to what matters in your field.</div>
</div>
</div>

<div class="myth-card">
<div class="myth-header">Myth 4</div>
<div class="myth-body">
<div class="myth-myth">"p = 0.049 and p = 0.051 lead to completely different conclusions."</div>
<div class="myth-truth"><strong>Reality:</strong> The p = 0.05 threshold is a
convention, not a physical law. The difference between p = 0.049 and p = 0.051
is no more meaningful than the difference between p = 0.38 and p = 0.40 — both
are well within sampling variation. Report exact p-values and confidence intervals;
do not use language that treats 0.05 as a sharp cliff. Many journals now
encourage reporting p values precisely (p = 0.048) rather than as thresholds
(p &lt; 0.05).</div>
</div>
</div>

<div class="myth-card">
<div class="myth-header">Myth 5</div>
<div class="myth-body">
<div class="myth-myth">"I can run multiple tests on the same dataset and report only the significant ones."</div>
<div class="myth-truth"><strong>Reality:</strong> Each test at α = 0.05 has a 5%
false-positive rate. Running 20 independent tests on a dataset where nothing is
truly different produces on average <em>one significant result by chance</em>.
This is the <strong>multiple comparisons problem</strong>. Pre-register your
hypotheses, correct for multiple comparisons (Bonferroni, Holm, or
false-discovery-rate correction), or treat exploratory tests as hypothesis
generation rather than confirmation.</div>
</div>
</div>

<div class="myth-card">
<div class="myth-header">Myth 6</div>
<div class="myth-body">
<div class="myth-myth">"A statistically significant result means my study will replicate."</div>
<div class="myth-truth"><strong>Reality:</strong> A single p &lt; 0.05 result in a
low-powered study has a poor replication probability — often below 50%. The
<em>positive predictive value</em> of a significant result depends on the prior
probability of the effect being real, study power, and the false-positive rate.
Small-n studies in ecology that just clear p = 0.05 are particularly likely to
produce inflated effect-size estimates that do not replicate (the "winner's curse").
Pre-register, report CIs, and treat replication as the standard of evidence.</div>
</div>
</div>

</div>

```{r multiple-comparisons, echo=FALSE, fig.cap="The multiple comparisons problem: 10,000 simulations each running 20 independent t-tests on null data. In ~64% of simulations, at least one test produced p < 0.05 by chance alone — far exceeding the nominal 5% error rate. The amber line marks 5% (one in twenty tests); the forest-green line marks the family-wise error rate after 20 tests.", fig.height=4}
set.seed(444)
n_exp     <- 10000
min_p_each <- replicate(n_exp, {
  ps <- replicate(20, {
    t.test(rnorm(10), rnorm(10))$p.value
  })
  min(ps)
})

par(mar = c(4.5, 4.5, 3, 1), las = 1)
hist(min_p_each, breaks = 60,
     col = "#d4e6da", border = "white", freq = FALSE,
     main = "Distribution of the smallest p-value across 20 simultaneous tests",
     xlab = "Minimum p-value in each experiment", ylab = "Density",
     cex.main = 0.9)

abline(v = 0.05,                         col = "#c8832a",  lwd = 2,   lty = 2)
abline(v = 1 - (1 - 0.05)^20,           col = "#1e3a2f",  lwd = 2,   lty = 1)

fwer <- mean(min_p_each < 0.05)
text(0.15, par("usr")[4] * 0.85,
     paste0("P(at least one false positive)\n= ", round(fwer*100, 0), "%"),
     col = "#7a1e1e", cex = 0.85, adj = 0)

legend("topright",
       legend = c("α = 0.05 (per-test)",
                  paste0("FWER = ", round((1-(1-0.05)^20)*100, 0),
                         "% (20 tests, Bonferroni)")),
       col  = c("#c8832a", "#1e3a2f"), lty = c(2, 1), lwd = 2,
       bty  = "n", cex = 0.82)
```

<div class="callout callout-tip">
<strong>💡 The Bonferroni correction</strong>

The simplest multiple-comparisons correction: divide α by the number of tests
performed. For 20 tests at α = 0.05, use α* = 0.05/20 = 0.0025 as the
per-test threshold. This controls the family-wise error rate (probability of
*any* false positive across all tests) at 5%. It is conservative — the
Holm-Bonferroni or Benjamini-Hochberg (FDR) procedures have more power while
still controlling errors. In R: `p.adjust(p_vector, method = "holm")`.
</div>

---

<div class="key-box">
<h4>🔑 Hypothesis testing — key takeaways</h4>
<ul>
<li><strong>Five steps, every time:</strong> state H₀ and H₁ → choose α → compute the test statistic → find the p-value → decide and report. The steps apply to every test in this module and beyond.</li>
<li><strong>Match test to data structure:</strong> one-sample t (single group vs. value) → two-sample t (independent groups) → paired t (matched / before-after) → Wilcoxon when normality fails → proportion tests for binary outcomes.</li>
<li><strong>Welch's t-test is the safe default</strong> for two-sample comparisons. It does not assume equal variances and performs as well as Student's t when variances are equal.</li>
<li><strong>Always report an effect size</strong> (Cohen's d, r, h) alongside the p-value. A significant p with d = 0.1 is a trivial effect detected in a large sample. An insignificant p with d = 0.8 is a large effect detected in a small sample. They require completely different interpretations.</li>
<li><strong>P-values are uniform under H₀.</strong> There is nothing sacred about 0.05 — it is a convention. A result with p = 0.048 is not more reliable than one with p = 0.052.</li>
<li><strong>Multiple comparisons inflate false positives</strong> exponentially. Pre-register hypotheses, use <code>p.adjust()</code>, and distinguish exploratory from confirmatory analysis.</li>
<li><strong>Failing to reject H₀ ≠ evidence for H₀.</strong> Report confidence intervals and power to communicate what the study was able (and unable) to detect.</li>
</ul>
</div>

---

### What's next?

- **[Four Frameworks](02_Four_Frameworks.html)** — ANOVA, regression, chi-square, logistic
- **[Signal Over Noise](03_Signal_Over_Noise.html)** — permutation tests and the t/F null distributions
- **[Describing Data](01_Describing_Data.html)** — mean, SD, SE, confidence intervals

```{r session-info, echo=FALSE}
# sessionInfo()
```

<footer class="page-footer">
EA030 Basic Statistics &mdash; Module 05
<span class="divider">|</span> Prof. Marc Los Huertos
<span class="divider">|</span> Pomona College
<span class="divider">|</span> Last updated: `r format(Sys.Date(), "%B %Y")`
<span class="divider">|</span> <a href="https://github.com/marclos/RTricks">GitHub</a>
</footer>
