---
title: "Simple Linear Regression"
author: "Marc Los Huertos"
date: "EA030 · Compiled `r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    self_contained: true
    theme: null
    highlight: null
    toc: false
knit: (function(inputFile, encoding) {
    rmarkdown::render(
      inputFile,
      encoding = encoding,
      output_file = file.path(dirname(inputFile),
                              "../docs/basic_statistics/07_Simple_Linear_Regression.html")
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 7,
  fig.height = 4
)
```

```{css, echo=FALSE}
@import url('https://fonts.googleapis.com/css2?family=Libre+Baskerville:ital,wght@0,400;0,700;1,400&family=Source+Code+Pro:wght@400;600&family=Lato:wght@300;400;700&display=swap');

*, *::before, *::after { box-sizing: border-box; margin: 0; padding: 0; }

:root {
  --forest:    #1e3a2f; --moss: #2d5a3d; --fern: #4a7c5e; --sage: #8aad96;
  --mist:      #d4e6da; --cream: #f7f3ed; --parchment: #ede8df;
  --ink:       #1a1a18; --charcoal: #3a3a38; --dust: #7a7a72;
  --amber:     #c8832a; --amber-lt: #e8a84a; --white: #ffffff;
  --card-shadow: 0 2px 10px rgba(30,58,47,.09);
}

html { scroll-behavior: smooth; }
body {
  font-family: 'Lato', sans-serif; background: var(--cream); color: var(--ink);
  font-size: 17px; line-height: 1.75;
  padding-left: clamp(14px, 4vw, 48px); padding-right: clamp(14px, 4vw, 48px);
  padding-bottom: 80px;
}
.main-container { max-width: 920px !important; margin: 0 auto !important; padding: 0 clamp(10px, 2vw, 32px) 80px !important; }
.page-header { margin-left: calc(-1 * clamp(14px, 4vw, 48px)); margin-right: calc(-1 * clamp(14px, 4vw, 48px)); }
h1.title, h4.author, h4.date { display: none; }

.page-header {
  background-image: repeating-linear-gradient(120deg, transparent, transparent 60px, rgba(255,255,255,.018) 60px, rgba(255,255,255,.018) 61px), linear-gradient(160deg, #1e3a2f 0%, #2d5a3d 60%, #1a3d30 100%);
  color: #fff; padding: 44px 4vw 36px; margin: 0 0 48px;
  border-bottom: 4px solid var(--amber); position: relative; overflow: hidden;
}
.page-header::after { content: ''; position: absolute; right: -40px; top: -40px; width: 260px; height: 260px; border-radius: 50%; background: radial-gradient(circle, rgba(138,173,150,.12) 0%, transparent 70%); pointer-events: none; }
.page-header .module-tag { display: inline-block; font-family: 'Source Code Pro', monospace; font-size: .68rem; font-weight: 600; letter-spacing: .18em; text-transform: uppercase; color: var(--amber-lt); border: 1px solid rgba(200,131,42,.45); padding: 3px 10px; border-radius: 2px; margin-bottom: 12px; }
.page-header h1 { font-family: 'Libre Baskerville', serif; font-size: clamp(1.6rem, 3vw, 2.3rem); font-weight: 700; color: #fff; line-height: 1.2; margin-bottom: 8px; }
.page-header h1 span { color: var(--sage); }
.page-header .sub { font-size: .9rem; color: rgba(255,255,255,.65); font-weight: 300; }
.page-header .sub strong { color: rgba(255,255,255,.9); font-weight: 700; }
.back-link { display: inline-block; margin-top: 16px; font-family: 'Source Code Pro', monospace; font-size: .7rem; letter-spacing: .08em; color: var(--amber-lt); text-decoration: none; border: 1px solid rgba(200,131,42,.4); padding: 4px 11px; border-radius: 3px; transition: background .2s; }
.back-link:hover { background: rgba(200,131,42,.18); }

.section-nav { display: flex; gap: 8px; flex-wrap: wrap; margin-bottom: 44px; padding-bottom: 20px; border-bottom: 1px solid var(--parchment); }
.section-nav a { font-family: 'Source Code Pro', monospace; font-size: .7rem; font-weight: 600; letter-spacing: .08em; text-transform: uppercase; color: var(--fern); text-decoration: none; border: 1px solid var(--mist); padding: 4px 12px; border-radius: 3px; background: var(--white); transition: background .2s, color .2s; }
.section-nav a:hover { background: var(--mist); color: var(--forest); }

h2 { font-family: 'Libre Baskerville', serif; font-size: 1.45rem; font-weight: 700; color: var(--forest); margin-top: 56px; margin-bottom: 6px; padding-bottom: 8px; border-bottom: 1px solid var(--parchment); }
h3 { font-family: 'Libre Baskerville', serif; font-size: 1.1rem; font-weight: 700; color: var(--moss); margin-top: 28px; margin-bottom: 4px; }

p { margin-bottom: 1em; }
strong { color: var(--forest); }
code { font-family: 'Source Code Pro', monospace; font-size: .84em; background: var(--mist); color: var(--moss); padding: 1px 5px; border-radius: 3px; }
pre { background: var(--ink); color: #e8e6df; border-radius: 5px; padding: 14px 18px; font-size: .83rem; line-height: 1.55; overflow-x: auto; border-left: 4px solid var(--amber); margin: 1.2em 0; }
pre code { background: transparent; color: inherit; padding: 0; font-size: 1em; }

.callout { border-left: 5px solid; padding: .65em 1.1em; margin: 1.4em 0; border-radius: 0 5px 5px 0; font-size: .92em; line-height: 1.6; }
.callout > strong:first-child { display: block; margin-bottom: .3em; }
.callout-concept { border-color: var(--fern);  background: #f0f7f2; }
.callout-tip     { border-color: #2e7d32;      background: #f1f8f1; }
.callout-warn    { border-color: #e65100;      background: #fff8f4; }
.callout-formula { border-color: var(--amber); background: #fdf8f0; font-family: 'Source Code Pro', monospace; font-size: .88em; }

table { font-size: .9rem; border-collapse: collapse; width: 100%; margin: 1.2em 0; }
thead th { background: var(--forest); color: #fff; font-family: 'Source Code Pro', monospace; font-size: .76rem; letter-spacing: .06em; text-transform: uppercase; padding: 8px 12px; text-align: left; }
tbody tr:nth-child(even) { background: var(--mist); }
tbody td { padding: 7px 12px; border-bottom: 1px solid var(--parchment); }

.key-box { background: var(--forest); color: #fff; border-radius: 6px; padding: 20px 24px; margin: 2em 0; font-size: .9rem; }
.key-box h4 { font-family: 'Libre Baskerville', serif; color: var(--sage); font-size: 1rem; margin-bottom: 10px; }
.key-box ul { padding-left: 1.3em; }
.key-box li { margin-bottom: 5px; color: rgba(255,255,255,.85); }
.key-box code { background: rgba(255,255,255,.12); color: var(--amber-lt); }

.page-footer { background: var(--ink); color: rgba(255,255,255,.42); text-align: center; padding: 24px 4vw; margin: 60px calc(-1 * clamp(14px, 4vw, 48px)) 0; font-size: .76rem; font-family: 'Source Code Pro', monospace; letter-spacing: .04em; }
.page-footer a { color: var(--sage); text-decoration: none; }
.page-footer .divider { margin: 0 8px; opacity: .3; }
```

<div class="page-header">
<div class="module-tag">EA030 &mdash; Basic Statistics &mdash; Module 07</div>
<h1>Simple <span>Linear Regression</span></h1>
<p class="sub"><strong>Prof. Marc Los Huertos</strong> &nbsp;&bull;&nbsp; Pomona College</p>
<a class="back-link" href="../index.html">&#8592; Back to RTricks</a>
</div>

<div class="section-nav">
<a href="#intuition">Intuition</a>
<a href="#model">The Model</a>
<a href="#fitting">Fitting in R</a>
<a href="#output">Reading Output</a>
<a href="#assumptions">Assumptions</a>
<a href="#diagnostics">Diagnostics</a>
<a href="#inference">Inference</a>
<a href="#prediction">Prediction</a>
</div>

> **What this module covers.** Simple linear regression describes the straight-line
> relationship between one continuous predictor and one continuous outcome. This
> tutorial builds the model from first principles, shows you how to fit and
> interpret `lm()` output, diagnose assumption violations with residual plots,
> and draw valid inference from slope confidence intervals and F-tests.

---

## 1. The Intuition {#intuition}

Regression asks: **if I know x, what is my best guess for y?** The word *simple* means
one predictor; *linear* means the relationship is a straight line.

Environmental applications are everywhere. Does PM2.5 increase as traffic volume
rises? Does soil carbon decrease with distance from a riparian corridor? Does tree
height predict biomass? In every case the method is the same: fit a line that minimizes
prediction error, then ask how confident we are in its slope.

```{r intuition-data}
# Ozone vs. solar radiation from the built-in airquality dataset
data(airquality)
aq <- na.omit(airquality[, c("Ozone", "Solar.R", "Temp", "Wind")])
head(aq)
```

```{r scatter-intuition, echo=FALSE, fig.cap="Ozone concentration vs. temperature in New York, summer 1973. Each point is one day. The dashed line is a visual guess; the solid line is the least-squares fit."}
par(las = 1, mar = c(4.5, 4.5, 2, 1))
plot(Ozone ~ Temp, data = aq, pch = 20,
     col = rgb(0.29, 0.49, 0.37, 0.65),
     xlab = "Temperature (°F)", ylab = "Ozone (ppb)",
     main = "Ozone vs. Temperature — New York 1973")
abline(lm(Ozone ~ Temp, data = aq), col = "#1e3a2f", lwd = 2)
abline(a = -200, b = 3.5, col = "#c8832a", lty = 2, lwd = 1.5)
legend("topleft", legend = c("Least-squares fit", "A reasonable guess"),
       col = c("#1e3a2f", "#c8832a"), lty = c(1, 2), lwd = 2, bty = "n")
```

---

## 2. The Model {#model}

The simple linear regression model is:

$$y_i = \beta_0 + \beta_1 x_i + \varepsilon_i$$

where:

- $y_i$ is the observed response for observation $i$
- $x_i$ is the predictor value
- $\beta_0$ is the **intercept** — the predicted value of y when x = 0
- $\beta_1$ is the **slope** — the change in y for a one-unit increase in x
- $\varepsilon_i \sim N(0, \sigma^2)$ is the **error** term — the part of y not explained by x

The **ordinary least-squares (OLS)** estimator chooses $\hat\beta_0$ and $\hat\beta_1$
to minimize the sum of squared residuals:

$$\text{SSR} = \sum_{i=1}^n (y_i - \hat y_i)^2 \quad \text{where} \quad \hat y_i = \hat\beta_0 + \hat\beta_1 x_i$$

The closed-form solution is:

$$\hat\beta_1 = \frac{\sum(x_i - \bar x)(y_i - \bar y)}{\sum(x_i - \bar x)^2} = \frac{S_{xy}}{S_{xx}}$$

$$\hat\beta_0 = \bar y - \hat\beta_1 \bar x$$

<div class="callout callout-formula">
<strong>Formula in words</strong>
The slope is the covariance of x and y divided by the variance of x.
The intercept is chosen so the line passes through (x̄, ȳ).
</div>

### Compute by hand first

```{r ols-by-hand}
x <- aq$Temp
y <- aq$Ozone

b1 <- sum((x - mean(x)) * (y - mean(y))) / sum((x - mean(x))^2)
b0 <- mean(y) - b1 * mean(x)

knitr::kable(
  data.frame(
    Coefficient = c("Slope (β₁)", "Intercept (β₀)"),
    Estimate    = round(c(b1, b0), 3),
    Interpretation = c("ppb per °F", "ppb when Temp = 0°F")
  ),
  caption = "OLS estimates computed by hand."
)
```

---

## 3. Fitting in R with `lm()` {#fitting}

In practice, always use `lm()`. The formula syntax `y ~ x` maps directly onto the
statistical model $y = \beta_0 + \beta_1 x + \varepsilon$.

```{r fit-lm}
mod <- lm(Ozone ~ Temp, data = aq)
```

The fitted object `mod` contains everything: coefficients, residuals, fitted values,
the variance–covariance matrix of the estimates, and more.

```{r fitted-coefs}
coef(mod)          # intercept and slope
fitted(mod)[1:6]   # predicted values yhat
residuals(mod)[1:6] # observed - predicted
```

---

## 4. Reading `summary()` Output {#output}

```{r lm-summary}
summary(mod)
```

The output has four important sections:

**Residuals.** A five-number summary of the residuals. Ideally symmetric around zero.
Large asymmetry suggests the normality assumption is violated.

**Coefficients table.** Each row is one model parameter.

| Column | Meaning |
|---|---|
| `Estimate` | $\hat\beta$ — the OLS coefficient |
| `Std. Error` | SE($\hat\beta$) — precision of the estimate |
| `t value` | $\hat\beta$ / SE — the test statistic for $H_0: \beta = 0$ |
| `Pr(>|t|)` | Two-tailed p-value for the t-test |

**Residual standard error (RSE).** $\hat\sigma$ — typical size of a residual.
Smaller is better; units match y.

**R-squared.** The proportion of variance in y explained by x.
$R^2 = 1 - \text{SS}_\text{res} / \text{SS}_\text{tot}$. Range 0–1; higher is better,
but high $R^2$ does not guarantee the model is correct.

**F-statistic.** Tests whether the overall model explains more variance than a
model with no predictors (intercept only). For simple regression, $F = t^2$.

<div class="callout callout-concept">
<strong>📌 The slope is the key result</strong>
In simple regression, the slope coefficient and its p-value are the primary result.
A slope of 2.43 ppb/°F means: for each 1 °F rise in temperature, ozone increases by
about 2.43 ppb on average. The p-value tests whether this relationship is
distinguishable from zero given sampling variability.
</div>

---

## 5. The Four OLS Assumptions {#assumptions}

Linear regression produces valid inference only when four assumptions hold. They are
easiest to remember with the acronym **LINE**:

| Letter | Assumption | What "violation" looks like |
|---|---|---|
| L | **L**inearity | Curved pattern in residuals vs. fitted |
| I | **I**ndependence | Residuals correlated over time or space |
| N | **N**ormality | Heavy-tailed or strongly skewed residuals |
| E | **E**qual variance (homoscedasticity) | Fan shape in residuals vs. fitted |

The most important are linearity and equal variance; the t-tests and confidence
intervals are fairly robust to modest normality violations when n is large.

---

## 6. Residual Diagnostics {#diagnostics}

`plot(mod)` produces four standard diagnostic plots. Always examine all four.

```{r diagnostic-plots, fig.height=7, fig.cap="Four standard regression diagnostic plots for the ozone model. (1) Residuals vs. Fitted checks linearity and equal variance. (2) Q-Q plot checks normality. (3) Scale-Location is a second variance check. (4) Residuals vs. Leverage identifies influential points."}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
plot(mod)
par(mfrow = c(1, 1))
```

### How to interpret each plot

**Residuals vs. Fitted (top-left).** Look for a horizontal band of points with no curve
or funnel. A curved pattern indicates the relationship is not linear; a funnel indicates
heteroscedasticity (unequal variance).

**Normal Q-Q (top-right).** Points should fall close to the diagonal line. Heavy tails
(points curving upward at both ends) indicate the residuals are more extreme than a
normal distribution predicts.

**Scale-Location (bottom-left).** The square root of the standardized residuals vs.
fitted values. The fitted line should be roughly horizontal. An upward slope indicates
that variance increases with the fitted value.

**Residuals vs. Leverage (bottom-right).** Identifies influential observations —
points whose removal would substantially change the regression line. Points outside
Cook's distance contours (dashed lines) deserve scrutiny.

<div class="callout callout-warn">
<strong>⚠️ High leverage vs. high influence</strong>
A point has high **leverage** if its x-value is far from x̄ — it has the potential
to pull the regression line. It has high **influence** if removing it actually changes
the fit substantially. Both Cook's distance and leverage are shown in plot 4.
</div>

### Checking for influential points

```{r influence}
# Cook's distance > 4/n is a common threshold
n  <- nrow(aq)
cd <- cooks.distance(mod)
influential <- which(cd > 4 / n)
```

`r length(influential)` observation(s) have Cook's D > 4/n (row indices:
`r if (length(influential) == 0) "none" else paste(influential, collapse = ", ")`).

```{r influential-table}
if (length(influential) > 0) {
  knitr::kable(
    aq[influential, ],
    caption = paste0("Observations with Cook's distance > 4/", n,
                     " (potentially influential).")
  )
}
```

---

## 7. Inference: Confidence Intervals and the F-test {#inference}

### Confidence interval for the slope

```{r slope-ci}
confint(mod, level = 0.95)
```

The 95% CI for the slope is more informative than the p-value alone: it shows the
range of plausible values for the true population slope. A narrow CI indicates the
estimate is precise; a wide CI indicates high uncertainty.

### Visualizing the fitted line with uncertainty band

```{r fitted-plot, fig.cap="Fitted regression line (dark) with 95% confidence band (grey) and 95% prediction interval (dashed). The CI captures uncertainty in the mean response; the PI captures uncertainty in a single future observation."}
new_x  <- data.frame(Temp = seq(min(aq$Temp), max(aq$Temp), length.out = 200))
ci_fit <- predict(mod, newdata = new_x, interval = "confidence")
pi_fit <- predict(mod, newdata = new_x, interval = "prediction")

par(las = 1, mar = c(4.5, 4.5, 2, 1))
plot(Ozone ~ Temp, data = aq, pch = 20,
     col = rgb(0.29, 0.49, 0.37, 0.55),
     xlab = "Temperature (°F)", ylab = "Ozone (ppb)",
     main = "Fitted regression with CI and PI")

# Prediction interval ribbon
polygon(c(new_x$Temp, rev(new_x$Temp)),
        c(pi_fit[, "lwr"], rev(pi_fit[, "upr"])),
        col = rgb(0.2, 0.35, 0.25, 0.08), border = NA)

# Confidence interval ribbon
polygon(c(new_x$Temp, rev(new_x$Temp)),
        c(ci_fit[, "lwr"], rev(ci_fit[, "upr"])),
        col = rgb(0.2, 0.35, 0.25, 0.20), border = NA)

lines(new_x$Temp, ci_fit[, "fit"], col = "#1e3a2f", lwd = 2)
lines(new_x$Temp, pi_fit[, "lwr"], col = "#4a7c5e", lwd = 1, lty = 2)
lines(new_x$Temp, pi_fit[, "upr"], col = "#4a7c5e", lwd = 1, lty = 2)

legend("topleft",
       legend = c("Fitted line", "95% CI (mean)", "95% PI (individual)"),
       col = c("#1e3a2f", rgb(0.2,.35,.25,.5), "#4a7c5e"),
       lty = c(1, 1, 2), lwd = c(2, 6, 1), bty = "n", cex = 0.85)
```

<div class="callout callout-concept">
<strong>📌 CI vs. PI — a critical distinction</strong>
The <strong>confidence interval</strong> captures uncertainty about the mean response
at a given x (the population average). The <strong>prediction interval</strong> captures
uncertainty about a single new observation. The PI is always wider because it must
account for both the uncertainty in the mean *and* the random scatter of individual
observations around that mean.
</div>

---

## 8. Making Predictions {#prediction}

```{r predictions}
# Predict ozone at specific temperatures
predict_data <- data.frame(Temp = c(65, 75, 85, 95))
preds <- predict(mod, newdata = predict_data, interval = "confidence")

result <- cbind(predict_data, round(preds, 1))
names(result) <- c("Temp (°F)", "Predicted Ozone (ppb)", "95% CI Lower", "95% CI Upper")
knitr::kable(result, caption = "Predicted mean ozone at selected temperatures.")
```

<div class="callout callout-warn">
<strong>⚠️ Do not extrapolate</strong>
Predictions are only reliable within the range of the observed x values. The model
has no information about the relationship outside that range — and the relationship
may be nonlinear beyond the observed data.
</div>

### The complete workflow in one block

```{r workflow-summary, eval=FALSE}
# 1. Explore
plot(y ~ x, data = mydata)

# 2. Fit
mod <- lm(y ~ x, data = mydata)

# 3. Summarize
summary(mod)
confint(mod)

# 4. Diagnose
par(mfrow = c(2, 2)); plot(mod); par(mfrow = c(1, 1))

# 5. Predict
predict(mod, newdata = data.frame(x = new_values), interval = "confidence")
```

<div class="key-box">
<h4>🔑 Key takeaways</h4>
<ul>
<li><code>lm(y ~ x)</code> fits the OLS line. The formula maps directly onto
$y = \beta_0 + \beta_1 x + \varepsilon$.</li>
<li>The <strong>slope</strong> is the primary result: it quantifies the change in y per
unit change in x. Always report it with a confidence interval.</li>
<li><strong>R²</strong> measures fit, not causation. A high R² does not mean x causes y.</li>
<li>Check the four <strong>LINE assumptions</strong> with <code>plot(mod)</code> before
reporting results. Curved residuals invalidate the model; heteroscedasticity inflates
standard errors.</li>
<li>Use <code>predict()</code> with <code>interval = "confidence"</code> for the mean
response and <code>interval = "prediction"</code> for individual observations.</li>
</ul>
</div>

---

### What's next?

- **[ANOVA](08_ANOVA.html)** — compare means across three or more groups
- **[Multiple Regression & GLMs](../advanced_stats/multiple_regression.html)** — add
  more predictors and handle count or binary outcomes

```{r session-info, echo=FALSE}
# sessionInfo()
```

<footer class="page-footer">
EA030 Basic Statistics — Module 07
<span class="divider">|</span> Prof. Marc Los Huertos
<span class="divider">|</span> Pomona College
<span class="divider">|</span> Last updated: `r format(Sys.Date(), "%B %Y")`
<span class="divider">|</span> <a href="https://github.com/marclos/RTricks">GitHub</a>
</footer>
