---
title: "ANOVA & Group Comparisons"
author: "Marc Los Huertos"
date: "EA030 · Compiled `r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    self_contained: true
    theme: null
    highlight: null
    toc: false
knit: (function(inputFile, encoding) {
    rmarkdown::render(
      inputFile,
      encoding = encoding,
      output_file = file.path(dirname(inputFile),
                              "../docs/basic_statistics/08_ANOVA.html")
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 7,
  fig.height = 4
)
library(ggplot2)
library(dplyr)
```

```{css, echo=FALSE}
@import url('https://fonts.googleapis.com/css2?family=Libre+Baskerville:ital,wght@0,400;0,700;1,400&family=Source+Code+Pro:wght@400;600&family=Lato:wght@300;400;700&display=swap');

*, *::before, *::after { box-sizing: border-box; margin: 0; padding: 0; }

:root {
  --forest: #1e3a2f; --moss: #2d5a3d; --fern: #4a7c5e; --sage: #8aad96;
  --mist: #d4e6da; --cream: #f7f3ed; --parchment: #ede8df;
  --ink: #1a1a18; --charcoal: #3a3a38; --dust: #7a7a72;
  --amber: #c8832a; --amber-lt: #e8a84a; --white: #ffffff;
  --card-shadow: 0 2px 10px rgba(30,58,47,.09);
}

html { scroll-behavior: smooth; }
body { font-family: 'Lato', sans-serif; background: var(--cream); color: var(--ink); font-size: 17px; line-height: 1.75; padding-left: clamp(14px, 4vw, 48px); padding-right: clamp(14px, 4vw, 48px); padding-bottom: 80px; }
.main-container { max-width: 920px !important; margin: 0 auto !important; padding: 0 clamp(10px, 2vw, 32px) 80px !important; }
.page-header { margin-left: calc(-1 * clamp(14px, 4vw, 48px)); margin-right: calc(-1 * clamp(14px, 4vw, 48px)); }
h1.title, h4.author, h4.date { display: none; }

.page-header { background-image: repeating-linear-gradient(120deg, transparent, transparent 60px, rgba(255,255,255,.018) 60px, rgba(255,255,255,.018) 61px), linear-gradient(160deg, #1e3a2f 0%, #2d5a3d 60%, #1a3d30 100%); color: #fff; padding: 44px 4vw 36px; margin: 0 0 48px; border-bottom: 4px solid var(--amber); position: relative; overflow: hidden; }
.page-header::after { content: ''; position: absolute; right: -40px; top: -40px; width: 260px; height: 260px; border-radius: 50%; background: radial-gradient(circle, rgba(138,173,150,.12) 0%, transparent 70%); pointer-events: none; }
.page-header .module-tag { display: inline-block; font-family: 'Source Code Pro', monospace; font-size: .68rem; font-weight: 600; letter-spacing: .18em; text-transform: uppercase; color: var(--amber-lt); border: 1px solid rgba(200,131,42,.45); padding: 3px 10px; border-radius: 2px; margin-bottom: 12px; }
.page-header h1 { font-family: 'Libre Baskerville', serif; font-size: clamp(1.6rem, 3vw, 2.3rem); font-weight: 700; color: #fff; line-height: 1.2; margin-bottom: 8px; }
.page-header h1 span { color: var(--sage); }
.page-header .sub { font-size: .9rem; color: rgba(255,255,255,.65); font-weight: 300; }
.page-header .sub strong { color: rgba(255,255,255,.9); font-weight: 700; }
.back-link { display: inline-block; margin-top: 16px; font-family: 'Source Code Pro', monospace; font-size: .7rem; letter-spacing: .08em; color: var(--amber-lt); text-decoration: none; border: 1px solid rgba(200,131,42,.4); padding: 4px 11px; border-radius: 3px; transition: background .2s; }
.back-link:hover { background: rgba(200,131,42,.18); }

.section-nav { display: flex; gap: 8px; flex-wrap: wrap; margin-bottom: 44px; padding-bottom: 20px; border-bottom: 1px solid var(--parchment); }
.section-nav a { font-family: 'Source Code Pro', monospace; font-size: .7rem; font-weight: 600; letter-spacing: .08em; text-transform: uppercase; color: var(--fern); text-decoration: none; border: 1px solid var(--mist); padding: 4px 12px; border-radius: 3px; background: var(--white); transition: background .2s, color .2s; }
.section-nav a:hover { background: var(--mist); color: var(--forest); }

h2 { font-family: 'Libre Baskerville', serif; font-size: 1.45rem; font-weight: 700; color: var(--forest); margin-top: 56px; margin-bottom: 6px; padding-bottom: 8px; border-bottom: 1px solid var(--parchment); }
h3 { font-family: 'Libre Baskerville', serif; font-size: 1.1rem; font-weight: 700; color: var(--moss); margin-top: 28px; margin-bottom: 4px; }
p { margin-bottom: 1em; }
strong { color: var(--forest); }
code { font-family: 'Source Code Pro', monospace; font-size: .84em; background: var(--mist); color: var(--moss); padding: 1px 5px; border-radius: 3px; }
pre { background: var(--ink); color: #e8e6df; border-radius: 5px; padding: 14px 18px; font-size: .83rem; line-height: 1.55; overflow-x: auto; border-left: 4px solid var(--amber); margin: 1.2em 0; }
pre code { background: transparent; color: inherit; padding: 0; font-size: 1em; }

.callout { border-left: 5px solid; padding: .65em 1.1em; margin: 1.4em 0; border-radius: 0 5px 5px 0; font-size: .92em; line-height: 1.6; }
.callout > strong:first-child { display: block; margin-bottom: .3em; }
.callout-concept { border-color: var(--fern);  background: #f0f7f2; }
.callout-tip     { border-color: #2e7d32;      background: #f1f8f1; }
.callout-warn    { border-color: #e65100;      background: #fff8f4; }
.callout-formula { border-color: var(--amber); background: #fdf8f0; font-family: 'Source Code Pro', monospace; font-size: .88em; }

table { font-size: .9rem; border-collapse: collapse; width: 100%; margin: 1.2em 0; }
thead th { background: var(--forest); color: #fff; font-family: 'Source Code Pro', monospace; font-size: .76rem; letter-spacing: .06em; text-transform: uppercase; padding: 8px 12px; text-align: left; }
tbody tr:nth-child(even) { background: var(--mist); }
tbody td { padding: 7px 12px; border-bottom: 1px solid var(--parchment); }

.key-box { background: var(--forest); color: #fff; border-radius: 6px; padding: 20px 24px; margin: 2em 0; font-size: .9rem; }
.key-box h4 { font-family: 'Libre Baskerville', serif; color: var(--sage); font-size: 1rem; margin-bottom: 10px; }
.key-box ul { padding-left: 1.3em; }
.key-box li { margin-bottom: 5px; color: rgba(255,255,255,.85); }
.key-box code { background: rgba(255,255,255,.12); color: var(--amber-lt); }

.page-footer { background: var(--ink); color: rgba(255,255,255,.42); text-align: center; padding: 24px 4vw; margin: 60px calc(-1 * clamp(14px, 4vw, 48px)) 0; font-size: .76rem; font-family: 'Source Code Pro', monospace; letter-spacing: .04em; }
.page-footer a { color: var(--sage); text-decoration: none; }
.page-footer .divider { margin: 0 8px; opacity: .3; }
```

<div class="page-header">
<div class="module-tag">EA030 &mdash; Basic Statistics &mdash; Module 08</div>
<h1>ANOVA & <span>Group Comparisons</span></h1>
<p class="sub"><strong>Prof. Marc Los Huertos</strong> &nbsp;&bull;&nbsp; Pomona College</p>
<a class="back-link" href="../index.html">&#8592; Back to RTricks</a>
</div>

<div class="section-nav">
<a href="#concept">The Concept</a>
<a href="#partition">Partitioning Variance</a>
<a href="#oneway">One-Way ANOVA</a>
<a href="#posthoc">Post-hoc Tests</a>
<a href="#assumptions">Assumptions</a>
<a href="#nonparametric">Nonparametric</a>
<a href="#twoway">Two-Way ANOVA</a>
<a href="#viz">Visualization</a>
</div>

> **What this module covers.** ANOVA (Analysis of Variance) tests whether any of
> three or more group means differ. It does this by asking whether the variation
> *between* groups is larger than the variation *within* groups — the signal-to-noise
> ratio introduced in Module 04. This tutorial covers one-way ANOVA, post-hoc
> comparisons, assumption checking, and an introduction to two-way ANOVA with
> interaction.

---

## 1. The Concept: Why Not Just Run More t-Tests? {#concept}

Suppose you measure PM2.5 at three sites (urban, suburban, rural) and want to know
whether air quality differs by location. You could run three t-tests: urban vs.
suburban, urban vs. rural, suburban vs. rural.

The problem: with three tests at α = 0.05, the probability of at least one false
positive is $1 - (1 - 0.05)^3 = 0.143$ — nearly 14%. With 10 groups the problem is
catastrophic. ANOVA tests *all* groups simultaneously with a single α level,
controlling the family-wise error rate.

<div class="callout callout-concept">
<strong>📌 ANOVA's null hypothesis</strong>
$H_0: \mu_1 = \mu_2 = \cdots = \mu_k$ (all group means are equal)

$H_A$: at least one pair of means differs

ANOVA tells you *whether* any group differs, not *which* ones. Post-hoc tests answer
the follow-up question.
</div>

---

## 2. Partitioning Variance {#partition}

ANOVA decomposes the total variation in y into two components:

$$\underbrace{SS_\text{total}}_{\text{all variation}} = \underbrace{SS_\text{between}}_{\text{group effect}} + \underbrace{SS_\text{within}}_{\text{random error}}$$

| Source | SS | df | MS | F |
|---|---|---|---|---|
| Between groups | $\sum_j n_j(\bar y_j - \bar y)^2$ | $k-1$ | $SS_B / (k-1)$ | $MS_B / MS_W$ |
| Within groups | $\sum_j \sum_i (y_{ij} - \bar y_j)^2$ | $N-k$ | $SS_W / (N-k)$ | — |
| Total | $\sum_j \sum_i (y_{ij} - \bar y)^2$ | $N-1$ | — | — |

The F-statistic is the ratio of between-group variance to within-group variance.
Under $H_0$, both estimate the same population variance, so F ≈ 1. A large F indicates
the group means are more spread apart than chance alone would produce.

---

## 3. One-Way ANOVA in R {#oneway}

### The dataset

We use plant weight data from a classic fertilizer experiment: three groups (control,
treatment 1, treatment 2) with 10 plants each.

```{r data-setup}
data(PlantGrowth)
str(PlantGrowth)
table(PlantGrowth$group)
```

### Visualize first

```{r group-plot, fig.cap="Dry weight of plants by treatment group. Points are individual plants; box = IQR; line = median. Showing raw data alongside summaries is always best practice.", fig.height=4}
ggplot(PlantGrowth, aes(x = group, y = weight, color = group)) +
  geom_boxplot(outlier.shape = NA, width = 0.4, linewidth = 0.8) +
  geom_jitter(width = 0.12, size = 2, alpha = 0.7) +
  scale_color_manual(values = c("#4a7c5e", "#c8832a", "#8b3a2a")) +
  labs(x = "Treatment group", y = "Dry weight (g)",
       title = "Plant growth by treatment") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

### Fit the ANOVA

The recommended workflow in R is: fit with `lm()`, then pass to `anova()`. This keeps
the output consistent with the regression framework you learned in Module 07.

```{r fit-anova}
mod_anova <- lm(weight ~ group, data = PlantGrowth)
anova(mod_anova)
```

Equivalently, you can use `aov()`:

```{r aov-equivalent}
mod_aov <- aov(weight ~ group, data = PlantGrowth)
summary(mod_aov)
```

Both give identical results. The ANOVA table shows:

- **Df**: degrees of freedom — 2 for between (3 groups − 1), 27 for within (30 obs − 3).
- **Sum Sq**: the between-group and within-group sums of squares.
- **Mean Sq**: SS / df for each source.
- **F value**: the F-statistic = MS_between / MS_within.
- **Pr(>F)**: p-value. If < α, reject $H_0$.

<div class="callout callout-concept">
<strong>📌 Effect size: η²</strong>
$\eta^2 = SS_\text{between} / SS_\text{total}$ is the proportion of total variance
explained by the group factor — analogous to R² in regression. Values of 0.01, 0.06,
and 0.14 are conventionally described as small, medium, and large.

```r
ss    <- anova(mod_anova)$"Sum Sq"
eta2  <- ss[1] / sum(ss)
```
</div>

---

## 4. Post-hoc Tests {#posthoc}

A significant ANOVA F-test tells you that *some* groups differ. Post-hoc tests
identify *which* pairs differ while controlling for multiple comparisons.

### Tukey's Honest Significant Difference (HSD)

Tukey's HSD is the most common choice when you want to compare all pairwise combinations.
It controls the family-wise error rate exactly at α.

```{r tukey}
tukey_res <- TukeyHSD(mod_aov, conf.level = 0.95)
tukey_res
```

```{r tukey-plot, fig.cap="Tukey 95% family-wise confidence intervals for all pairwise group differences. Intervals that do not cross zero indicate a statistically significant difference.", fig.height=3.5}
par(las = 1, mar = c(4.5, 8, 3, 1))
plot(tukey_res, col = "#4a7c5e")
```

### Interpreting the output

Each row shows one pairwise comparison. The columns are:

| Column | Meaning |
|---|---|
| `diff` | Estimated difference in group means |
| `lwr`, `upr` | 95% confidence interval for the difference |
| `p adj` | Tukey-adjusted p-value |

An interval that does not include zero, or a p-value < α, indicates the two groups
are statistically different.

### Bonferroni correction (manual)

For a smaller set of planned comparisons, the Bonferroni correction is
straightforward: divide α by the number of comparisons.

```{r bonferroni}
pairwise.t.test(PlantGrowth$weight, PlantGrowth$group,
                p.adjust.method = "bonferroni")
```

<div class="callout callout-warn">
<strong>⚠️ Planned vs. unplanned comparisons</strong>
If you specified your hypotheses *before* seeing the data, a smaller set of
planned contrasts is appropriate and more powerful than Tukey HSD. If you are
doing exploratory pairwise comparisons after a significant F-test, use Tukey HSD
or Bonferroni to control inflation of the false positive rate.
</div>

---

## 5. Checking ANOVA Assumptions {#assumptions}

One-way ANOVA requires: (1) independence of observations, (2) normally distributed
residuals within each group, and (3) equal variances across groups (homoscedasticity).

### Residual normality

```{r qq-anova, fig.cap="Q-Q plot of ANOVA residuals. Points close to the diagonal support the normality assumption.", fig.height=3.5}
par(las = 1, mar = c(4.5, 4.5, 3, 1))
qqnorm(residuals(mod_anova), main = "Q-Q Plot of ANOVA Residuals",
       col = "#4a7c5e", pch = 20)
qqline(residuals(mod_anova), col = "#c8832a", lwd = 2)
```

```{r shapiro}
shapiro.test(residuals(mod_anova))
```

### Equal variances: Levene's test

```{r levene}
# Base R: Bartlett's test (sensitive to non-normality)
bartlett.test(weight ~ group, data = PlantGrowth)
```

<div class="callout callout-tip">
<strong>💡 ANOVA is robust to mild violations</strong>
ANOVA is reasonably robust to moderate departures from normality, especially with
equal group sizes. It is more sensitive to unequal variances when group sizes differ.
If Levene's or Bartlett's test is significant, consider Welch's one-way ANOVA
(`oneway.test()`) which does not assume equal variances.
</div>

### Welch's ANOVA (robust to unequal variances)

```{r welch-anova}
oneway.test(weight ~ group, data = PlantGrowth, var.equal = FALSE)
```

---

## 6. Nonparametric Alternative: Kruskal–Wallis Test {#nonparametric}

When the normality assumption is seriously violated and transformations do not help,
the Kruskal–Wallis test is the nonparametric counterpart of one-way ANOVA. It ranks
all observations and tests whether the rank distributions differ across groups.

```{r kruskal}
kruskal.test(weight ~ group, data = PlantGrowth)
```

Post-hoc: pairwise Wilcoxon tests with Bonferroni or BH correction.

```{r kruskal-posthoc}
pairwise.wilcox.test(PlantGrowth$weight, PlantGrowth$group,
                     p.adjust.method = "BH")
```

---

## 7. Two-Way ANOVA {#twoway}

Two-way ANOVA tests the effects of two categorical predictors simultaneously and
their **interaction** — whether the effect of one factor depends on the level of
the other.

### Model structure

```{r twoway-data}
# Use ToothGrowth: tooth length by supplement type and dose
data(ToothGrowth)
ToothGrowth$dose <- factor(ToothGrowth$dose)
str(ToothGrowth)
```

```{r twoway-fit}
mod_2way <- lm(len ~ supp * dose, data = ToothGrowth)
anova(mod_2way)
```

The ANOVA table now has three rows: main effect of `supp`, main effect of `dose`, and
the `supp:dose` interaction. A significant interaction means the effect of supplement
type differs across dose levels — the two factors are not independent.

### Interaction plot

```{r interaction-plot, fig.cap="Interaction plot: tooth length by dose and supplement. Non-parallel lines indicate an interaction between the two factors — the supplement effect is larger at lower doses.", fig.height=4}
ToothGrowth |>
  group_by(supp, dose) |>
  summarise(mean_len = mean(len), se_len = sd(len) / sqrt(n()), .groups = "drop") |>
  ggplot(aes(x = dose, y = mean_len, color = supp, group = supp)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_len - se_len, ymax = mean_len + se_len),
                width = 0.12) +
  scale_color_manual(values = c("#4a7c5e", "#c8832a")) +
  labs(x = "Dose (mg/day)", y = "Tooth length (mm)",
       title = "Tooth length by supplement and dose",
       color = "Supplement") +
  theme_minimal(base_size = 13)
```

<div class="callout callout-warn">
<strong>⚠️ If the interaction is significant, do not interpret main effects alone</strong>
A significant interaction means the effect of each factor depends on the level of the
other. Report and interpret the interaction first; the main effects become secondary
descriptions. Ignoring a significant interaction and reporting only main effects is a
common and serious misinterpretation.
</div>

---

## 8. Visualizing Group Differences {#viz}

A well-designed group comparison plot always shows: (1) the raw data, (2) a measure
of center, and (3) a measure of uncertainty.

```{r final-viz, fig.cap="Recommended visualization for group comparisons: raw jittered points (low alpha), group mean (filled point), and 95% confidence interval (error bar).", fig.height=4.5}
PlantGrowth |>
  group_by(group) |>
  summarise(
    mean_w = mean(weight),
    ci_lo  = mean_w - qt(0.975, n() - 1) * sd(weight) / sqrt(n()),
    ci_hi  = mean_w + qt(0.975, n() - 1) * sd(weight) / sqrt(n()),
    .groups = "drop"
  ) -> pg_ci

ggplot(PlantGrowth, aes(x = group, y = weight)) +
  geom_jitter(width = 0.1, alpha = 0.45, size = 2,
              color = "#4a7c5e") +
  geom_point(data = pg_ci, aes(y = mean_w),
             size = 4, color = "#1e3a2f") +
  geom_errorbar(data = pg_ci, aes(y = mean_w, ymin = ci_lo, ymax = ci_hi),
                width = 0.18, linewidth = 1, color = "#1e3a2f") +
  labs(x = "Treatment group", y = "Dry weight (g)",
       title = "Plant growth: mean ± 95% CI with raw data") +
  theme_minimal(base_size = 13)
```

<div class="key-box">
<h4>🔑 Key takeaways</h4>
<ul>
<li>ANOVA tests whether any of k group means differ using a single F-test, avoiding
the multiple-comparison inflation of repeated t-tests.</li>
<li>A significant F only tells you *some* groups differ. Use <strong>Tukey HSD</strong>
to identify which pairs are different.</li>
<li>Check the <strong>LINE assumptions</strong> as you would for regression — ANOVA is
the same model, just with a categorical predictor.</li>
<li>If variances are unequal, use <code>oneway.test(var.equal = FALSE)</code>.
If normality is seriously violated, use the <strong>Kruskal–Wallis test</strong>.</li>
<li>A significant <strong>interaction</strong> in two-way ANOVA must be reported and
interpreted before main effects.</li>
</ul>
</div>

---

### What's next?

- **[Simple Linear Regression](07_Simple_Linear_Regression.html)** — continuous predictor
  version of the same F-test framework
- **[Logistic Regression](09_Logistic_Regression.html)** — extend to binary outcomes
- **[Mixed-Effects Models](../advanced_stats/mixed_effects.html)** — handle repeated
  measures and nested designs

```{r session-info, echo=FALSE}
# sessionInfo()
```

<footer class="page-footer">
EA030 Basic Statistics — Module 08
<span class="divider">|</span> Prof. Marc Los Huertos
<span class="divider">|</span> Pomona College
<span class="divider">|</span> Last updated: `r format(Sys.Date(), "%B %Y")`
<span class="divider">|</span> <a href="https://github.com/marclos/RTricks">GitHub</a>
</footer>
