---
title: "ANOVA Example — Treatment Systems and E. coli"
author: "Your Name"
date: "`r Sys.Date()`"
output: pdf_document
---

## Introduction

Water-treatment systems are designed to reduce microbial contamination in wastewater.  
In this study, we compare **four different water-management systems** for their effect on *E. coli* concentration (log CFU/100 mL) after treatment:

1. Raw Sewage  
2. Treatment Plant  
3. Treatment Wetland  
4. Plant + Wetland  

Each treatment was replicated 4 times (16 observations total).

---

## Research Question

Do mean *E. coli* levels (log CFU/100 mL) differ among the four treatment systems?

### Hypotheses

- **Null hypothesis (H):**  
  \( \mu_{Raw} = \mu_{Plant} = \mu_{Wetland} = \mu_{Plant+Wetland} \)

- **Alternative hypothesis (H):**  
  At least one group mean differs.

We will use a **one-way ANOVA** to test for differences among the treatment means.

---

## Data

The data are simulated log-transformed *E. coli* concentrations for illustration.

| ID | Treatment | Ecoli_LogCFU |
|----|------------|--------------|
| 1 | Raw_Sewage | 6.9 |
| 2 | Raw_Sewage | 6.7 |
| 3 | Raw_Sewage | 6.8 |
| 4 | Raw_Sewage | 7.0 |
| 5 | Treatment_Plant | 3.7 |
| 6 | Treatment_Plant | 3.9 |
| 7 | Treatment_Plant | 3.6 |
| 8 | Treatment_Plant | 3.8 |
| 9 | Treatment_Wetland | 2.6 |
| 10 | Treatment_Wetland | 2.4 |
| 11 | Treatment_Wetland | 2.7 |
| 12 | Treatment_Wetland | 2.5 |
| 13 | Plant_and_Wetland | 1.9 |
| 14 | Plant_and_Wetland | 2.0 |
| 15 | Plant_and_Wetland | 1.8 |
| 16 | Plant_and_Wetland | 2.1 |

---

## ANOVA Analysis

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Create the dataset
```{r}
ID <- 1:16
Treatment <- c(rep("Raw_Sewage", 4),
               rep("Treatment_Plant", 4),
               rep("Treatment_Wetland", 4),
               rep("Plant_and_Wetland", 4))
Ecoli_LogCFU <- c(6.9, 6.7, 6.8, 7.0,
                  3.7, 3.9, 3.6, 3.8,
                  2.6, 2.4, 2.7, 2.5,
                  1.9, 2.0, 1.8, 2.1)

data <- data.frame(ID, Treatment, Ecoli_LogCFU)

# Display first few rows
head(data)

library(ggplot2)

ggplot(data, aes(x = Treatment, y = Ecoli_LogCFU, fill = Treatment)) +
  geom_boxplot() +
  labs(title = "E. coli (log CFU/100 mL) Across Treatment Systems",
       x = "Treatment System",
       y = "E. coli (log CFU/100 mL)") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 20, hjust = 1))


anova_model <- aov(Ecoli_LogCFU ~ Treatment, data = data)
summary(anova_model)

```