---
title: "Problem Results"
format: revealjs
editor: visual
---

## Some Dataset that "beg the question"

## The problems of data analysis and interpretation

Looking into **Using Models** might be dangerous

-   models are not reality
-   models are not the data
-   models are not the truth

## Anscombe Data

```{r}
Anscombe = as.data.frame(
  matrix(c(
    1, 10, 10, 10, 8, 8.04, 9.14, 7.46, 6.58, 
    2, 8, 8, 8, 8, 6.95, 8.14, 6.77, 5.76,
    3, 13, 13, 13, 8, 7.58, 8.74, 12.74, 7.71,
    4,  9, 9, 9, 8, 8.81, 8.77, 7.11, 8.84,
    5, 11, 11, 11, 8, 8.33, 9.26, 7.81, 8.47,
    6, 14, 14, 14, 8, 9.96, 8.10, 8.84, 7.04,
    7,  6, 6, 6, 8, 7.24, 6.13, 6.08, 5.25,
    8,  4, 4, 4, 19, 4.26, 3.10, 5.39, 12.50,
    9, 12, 12, 12, 8, 10.84, 9.13, 8.15, 5.56,
    10, 7, 7, 7, 8, 4.82, 7.26, 6.42, 7.91,
    11, 5, 5, 5, 8, 5.68, 4.74, 5.73, 6.89), 
    nrow=11,  byrow=TRUE)) 

names(Anscombe) = c("Obs.", "x1", "x2", "x3", "x4", "y1", "y2", "y3", "y4")

write.csv(file="/home/mwl04747/RTricks/00_Stats/Anscombe_files/ansombe.csv", Anscombe)

```

## Results -- Visualizations

```{r}
par(mfrow=c(2,2))
plot(y1 ~ x1, data = Anscombe, main="I")
plot(Anscombe$x2, Anscombe$y2, main="II")
plot(Anscombe$x3, Anscombe$y3, main="III")
plot(Anscombe$x4, Anscombe$y4, main="IV")
```

## Linear Model Coefficents
```{r}
lm1 <- lm(y1 ~ x1, data = Anscombe)
lm2 <- lm(y2 ~ x2, data = Anscombe)
lm3 <- lm(y3 ~ x3, data = Anscombe)
lm4 <- lm(y4 ~ x4, data = Anscombe)
```

## Linear Model Summaries

```{r}
summary(lm1)
```

## Linear Model Summaries

```{r}
summary(lm2)
```

## Linear Model Summaries

```{r}
summary(lm3)
```

## Linear Model Summaries

```{r}
summary(lm4)
```

## Plotting Results

```{r, echo=FALSE}
par(mfrow=c(2,2))
plot(y1 ~ x1, data = Anscombe, main="I")
abline(coef(lm1), col="red")
plot(y1 ~ x1, data = Anscombe, main="II")
abline(coef(lm2), col="red")
plot(y1 ~ x1, data = Anscombe, main="III")
abline(coef(lm3), col="red")
plot(y1 ~ x1, data = Anscombe, main="IV")
abline(coef(lm4), col="red")
```

## Plotting Diagnostics Plots

```{r}
par(mfrow=c(2,2))
plot(lm1)
```

```{r}
par(mfrow=c(2,2))
plot(lm2)
```

```{r}
par(mfrow=c(2,2))
plot(lm3)
```

```{r}
par(mfrow=c(2,2))
plot(lm4)
```

