---
title: "Course Time"
author: "Marc Los Huertos"
date: "`r Sys.Date()`"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## General Expections

Based on the "accepted" expectations, we should be a spending a minimum of 2 hours outside of class for every hour in class (2:1 ratio). Counting only the lecture and 1/2 of lab time that would be about atleast 6 hours outside of class. 

In general, I get worried if a class requires more than 8 hours outside of class for every hour in class on a consistent basis.

## Time Spent Outside of Class

```{r, echo=TRUE}

week2 <- c(1.5, 1, 15, 40, 3, 2, 10, 1.5, 
           0, 20, 2, 45, 7, 10, 2, 0, 2, 2, 2.5, 5, 0)
week3 <- c(0.5, 4, 2, 1.5, 6, 5, 3, 5, 3, 5, 3, 6, 2, 5, 5, 3, 3, 4.5, 5, 8)
week4 <- c(1.5, 3,3, 2, 7, 8, 4, 3, 3, 3, 6, 1,2, 8, 6, 2, 6, 10, 2, 7, 5.5, 5, 8)
week5 <- c(1, 3, 6, 3, 5, 7, 4, 2, 4, 3, 7, 4, 8, 4, 1, 6, 13, 5, 6, 7, 7)
week6 <- c(6, 3, 6, 6, 10, 5, 8, 6, 4, 7, 6, 4, 3, 7, 5, 6, 9, 7)
```

## Create a Data Frame

```{r}
TimeOutSideClass <- data.frame(Week = c(rep("Week 2", length=length(week2)),
                                        rep("Week 3", length=length(week3)),
                                        rep("Week 4", length=length(week4)),
                                        rep("Week 5", length=length(week5)),
                                        rep("Week 6", length=length(week6))),
                               Time = c(week2, week3, week4, week5, week6))
```

## Boxplot

```{r}
par(mar=c(5,5,2,2))
boxplot(TimeOutSideClass$Time ~ TimeOutSideClass$Week, 
        col=c("lightblue", "lightgreen", "lightyellow", "lightpink", "lightgrey"),
        ylab="Time (hours)", xlab="Week", main="Time spent outside in a week", las=1)
abline(h=mean(TimeOutSideClass$Time), lwd=2, lty=2, col="red")

```




## Skewed Distribution

- Outliers
- Data Entry Mistakes, maybe?
- Excessive Readings, likely!
```{r}
par(mfrow=c(3,2), las=1)
hist(week2, breaks=20, col="lightblue", border="black", xlim=c(0,50), xlab="Time (hours)", main="Week 2", ylab="Frequency")
hist(week3, breaks=20, col="lightblue", border="black", xlim=c(0,13), xlab="Time (hours)", main="Week 3", ylab="Frequency")
hist(week4, breaks=20, col="lightblue", border="black", xlim=c(0,13), xlab="Time (hours)", main="Week 4", ylab="Frequency")
hist(week5, breaks=20, col="lightblue", border="black", xlim=c(0,13), xlab="Time (hours)", main="Week 5", ylab="Frequency")
hist(week6, breaks=20, col="lightblue", border="black", xlim=c(0,13), xlab="Time (hours)", main="Week 6", ylab="Frequency")


```

## t-test a hypothesis: Different than 6 hours

```{r ttest2, echo = TRUE}
summary(aov(Time ~ Week, data=TimeOutSideClass))
```


```{r}

TimeOutsideClass2 <- subset(TimeOutSideClass, Week!="Week 2")

summary(aov(Time ~ Week, data=TimeOutsideClass2))


```
## Boxplot

```{r}
par(mar=c(5,5,2,2))
boxplot(Time ~ Week, data = TimeOutsideClass2,
        col=c("lightblue", "lightgreen", "lightyellow", "lightpink", "lightgrey"),
        ylab="Time (hours)", xlab="Week", main="Time spent outside in a week", las=1)
abline(h=mean(TimeOutsideClass2$Time), lwd=2, lty=2, col="red")
abline(h=6, lwd=2, lty=2, col="blue")


```


## t-test a hypothesis: Different than 6 hours (Week 3)

```{r test, echo = TRUE}
t.test(week6, mu=6)
```


## t-test: Greater than 6 hours

```{r testcompare, echo = TRUE}
week2.ttest =t.test(week5, mu=6, alternative="greater")
week3.ttest = t.test(week6, mu=6, alternative="less" )
```

```{r, echo=FALSE}
# str(week2.ttest)

print(paste0("H1 NULL: Week 5 (!>6): ", round(week2.ttest$p.value,3), " (", week2.ttest$method, " Greater than mu)")
)
print(paste0("H2 NULL: Week 6 (!>6): ", round(week3.ttest$p.value,3), " (", week2.ttest$method, " (Greater than mu)")
)
```

## Marc's Prior Conclusions

- Make sure the quiz directions are clear.
- Reduced readings are working, perhaps too well!
- Offer additional support: If you find the readings to take more than 6-8 hours.
- Determine how lab work might be used to augment weekly hour workload. Project will take up more time. Will gauge next week.

## New Hypotheses ?





