---
title: "Contigency"
author: "Marc Los Huertos"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Freshwater contamination is a major public health concern, particularly in regions where people rely on untreated water sources. Previous studies have shown that exposure to contaminated water may increase the risk of gastrointestinal disease.  

This analysis examines whether there is a **statistically significant association** between freshwater contamination and disease occurrence using hypothetical survey data.

## Data

The following table summarizes household survey results:

| Water Contamination | Disease Reported | No Disease Reported | Total |
|----------------------|------------------|---------------------|--------|
| Contaminated         | 35               | 15                  | 50     |
| Clean                | 10               | 40                  | 50     |
| **Total**            | **45**           | **55**              | **100**|

We will use a **Chi-square test of independence** to determine whether there is a significant association between water contamination and disease occurrence.

# Create contingency table
```{r}
data <- matrix(c(35, 15, 10, 40),
               nrow = 2,
               byrow = TRUE)

colnames(data) <- c("Disease", "No_Disease")
rownames(data) <- c("Contaminated", "Clean")

# Display the table
data


# Perform chi-square test
chisq_result <- chisq.test(data)

# Display results
chisq_result

```
