---
title: "Regression Analysis: Distance from CO Source and Asthma Severity"
author: "Marc"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Introduction

Air pollution, especially **carbon monoxide (CO)**, has been linked to a variety of negative respiratory health outcomes.  
In this analysis, we test whether **distance from a CO source** can predict **asthma severity** among households.

We expect that people living **closer to a CO source** may experience **more severe asthma symptoms** than those living farther away.

---

## Data

The following table presents the study data:

| Household | Distance_from_CO (m) | Asthma_Severity (1–10) |
|------------|----------------------|------------------------|
| 1 | 50 | 9.0 |
| 2 | 75 | 8.5 |
| 3 | 100 | 8.0 |
| 4 | 125 | 7.4 |
| 5 | 150 | 7.0 |
| 6 | 200 | 6.5 |
| 7 | 250 | 6.0 |
| 8 | 300 | 5.5 |
| 9 | 350 | 5.2 |
| 10 | 400 | 5.0 |

---

## Regression Model

We will fit a **simple linear regression model** to examine whether *distance from CO source* predicts *asthma severity*.

### Model:
\[
\text{Asthma Severity} = \beta_0 + \beta_1 (\text{Distance from CO}) + \epsilon
\]

# Create the dataset
```{r}
household <- 1:10
distance <- c(50, 75, 100, 125, 150, 200, 250, 300, 350, 400)
asthma <- c(9.0, 8.5, 8.0, 7.4, 7.0, 6.5, 6.0, 5.5, 5.2, 5.0)

data <- data.frame(Household = household,
                   Distance_from_CO = distance,
                   Asthma_Severity = asthma)

# Fit linear regression model
model <- lm(Asthma_Severity ~ Distance_from_CO, data = data)

# Show model summary
summary(model)

coef(model)


library(ggplot2)

ggplot(data, aes(x = Distance_from_CO, y = Asthma_Severity)) +
  geom_point(size = 3, color = "blue") +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(title = "Asthma Severity vs. Distance from CO Source",
       x = "Distance from CO Source (m)",
       y = "Asthma Severity (1-10)") +
  theme_minimal()
```
```

