---
title: "Climate Change Project"
author: "Marc Los Huertos"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Updated: 2024-02-16


# Guide 1

## Chunk 1 Source Guide1fuctions

```{r, chunk1_1}
source("/home/mwl04747/RTricks/04_Regional_Climate_Trends/Guides/Guide1functions.R")
```

## Chunk 2 Read Inventory
```{r, chunk1_2}
my.state = "IL"
filename.csv <- "/home/mwl04747/RTricks/04_Regional_Climate_Trends/stations.active.oldest.csv"
my.inventory <- readInventory.fun(filename.csv, my.state)
```

## Chunk 3 Download Stations
```{r, chunk1_3, results='hide'}
datapath = "/home/mwl04747/RTricks/04_Regional_Climate_Trends/Data/SP24/"
downloadStations.fun(datapath, my.inventory)
```

# Guide 2

## Chunk 1 Source Guide1fuctions

```{r chunk2_1}
source("/home/mwl04747/RTricks/04_Regional_Climate_Trends/Guides/Guide2functions.R")
```

## Chunk 2 Read Stations
```{r, chunk2_2}
datafolder = "/home/mwl04747/RTricks/04_Regional_Climate_Trends/Data/SP24/"
ReadStations2.fun(datafolder)
```

## Chunk 3 Sort Stations --> Skip
```{r}
# sortstations.fun() Nor working yet. 
```

## Chunk 4 (Fix Date)
```{r, chunk2_4}
ls()
USC00117391a <- fixDates.fun(USC00117391)
USC00116526a <- fixDates.fun(USC00116526)
USC00110338a <- fixDates.fun(USC00110338)
USC00111280a <- fixDates.fun(USC00111280)
USC00111329a <- fixDates.fun(USC00111329)
USC00112193a <- fixDates.fun(USC00112193)
USC00112348a <- fixDates.fun(USC00112348)
USC00113335a <- fixDates.fun(USC00113335)
USC00115768a <- fixDates.fun(USC00115768)
USC00116610a <- fixDates.fun(USC00116610)
USC00118684a <- fixDates.fun(USC00118684)
USC00118916a <- fixDates.fun(USC00118916)
USC00119241a <- fixDates.fun(USC00119241)
USC00110187a <- fixDates.fun(USC00110187)
USC00112140a <- fixDates.fun(USC00112140)


```

## Chunk 5 Check Coverage
```{r, chunk2_5}
coverage.fun(USC00331592a)
coverage.fun(USC00330563a)
coverage.fun(USC00330862a)
coverage.fun(USC00331072a)
coverage.fun(USC00332098a)
coverage.fun(USC00332791a)
coverage.fun(USC00333375a)
coverage.fun(USC00333758a)
coverage.fun(USC00333780a)
coverage.fun(USC00335438a)
coverage.fun(USC00336781a)
coverage.fun(USC00338609a)
coverage.fun(USC00338769a)
coverage.fun(USC00338822a)
coverage.fun(USC00338830a)


```

## Chunk 6 Fix Values
```{r, chunk2_6}
USC00331592b <- fixValues.fun(USC00331592a)
USC00330563b <- fixValues.fun(USC00330563a)
USC00330862b <- fixValues.fun(USC00330862a)
USC00331072b <- fixValues.fun(USC00331072a)
USC00332098b <- fixValues.fun(USC00332098a)
USC00332791b <- fixValues.fun(USC00332791a)
USC00333375b <- fixValues.fun(USC00333375a)
USC00333758b <- fixValues.fun(USC00333758a)
USC00333780b <- fixValues.fun(USC00333780a)
USC00335438b <- fixValues.fun(USC00335438a)
USC00336781b <- fixValues.fun(USC00336781a)
USC00338609b <- fixValues.fun(USC00338609a)
USC00338769b <- fixValues.fun(USC00338769a)
USC00338822b <- fixValues.fun(USC00338822a)
USC00338830b <- fixValues.fun(USC00338830a)

```

## Chunk 7 QAQC
```{r, chunk2_7}
QAQC.fun(USC00042294b)
```

## Chunk 8 # Monthly Means and Normals
```{r, chunk2_8}
USC00331592.monthly <- MonthlyValues.fun(USC00331592b)
USC00331592.normals <- MonthlyNormals.fun(USC00331592b)

USC00330563.monthly <- MonthlyValues.fun(USC00330563b)
USC00330563.normals <- MonthlyNormals.fun(USC00330563b)

USC00330862.monthly <- MonthlyValues.fun(USC00330862b)
USC00330862.normals <- MonthlyNormals.fun(USC00330862b)

USC00331072.monthly <- MonthlyValues.fun(USC00331072b)
USC00331072.normals <- MonthlyNormals.fun(USC00331072b)

USC00332098.monthly <- MonthlyValues.fun(USC00332098b)
USC00332098.normals <- MonthlyNormals.fun(USC00332098b)

USC00332791.monthly <- MonthlyValues.fun(USC00332791b)
USC00332791.normals <- MonthlyNormals.fun(USC00332791b)

USC00333375.monthly <- MonthlyValues.fun(USC00333375b)
USC00333375.normals <- MonthlyNormals.fun(USC00333375b)

USC00333758.monthly <- MonthlyValues.fun(USC00333758b)
USC00333758.normals <- MonthlyNormals.fun(USC00333758b)

USC00333780.monthly <- MonthlyValues.fun(USC00333780b)
USC00333780.normals <- MonthlyNormals.fun(USC00333780b)

USC00335438.monthly <- MonthlyValues.fun(USC00335438b)
USC00335438.normals <- MonthlyNormals.fun(USC00335438b)

USC00336781.monthly <- MonthlyValues.fun(USC00336781b)
USC00336781.normals <- MonthlyNormals.fun(USC00336781b)

USC00338609.monthly <- MonthlyValues.fun(USC00338609b)
USC00338609.normals <- MonthlyNormals.fun(USC00338609b)

USC00338769.monthly <- MonthlyValues.fun(USC00338769b)
USC00338769.normals <- MonthlyNormals.fun(USC00338769b)

USC00338822.monthly <- MonthlyValues.fun(USC00338822b)
USC00338822.normals <- MonthlyNormals.fun(USC00338822b)

USC00338830.monthly <- MonthlyValues.fun(USC00338830b)
USC00338830.normals <- MonthlyNormals.fun(USC00338830b)


```


## Chunk 9 Anomalies
```{r, chunk2_9}
USC00331592.anomalies <- MonthlyAnomalies.fun(USC00331592.monthly, USC00331592.normals)
USC00330563.anomalies <- MonthlyAnomalies.fun(USC00330563.monthly, USC00330563.normals)
USC00330862.anomalies <- MonthlyAnomalies.fun(USC00330862.monthly, USC00330862.normals)
USC00331072.anomalies <- MonthlyAnomalies.fun(USC00331072.monthly, USC00331072.normals)
USC00332098.anomalies <- MonthlyAnomalies.fun(USC00332098.monthly, USC00332098.normals)
USC00332791.anomalies <- MonthlyAnomalies.fun(USC00332791.monthly, USC00332791.normals)
USC00333375.anomalies <- MonthlyAnomalies.fun(USC00333375.monthly, USC00333375.normals)
USC00333758.anomalies <- MonthlyAnomalies.fun(USC00333758.monthly, USC00333758.normals)
USC00333780.anomalies <- MonthlyAnomalies.fun(USC00333780.monthly, USC00333780.normals)
USC00335438.anomalies <- MonthlyAnomalies.fun(USC00335438.monthly, USC00335438.normals)
USC00336781.anomalies <- MonthlyAnomalies.fun(USC00336781.monthly, USC00336781.normals)
USC00338609.anomalies <- MonthlyAnomalies.fun(USC00338609.monthly, USC00338609.normals)
USC00338769.anomalies <- MonthlyAnomalies.fun(USC00338769.monthly, USC00338769.normals)
USC00338822.anomalies <- MonthlyAnomalies.fun(USC00338822.monthly, USC00338822.normals)
USC00338830.anomalies <- MonthlyAnomalies.fun(USC00338830.monthly, USC00338830.normals)


```



# Guide 3

## Chunk 1 Source Guide3fuctions

You might see a note about file not found, this is for Marc, you can ignore.
```{r, chunk3_1}
source("/home/mwl04747/RTricks/04_Regional_Climate_Trends/Guides/Guide3functions.R")
```

## Chunk 2 Analyze for Trends by Month
```{r chunk3_2}
USC00331592.trends <- monthlyTrend.fun(USC00331592.anomalies)
USC00330563.trends <- monthlyTrend.fun(USC00330563.anomalies)
USC00330862.trends <- monthlyTrend.fun(USC00330862.anomalies)
USC00331072.trends <- monthlyTrend.fun(USC00331072.anomalies)
USC00332098.trends <- monthlyTrend.fun(USC00332098.anomalies)
USC00332791.trends <- monthlyTrend.fun(USC00332791.anomalies)
USC00333375.trends <- monthlyTrend.fun(USC00333375.anomalies)
USC00333758.trends <- monthlyTrend.fun(USC00333758.anomalies)
USC00333780.trends <- monthlyTrend.fun(USC00333780.anomalies)
USC00335438.trends <- monthlyTrend.fun(USC00335438.anomalies)
USC00336781.trends <- monthlyTrend.fun(USC00336781.anomalies)
USC00338609.trends <- monthlyTrend.fun(USC00338609.anomalies)
USC00338769.trends <- monthlyTrend.fun(USC00338769.anomalies)
USC00338822.trends <- monthlyTrend.fun(USC00338822.anomalies)
USC00338830.trends <- monthlyTrend.fun(USC00338830.anomalies)


```

## Chunk 3 TBD

Camille's Analysis

Let's find rank TMIN trends for each station. 
```{r}
library(dplyr)

temp1 <- USC00331592.trends %>%
  filter(ELEMENT == "TMIN") %>%
  filter(`Pr(>|t|)` < 0.05) %>%
  arrange(-Estimate) %>%
  mutate(ID = "USC00331592")
temp1

temp2 <- USC00330563.trends %>%
  filter(ELEMENT == "TMIN") %>%
  filter(`Pr(>|t|)` < 0.05) %>%
  arrange(-Estimate) %>%
  mutate(ID = "USC00330563")
temp2

temp3 <- USC00330862.trends %>%
  filter(ELEMENT == "TMIN") %>%
  filter(`Pr(>|t|)` < 0.05) %>%
  arrange(-Estimate) %>%
  mutate(ID = "USC00330862")
  

temp4 <- USC00331072.trends %>%
  filter(ELEMENT == "TMIN") %>%
  filter(`Pr(>|t|)` < 0.05) %>%
  arrange(-Estimate) %>%
  mutate(ID = "USC00331072")

temp5 <- USC00332098.trends %>%
  filter(ELEMENT == "TMIN") %>%
  filter(`Pr(>|t|)` < 0.05) %>%
  arrange(-Estimate) %>%
  mutate(ID = "USC00332098")

temp6 <- USC00332791.trends %>%
  filter(ELEMENT == "TMIN") %>%
  filter(`Pr(>|t|)` < 0.05) %>%
  arrange(-Estimate) %>%
  mutate(ID = "USC00332791")

temp7 <- USC00333375.trends %>%
  filter(ELEMENT == "TMIN") %>%
  filter(`Pr(>|t|)` < 0.05) %>%
  arrange(-Estimate) %>%
  mutate(ID = "USC00333375")

temp8 <- USC00333758.trends %>%
  filter(ELEMENT == "TMIN") %>%
  filter(`Pr(>|t|)` < 0.05) %>%
  arrange(-Estimate) %>%
  mutate(ID = "USC00333758")

temp9 <- USC00333780.trends %>%
  filter(ELEMENT == "TMIN") %>%
  filter(`Pr(>|t|)` < 0.05) %>%
  arrange(-Estimate) %>%
  mutate(ID = "USC00333780")

temp10 <- USC00335438.trends %>%
  filter(ELEMENT == "TMIN") %>%
  filter(`Pr(>|t|)` < 0.05) %>%
  arrange(-Estimate) %>%
  mutate(ID = "USC00335438")

temp11 <- USC00336781.trends %>%
  filter(ELEMENT == "TMIN") %>%
  filter(`Pr(>|t|)` < 0.05) %>%
  arrange(-Estimate) %>%
  mutate(ID = "USC00336781")


temp12 <- USC00338609.trends %>%
  filter(ELEMENT == "TMIN") %>%
  filter(`Pr(>|t|)` < 0.05) %>%
  arrange(-Estimate) %>%
  mutate(ID = "USC00338609")

temp13 <- USC00338769.trends %>%
  filter(ELEMENT == "TMIN") %>%
  filter(`Pr(>|t|)` < 0.05) %>%
  arrange(-Estimate) %>%
  mutate(ID = "USC00338769")

temp14 <- USC00338822.trends %>%
  filter(ELEMENT == "TMIN") %>%
  filter(`Pr(>|t|)` < 0.05) %>%
  arrange(-Estimate) %>%
  mutate(ID = "USC00338822")

temp15 <- USC00338830.trends %>%
  filter(ELEMENT == "TMIN") %>%
  filter(`Pr(>|t|)` < 0.05) %>%
  arrange(-Estimate) %>%
  mutate(ID = "USC00338830")
```

## Suggestions for Next Steps

Create a dataframe with each of the stations' locations and the TMIN trend for June that includes the p-value and slope.

```{r}
# example
# put stations together
temp.trends = rbind(temp1, temp2, temp3, temp4, temp5, temp6, temp7, temp8, temp10, temp11, temp12, temp13, temp14)
# 9, 15 no obs
temp.trends

# join with my.inventory
temp.trend2 <- full_join(temp.trends, 
        subset(my.inventory, select=c("ID", "LATITUDE", "LONGITUDE", "STATE")), by='ID')
names(temp.trend2)

# select june and select columns
june.trend3 <- temp.trend2 %>%
  filter(MONTH == 6) %>%
  select(ID, STATE, LATITUDE, LONGITUDE, Estimate, `Pr(>|t|)`, r.squared) %>%
  mutate(Estimate = round(Estimate*100, 0))

june.trend3

```


Then we create a map with the size of the point representing the magnitude of the trend and the color representing the p-value.

using 

```{r}
# map ID for LONGITUDE and LATITUDE USING ESTIMATE AS SIZE AND P-VALUE AS COLOR
# ohio boundaries
ohio <- map_data("state", region = "ohio")
library(ggplot2)
library(ggmap)
library(maps)
library(mapdata)
ggplot(june.trend3, aes(x=LONGITUDE, y=LATITUDE, size=Estimate, color=r.squared)) +
#  geom_polygon(data = ohio, fill = "white", color = "black") +
  geom_point() +
  scale_size_continuous(range = c(1, 4)) +
  scale_color_gradient(low = "blue", high = "red") +
  theme_minimal() +
  labs(title = "TMIN June Trends", x = "Longitude", y = "Latitude")


```



# Guide 4

## Chunk 1 Source Guide4functions

You might see a note about file note found, this is for Marc, you can ignore.
```{r chunk4_1}
source("/home/mwl04747/RTricks/04_Regional_Climate_Trends/Guides/Guide4functions.R")
```

## Chunk 2 Plot Trend
```{r}
plotTrend2.fun(USC00042294.anomalies, "TMIN", 7)
```

almost done!


# Zip Files for Class

Marc's has created a function to zip all the costum functions and active stations file into a zip file for easy download/upload: Guidefunctions.zip.
```{r, echo=FALSE}
zip(zipfile = "/home/mwl04747/RTricks/04_Regional_Climate_Trends/Guides/Guidefunctions.zip", 
    files = c("/home/mwl04747/RTricks/04_Regional_Climate_Trends/Guides/Guide1functions.R",
              "/home/mwl04747/RTricks/04_Regional_Climate_Trends/Guides/Guide2functions.R",
              "/home/mwl04747/RTricks/04_Regional_Climate_Trends/Guides/Guide3functions.R",
              "/home/mwl04747/RTricks/04_Regional_Climate_Trends/Guides/Guide4functions.R",
              "/home/mwl04747/RTricks/04_Regional_Climate_Trends/Guides/MarcsTemplate.Rmd",
              "/home/mwl04747/RTricks/04_Regional_Climate_Trends/stations.active.oldest.csv"),
    flags= "-j")
```


# Putting it together 
```{r, eval=FALSE}
# Guide 1
source("/home/mwl04747/RTricks/04_Regional_Climate_Trends/Guides/Guide1functions.R") # 1_1

my.state = "MT"
filename.csv <- "/home/mwl04747/RTricks/04_Regional_Climate_Trends/stations.active.oldest.csv" # 1_1
my.inventory <- readInventory.fun(filename.csv, my.state) # 1_2
datapath = "/home/mwl04747/RTricks/04_Regional_Climate_Trends/Data/SP24/"
downloadStations.fun(datapath, my.inventory) # 1_3

# Guide 2
source("/home/mwl04747/RTricks/04_Regional_Climate_Trends/Guides/Guide2functions.R") # 2_1

datafolder = "/home/mwl04747/RTricks/04_Regional_Climate_Trends/Data/SP24/"
ReadStations2.fun(datafolder) # 2_2

#station = 	USC00246918
#stationID = "USW00054797"

	
USC00246918a <- fixDates.fun(USC00246918) # 2_4
coverage.fun(USC00246918a) # 2_5
USC00246918b <- fixValues.fun(USC00246918a) # 2_6
QAQC.fun(USC00246918b) # 2_7
USC00246918.monthly <- MonthlyValues.fun(USC00246918b) # 2_8
USC00246918.normals <- MonthlyNormals.fun(USC00246918b) # 2_8

USC00246918.anomalies <- MonthlyAnomalies.fun(USC00246918.monthly, USC00246918.normals) # 2_9

# Guide 3
source("/home/mwl04747/RTricks/04_Regional_Climate_Trends/Guides/Guide3functions.R") # 3_1
USC00246918.trends <- monthlyTrend.fun(USC00246918.anomalies) # 3_2
plotTrend2.fun(USC00246918.anomalies, "TMIN", 7) # 4_2


#Guide 4
#source("/home/mwl04747/RTricks/04_Regional_Climate_Trends/Guides/Guide4functions.R") # 4_1

```

















