---
title: "Camera Trap Analysis"
author: "Marc Los Huertos"
date: "EA030 · Compiled `r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    self_contained: true
    theme: null
    highlight: null
    toc: false
knit: (function(inputFile, encoding) {
    rmarkdown::render(
      inputFile,
      encoding = encoding,
      output_file = file.path(dirname(inputFile),
                              "../docs/Unit_Biodiversity/P05_Camera_Trap.html")
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 8,
  fig.height = 5,
  fig.align  = "center"
)
```

```{css, echo=FALSE}
@import url('https://fonts.googleapis.com/css2?family=Libre+Baskerville:ital,wght@0,400;0,700;1,400&family=Source+Code+Pro:wght@400;600&family=Lato:wght@300;400;700&display=swap');

*, *::before, *::after { box-sizing: border-box; margin: 0; padding: 0; }

:root {
  --forest:    #1e3a2f;
  --moss:      #2d5a3d;
  --fern:      #4a7c5e;
  --sage:      #8aad96;
  --mist:      #d4e6da;
  --cream:     #f7f3ed;
  --parchment: #ede8df;
  --ink:       #1a1a18;
  --charcoal:  #3a3a38;
  --dust:      #7a7a72;
  --amber:     #c8832a;
  --amber-lt:  #e8a84a;
  --white:     #ffffff;
  --card-shadow: 0 2px 10px rgba(30,58,47,.09);
}

html { scroll-behavior: smooth; }
body { font-family: 'Lato', sans-serif; background: var(--cream);
       color: var(--ink); font-size: 17px; line-height: 1.75; }
.main-container { max-width: 960px !important; margin: 0 auto !important;
                  padding: 0 4vw 80px !important; }
h1.title, h4.author, h4.date { display: none; }

/* ── Page header ─────────────────────────────────────────────────────────── */
.page-header {
  background-image:
    repeating-linear-gradient(120deg, transparent, transparent 60px,
      rgba(255,255,255,.018) 60px, rgba(255,255,255,.018) 61px),
    linear-gradient(160deg, #1e3a2f 0%, #2d5a3d 60%, #1a3d30 100%);
  color: #fff; padding: 44px 4vw 36px; margin: 0 -4vw 48px;
  border-bottom: 4px solid var(--amber); position: relative; overflow: hidden;
}
.page-header::after {
  content: ''; position: absolute; right: -40px; top: -40px;
  width: 260px; height: 260px; border-radius: 50%;
  background: radial-gradient(circle, rgba(138,173,150,.12) 0%, transparent 70%);
  pointer-events: none;
}
.page-header .module-tag {
  display: inline-block; font-family: 'Source Code Pro', monospace;
  font-size: .68rem; font-weight: 600; letter-spacing: .18em; text-transform: uppercase;
  color: var(--amber-lt); border: 1px solid rgba(200,131,42,.45);
  padding: 3px 10px; border-radius: 2px; margin-bottom: 12px;
}
.page-header h1 {
  font-family: 'Libre Baskerville', serif;
  font-size: clamp(1.6rem, 3vw, 2.3rem);
  font-weight: 700; color: #fff; line-height: 1.2; margin-bottom: 8px;
}
.page-header h1 span { color: var(--sage); }
.page-header .sub { font-size: .9rem; color: rgba(255,255,255,.65); font-weight: 300; }
.page-header .sub strong { color: rgba(255,255,255,.9); font-weight: 700; }
.back-link {
  display: inline-block; margin-top: 16px;
  font-family: 'Source Code Pro', monospace; font-size: .7rem; letter-spacing: .08em;
  color: var(--amber-lt); text-decoration: none;
  border: 1px solid rgba(200,131,42,.4); padding: 4px 11px; border-radius: 3px;
  transition: background .2s;
}
.back-link:hover { background: rgba(200,131,42,.18); }

/* ── Section nav ─────────────────────────────────────────────────────────── */
.section-nav {
  display: flex; gap: 8px; flex-wrap: wrap; margin-bottom: 44px;
  padding-bottom: 20px; border-bottom: 1px solid var(--parchment);
}
.section-nav a {
  font-family: 'Source Code Pro', monospace; font-size: .7rem; font-weight: 600;
  letter-spacing: .08em; text-transform: uppercase; color: var(--fern);
  text-decoration: none; border: 1px solid var(--mist); padding: 4px 12px;
  border-radius: 3px; background: var(--white); transition: background .2s, color .2s;
}
.section-nav a:hover { background: var(--mist); color: var(--forest); }

/* ── Headings ────────────────────────────────────────────────────────────── */
h2 {
  font-family: 'Libre Baskerville', serif; font-size: 1.45rem; font-weight: 700;
  color: var(--forest); margin-top: 56px; margin-bottom: 6px;
  padding-bottom: 8px; border-bottom: 1px solid var(--parchment);
}
h3 {
  font-family: 'Libre Baskerville', serif; font-size: 1.1rem; font-weight: 700;
  color: var(--moss); margin-top: 28px; margin-bottom: 4px;
}

p { margin-bottom: 1em; }
strong { color: var(--forest); }

/* ── Code ────────────────────────────────────────────────────────────────── */
code {
  font-family: 'Source Code Pro', monospace; font-size: .84em;
  background: var(--mist); color: var(--moss); padding: 1px 5px; border-radius: 3px;
}
pre {
  background: var(--ink); color: #e8e6df; border-radius: 5px;
  padding: 14px 18px; font-size: .83rem; line-height: 1.55;
  overflow-x: auto; border-left: 4px solid var(--amber); margin: 1.2em 0;
}
pre code { background: transparent; color: inherit; padding: 0; font-size: 1em; }

/* ── Callouts ────────────────────────────────────────────────────────────── */
.callout {
  border-left: 5px solid; padding: .65em 1.1em;
  margin: 1.4em 0; border-radius: 0 5px 5px 0; font-size: .92em; line-height: 1.6;
}
.callout > strong:first-child { display: block; margin-bottom: .3em; }
.callout-concept { border-color: var(--fern);  background: #f0f7f2; }
.callout-tip     { border-color: #2e7d32;      background: #f1f8f1; }
.callout-warn    { border-color: #e65100;      background: #fff8f4; }
.callout-new     { border-color: #1565c0;      background: #f0f4ff; }
.callout-formula { border-color: var(--amber); background: #fdf8f0;
                   font-family: 'Source Code Pro', monospace; font-size: .88em; }

/* ── Pipeline summary table ──────────────────────────────────────────────── */
.pipeline-grid {
  display: grid; grid-template-columns: repeat(4, 1fr);
  gap: 10px; margin: 2em 0 2.5em;
}
.pipeline-card {
  background: var(--white); border: 1px solid var(--parchment);
  border-radius: 6px; padding: 14px 16px; box-shadow: var(--card-shadow);
  border-top: 3px solid var(--fern);
}
.pipeline-card.part2 { border-top-color: var(--amber); }
.pipeline-card.part3 { border-top-color: #2e7d94; }
.pipeline-card.part4 { border-top-color: #5d4037; }
.pipeline-card.part5 { border-top-color: #6a1b9a; }
.pc-num  { font-family: 'Source Code Pro', monospace; font-size: .68rem;
           font-weight: 700; letter-spacing: .14em; text-transform: uppercase;
           color: var(--dust); margin-bottom: 4px; }
.pc-name { font-family: 'Libre Baskerville', serif; font-size: .95rem;
           font-weight: 700; color: var(--forest); margin-bottom: 3px; }
.pc-desc { font-size: .82rem; color: var(--charcoal); line-height: 1.45; }

/* ── Step badges ─────────────────────────────────────────────────────────── */
.step-badge {
  display: inline-block; background: var(--forest); color: #fff;
  font-family: 'Source Code Pro', monospace; font-size: .68rem; font-weight: 700;
  letter-spacing: .1em; text-transform: uppercase;
  padding: 3px 10px; border-radius: 3px; margin-right: 8px; vertical-align: middle;
}

/* ── Key box ─────────────────────────────────────────────────────────────── */
.key-box {
  background: var(--forest); color: #fff; border-radius: 6px;
  padding: 20px 24px; margin: 2em 0; font-size: .9rem;
}
.key-box h4 {
  font-family: 'Libre Baskerville', serif; color: var(--sage);
  font-size: 1rem; margin-bottom: 10px;
}
.key-box ul { padding-left: 1.3em; }
.key-box li { margin-bottom: 5px; color: rgba(255,255,255,.85); }
.key-box code { background: rgba(255,255,255,.12); color: var(--amber-lt); }

/* ── Discussion box ──────────────────────────────────────────────────────── */
.discuss-box {
  background: var(--parchment); border: 1px solid var(--mist);
  border-radius: 6px; padding: 18px 22px; margin: 2em 0; font-size: .92em;
}
.discuss-box h4 {
  font-family: 'Libre Baskerville', serif; color: var(--moss);
  font-size: 1rem; margin-bottom: 10px;
}
.discuss-box ol { padding-left: 1.4em; }
.discuss-box li { margin-bottom: .5em; }

/* ── Tables ──────────────────────────────────────────────────────────────── */
table { font-size: .9rem; border-collapse: collapse; width: 100%; margin: 1.2em 0; }
thead th {
  background: var(--forest); color: #fff;
  font-family: 'Source Code Pro', monospace; font-size: .76rem;
  letter-spacing: .06em; text-transform: uppercase; padding: 8px 12px; text-align: left;
}
tbody tr:nth-child(even) { background: var(--mist); }
tbody td { padding: 7px 12px; border-bottom: 1px solid var(--parchment); }

/* ── Footer ──────────────────────────────────────────────────────────────── */
.page-footer {
  background: var(--ink); color: rgba(255,255,255,.42); text-align: center;
  padding: 24px 4vw; margin: 60px -4vw 0;
  font-size: .76rem; font-family: 'Source Code Pro', monospace; letter-spacing: .04em;
}
.page-footer a { color: var(--sage); text-decoration: none; }
.page-footer .divider { margin: 0 8px; opacity: .3; }
```

<div class="page-header">
<div class="module-tag">EA030 &mdash; RTricks &mdash; Project 05</div>
<h1>Camera Trap Analysis: <span>Wildlife Detection and Behavior</span></h1>
<p class="sub"><strong>Prof. Marc Los Huertos</strong> &nbsp;&bull;&nbsp; Pomona College &nbsp;&bull;&nbsp; Environmental Analysis</p>
<a class="back-link" href="../index.html">&#8592; Back to Index</a>
</div>

<div class="section-nav">
<a href="#intro">Introduction</a>
<a href="#setup">Setup</a>
<a href="#import">Import & Clean</a>
<a href="#events">Detection Events</a>
<a href="#descriptive">Descriptive</a>
<a href="#behavior-diel">Behavior × Diel</a>
<a href="#overlap">Activity Overlap</a>
<a href="#diversity">Diversity</a>
<a href="#occupancy">Occupancy</a>
<a href="#summary">Summary</a>
</div>

> **What this project covers.** Camera traps record *who* was present, *when*,
> and *what they were doing* without a researcher present. This tutorial walks
> the complete analytical pipeline — from raw Excel export through cleaning,
> independent-event definition, descriptive ethograms, and four quantitative
> methods: chi-square/Fisher tests of diel behavior shifts, kernel-density
> activity overlap, rarefaction-corrected diversity, and single-season
> occupancy modeling.

---

## The Analytical Pipeline {#intro}

Camera-trap data require several decisions before any statistics can be run.
The diagram below maps the full workflow covered in this project.

<div class="pipeline-grid">
<div class="pipeline-card">
  <div class="pc-num">Stage 1</div>
  <div class="pc-name">Import & Clean</div>
  <div class="pc-desc">Read Excel, standardize columns, remove ambiguous taxa, collapse behavior codes.</div>
</div>
<div class="pipeline-card">
  <div class="pc-num">Stage 2</div>
  <div class="pc-name">Independent Events</div>
  <div class="pc-desc">Apply the 30-minute rule to collapse burst detections into single events. Prevents pseudo-replication.</div>
</div>
<div class="pipeline-card part2">
  <div class="pc-num">Stage 3</div>
  <div class="pc-name">Descriptive Stats</div>
  <div class="pc-desc">Ethogram, hourly heat map, camera-level activity profiles.</div>
</div>
<div class="pipeline-card part3">
  <div class="pc-num">Stage 4</div>
  <div class="pc-name">Hypothesis Testing</div>
  <div class="pc-desc">Chi-square / Fisher: does behavior composition differ between day and night?</div>
</div>
<div class="pipeline-card part4">
  <div class="pc-num">Stage 5</div>
  <div class="pc-name">Activity Overlap</div>
  <div class="pc-desc">Kernel-density overlap coefficient Δ quantifies temporal co-occurrence between species pairs.</div>
</div>
<div class="pipeline-card part5">
  <div class="pc-num">Stage 6</div>
  <div class="pc-name">Occupancy Model</div>
  <div class="pc-desc">MacKenzie et al. (2002) separates true absence from non-detection to estimate ψ and p.</div>
</div>
</div>

<div class="callout callout-new">
<strong>🆕 New to R?</strong>
Don't worry if every line of code is not immediately clear. Each section
explains <em>what</em> the code does and <em>why</em> before showing how to
run it. Focus on understanding the logic; the syntax will become familiar
with practice. Sections marked with ⚙️ go deeper into the statistical
machinery — they are optional on a first reading.
</div>

---

## 1. Setup {#setup}

### Packages

R's base functionality is extended by **packages** — collections of functions
written by the community. Install each package once with `install.packages()`,
then load it every session with `library()`.

```{r packages}
# Run the install lines once — remove the # to uncomment:
# install.packages(c("tidyverse", "lubridate", "janitor", "readxl",
#                    "vegan", "overlap", "unmarked", "RColorBrewer"))

library(tidyverse)   # data wrangling + ggplot2
library(lubridate)   # date-time helpers
library(janitor)     # clean_names() standardises column headers
library(readxl)      # read Excel files
library(vegan)       # diversity indices and rarefaction
library(overlap)     # kernel-density activity-overlap (Ridout & Linkie 2009)
library(unmarked)    # occupancy models (MacKenzie et al. 2002)
```

<div class="callout callout-formula">
<strong>⚙️ Why these packages?</strong><br>
<code>tidyverse</code> — a suite of packages sharing a consistent grammar
for data manipulation (<code>dplyr</code>, <code>tidyr</code>) and
visualisation (<code>ggplot2</code>).<br>
<code>overlap</code> — implements the Δ estimator of Ridout & Linkie (2009)
for circular (time-of-day) kernel densities.<br>
<code>unmarked</code> — fits hierarchical models that separate ecological
state (occupancy) from the observation process (detection probability).
</div>

---

## 2. Data Import and Cleaning {#import}

### <span class="step-badge">Step 1</span> Load the data

`read_excel()` reads a named sheet from an `.xlsx` file into a **data frame** —
R's standard rectangular data structure (rows = observations, columns =
variables). Update `file_path` to match your own file location.

```{r load-data}
file_path <- "/home/mwl04747/RTricks/15_Camera_Traps/Camera Trap Data Entry.xlsx"

cam_raw <- read_excel(path = file_path, sheet = "Combined")

glimpse(cam_raw)
```

<div class="callout callout-tip">
<strong>💡 `glimpse()` is your first-look friend.</strong>
It shows every column name, its data type (<code>&lt;chr&gt;</code> = text,
<code>&lt;dbl&gt;</code> = number, <code>&lt;dttm&gt;</code> = date-time),
and the first few values — far more useful than <code>head()</code> for wide
data frames.
</div>

### <span class="step-badge">Step 2</span> Standardize column names and parse dates

Column names from Excel often contain spaces and mixed capitalisation.
`clean_names()` converts them to `snake_case` (lower case, underscores instead
of spaces) so they are easier to type and less error-prone.

```{r clean}
cam_raw$datetime <- cam_raw$Datetime_corrected

cam <- cam_raw %>%
  clean_names() %>%
  mutate(
    date = as.Date(datetime),   # strip the time component; keep date only
    hour = hour(datetime)       # extract hour of day as integer (0–23)
  )

summary(cam$date)
```

### <span class="step-badge">Step 3</span> Remove ambiguous taxa

Some records contain notes like `"bird"`, `"unknown"`, or combined entries
(`"squirrel/rat"`). These are genuinely unidentifiable and must be excluded
before analysis — their presence would inflate richness estimates and
contaminate behavior summaries.

```{r filter-taxa}
ambiguous_patterns <- c(
  " and ", " or ", "/",
  "unknown", "unidentified", "n/a", "^na$",
  "bird$", "small brown bird"
)

cam_clean <- cam %>%
  filter(
    !str_detect(
      str_to_lower(taxa_code),
      paste(ambiguous_patterns, collapse = "|")
    )
  )

cat("Taxa retained after filtering:\n")
unique(cam_clean$taxa_code)
```

<div class="callout callout-concept">
<strong>📌 Regular expressions</strong>

<code>str_detect()</code> matches text against a **pattern**.
<code>paste(..., collapse = "|")</code> joins our list of ambiguous strings
with the "or" operator (<code>|</code>), so one call checks all of them
simultaneously. The <code>!</code> at the front keeps only rows that do
<em>not</em> match any ambiguous pattern.
</div>

### <span class="step-badge">Step 4</span> Collapse behaviors into functional types

Raw behavior codes (`"walk"`, `"run"`, `"hop"`) are too granular for most
analyses. `case_when()` groups them into five **functional categories** — think
of it as a multi-way if/else evaluated in order.

```{r behaviors}
cam <- cam_clean %>%
  mutate(
    behavior_function = case_when(
      str_detect(behavior_code, "walk|run|hop|fly|climb|crawl") ~ "movement",
      str_detect(behavior_code, "eat|dig|cache")                ~ "foraging",
      str_detect(behavior_code, "sniff")                        ~ "investigation",
      str_detect(behavior_code, "stare")                        ~ "vigilance",
      str_detect(behavior_code, "stop|stand|sit|lay")           ~ "stationary",
      TRUE                                                      ~ "other"
    )
  )

table(cam$behavior_function)
```

---

## 3. Defining Independent Detection Events {#events}

### The pseudo-replication problem

A single animal may trigger the same camera dozens of times over a few minutes.
Treating each trigger as an independent observation would **pseudo-replicate**
the data — inflating sample sizes and biasing every statistic downstream.

### The 30-minute rule

The standard solution: any detections of the **same species at the same
camera** within 30 minutes of the previous detection are treated as part of
the *same* event. Only a detection after a ≥ 30-minute gap counts as a new
independent event.

```{r events}
cam_events <- cam %>%
  arrange(camera_id, taxa_code, datetime) %>%
  group_by(camera_id, taxa_code) %>%
  mutate(
    # Time gap in minutes between consecutive rows within each camera × species group
    minutes_since_last = as.numeric(
      difftime(datetime, lag(datetime), units = "mins")
    ),
    # A new event begins if gap > 30 min OR this is the very first record
    new_event = if_else(
      is.na(minutes_since_last) | minutes_since_last > 30, 1L, 0L
    ),
    # Cumulative sum gives a running event counter within each group
    event_id = cumsum(new_event)
  ) %>%
  ungroup() %>%
  mutate(diel = if_else(hour >= 6 & hour < 18, "day", "night"))

# How many independent events per species?
cam_events %>%
  distinct(camera_id, taxa_code, event_id) %>%
  count(taxa_code, name = "independent_events") %>%
  arrange(desc(independent_events))
```

<div class="callout callout-concept">
<strong>📌 Why `lag()` is the right tool here</strong>

<code>lag(datetime)</code> returns the value from the <em>previous row</em>.
Because we first <code>group_by(camera_id, taxa_code)</code>, "previous"
means the last detection of that species at that specific camera — exactly
what we need to measure within-burst gaps. Without grouping,
<code>lag()</code> would compare across species and cameras, giving
meaningless gaps.
</div>

<div class="callout callout-warn">
<strong>⚠️ Choosing a time threshold</strong>

30 minutes is a widely used convention, but the right threshold depends on
your study organism's typical bout length and movement speed. For a chipmunk
foraging in a small area, 30 minutes may be too long. For a wide-ranging
carnivore like a mountain lion, 60–120 minutes may be more appropriate.
Whatever threshold you choose, document it clearly in your methods.
</div>

---

## 4. Descriptive Statistics {#descriptive}

### 4.1 Detection event totals per taxon

```{r detection-summary}
event_summary <- cam_events %>%
  distinct(camera_id, taxa_code, event_id) %>%
  count(taxa_code, name = "detection_events") %>%
  arrange(desc(detection_events))

print(event_summary)

ggplot(event_summary,
       aes(x = reorder(taxa_code, detection_events),
           y = detection_events,
           fill = reorder(taxa_code, detection_events))) +
  geom_col(color = "#1e3a2f", alpha = 0.88) +
  geom_text(aes(label = detection_events), hjust = -0.2,
            family = "mono", size = 3.4, color = "#1e3a2f") +
  coord_flip() +
  scale_fill_manual(
    values = colorRampPalette(c("#d4e6da", "#1e3a2f"))(nrow(event_summary))
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(title = "Independent detection events by taxon",
       x = NULL, y = "Detection events") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title      = element_text(family = "serif", face = "bold", color = "#1e3a2f"),
    legend.position = "none",
    panel.grid.major.y = element_blank()
  )
```

### 4.2 Behavioral ethogram (stacked bar)

An **ethogram** documents the relative frequency of each behavior.
Proportional bars enable cross-species comparison even when total event counts
differ substantially between species.

```{r ethogram}
ethogram <- cam_events %>%
  count(taxa_code, behavior_function) %>%
  group_by(taxa_code) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

ggplot(ethogram, aes(x = taxa_code, y = prop, fill = behavior_function)) +
  geom_col(color = "white", linewidth = 0.3) +
  scale_fill_manual(
    values = c(
      "movement"      = "#4a7c5e",
      "foraging"      = "#c8832a",
      "investigation" = "#2e7d94",
      "vigilance"     = "#8aad96",
      "stationary"    = "#d4e6da",
      "other"         = "#7a7a72"
    )
  ) +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title   = element_text(family = "serif", face = "bold", color = "#1e3a2f"),
    axis.text.x  = element_text(angle = 40, hjust = 1),
    legend.title = element_blank()
  ) +
  labs(title = "Behavioral composition by taxon",
       x = NULL, y = "Proportion of observations")
```

### 4.3 Hourly activity heat map

A **heat map** encodes counts as color intensity. Each cell is a
taxon × hour combination — a fast way to spot nocturnal, diurnal, or
crepuscular patterns across all taxa simultaneously.

```{r heatmap}
taxon_hour <- cam_events %>%
  distinct(taxa_code, event_id, hour) %>%
  count(taxa_code, hour, name = "events")

ggplot(taxon_hour, aes(x = hour, y = taxa_code, fill = events)) +
  geom_tile(color = "white", linewidth = 0.4) +
  scale_fill_gradientn(
    colors = c(var_cream = "#f7f3ed", "#d4e6da", "#4a7c5e", "#1e3a2f"),
    name    = "Events"
  ) +
  scale_x_continuous(breaks = seq(0, 23, 3),
                     labels = paste0(seq(0, 23, 3), ":00")) +
  labs(title = "Hourly detection patterns by taxon",
       x = "Hour of day", y = NULL) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title  = element_text(family = "serif", face = "bold", color = "#1e3a2f"),
    axis.text.x = element_text(size = 9)
  )
```

### 4.4 Camera-level activity profiles

Different cameras are placed in different micro-habitats. This plot reveals
whether behavioral composition varies by camera location — an early signal
of habitat-driven behavior differences.

```{r camera-profiles}
camera_behavior <- cam_events %>%
  distinct(camera_id, taxa_code, event_id, behavior_function) %>%
  count(camera_id, behavior_function, name = "events")

ggplot(camera_behavior,
       aes(x = camera_id, y = events, fill = behavior_function)) +
  geom_col(color = "white", linewidth = 0.3) +
  scale_fill_manual(
    values = c(
      "movement"      = "#4a7c5e",
      "foraging"      = "#c8832a",
      "investigation" = "#2e7d94",
      "vigilance"     = "#8aad96",
      "stationary"    = "#d4e6da",
      "other"         = "#7a7a72"
    )
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title  = element_text(family = "serif", face = "bold", color = "#1e3a2f"),
    axis.text.x = element_text(angle = 40, hjust = 1),
    legend.title = element_blank()
  ) +
  labs(title = "Camera activity profiles by behavior function",
       x = "Camera ID", y = "Detection events")
```

---

## 5. Hypothesis Testing: Behavior × Diel Period {#behavior-diel}

### Rationale

**Question:** Does the *proportion* of behaviors (e.g., foraging vs.
vigilance) differ between day and night?

We use a **chi-square test of independence** — or Fisher's exact test when
expected cell counts are small — to test whether behavior type is statistically
independent of diel period.

**H₀:** Behavior composition is the same during the day and at night.<br>
**H₁:** At least one behavior type is more common at one time of day.

<div class="callout callout-formula">
<strong>⚙️ Chi-square vs. Fisher's exact test</strong><br>
The chi-square test assumes all <em>expected</em> cell counts ≥ 5. With
small samples or rare behaviors, this fails. Fisher's exact test has no
such requirement — it computes an exact p-value by enumerating all possible
contingency tables with the same marginal totals. Use Fisher when any
expected count is below 5. The code below checks automatically.
</div>

### Run tests per taxon

```{r chi-fisher}
test_behavior_diel <- function(df) {

  # 1. Drop rows where diel is NA (records with hour == NA or bad datetime)
  df <- df %>% filter(!is.na(diel), !is.na(behavior_function))

  # 2. Guard: need observations in BOTH diel periods
  if (length(unique(df$diel)) < 2)
    return(tibble(test = "insufficient data — one diel period only",
                  stat_type = "—", statistic = NA_real_, p_value = NA_real_))

  # 3. Build the contingency table
  wide <- df %>%
    count(diel, behavior_function) %>%
    pivot_wider(names_from  = behavior_function,
                values_from = n,
                values_fill = 0)

  # 4. Guard: diel column must have no NAs before using as row names
  wide <- wide %>% filter(!is.na(diel))
  if (nrow(wide) < 2)
    return(tibble(test = "insufficient data — pivot produced < 2 rows",
                  stat_type = "—", statistic = NA_real_, p_value = NA_real_))

  tbl <- wide %>%
    column_to_rownames("diel") %>%
    as.matrix()

  # 5. Guard: need at least 2 behavior columns
  if (ncol(tbl) < 2)
    return(tibble(test = "insufficient data — fewer than 2 behavior types",
                  stat_type = "—", statistic = NA_real_, p_value = NA_real_))

  # 6. Choose chi-square vs. Fisher based on minimum expected count
  exp_counts <- suppressWarnings(chisq.test(tbl)$expected)
  if (any(exp_counts < 5)) {
    # Fisher's exact test has no single test statistic (no chi-square value).
    # For 2x2 tables it returns an odds ratio; for larger tables use NA.
    res <- fisher.test(tbl, simulate.p.value = TRUE, B = 2000)
    or_val <- if (!is.null(res$estimate)) round(as.numeric(res$estimate), 3)
              else NA_real_
    tibble(test      = "Fisher",
           statistic = or_val,      # odds ratio when 2x2; NA for larger tables
           stat_type = if (!is.na(or_val)) "odds ratio" else "—",
           p_value   = res$p.value)
  } else {
    res <- chisq.test(tbl)
    tibble(test      = "Chi-square",
           statistic = round(as.numeric(res$statistic), 3),
           stat_type = "X²",
           p_value   = res$p.value)
  }
}

behavior_tests <- cam_events %>%
  group_by(taxa_code) %>%
  group_modify(~ test_behavior_diel(.x)) %>%
  ungroup() %>%
  mutate(
    significant = ifelse(is.na(p_value), NA, p_value < 0.05),
    p_value     = round(p_value, 4)
  )

# Re-order columns for a readable table
behavior_tests <- behavior_tests %>%
  select(taxa_code, test, stat_type, statistic, p_value, significant)

knitr::kable(
  behavior_tests,
  col.names = c("Taxon", "Test used", "Statistic type",
                "Value", "p-value", "p < 0.05?"),
  caption   = paste0(
    "Diel behavior tests by taxon. ",
    "For chi-square the value is X²; ",
    "for Fisher it is the odds ratio (2×2 tables only; — for larger tables). ",
    "Fisher\'s exact test has no chi-square statistic — ",
    "NA in earlier versions was correct but misleadingly labeled."
  )
)
```

### Visualize the diel behavior shift

```{r diel-behavior-plot, fig.height=6}
behavior_diel <- cam_events %>%
  # Remove records with no diel assignment or unclassified behavior
  filter(!is.na(diel), !is.na(behavior_function)) %>%
  count(taxa_code, diel, behavior_function) %>%
  group_by(taxa_code, diel) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup() %>%
  # Convert diel to a factor with clean labels so NA never appears as a panel
  mutate(
    diel = factor(diel, levels = c("day", "night"),
                  labels = c("Day", "Night"))
  )

# Only keep taxa that have observations in BOTH diel periods
both_periods <- behavior_diel %>%
  distinct(taxa_code, diel) %>%
  count(taxa_code) %>%
  filter(n == 2) %>%
  pull(taxa_code)

behavior_diel <- behavior_diel %>% filter(taxa_code %in% both_periods)

ggplot(behavior_diel,
       aes(x = diel, y = prop, fill = behavior_function)) +
  geom_col(color = "white", linewidth = 0.3) +
  facet_wrap(~ taxa_code, ncol = 3) +
  scale_fill_manual(
    values = c(
      "movement"      = "#4a7c5e",
      "foraging"      = "#c8832a",
      "investigation" = "#2e7d94",
      "vigilance"     = "#8aad96",
      "stationary"    = "#d4e6da",
      "other"         = "#7a7a72"
    )
  ) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Behavioral composition: day vs. night",
       x = NULL, y = "Proportion", fill = "Behavior") +
  theme_minimal(base_size = 11) +
  theme(
    plot.title   = element_text(family = "serif", face = "bold", color = "#1e3a2f"),
    strip.text   = element_text(family = "mono",  face = "bold", color = "#2d5a3d"),
    legend.title = element_blank(),
    panel.spacing = unit(0.8, "lines")
  )
```

<div class="callout callout-tip">
<strong>💡 Interpreting a significant test</strong>
A significant p-value tells you <em>that</em> the behavior mix differs
between day and night, but not <em>which</em> behaviors drive the difference.
Use the faceted stacked bars above to identify which functional types shift
most. The chi-square **standardized residuals** (available in the test object
as <code>chisq.test(tbl)$stdres</code>) also pinpoint which cells deviate
most from independence.
</div>

---

## 6. Activity Overlap (Kernel Density) {#overlap}

### Rationale

Two species active at the same hours are more likely to interact — as
competitors, predators, or prey. The **overlap coefficient Δ** (Ridout &
Linkie 2009) quantifies the similarity of two diel activity curves on a 0–1
scale (0 = no temporal overlap, 1 = identical schedules).

The `overlap` package fits smooth **kernel density** curves to detection
times treated as a circular variable — because 23:00 and 01:00 are only
2 hours apart, not 22.

### Why radians? Putting time on a circle

A standard 24-hour clock has an awkward edge: 23:45 and 00:15 are only
30 minutes apart in reality, but subtracting their decimal hours gives
23.5 — a huge apparent gap. The kernel density estimator would treat those
two detections as nearly opposite ends of the day.

The solution is to **map the 24-hour cycle onto a circle** where midnight
wraps seamlessly back to midnight. We do this by converting each detection
time to **radians** (0 to 2π), the natural unit for circular data.

The conversion has three steps:

1. **Decimal hours** — combine the hour and fractional minutes into a single
   number: `hour + minute/60`. A detection at 14:45 becomes 14.75.
2. **Fraction of the day** — divide by 24 to get a value between 0 and 1.
   14.75 / 24 = 0.615.
3. **Scale to a full circle** — multiply by 2π (≈ 6.283).
   0.615 × 2π ≈ 3.86 radians.

<div class="callout callout-formula">
<strong>⚙️ The full formula</strong><br>
<code>time_rad = (hour + minute/60) / 24 × 2π</code><br><br>
Key reference points on the circle:<br>
midnight (00:00) = <strong>0</strong> rad &nbsp;|&nbsp;
6 am = <strong>π/2</strong> ≈ 1.57 rad &nbsp;|&nbsp;
noon (12:00) = <strong>π</strong> ≈ 3.14 rad &nbsp;|&nbsp;
6 pm = <strong>3π/2</strong> ≈ 4.71 rad &nbsp;|&nbsp;
midnight (24:00) = <strong>2π</strong> ≈ 6.28 rad = 0 again<br><br>
Because 2π and 0 are the <em>same point</em> on the circle, a detection at
23:45 (≈ 6.25 rad) and one at 00:15 (≈ 0.13 rad) are only 0.13 rad apart —
correctly close — rather than 6.12 rad apart as a linear scale would report.
</div>

```{r radians}
cam_events <- cam_events %>%
  mutate(
    # Step 1: decimal hours  →  Step 2: fraction of day  →  Step 3: radians
    time_rad = (hour(datetime) + minute(datetime) / 60) / 24 * 2 * pi
  )

# Sanity check: values should span 0 to ~2π (6.28)
cat(sprintf("time_rad range: %.3f to %.3f  (expected 0 to %.3f)\n",
            min(cam_events$time_rad, na.rm = TRUE),
            max(cam_events$time_rad, na.rm = TRUE),
            2 * pi))
```

### Single pair: estimate Δ and bootstrap CI

```{r activity-overlap, fig.height=5}
# ── Minimum detections needed for bandwidth estimation ────────────────────────
MIN_N_OVERLAP <- 10   # raise further if bandwidth errors persist

# Helper: attempt overlap analysis, return NULL gracefully on failure
run_overlap <- function(ta, tb, sp1, sp2) {
  if (length(ta) < MIN_N_OVERLAP || length(tb) < MIN_N_OVERLAP) {
    message(sprintf("Skipping %s vs %s: too few detections (%d, %d).",
                    sp1, sp2, length(ta), length(tb)))
    return(NULL)
  }
  if (length(unique(ta)) < 5 || length(unique(tb)) < 5) {
    message(sprintf("Skipping %s vs %s: too few unique time values.", sp1, sp2))
    return(NULL)
  }
  estimator <- if (min(length(ta), length(tb)) < 75) "Dhat4" else "Dhat1"
  tryCatch({
    delta_val <- overlapEst(ta, tb, type = estimator)[estimator]
    set.seed(42)
    bs_val <- bootstrap(ta, tb, nb = 999, type = estimator)
    ci_val <- bootCI(delta_val, bs_val)
    list(estimator = estimator, delta = delta_val,
         ci_lower  = ci_val["norm0", "lower"],
         ci_upper  = ci_val["norm0", "upper"])
  }, error = function(e) {
    message(sprintf("Overlap failed for %s vs %s: %s", sp1, sp2, conditionMessage(e)))
    NULL
  })
}

# ── Pick the two species with the most independent detections ─────────────────
top2 <- cam_events %>%
  distinct(taxa_code, event_id) %>%
  count(taxa_code, sort = TRUE) %>%
  slice_head(n = 2) %>%
  pull(taxa_code)

sp1_code <- top2[1]
sp2_code <- top2[2]

t1 <- cam_events %>% filter(taxa_code == sp1_code) %>% pull(time_rad)
t2 <- cam_events %>% filter(taxa_code == sp2_code) %>% pull(time_rad)

ov <- run_overlap(t1, t2, sp1_code, sp2_code)

if (!is.null(ov)) {
  cat(sprintf("Activity overlap (%s):  %s vs. %s  \u2192  \u0394 = %.3f\n",
              ov$estimator, sp1_code, sp2_code, ov$delta))
  cat(sprintf("Bootstrap 95%% CI: [%.3f, %.3f]\n", ov$ci_lower, ov$ci_upper))
  overlapPlot(
    t1, t2,
    rug        = TRUE,
    main       = paste("Activity overlap:", sp1_code, "vs.", sp2_code),
    xlab       = "Time of day",
    linecol    = "#4a7c5e",
    linecol2   = "#c8832a",
    linewidth  = 2,
    linewidth2 = 2
  )
  legend("topright", legend = c(sp1_code, sp2_code),
         col = c("#4a7c5e", "#c8832a"), lwd = 2, bty = "n")
} else {
  # Fallback: plain hourly histograms when KDE cannot run
  par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
  hist(t1 / (2*pi) * 24, breaks = 24, col = adjustcolor("#4a7c5e", 0.7),
       border = "white", xlab = "Hour of day", main = sp1_code, xlim = c(0,24))
  hist(t2 / (2*pi) * 24, breaks = 24, col = adjustcolor("#c8832a", 0.7),
       border = "white", xlab = "Hour of day", main = sp2_code, xlim = c(0,24))
  par(mfrow = c(1, 1))
  message("KDE overlap unavailable — showing hourly histograms instead.")
}
```

<div class="callout callout-concept">
<strong>📌 Interpreting Δ</strong>

Δ &gt; 0.75 = high overlap; 0.50–0.75 = moderate; &lt; 0.50 = low.
If the bootstrap CI excludes a reference value of interest (e.g., Δ = 0.5),
the difference is statistically supported. See Ridout & Linkie (2009) for
formal guidance on estimator choice and CI interpretation.
</div>

### All pairwise overlaps (heat map)

```{r pairwise-overlap, fig.height=6}
taxa_list <- unique(cam_events$taxa_code)
n_sp      <- length(taxa_list)

time_by_sp <- setNames(
  lapply(taxa_list, function(sp)
    cam_events %>% filter(taxa_code == sp) %>% pull(time_rad)),
  taxa_list
)

# Helper: safe single-pair overlap (returns NA on any failure)
safe_overlap <- function(ta, tb) {
  if (length(ta) < MIN_N_OVERLAP || length(tb) < MIN_N_OVERLAP) return(NA_real_)
  if (length(unique(ta)) < 5    || length(unique(tb)) < 5)    return(NA_real_)
  est <- if (min(length(ta), length(tb)) < 75) "Dhat4" else "Dhat1"
  tryCatch(
    overlapEst(ta, tb, type = est)[est],
    error = function(e) NA_real_
  )
}

# Fill the symmetric overlap matrix, computing each pair only once
overlap_mat <- matrix(NA, n_sp, n_sp,
                      dimnames = list(taxa_list, taxa_list))
for (i in seq_len(n_sp)) {
  for (j in seq_len(n_sp)) {
    if (i == j) {
      overlap_mat[i, j] <- 1
    } else if (!is.na(overlap_mat[j, i])) {
      overlap_mat[i, j] <- overlap_mat[j, i]   # mirror already computed
    } else {
      overlap_mat[i, j] <- safe_overlap(time_by_sp[[i]], time_by_sp[[j]])
    }
  }
}

overlap_df <- as.data.frame(overlap_mat) %>%
  rownames_to_column("sp1") %>%
  pivot_longer(-sp1, names_to = "sp2", values_to = "delta")

ggplot(overlap_df, aes(x = sp1, y = sp2, fill = delta)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = round(delta, 2)), size = 3,
            color = ifelse(overlap_df$delta > 0.6, "white", "#1a1a18")) +
  scale_fill_gradientn(
    colors = c("#f7f3ed", "#d4e6da", "#4a7c5e", "#1e3a2f"),
    limits  = c(0, 1), name = "Δ"
  ) +
  labs(title = "Pairwise activity overlap (Δ)", x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title  = element_text(family = "serif", face = "bold", color = "#1e3a2f"),
    axis.text.x = element_text(angle = 40, hjust = 1)
  )
```

---

## 7. Species Diversity and Rarefaction {#diversity}

### Rationale

Some cameras may be in richer habitats and detect more species. But cameras
also vary in **effort** — a busier camera accumulates more detections by
chance, inflating apparent diversity. **Rarefaction** corrects for this by
asking: *how many species would we expect if all cameras had the same total
event count?*

### Build the site × species matrix

```{r site-sp-matrix}
# One row per camera, one column per taxon, cell = independent event count
site_sp <- cam_events %>%
  distinct(camera_id, taxa_code, event_id) %>%
  count(camera_id, taxa_code, name = "events") %>%
  pivot_wider(names_from  = taxa_code,
              values_from = events,
              values_fill = 0) %>%
  column_to_rownames("camera_id")

print(site_sp)
```

### Diversity indices

```{r diversity}
div_df <- tibble(
  camera_id    = rownames(site_sp),
  richness     = specnumber(site_sp),
  shannon      = diversity(site_sp, "shannon"),
  simpson      = diversity(site_sp, "simpson"),
  total_events = rowSums(site_sp)
) %>%
  arrange(desc(shannon))

knitr::kable(div_df,
             digits  = 3,
             col.names = c("Camera", "Richness (S)", "Shannon (H′)",
                           "Simpson (1−D)", "Total events"),
             caption = "Diversity indices by camera. Rarefied richness (below) controls for effort differences.")

div_df %>%
  pivot_longer(c(richness, shannon, simpson),
               names_to = "index", values_to = "value") %>%
  ggplot(aes(x = camera_id, y = value, fill = camera_id)) +
  geom_col(color = "#1e3a2f", alpha = 0.85, show.legend = FALSE) +
  facet_wrap(~ index, scales = "free_y",
             labeller = labeller(index = c(
               richness = "Richness (S)",
               shannon  = "Shannon (H′)",
               simpson  = "Simpson (1−D)"
             ))) +
  scale_fill_manual(
    values = colorRampPalette(c("#d4e6da", "#1e3a2f"))(nrow(div_df))
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title  = element_text(family = "serif", face = "bold", color = "#1e3a2f"),
    axis.text.x = element_text(angle = 40, hjust = 1),
    strip.text  = element_text(family = "mono", face = "bold", color = "#2d5a3d")
  ) +
  labs(title = "Diversity indices by camera", x = "Camera ID", y = "Index value")
```

<div class="callout callout-concept">
<strong>📌 Which diversity index to report?</strong>

**Richness (S)** — counts species, ignores abundance; sensitive to sampling effort.<br>
**Shannon (H′)** — weights common and rare species equally on a log scale; the
most widely reported index in ecology.<br>
**Simpson (1−D)** — weighted toward dominant species; less sensitive to rare taxa.<br><br>
Report all three and discuss what differences reveal about community structure.
High S with low H′ indicates one highly dominant taxon with many rare ones.
</div>

### Rarefaction curves

**The problem rarefaction solves.** Camera A recorded 80 independent events
and detected 6 species. Camera B recorded 12 events and detected 4 species.
Is Camera A genuinely richer, or did it simply run longer and accumulate more
detections by chance? Raw species counts cannot answer this — they confound
ecological richness with sampling effort.

**What a rarefaction curve does.** For each camera, the `vegan` function
`rarecurve()` randomly sub-samples its detection records 1, 2, 3 … n times
and asks: *on average, how many species had we seen after exactly k detections?*
The result is a curve of *expected* richness as a function of effort. Plotting
all cameras on the same axes — with the same x-scale — makes a fair comparison
possible: read off each curve at a common x-value (the **rarefaction target**,
set to the smallest total across all cameras) to get effort-controlled richness.

**How to read the plot.** Each coloured line is one camera. The vertical dashed
line marks the rarefaction target — the point where all cameras can be compared
fairly. At that x-value, a higher curve means genuinely greater local richness,
not just more deployment time.

- **Steep and still rising at the target** → the camera is under-sampled;
  more events would likely reveal additional species.
- **Levelling off (asymptote) before the target** → the local species pool
  is well-characterized; additional effort would add little.
- **Short curves that barely reach the target** → those cameras had very few
  total events and their richness estimates are highly uncertain.

```{r rarefaction, fig.height=6}
min_n <- min(rowSums(site_sp))
cat(sprintf(
  "Rarefaction target: %d detections (the minimum across %d cameras)\n",
  min_n, nrow(site_sp)
))
cat(sprintf(
  "Cameras with more than 2× the target: %d\n",
  sum(rowSums(site_sp) > 2 * min_n)
))

# Build a color palette with one color per camera
cam_cols <- colorRampPalette(
  RColorBrewer::brewer.pal(8, "Dark2")
)(nrow(site_sp))

rarecurve(
  site_sp,
  step    = 1,
  col     = cam_cols,
  lwd     = 1.8,
  xlab    = "Detection events (sampling effort)",
  ylab    = "Expected species richness",
  main    = "Rarefaction curves by camera",
  label   = TRUE,
  cex     = 0.75
)
abline(v = min_n, lty = 2, col = "#7a7a72", lwd = 1.5)
text(min_n, par("usr")[3] + 0.3,
     paste0("Rarefaction target = ", min_n, " events"),
     adj = c(-0.08, 0), cex = 0.8, col = "#7a7a72")
```

```{r rarefaction-table}
# Effort-controlled richness at the rarefaction target for each camera
rare_vals <- rarefy(site_sp, sample = min_n)

rarefied_df <- tibble(
  Camera            = names(rare_vals),
  `Raw richness`    = specnumber(site_sp),
  `Total events`    = rowSums(site_sp),
  `Rarefied S`      = round(rare_vals, 2),
  `Rank (raw)`      = rank(-specnumber(site_sp), ties.method = "min"),
  `Rank (rarefied)` = rank(-rare_vals, ties.method = "min")
) %>% arrange(`Rank (rarefied)`)

knitr::kable(
  rarefied_df,
  caption = paste0(
    "Effort-controlled species richness rarefied to ", min_n,
    " detection events per camera. ",
    "Cameras whose rarefied rank differs from their raw rank had ",
    "apparent richness inflated (or deflated) by effort differences."
  ),
  align = "lrrrrr"
)
```

<div class="callout callout-tip">
<strong>💡 Rarefied S vs. raw richness — when do they disagree?</strong>

If a camera's raw rank is higher than its rarefied rank, its raw species count
was inflated by extra effort — it detected more species simply because it ran
longer, not because its habitat is richer. The rarefied value is the fairer
number to report in any cross-camera comparison. Note that rarefaction always
<em>reduces</em> or maintains richness relative to the raw count — it can never
add species that were not observed.
</div>

---

## 8. Single-Season Occupancy Modeling {#occupancy}

### Rationale and key insight

**Which cameras does species X truly occupy?** The complication is
**imperfect detection**: even when a species is present, a camera will not
always fire. Treating "no detection" as "truly absent" underestimates
occupancy — sometimes dramatically.

The MacKenzie et al. (2002) model simultaneously estimates:

- **ψ (psi)** — the probability a site is occupied.
- **p** — the probability of detecting the species on a given survey
  occasion, *given it is present*.

We split the monitoring week into daily occasions. For each camera × day,
the detection history is **1** (detected) or **0** (not detected).

<div class="callout callout-formula">
<strong>⚙️ The occupancy likelihood</strong><br>
For a site with detection history [1, 0, 1]:<br>
<code>P(data) = ψ × p(1−p)p  +  (1−ψ) × 0</code><br>
For a history of all zeros [0, 0, 0]:<br>
<code>P(data) = ψ × (1−p)³  +  (1−ψ)</code><br>
The second term captures the possibility the site was unoccupied — this
is what separates a true absence from a detection failure.
</div>

### Build detection histories

```{r det-history}
focal_sp    <- unique(cam_events$taxa_code)[1]
cat("Focal species:", focal_sp, "\n")

all_dates   <- sort(unique(cam_events$date))
all_cameras <- sort(unique(cam_events$camera_id))

# 1 = species detected that camera × day; 0 = not detected
det_wide <- cam_events %>%
  filter(taxa_code == focal_sp) %>%
  distinct(camera_id, date) %>%
  mutate(detected = 1) %>%
  complete(camera_id = all_cameras, date = all_dates,
           fill = list(detected = 0)) %>%
  pivot_wider(names_from = date, values_from = detected) %>%
  column_to_rownames("camera_id")

print(det_wide)
```

### Fit the null occupancy model

```{r occupancy}
umf      <- unmarkedFrameOccu(y = as.matrix(det_wide))
fit_null <- occu(~ 1 ~ 1, data = umf)
summary(fit_null)

# Back-transform from logit scale to probability
occ_est <- backTransform(fit_null, type = "state")
det_est <- backTransform(fit_null, type = "det")

cat(sprintf(
  "\nEstimated occupancy  ψ = %.3f  (95%% CI: %.3f – %.3f)\n",
  occ_est@estimate,
  confint(occ_est)[1], confint(occ_est)[2]
))
cat(sprintf(
  "Estimated detection  p = %.3f  (95%% CI: %.3f – %.3f)\n",
  det_est@estimate,
  confint(det_est)[1], confint(det_est)[2]
))
```

### Compare naïve vs. model-estimated occupancy

```{r naive-vs-model}
naive_occ <- mean(rowSums(det_wide) > 0)

cat(sprintf(
  "Naïve occupancy (fraction of cameras with ≥1 detection): %.3f\n", naive_occ
))
cat(sprintf(
  "Model-estimated occupancy ψ:                             %.3f\n",
  occ_est@estimate
))
cat(sprintf(
  "Inflation due to imperfect detection:                    +%.3f\n",
  occ_est@estimate - naive_occ
))
```

### Covariate model: does vegetation density affect occupancy?

If you have a site-level covariate (e.g., canopy cover measured at each
camera), you can test whether occupancy varies with habitat quality.
Replace the simulated `veg_density` below with your real covariate.

```{r occupancy-covariate}
# ── Replace this block with your real covariate data ────────────────────────
set.seed(7)
site_covs <- data.frame(
  veg_density = scale(runif(nrow(det_wide), 0, 100))
)
# ────────────────────────────────────────────────────────────────────────────

umf_cov <- unmarkedFrameOccu(
  y        = as.matrix(det_wide),
  siteCovs = site_covs
)

fit_cov <- occu(~ 1 ~ veg_density, data = umf_cov)
summary(fit_cov)

# AIC comparison
aic_df <- data.frame(
  Model = c("Null  (ψ(.), p(.))",
            "Veg   (ψ(veg), p(.))"),
  AIC   = round(c(fit_null@AIC, fit_cov@AIC), 2),
  dAIC  = round(c(0, fit_cov@AIC - fit_null@AIC), 2)
)
knitr::kable(
  aic_df,
  col.names = c("Model", "AIC", "ΔAIC"),
  align     = "lcc",
  caption   = paste0(
    "AIC model comparison. The model with lower AIC is better supported. ",
    "ΔAIC > 2 is the conventional threshold for meaningful improvement. ",
    "ΔAIC < 2 means the simpler null model is preferred."
  )
)
```

<div class="callout callout-concept">
<strong>📌 Interpreting AIC</strong>

The model with the <strong>lower AIC</strong> has better predictive support.
ΔAIC &gt; 2 is the conventional threshold for "meaningfully better".
A positive coefficient on <code>veg_density</code> in the occupancy sub-model
means the species is more likely to occupy densely vegetated sites.
AIC penalizes complexity — a more parameterized model must fit considerably
better to justify its extra parameters.
</div>

### Predicted occupancy across the vegetation gradient

The covariate model estimated a single coefficient for `veg_density` on the
**logit** (log-odds) scale. That number is hard to interpret directly —
a coefficient of +0.8 does not mean "occupancy increases by 0.8." The logit
link is non-linear, so the same unit increase in vegetation density produces
a larger change in ψ near 0.50 than it does near 0 or 1.

`predict()` translates the coefficient back into probability space across the
full observed range of the covariate, producing the **response curve** — the
most interpretable summary of the model. The code below:

1. **Creates a prediction grid** — 100 evenly spaced values of `veg_density`
   spanning the observed range (from the most open to the densest site).
2. **Calls `predict(type = "state")`** — `"state"` means the occupancy
   sub-model (ψ), as opposed to `"det"` which would give the detection
   sub-model (p). The function returns four columns: `Predicted` (ψ̂),
   `SE` (standard error on the probability scale), and `lower`/`upper`
   (95% confidence interval bounds).
3. **Plots the curve and ribbon** — the shaded band is the 95% CI, reflecting
   uncertainty in the estimated coefficient. Wide bands indicate either a
   small sample of cameras or weak signal in the data.

```{r occ-predictions, fig.height=4.8}
# Step 1: prediction grid across the full covariate range
newdat <- data.frame(
  veg_density = seq(min(site_covs$veg_density),
                    max(site_covs$veg_density),
                    length.out = 100)
)

# Step 2: predict ψ at each grid point (back-transformed to probability scale)
pred <- predict(fit_cov, type = "state", newdata = newdat)

# Show what predict() returns — four columns
head(round(pred, 3))
```

```{r occ-predictions-plot, fig.height=4.5}
# Step 3: plot the response curve
pred_df <- bind_cols(newdat, pred)

# Annotate where ψ = 0.5 (the inflection point of the logistic curve)
psi50_x <- pred_df$veg_density[which.min(abs(pred_df$Predicted - 0.5))]
psi50_y <- 0.5

ggplot(pred_df, aes(x = veg_density, y = Predicted)) +
  # 95% CI ribbon
  geom_ribbon(aes(ymin = lower, ymax = upper),
              fill = "#d4e6da", alpha = 0.75) +
  # Response curve
  geom_line(color = "#1e3a2f", linewidth = 1.4) +
  # ψ = 0.5 reference lines
  geom_hline(yintercept = 0.5, linetype = "dashed",
             color = "#c8832a", linewidth = 0.7) +
  geom_vline(xintercept = psi50_x, linetype = "dashed",
             color = "#c8832a", linewidth = 0.7) +
  annotate("point", x = psi50_x, y = psi50_y,
           shape = 21, size = 3.5, fill = "#c8832a", color = "white") +
  annotate("text",
           x     = psi50_x + 0.08 * diff(range(pred_df$veg_density)),
           y     = 0.53,
           label = sprintf("ψ = 0.50
at veg = %.2f", psi50_x),
           color = "#c8832a", size = 3.2, hjust = 0, fontface = "italic") +
  # Rug: actual camera covariate values
  geom_rug(data = data.frame(veg_density = site_covs$veg_density),
           aes(x = veg_density), y = NULL,
           sides = "b", color = "#4a7c5e", alpha = 0.6, linewidth = 0.5) +
  scale_y_continuous(limits = c(0, 1),
                     labels = scales::percent_format(),
                     breaks = seq(0, 1, 0.25)) +
  labs(
    title    = paste("Predicted occupancy (ψ) for", focal_sp),
    subtitle = "Shaded band = 95% CI  |  Rug marks = actual camera locations  |  Orange lines = ψ = 0.50",
    x        = "Vegetation density (standardized z-score)",
    y        = "Estimated occupancy probability (ψ)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(family = "serif", face = "bold", color = "#1e3a2f"),
    plot.subtitle = element_text(size = 9, color = "#7a7a72")
  )
```

<div class="callout callout-concept">
<strong>📌 Reading the response curve — four questions to ask</strong>

**1. Direction:** Does ψ increase or decrease with the covariate?
A rising curve means the species is more likely to occupy sites with more
vegetation; a falling curve means the opposite.

**2. Magnitude:** How much does ψ change across the full covariate range?
If ψ only moves from 0.42 to 0.51, the effect is statistically detectable
but ecologically trivial. If it moves from 0.10 to 0.85, habitat selection
is strong.

**3. Uncertainty:** How wide is the shaded band? Wide CIs at the
covariate extremes mean few cameras were sampled there — interpret the
predicted ψ at those values with caution.

**4. Inflection point:** The orange dot marks ψ = 0.50, the covariate
value at which occupancy and non-occupancy are equally likely. This is
the logistic regression analogue of an LC50 — a useful threshold for
management (e.g., "sites with vegetation density above X are more likely
than not to be occupied").
</div>

<div class="callout callout-warn">
<strong>⚠️ This example uses simulated covariate data</strong>

The <code>veg_density</code> values above were generated with
<code>runif()</code> and have no relationship to the actual cameras.
The curve shape and inflection point are therefore meaningless for your
real focal species. Replace the simulation block with measured field
covariates — canopy cover from a densiometer, distance to water,
trail-camera height class, or any site-level variable you recorded —
before drawing ecological conclusions.
</div>

---

## 9. Summary and Discussion {#summary}

```{r pipeline-table, echo=FALSE}
knitr::kable(
  data.frame(
    Stage = c("Import & clean", "Independent events",
              "Descriptive stats", "Chi-square / Fisher",
              "Activity overlap Δ", "Rarefaction",
              "Occupancy model"),
    `Key idea` = c(
      "Remove ambiguous taxa; standardize columns with clean_names()",
      "30-minute rule collapses burst detections; prevents pseudo-replication",
      "Ethograms, heat maps, and camera profiles reveal community-level patterns",
      "Tests whether behavior composition shifts between day and night",
      "Kernel density Δ quantifies temporal co-occurrence on a 0–1 scale",
      "Controls for unequal detection effort before comparing camera richness",
      "Separates true absence from non-detection to estimate ψ and p"
    ),
    `R functions` = c(
      "read_excel(), clean_names(), case_when()",
      "lag(), difftime(), cumsum()",
      "ggplot2: geom_col(), geom_tile()",
      "chisq.test(), fisher.test()",
      "overlapEst(), bootstrap(), bootCI()",
      "rarecurve(), specnumber(), diversity()",
      "occu(), backTransform(), predict()"
    )
  ),
  col.names = c("Stage", "Key idea", "R functions"),
  caption   = "Full camera-trap analysis pipeline."
)
```

<div class="discuss-box">
<h4>💬 Discussion Questions</h4>
<ol>
<li>Which taxon showed the greatest day–night behavioral shift? What ecological
    mechanism might explain it — predation pressure, thermoregulation, or
    resource availability?</li>
<li>Do species with high activity overlap (Δ &gt; 0.75) tend to be
    ecologically similar (same guild) or different? What does this suggest
    about competitive exclusion vs. temporal partitioning?</li>
<li>Examine the rarefaction curves. Which cameras appear under-sampled
    (curves not levelled off)? How would you redesign the study to
    address this — longer deployment, more cameras, or both?</li>
<li>Compare the naïve detection rate for your focal species with the
    model-estimated occupancy ψ. How large is the imperfect-detection
    bias? Does it change any substantive conclusions about habitat use?</li>
<li>The covariate model used a simulated vegetation density variable.
    What real site covariates would you collect in the field, and how
    would you measure them consistently across cameras?</li>
</ol>
</div>

<div class="key-box">
<h4>🔑 Key takeaways</h4>
<ul>
<li><strong>Always define independent events</strong> before analysis. The 30-minute rule is a convention — document your threshold and justify it relative to your study organism.</li>
<li><strong>Descriptive plots come first.</strong> An ethogram or heat map often surfaces the most important ecological patterns before any test is run.</li>
<li><strong>Chi-square vs. Fisher:</strong> check expected counts with <code>chisq.test()$expected</code>; switch to <code>fisher.test()</code> automatically when any cell is below 5.</li>
<li><strong>Activity overlap Δ is sample-size sensitive.</strong> Use <code>Dhat4</code> when either species has fewer than 75 detections; always report a bootstrap CI alongside the point estimate.</li>
<li><strong>Rarefaction, not raw richness,</strong> is the correct basis for comparing cameras with unequal total events.</li>
<li><strong>Occupancy models separate biology from observation.</strong> Naïve detection rate systematically underestimates true occupancy when detection is imperfect — which it always is with camera traps.</li>
</ul>
</div>

---

### References

- MacKenzie, D. I., Nichols, J. D., Lachman, G. B., Droege, S., Royle, J. A.,
  & Langtimm, C. A. (2002). Estimating site occupancy rates when detection
  probabilities are less than one. *Ecology*, 83(8), 2248–2255.
- Ridout, M. S., & Linkie, M. (2009). Estimating overlap of daily activity
  patterns from camera trap data. *Journal of Agricultural, Biological, and
  Environmental Statistics*, 14(3), 322–337.
- Fiske, I., & Chandler, R. (2011). unmarked: An R package for fitting
  hierarchical models of wildlife occurrence and abundance. *Journal of
  Statistical Software*, 43(10), 1–23.
- Oksanen, J., et al. (2022). vegan: Community Ecology Package. R package
  version 2.6-4.

```{r session-info, echo=FALSE}
# sessionInfo()
```

<footer class="page-footer">
EA030 RTricks &mdash; Project 05: Camera Trap Analysis
<span class="divider">|</span> Prof. Marc Los Huertos
<span class="divider">|</span> Pomona College
<span class="divider">|</span> Last updated: `r format(Sys.Date(), "%B %Y")`
<span class="divider">|</span> <a href="https://github.com/marclos/RTricks">GitHub</a>
</footer>
