---
title: "Urban Canopy and Temperature"
author: "Marc Los Huertos"
date: "EA030 · Compiled `r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    self_contained: true
    theme: null
    highlight: null
    toc: false
knit: (function(inputFile, encoding) {
    rmarkdown::render(
      inputFile,
      encoding = encoding,
      output_file = file.path(dirname(inputFile),
                              "../docs/Activities/A13_Urban_Canopy_Temperature_Analysis.html")
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 8,
  fig.height = 4.5
)

# ── Packages ──────────────────────────────────────────────────────────────────
library(tidyverse)
library(knitr)
library(kableExtra)
library(patchwork)
library(broom)
library(emmeans)
library(car)
library(moments)

# ── Shared palette (income groups) ────────────────────────────────────────────
income_pal  <- c("Low-income"    = "#c0392b",
                 "Middle-income" = "#e67e22",
                 "High-income"   = "#27ae60")

setting_pal <- c("School"      = "#1e3a2f",
                 "Park"        = "#4a7c5e",
                 "Commercial"  = "#c8832a",
                 "Parking Lot" = "#7f8c8d",
                 "Playground"  = "#8e44ad")

# ── Factor level order (used throughout) ──────────────────────────────────────
income_lvl  <- c("Low-income", "Middle-income", "High-income")
setting_lvl <- c("School", "Park", "Commercial", "Parking Lot", "Playground")

theme_set(
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(family = "serif", face = "bold",
                                 color  = "#1e3a2f", size = 13),
    plot.subtitle = element_text(color = "#7a7a72", size = 10),
    axis.title    = element_text(color = "#3a3a38", size = 11),
    strip.text    = element_text(face = "bold", color = "#1e3a2f"),
    legend.position = "top"
  )
)
```

```{css, echo=FALSE}
@import url('https://fonts.googleapis.com/css2?family=Libre+Baskerville:ital,wght@0,400;0,700;1,400&family=Source+Code+Pro:wght@400;600&family=Lato:wght@300;400;700&display=swap');

*, *::before, *::after { box-sizing: border-box; margin: 0; padding: 0; }

:root {
  --forest:    #1e3a2f; --moss: #2d5a3d; --fern: #4a7c5e; --sage: #8aad96;
  --mist:      #d4e6da; --cream: #f7f3ed; --parchment: #ede8df;
  --ink:       #1a1a18; --charcoal: #3a3a38; --dust: #7a7a72;
  --amber:     #c8832a; --amber-lt: #e8a84a; --white: #ffffff;
  --card-shadow: 0 2px 10px rgba(30,58,47,.09);
  --part1: #1565c0; --part2: #7b1fa2;
}

html { scroll-behavior: smooth; }
body {
  font-family: 'Lato', sans-serif; background: var(--cream); color: var(--ink);
  font-size: 17px; line-height: 1.75;
  padding-left: clamp(14px,4vw,48px); padding-right: clamp(14px,4vw,48px);
  padding-bottom: 80px;
}
.main-container { max-width: 940px !important; margin: 0 auto !important;
                  padding: 0 clamp(10px,2vw,32px) 80px !important; }
.page-header { margin-left: calc(-1 * clamp(14px,4vw,48px));
               margin-right: calc(-1 * clamp(14px,4vw,48px)); }
h1.title, h4.author, h4.date { display: none; }

/* ── Main header ─────────────────────────────────────────────────────────── */
.page-header {
  background-image:
    repeating-linear-gradient(120deg, transparent, transparent 60px,
      rgba(255,255,255,.018) 60px, rgba(255,255,255,.018) 61px),
    linear-gradient(160deg, #1e3a2f 0%, #2d5a3d 60%, #1a3d30 100%);
  color: #fff; padding: 44px 4vw 36px; margin: 0 -4vw 48px;
  border-bottom: 4px solid var(--amber); position: relative; overflow: hidden;
}
.page-header::after {
  content: ''; position: absolute; right: -40px; top: -40px;
  width: 260px; height: 260px; border-radius: 50%;
  background: radial-gradient(circle, rgba(138,173,150,.12) 0%, transparent 70%);
  pointer-events: none;
}
.page-header .module-tag {
  display: inline-block; font-family: 'Source Code Pro', monospace;
  font-size: .68rem; font-weight: 600; letter-spacing: .18em; text-transform: uppercase;
  color: var(--amber-lt); border: 1px solid rgba(200,131,42,.45);
  padding: 3px 10px; border-radius: 2px; margin-bottom: 12px;
}
.page-header h1 { font-family: 'Libre Baskerville', serif;
  font-size: clamp(1.6rem,3vw,2.3rem); font-weight: 700; color: #fff;
  line-height: 1.2; margin-bottom: 8px; }
.page-header h1 span { color: var(--sage); }
.page-header .sub { font-size: .9rem; color: rgba(255,255,255,.65); font-weight: 300; }
.page-header .sub strong { color: rgba(255,255,255,.9); }
.back-link { display: inline-block; margin-top: 16px;
  font-family: 'Source Code Pro', monospace; font-size: .7rem; letter-spacing: .08em;
  color: var(--amber-lt); text-decoration: none;
  border: 1px solid rgba(200,131,42,.4); padding: 4px 11px; border-radius: 3px; }
.back-link:hover { background: rgba(200,131,42,.18); }

/* ── Part banners ────────────────────────────────────────────────────────── */
.part-banner {
  padding: 18px 24px; border-radius: 6px; margin: 48px 0 28px;
  color: #fff; position: relative; overflow: hidden;
}
.part-banner::after {
  content: ''; position: absolute; right: -20px; top: -20px;
  width: 120px; height: 120px; border-radius: 50%;
  background: rgba(255,255,255,.08); pointer-events: none;
}
.part1-banner { background: linear-gradient(135deg, #1565c0, #1976d2); }
.part2-banner { background: linear-gradient(135deg, #6a1b9a, #8e24aa); }
.part-banner .part-num {
  font-family: 'Source Code Pro', monospace; font-size: .68rem;
  letter-spacing: .2em; text-transform: uppercase; opacity: .75; margin-bottom: 4px;
}
.part-banner h2 { font-family: 'Libre Baskerville', serif; font-size: 1.4rem;
  font-weight: 700; color: #fff; border: none; margin: 0; padding: 0; }
.part-banner .part-desc { font-size: .88rem; opacity: .8; margin-top: 4px; }

/* ── Navigation ─────────────────────────────────────────────────────────── */
.section-nav { display: flex; gap: 8px; flex-wrap: wrap; margin-bottom: 44px;
  padding-bottom: 20px; border-bottom: 1px solid var(--parchment); }
.section-nav a { font-family: 'Source Code Pro', monospace; font-size: .7rem;
  font-weight: 600; letter-spacing: .08em; text-transform: uppercase;
  color: var(--fern); text-decoration: none; border: 1px solid var(--mist);
  padding: 4px 12px; border-radius: 3px; background: var(--white);
  transition: background .2s, color .2s; }
.section-nav a:hover { background: var(--mist); color: var(--forest); }
.section-nav .nav-p1 { border-color: #bbdefb; color: #1565c0; }
.section-nav .nav-p1:hover { background: #e3f2fd; }
.section-nav .nav-p2 { border-color: #e1bee7; color: #6a1b9a; }
.section-nav .nav-p2:hover { background: #f3e5f5; }

/* ── Headings ────────────────────────────────────────────────────────────── */
h2 { font-family: 'Libre Baskerville', serif; font-size: 1.35rem; font-weight: 700;
  color: var(--forest); margin-top: 48px; margin-bottom: 6px;
  padding-bottom: 8px; border-bottom: 1px solid var(--parchment); }
h3 { font-family: 'Libre Baskerville', serif; font-size: 1.05rem; font-weight: 700;
  color: var(--moss); margin-top: 26px; margin-bottom: 4px; }

p { margin-bottom: 1em; }
strong { color: var(--forest); }
blockquote { border-left: 4px solid var(--sage); margin: 1.2em 0; padding: .5em 1em;
  background: var(--white); border-radius: 0 5px 5px 0; font-size: .94em;
  color: var(--charcoal); }

code { font-family: 'Source Code Pro', monospace; font-size: .84em;
  background: var(--mist); color: var(--moss); padding: 1px 5px; border-radius: 3px; }
pre { background: var(--ink); color: #e8e6df; border-radius: 5px;
  padding: 14px 18px; font-size: .83rem; line-height: 1.55; overflow-x: auto;
  border-left: 4px solid var(--amber); margin: 1.2em 0; }
pre code { background: transparent; color: inherit; padding: 0; }

.callout { border-left: 5px solid; padding: .65em 1.1em; margin: 1.4em 0;
  border-radius: 0 5px 5px 0; font-size: .92em; line-height: 1.6; }
.callout > strong:first-child { display: block; margin-bottom: .3em; }
.callout-concept { border-color: var(--fern);  background: #f0f7f2; }
.callout-tip     { border-color: #2e7d32;      background: #f1f8f1; }
.callout-warn    { border-color: #e65100;      background: #fff8f4; }
.callout-interp  { border-color: var(--part1); background: #e8f4fd; }
.callout-interp2 { border-color: var(--part2); background: #f5eafa; }

table { font-size: .88rem; border-collapse: collapse; width: 100%; margin: 1.2em 0; }
thead th { background: var(--forest); color: #fff; font-family: 'Source Code Pro', monospace;
  font-size: .74rem; letter-spacing: .06em; text-transform: uppercase;
  padding: 8px 12px; text-align: left; }
tbody tr:nth-child(even) { background: var(--mist); }
tbody td { padding: 7px 12px; border-bottom: 1px solid var(--parchment); }

.key-box { background: var(--forest); color: #fff; border-radius: 6px;
  padding: 20px 24px; margin: 2em 0; font-size: .9rem; }
.key-box h4 { font-family: 'Libre Baskerville', serif; color: var(--sage);
  font-size: 1rem; margin-bottom: 10px; }
.key-box ul { padding-left: 1.3em; }
.key-box li { margin-bottom: 5px; color: rgba(255,255,255,.85); }
.key-box code { background: rgba(255,255,255,.12); color: var(--amber-lt); }

.page-footer { background: var(--ink); color: rgba(255,255,255,.42); text-align: center;
  padding: 24px 4vw; margin: 60px -4vw 0;
  font-size: .76rem; font-family: 'Source Code Pro', monospace; letter-spacing: .04em; }
.page-footer a { color: var(--sage); text-decoration: none; }
.page-footer .divider { margin: 0 8px; opacity: .3; }
```

<div class="page-header">
<div class="module-tag">EA030 &mdash; Urban Ecology &mdash; Field Analysis</div>
<h1>Urban Canopy Cover and <span>Heat Exposure</span></h1>
<p class="sub"><strong>Prof. Marc Los Huertos</strong> &nbsp;&bull;&nbsp; Pomona College &nbsp;&bull;&nbsp; Spring 2026</p>
<a class="back-link" href="index.html">&#8592; Back to Project Index</a>
</div>

<div class="section-nav">
<a href="#background">Background</a>
<a class="nav-p1" href="#p1-desc">P1 · Descriptive</a>
<a class="nav-p1" href="#p1-viz">P1 · Visualization</a>
<a class="nav-p1" href="#p1-stats">P1 · Statistics</a>
<a class="nav-p2" href="#p2-desc">P2 · Descriptive</a>
<a class="nav-p2" href="#p2-viz">P2 · Visualization</a>
<a class="nav-p2" href="#p2-stats">P2 · Statistics</a>
<a href="#synthesis">Synthesis</a>
</div>

> **Study overview.** This analysis examines how urban tree canopy cover affects heat
> exposure across five urban land-use settings and three socioeconomic groups in the
> Inland Empire. Two complementary instrument types were deployed: **Part I** uses
> 24-hour infield temperature loggers (continuous air temperature) and **Part II**
> uses handheld IR thermometers (point-in-time surface temperature). Together they
> reveal both chronic heat burden and instantaneous thermal conditions.

---

## Background and Research Context {#background}

Urban heat islands (UHIs) emerge when impervious surfaces, reduced vegetation, and
waste heat from buildings and vehicles elevate temperatures above surrounding
rural areas. Tree canopy provides one of the most cost-effective mitigation
strategies — shade reduces incoming radiation while transpiration cools surfaces
through latent heat flux.

Environmental justice concerns arise because canopy cover is unevenly distributed:
wealthier neighborhoods historically receive more urban greening investment. If
canopy reduces heat exposure and low-income communities have less canopy, those
communities bear a disproportionate thermal burden.

```{r data-load}
# ── Set paths — update if your CSVs live elsewhere ───────────────────────────
log_path <- "/home/mwl04747/RTricks/A13_Heat_Islands/Data/infield_logger_canopy_urban.csv"
ir_path  <- "/home/mwl04747/RTricks/A13_Heat_Islands/Data/ir_spot_temperature_canopy_urban.csv"

# ── Part I: infield loggers ───────────────────────────────────────────────────
log_raw <- read_csv(log_path, show_col_types = FALSE) |>
  mutate(
    urban_setting    = factor(urban_setting,    levels = setting_lvl),
    demographic_group = factor(demographic_group, levels = income_lvl)
  )

# ── Part II: IR spot temperatures ────────────────────────────────────────────
ir_raw <- read_csv(ir_path, show_col_types = FALSE) |>
  mutate(
    urban_setting    = factor(urban_setting,    levels = setting_lvl),
    demographic_group = factor(demographic_group, levels = income_lvl),
    time_of_day = cut(measurement_hour,
                      breaks = c(8, 11, 14, 17),
                      labels = c("Morning (9–11)", "Midday (12–14)",
                                 "Afternoon (15–17)"),
                      include.lowest = TRUE)
  )

cat("Logger dataset:  ", nrow(log_raw), "rows ×", ncol(log_raw), "columns\n")
cat("IR spot dataset: ", nrow(ir_raw),  "rows ×", ncol(ir_raw),  "columns\n")
```

**Two research questions guide this analysis:**

1. Does mean 24-hour air temperature differ by urban setting, income group, and canopy
   cover? (*Part I — loggers*)
2. Does instantaneous surface temperature vary by canopy cover, time of day, urban
   setting, and income group? (*Part II — IR spot*)

<hr>

<div class="part-banner part1-banner">
<div class="part-num">Part I &mdash; Dataset 1</div>
<h2>Infield Temperature Loggers</h2>
<p class="part-desc">Cold-chain food logistics loggers deployed for 24 hours · 180 observations · 5 settings × 3 income groups</p>
</div>

## 1. Descriptive Statistics — Logger Data {#p1-desc}

### 1.1 Overall summary

```{r p1-summary-overall}
log_raw |>
  summarise(
    N          = n(),
    Mean_temp  = round(mean(mean_air_temperature_c), 2),
    SD_temp    = round(sd(mean_air_temperature_c),   2),
    Median_temp = round(median(mean_air_temperature_c), 2),
    IQR_temp   = round(IQR(mean_air_temperature_c),  2),
    Min_temp   = round(min(mean_air_temperature_c),  2),
    Max_temp   = round(max(mean_air_temperature_c),  2),
    Skewness   = round(moments::skewness(mean_air_temperature_c), 3),
    Mean_canopy = round(mean(canopy_cover_percent), 1),
    SD_canopy   = round(sd(canopy_cover_percent),   1)
  ) |>
  kable(caption = "Overall descriptive statistics — logger air temperature and canopy cover") |>
  kable_styling(full_width = FALSE)
```

### 1.2 By urban setting and income group

```{r p1-summary-by-group}
log_raw |>
  group_by(urban_setting, demographic_group) |>
  summarise(
    N    = n(),
    Mean = round(mean(mean_air_temperature_c), 2),
    SD   = round(sd(mean_air_temperature_c),   2),
    Med  = round(median(mean_air_temperature_c), 2),
    Canopy_mean = round(mean(canopy_cover_percent), 1),
    .groups = "drop"
  ) |>
  rename(`Urban setting` = urban_setting,
         `Income group`  = demographic_group,
         `Mean temp (°C)` = Mean, `SD` = SD, `Median` = Med,
         `Mean canopy (%)` = Canopy_mean) |>
  kable(caption = "Mean 24-hour air temperature (°C) by urban setting and income group") |>
  kable_styling(full_width = TRUE) |>
  column_spec(1, bold = TRUE)
```

### 1.3 Normality check

```{r p1-normality}
set.seed(42)
norm_tbl <- log_raw |>
  group_by(urban_setting) |>
  summarise(
    n  = n(),
    W  = round(shapiro.test(mean_air_temperature_c)$statistic, 4),
    p  = round(shapiro.test(mean_air_temperature_c)$p.value,   4),
    Skew = round(moments::skewness(mean_air_temperature_c), 3),
    .groups = "drop"
  ) |>
  mutate(Normal = ifelse(p > 0.05, "Yes", "No"))

kable(norm_tbl,
      caption = "Shapiro-Wilk normality test per urban setting (logger data)",
      col.names = c("Setting","N","W","p-value","Skewness","Normal?")) |>
  kable_styling(full_width = FALSE)
```

<div class="callout callout-concept">
<strong>💡 Why normality matters here</strong>
One-way ANOVA assumes approximately normal residuals within each group.
With n = 36 per setting, the Central Limit Theorem provides reasonable
robustness — but we will verify residual normality after fitting and use
Kruskal-Wallis as a non-parametric check.
</div>

<hr>

## 2. Visualization — Logger Data {#p1-viz}

### 2.1 Temperature distributions by urban setting

```{r p1-boxplot-setting, fig.height=4.5}
ggplot(log_raw, aes(x = urban_setting, y = mean_air_temperature_c,
                    fill = urban_setting)) +
  geom_boxplot(outlier.alpha = 0.4, outlier.size = 1.5, width = 0.55) +
  geom_jitter(aes(color = demographic_group), width = 0.12,
              alpha = 0.6, size = 2) +
  stat_summary(fun = mean, geom = "point",
               shape = 23, size = 3, fill = "white", color = "#1e3a2f") +
  scale_fill_manual(values  = setting_pal, guide = "none") +
  scale_color_manual(values = income_pal, name = "Income group") +
  labs(title    = "24-Hour Mean Air Temperature by Urban Setting",
       subtitle = "Points colored by income group · ◇ = group mean",
       x = NULL, y = "Mean air temperature (°C)") +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

### 2.2 Temperature by income group across settings

```{r p1-income-facet, fig.height=5}
ggplot(log_raw, aes(x = demographic_group, y = mean_air_temperature_c,
                    fill = demographic_group)) +
  geom_boxplot(outlier.alpha = 0.35, width = 0.5) +
  stat_summary(fun = mean, geom = "point",
               shape = 23, size = 2.5, fill = "white") +
  facet_wrap(~ urban_setting, nrow = 1) +
  scale_fill_manual(values = income_pal, guide = "none") +
  labs(title    = "Air Temperature by Income Group within Each Urban Setting",
       subtitle = "◇ = mean",
       x = NULL, y = "Mean air temperature (°C)") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1, size = 9))
```

### 2.3 Canopy cover vs air temperature

```{r p1-scatter, fig.height=4.5}
ggplot(log_raw, aes(x = canopy_cover_percent, y = mean_air_temperature_c,
                    color = demographic_group, shape = urban_setting)) +
  geom_point(alpha = 0.75, size = 2.5) +
  geom_smooth(aes(group = 1), method = "lm", se = TRUE,
              color = "#1e3a2f", linewidth = 1, linetype = "dashed") +
  scale_color_manual(values = income_pal, name = "Income group") +
  scale_shape_manual(values = c(16, 17, 15, 18, 8), name = "Urban setting") +
  labs(title    = "Canopy Cover vs Mean Air Temperature (24-hour loggers)",
       subtitle = "Dashed line = overall OLS trend",
       x = "Canopy cover (%)", y = "Mean air temperature (°C)") +
  guides(color = guide_legend(nrow = 1),
         shape = guide_legend(nrow = 1))
```

### 2.4 Canopy cover distribution by income group

```{r p1-canopy-income, fig.height=3.5}
ggplot(log_raw, aes(x = canopy_cover_percent, fill = demographic_group,
                    color = demographic_group)) +
  geom_density(alpha = 0.30, linewidth = 0.8) +
  scale_fill_manual(values  = income_pal, name = "Income group") +
  scale_color_manual(values = income_pal, name = "Income group") +
  labs(title    = "Canopy Cover Distribution by Income Group — Logger Sites",
       subtitle = "Are low-income sites systematically under-canopied?",
       x = "Canopy cover (%)", y = "Density")
```

<div class="callout callout-interp">
<strong>📊 What to look for in these plots</strong>
If environmental justice patterns are present, you expect: (a) low-income sites
cluster at <em>lower canopy cover</em> and <em>higher temperatures</em>,
(b) parking lots and commercial areas show the highest temperatures regardless
of income group, and (c) the scatter plot shows a negative slope (more canopy
→ cooler air).
</div>

<hr>

## 3. Statistical Analysis — Logger Data {#p1-stats}

### 3.1 One-way ANOVA: does temperature differ by urban setting?

**H₀:** Mean 24-hour air temperature is equal across all five urban settings.
**H₁:** At least one setting has a different mean temperature.

```{r p1-anova-setting}
aov_setting <- aov(mean_air_temperature_c ~ urban_setting, data = log_raw)
summary(aov_setting)
```

```{r p1-anova-setting-tidy}
tidy(aov_setting) |>
  mutate(across(where(is.numeric), ~ round(.x, 4))) |>
  kable(caption = "ANOVA table: air temperature ~ urban setting") |>
  kable_styling(full_width = FALSE)
```

### 3.2 Post-hoc comparisons (Tukey HSD)

```{r p1-tukey, fig.height=4}
em_setting <- emmeans(aov_setting, ~ urban_setting)
pairs_setting <- pairs(em_setting, adjust = "tukey")
summary(pairs_setting) |>
  as.data.frame() |>
  mutate(across(where(is.numeric), ~ round(.x, 4)),
         sig = case_when(p.value < 0.001 ~ "***",
                         p.value < 0.01  ~ "**",
                         p.value < 0.05  ~ "*",
                         TRUE            ~ "")) |>
  kable(caption = "Tukey HSD pairwise comparisons — urban setting") |>
  kable_styling(full_width = FALSE)
```

```{r p1-emmeans-plot, fig.height=3.5}
as.data.frame(em_setting) |>
  ggplot(aes(x = reorder(urban_setting, emmean), y = emmean,
             ymin = lower.CL, ymax = upper.CL, color = urban_setting)) +
  geom_pointrange(size = 0.9, linewidth = 1) +
  scale_color_manual(values = setting_pal, guide = "none") +
  coord_flip() +
  labs(title    = "Estimated Marginal Means ± 95% CI",
       subtitle = "Urban setting effect on mean air temperature",
       x = NULL, y = "Estimated mean temperature (°C)")
```

### 3.3 Two-way ANOVA: setting × income group interaction

```{r p1-anova-2way}
aov_2way <- aov(mean_air_temperature_c ~ urban_setting * demographic_group,
                data = log_raw)
tidy(aov_2way) |>
  mutate(across(where(is.numeric), ~ round(.x, 4)),
         sig = case_when(p.value < 0.001 ~ "***",
                         p.value < 0.01  ~ "**",
                         p.value < 0.05  ~ "*",
                         TRUE            ~ "")) |>
  kable(caption = "Two-way ANOVA: temperature ~ setting * income group") |>
  kable_styling(full_width = FALSE)
```

### 3.4 ANOVA assumptions

```{r p1-residuals, fig.height=5, fig.width=9}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
plot(aov_2way)
par(mfrow = c(1, 1))
```

```{r p1-levene}
leveneTest(mean_air_temperature_c ~ urban_setting * demographic_group,
           data = log_raw) |>
  tidy() |>
  mutate(across(where(is.numeric), ~ round(.x, 4))) |>
  kable(caption = "Levene's test for homogeneity of variance") |>
  kable_styling(full_width = FALSE)
```

### 3.5 Simple linear regression: canopy cover predicts temperature

```{r p1-regression}
lm_canopy <- lm(mean_air_temperature_c ~ canopy_cover_percent, data = log_raw)

tidy(lm_canopy) |>
  mutate(across(where(is.numeric), ~ round(.x, 4))) |>
  kable(caption = "OLS regression: air temperature ~ canopy cover (%)") |>
  kable_styling(full_width = FALSE)

glance(lm_canopy) |>
  select(r.squared, adj.r.squared, statistic, p.value, nobs) |>
  mutate(across(where(is.numeric), ~ round(.x, 4))) |>
  kable(caption = "Model fit statistics") |>
  kable_styling(full_width = FALSE)
```

```{r p1-regression-plot, fig.height=4}
ggplot(log_raw, aes(x = canopy_cover_percent, y = mean_air_temperature_c)) +
  geom_point(aes(color = urban_setting, shape = demographic_group),
             alpha = 0.7, size = 2.5) +
  geom_smooth(method = "lm", se = TRUE,
              color = "#1e3a2f", linewidth = 1) +
  scale_color_manual(values = setting_pal, name = "Setting") +
  scale_shape_manual(values = c(16, 17, 8), name = "Income") +
  annotate("text",
           x = max(log_raw$canopy_cover_percent) * 0.85,
           y = max(log_raw$mean_air_temperature_c) * 0.99,
           label = paste0("R² = ", round(glance(lm_canopy)$r.squared, 3)),
           color = "#1e3a2f", fontface = "bold", size = 4) +
  labs(title    = "OLS Regression: Canopy Cover Predicts Air Temperature",
       x = "Canopy cover (%)", y = "Mean air temperature (°C)")
```

### 3.6 Multiple regression: canopy + setting + income group

```{r p1-mlr}
lm_multi <- lm(mean_air_temperature_c ~
                 canopy_cover_percent + urban_setting + demographic_group,
               data = log_raw)

tidy(lm_multi) |>
  mutate(
    sig = case_when(p.value < 0.001 ~ "***", p.value < 0.01 ~ "**",
                    p.value < 0.05  ~ "*",   TRUE ~ ""),
    across(where(is.numeric), ~ round(.x, 4))
  ) |>
  kable(caption = "Multiple regression: temperature ~ canopy + setting + income") |>
  kable_styling(full_width = FALSE)

glance(lm_multi) |>
  select(r.squared, adj.r.squared, AIC, BIC, nobs) |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  kable(caption = "Multiple regression model fit") |>
  kable_styling(full_width = FALSE)
```

### 3.7 Non-parametric check: Kruskal-Wallis test

```{r p1-kruskal}
kw_setting <- kruskal.test(mean_air_temperature_c ~ urban_setting, data = log_raw)
kw_income  <- kruskal.test(mean_air_temperature_c ~ demographic_group, data = log_raw)

tibble(
  Test      = c("Urban setting", "Income group"),
  Chi_sq    = round(c(kw_setting$statistic, kw_income$statistic), 3),
  df        = c(kw_setting$parameter, kw_income$parameter),
  p_value   = round(c(kw_setting$p.value, kw_income$p.value), 4),
  Sig       = ifelse(c(kw_setting$p.value, kw_income$p.value) < 0.05, "✓", "—")
) |>
  kable(caption = "Kruskal-Wallis tests (non-parametric alternative to ANOVA)") |>
  kable_styling(full_width = FALSE)
```

<div class="callout callout-interp">
<strong>📊 Interpreting the logger results</strong>
A significant ANOVA F-test tells you that at least one setting has a different
mean temperature. The Tukey HSD post-hoc narrows down <em>which pairs</em>
differ. The regression slope on canopy cover quantifies how much each additional
percent of canopy is associated with cooler (or warmer) air. If the income group
coefficient is significant in the multiple regression, it suggests that income
group predicts temperature even after accounting for canopy and setting —
consistent with an environmental justice pattern.
</div>

<hr>

<div class="part-banner part2-banner">
<div class="part-num">Part II &mdash; Dataset 2</div>
<h2>IR Spot Surface Temperature</h2>
<p class="part-desc">Handheld IR thermometer · 600 observations · 5 settings × 3 income groups × 40 measurements · Diurnal variation (9 AM – 5 PM)</p>
</div>

## 4. Descriptive Statistics — IR Spot Data {#p2-desc}

### 4.1 Overall summary

```{r p2-summary-overall}
ir_raw |>
  summarise(
    N            = n(),
    Mean_surf    = round(mean(ir_surface_temperature_c),   2),
    SD_surf      = round(sd(ir_surface_temperature_c),     2),
    Median_surf  = round(median(ir_surface_temperature_c), 2),
    IQR_surf     = round(IQR(ir_surface_temperature_c),    2),
    Min_surf     = round(min(ir_surface_temperature_c),    2),
    Max_surf     = round(max(ir_surface_temperature_c),    2),
    Skewness     = round(moments::skewness(ir_surface_temperature_c), 3),
    Mean_canopy  = round(mean(canopy_cover_percent), 1),
    Hour_range   = paste(min(measurement_hour), "–", max(measurement_hour))
  ) |>
  kable(caption = "Overall descriptive statistics — IR surface temperature") |>
  kable_styling(full_width = FALSE)
```

### 4.2 By urban setting and income group

```{r p2-summary-group}
ir_raw |>
  group_by(urban_setting, demographic_group) |>
  summarise(
    N    = n(),
    Mean = round(mean(ir_surface_temperature_c),   2),
    SD   = round(sd(ir_surface_temperature_c),     2),
    Med  = round(median(ir_surface_temperature_c), 2),
    Max  = round(max(ir_surface_temperature_c),    2),
    .groups = "drop"
  ) |>
  rename(`Urban setting` = urban_setting, `Income group` = demographic_group,
         `Mean (°C)` = Mean, `Max (°C)` = Max) |>
  kable(caption = "IR surface temperature summary by urban setting and income group") |>
  kable_styling(full_width = TRUE) |>
  column_spec(1, bold = TRUE)
```

### 4.3 By time of day

```{r p2-summary-tod}
ir_raw |>
  group_by(time_of_day, urban_setting) |>
  summarise(
    N    = n(),
    Mean = round(mean(ir_surface_temperature_c), 2),
    SD   = round(sd(ir_surface_temperature_c),   2),
    .groups = "drop"
  ) |>
  pivot_wider(names_from = time_of_day,
              values_from = c(Mean, SD),
              names_glue  = "{time_of_day}_{.value}") |>
  kable(caption = "Mean IR surface temperature (°C) by time of day and urban setting") |>
  kable_styling(full_width = TRUE)
```

### 4.4 Normality check

```{r p2-normality}
ir_raw |>
  group_by(urban_setting) |>
  summarise(
    n    = n(),
    W    = round(shapiro.test(sample(ir_surface_temperature_c,
                                     min(n(), 5000)))$statistic, 4),
    p    = round(shapiro.test(sample(ir_surface_temperature_c,
                                     min(n(), 5000)))$p.value, 4),
    Skew = round(moments::skewness(ir_surface_temperature_c), 3),
    .groups = "drop"
  ) |>
  mutate(Normal = ifelse(p > 0.05, "Yes", "No")) |>
  kable(caption = "Shapiro-Wilk normality test per urban setting (IR data)") |>
  kable_styling(full_width = FALSE)
```

<hr>

## 5. Visualization — IR Spot Data {#p2-viz}

### 5.1 Surface temperature distributions

```{r p2-dist, fig.height=4.5}
p_box2 <- ggplot(ir_raw,
                 aes(x = urban_setting, y = ir_surface_temperature_c,
                     fill = urban_setting)) +
  geom_boxplot(outlier.alpha = 0.2, outlier.size = 0.8, width = 0.55) +
  geom_jitter(aes(color = demographic_group), width = 0.1,
              alpha = 0.25, size = 1) +
  stat_summary(fun = mean, geom = "point",
               shape = 23, size = 2.5, fill = "white", color = "#1e3a2f") +
  scale_fill_manual(values  = setting_pal, guide = "none") +
  scale_color_manual(values = income_pal,  name  = "Income group") +
  labs(title = "IR Surface Temperature by Urban Setting",
       subtitle = "◇ = mean; points colored by income group",
       x = NULL, y = "Surface temperature (°C)") +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

p_dens2 <- ggplot(ir_raw,
                  aes(x = ir_surface_temperature_c, fill = urban_setting,
                      color = urban_setting)) +
  geom_density(alpha = 0.25, linewidth = 0.7) +
  geom_vline(xintercept = 35, linetype = "dashed",
             color = "firebrick", linewidth = 0.8) +
  scale_fill_manual(values  = setting_pal, name = "Setting") +
  scale_color_manual(values = setting_pal, name = "Setting") +
  labs(title    = "Density of Surface Temperatures",
       subtitle = "Dashed = 35°C heat stress threshold",
       x = "Surface temperature (°C)", y = "Density")

p_box2 + p_dens2
```

### 5.2 Diurnal pattern by urban setting

```{r p2-diurnal, fig.height=4.5}
ir_raw |>
  group_by(urban_setting, measurement_hour) |>
  summarise(
    mean_temp = mean(ir_surface_temperature_c),
    se_temp   = sd(ir_surface_temperature_c) / sqrt(n()),
    .groups   = "drop"
  ) |>
  ggplot(aes(x = measurement_hour, y = mean_temp,
             color = urban_setting, fill = urban_setting)) +
  geom_ribbon(aes(ymin = mean_temp - se_temp, ymax = mean_temp + se_temp),
              alpha = 0.15, color = NA) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5) +
  scale_color_manual(values = setting_pal, name = "Urban setting") +
  scale_fill_manual(values  = setting_pal, guide = "none") +
  scale_x_continuous(breaks = 9:17,
                     labels = paste0(9:17, ":00")) +
  labs(title    = "Diurnal Profile of IR Surface Temperature",
       subtitle = "Mean ± 1 SE by hour of day",
       x = "Hour of day", y = "Surface temperature (°C)") +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))
```

### 5.3 Diurnal pattern by income group

```{r p2-diurnal-income, fig.height=4}
ir_raw |>
  group_by(demographic_group, measurement_hour) |>
  summarise(
    mean_temp = mean(ir_surface_temperature_c),
    se_temp   = sd(ir_surface_temperature_c) / sqrt(n()),
    .groups   = "drop"
  ) |>
  ggplot(aes(x = measurement_hour, y = mean_temp,
             color = demographic_group, fill = demographic_group)) +
  geom_ribbon(aes(ymin = mean_temp - se_temp, ymax = mean_temp + se_temp),
              alpha = 0.18, color = NA) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5) +
  scale_color_manual(values = income_pal, name = "Income group") +
  scale_fill_manual(values  = income_pal, guide = "none") +
  scale_x_continuous(breaks = 9:17, labels = paste0(9:17, ":00")) +
  labs(title    = "Diurnal Profile by Income Group",
       subtitle = "Do low-income sites stay hotter throughout the day?",
       x = "Hour of day", y = "Surface temperature (°C)") +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))
```

### 5.4 Canopy cover vs surface temperature by time of day

```{r p2-scatter-tod, fig.height=5}
ggplot(ir_raw,
       aes(x = canopy_cover_percent, y = ir_surface_temperature_c,
           color = demographic_group)) +
  geom_point(alpha = 0.4, size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.1) +
  facet_wrap(~ time_of_day) +
  scale_color_manual(values = income_pal, name = "Income group") +
  labs(title    = "Canopy Cover vs Surface Temperature by Time of Day",
       subtitle = "Separate regression lines per income group",
       x = "Canopy cover (%)", y = "Surface temperature (°C)") +
  guides(color = guide_legend(nrow = 1))
```

### 5.5 Income group × urban setting heatmap

```{r p2-heatmap, fig.height=3.5}
ir_raw |>
  group_by(urban_setting, demographic_group) |>
  summarise(mean_temp = mean(ir_surface_temperature_c), .groups = "drop") |>
  ggplot(aes(x = demographic_group, y = fct_rev(urban_setting),
             fill = mean_temp)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = round(mean_temp, 1)),
            color = "white", fontface = "bold", size = 4) +
  scale_fill_gradient2(low = "#2e86ab", mid = "#f6ae2d", high = "#f26419",
                       midpoint = 42, name = "°C") +
  labs(title = "Mean IR Surface Temperature (°C)",
       subtitle = "Urban setting × income group",
       x = NULL, y = NULL) +
  theme(legend.position = "right")
```

<hr>

## 6. Statistical Analysis — IR Spot Data {#p2-stats}

### 6.1 One-way ANOVA: surface temperature by urban setting

```{r p2-anova-setting}
aov2_setting <- aov(ir_surface_temperature_c ~ urban_setting, data = ir_raw)
tidy(aov2_setting) |>
  mutate(across(where(is.numeric), ~ round(.x, 4)),
         sig = case_when(p.value < 0.001 ~ "***", p.value < 0.01 ~ "**",
                         p.value < 0.05 ~ "*", TRUE ~ "")) |>
  kable(caption = "ANOVA: IR surface temperature ~ urban setting") |>
  kable_styling(full_width = FALSE)
```

### 6.2 One-way ANOVA: surface temperature by income group

```{r p2-anova-income}
aov2_income <- aov(ir_surface_temperature_c ~ demographic_group, data = ir_raw)
tidy(aov2_income) |>
  mutate(across(where(is.numeric), ~ round(.x, 4)),
         sig = case_when(p.value < 0.001 ~ "***", p.value < 0.01 ~ "**",
                         p.value < 0.05 ~ "*", TRUE ~ "")) |>
  kable(caption = "ANOVA: IR surface temperature ~ income group") |>
  kable_styling(full_width = FALSE)
```

### 6.3 Tukey HSD post-hoc: urban setting

```{r p2-tukey}
em2 <- emmeans(aov2_setting, ~ urban_setting)
pairs(em2, adjust = "tukey") |>
  as.data.frame() |>
  mutate(across(where(is.numeric), ~ round(.x, 4)),
         sig = case_when(p.value < 0.001 ~ "***", p.value < 0.01 ~ "**",
                         p.value < 0.05 ~ "*", TRUE ~ "")) |>
  kable(caption = "Tukey HSD: urban setting pairwise comparisons (IR data)") |>
  kable_styling(full_width = FALSE)
```

### 6.4 Kruskal-Wallis: time of day effect

```{r p2-kruskal-tod}
kw_tod <- kruskal.test(ir_surface_temperature_c ~ time_of_day, data = ir_raw)

tibble(
  Test    = "Time of day (Morning / Midday / Afternoon)",
  Chi_sq  = round(kw_tod$statistic, 3),
  df      = kw_tod$parameter,
  p_value = round(kw_tod$p.value,   4),
  Sig     = ifelse(kw_tod$p.value < 0.05, "✓", "—")
) |>
  kable(caption = "Kruskal-Wallis: does surface temperature vary by time of day?") |>
  kable_styling(full_width = FALSE)
```

### 6.5 Simple regression: canopy cover predicts surface temperature

```{r p2-simple-lm}
lm2_canopy <- lm(ir_surface_temperature_c ~ canopy_cover_percent, data = ir_raw)

tidy(lm2_canopy) |>
  mutate(across(where(is.numeric), ~ round(.x, 4))) |>
  kable(caption = "OLS regression: surface temperature ~ canopy cover (%)") |>
  kable_styling(full_width = FALSE)

glance(lm2_canopy) |>
  select(r.squared, adj.r.squared, statistic, p.value, nobs) |>
  mutate(across(where(is.numeric), ~ round(.x, 4))) |>
  kable(caption = "Model fit (simple regression)") |>
  kable_styling(full_width = FALSE)
```

### 6.6 Multiple regression: canopy + hour + setting + income

```{r p2-mlr}
lm2_multi <- lm(ir_surface_temperature_c ~
                  canopy_cover_percent +
                  measurement_hour     +
                  urban_setting        +
                  demographic_group,
                data = ir_raw)

tidy(lm2_multi) |>
  mutate(
    sig = case_when(p.value < 0.001 ~ "***", p.value < 0.01 ~ "**",
                    p.value < 0.05  ~ "*",   TRUE ~ ""),
    across(where(is.numeric), ~ round(.x, 4))
  ) |>
  kable(caption = "Multiple regression: surface temp ~ canopy + hour + setting + income") |>
  kable_styling(full_width = FALSE)

glance(lm2_multi) |>
  select(r.squared, adj.r.squared, AIC, BIC, nobs) |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  kable(caption = "Multiple regression model fit (IR data)") |>
  kable_styling(full_width = FALSE)
```

### 6.7 Model comparison: does adding hour improve fit?

```{r p2-model-compare}
lm2_base   <- lm(ir_surface_temperature_c ~
                   canopy_cover_percent + urban_setting + demographic_group,
                 data = ir_raw)
lm2_withhour <- lm2_multi   # already includes measurement_hour

anova(lm2_base, lm2_withhour) |>
  tidy() |>
  mutate(across(where(is.numeric), ~ round(.x, 4))) |>
  kable(caption = "Likelihood ratio test: does measurement hour improve model fit?") |>
  kable_styling(full_width = FALSE)

tibble(
  Model   = c("Base (canopy + setting + income)",
              "Full (+ measurement hour)"),
  AIC     = round(c(AIC(lm2_base), AIC(lm2_multi)), 2),
  adj_R2  = round(c(summary(lm2_base)$adj.r.squared,
                    summary(lm2_multi)$adj.r.squared), 4)
) |>
  kable(caption = "AIC and adjusted R² comparison") |>
  kable_styling(full_width = FALSE)
```

### 6.8 Interaction: does canopy effect vary by income group?

```{r p2-interaction}
lm2_interact <- lm(ir_surface_temperature_c ~
                     canopy_cover_percent * demographic_group +
                     measurement_hour + urban_setting,
                   data = ir_raw)

tidy(lm2_interact) |>
  filter(str_detect(term, "canopy|demographic|:")) |>
  mutate(
    sig = case_when(p.value < 0.001 ~ "***", p.value < 0.01 ~ "**",
                    p.value < 0.05  ~ "*",   TRUE ~ ""),
    across(where(is.numeric), ~ round(.x, 4))
  ) |>
  kable(caption = "Interaction model: canopy × income group coefficients (selected terms)") |>
  kable_styling(full_width = FALSE)
```

```{r p2-interaction-plot, fig.height=4}
newdata <- expand.grid(
  canopy_cover_percent = seq(0, 95, by = 5),
  demographic_group    = income_lvl,
  measurement_hour     = 13,          # midday
  urban_setting        = "Park"       # reference setting
) |>
  mutate(demographic_group = factor(demographic_group, levels = income_lvl),
         urban_setting     = factor(urban_setting,     levels = setting_lvl))

newdata$pred_temp <- predict(lm2_interact, newdata = newdata)

ggplot(newdata, aes(x = canopy_cover_percent, y = pred_temp,
                   color = demographic_group)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = income_pal, name = "Income group") +
  labs(title    = "Predicted Surface Temperature by Canopy Cover and Income Group",
       subtitle = "From interaction model · Midday · Park setting · 95% CI not shown",
       x = "Canopy cover (%)", y = "Predicted surface temperature (°C)")
```

### 6.9 Residual diagnostics

```{r p2-residuals, fig.height=5, fig.width=9}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
plot(lm2_multi, main = "IR Multiple Regression Diagnostics")
par(mfrow = c(1, 1))
```

<div class="callout callout-interp2">
<strong>📊 Key interpretations — IR surface temperature</strong>
The multiple regression coefficient on <code>canopy_cover_percent</code> gives
the average surface temperature reduction (°C) per additional percent of canopy,
<em>holding hour, setting, and income constant</em>. A significant interaction
between canopy and income group would mean the canopy benefit is distributed
unequally — e.g., the cooling effect per percent canopy differs between
high- and low-income sites. Compare model AIC to confirm whether adding
measurement hour materially improves fit.
</div>

<hr>

## Synthesis {#synthesis}

```{r synthesis-compare, fig.height=4}
# Side-by-side: income group means for both instruments
p_log_inc <- log_raw |>
  group_by(demographic_group) |>
  summarise(mean = mean(mean_air_temperature_c),
            se   = sd(mean_air_temperature_c) / sqrt(n()), .groups = "drop") |>
  ggplot(aes(x = demographic_group, y = mean, ymin = mean - se, ymax = mean + se,
             color = demographic_group)) +
  geom_pointrange(size = 1) +
  scale_color_manual(values = income_pal, guide = "none") +
  labs(title = "Logger: air temp by income", x = NULL, y = "Mean temp (°C)") +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

p_ir_inc <- ir_raw |>
  group_by(demographic_group) |>
  summarise(mean = mean(ir_surface_temperature_c),
            se   = sd(ir_surface_temperature_c) / sqrt(n()), .groups = "drop") |>
  ggplot(aes(x = demographic_group, y = mean, ymin = mean - se, ymax = mean + se,
             color = demographic_group)) +
  geom_pointrange(size = 1) +
  scale_color_manual(values = income_pal, guide = "none") +
  labs(title = "IR spot: surface temp by income", x = NULL, y = "Mean temp (°C)") +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

p_log_inc + p_ir_inc
```

```{r synthesis-canopy-compare, fig.height=4}
# Canopy slopes comparison
p_slope_log <- ggplot(log_raw,
                      aes(x = canopy_cover_percent,
                          y = mean_air_temperature_c,
                          color = demographic_group)) +
  geom_point(alpha = 0.4, size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  scale_color_manual(values = income_pal, name = "Income") +
  labs(title = "Logger: canopy vs air temp",
       x = "Canopy (%)", y = "Air temp (°C)")

p_slope_ir <- ggplot(ir_raw,
                     aes(x = canopy_cover_percent,
                         y = ir_surface_temperature_c,
                         color = demographic_group)) +
  geom_point(alpha = 0.25, size = 1.2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  scale_color_manual(values = income_pal, name = "Income") +
  labs(title = "IR: canopy vs surface temp",
       x = "Canopy (%)", y = "Surface temp (°C)")

p_slope_log + p_slope_ir +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

<div class="key-box">
<h4>🔑 Key takeaways</h4>
<ul>
<li><strong>Logger data (air temperature)</strong> captures the chronic 24-hour thermal burden. ANOVA tests whether urban settings differ in mean air temperature; regression quantifies how much canopy cover predicts that temperature.</li>
<li><strong>IR spot data (surface temperature)</strong> captures instantaneous peak heat exposure and diurnal dynamics. Surface temperatures are substantially higher than air temperatures and are more sensitive to canopy cover, time of day, and impervious surface type.</li>
<li><strong>Environmental justice signal</strong>: if low-income sites systematically have lower canopy cover and higher temperatures — and if the income group coefficient remains significant after controlling for canopy and setting — that is evidence of disproportionate heat burden.</li>
<li><strong>Canopy effect</strong>: a negative coefficient on <code>canopy_cover_percent</code> in both regressions confirms cooling. Compare the coefficient magnitude between the two instruments and across income groups to assess equity in the cooling benefit.</li>
<li><strong>Interaction terms</strong>: a significant canopy × income interaction means the cooling benefit per percent canopy differs by income group — the most direct test of whether greening investments would have equitable outcomes.</li>
</ul>
</div>

```{r session-info, echo=FALSE}
# sessionInfo()
```

<footer class="page-footer">
EA030 Urban Ecology &mdash; Field Analysis
<span class="divider">|</span> Prof. Marc Los Huertos
<span class="divider">|</span> Pomona College
<span class="divider">|</span> Last updated: `r format(Sys.Date(), "%B %Y")`
<span class="divider">|</span> <a href="https://github.com/marclos/RTricks">GitHub</a>
</footer>
