---
title: "Visualizating pHake Lake Temperature Profiles"
author: "Marc Los Huertos"
date: "2024-01-18"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Downloading and Uploading Data

1. Download the data from Google Sheets as a csv file (Note where on your computer the file is saved!)
2. Start Rstudio Server (rstudio.pomona.edu)
3. Create New Folder called pHake_Lake
4. Upload data into pHake_Lake folder

## Import into R

1. Create an R Markdown file and give it a title and select pdf for the output.
2. Save the file as R_pHakeLakeTempProfile.Rmd
3. Use file.choose() in the console to select the file and path of the csv.
4. Using your path, change the code below to match your path and file name, where you are specifying path to the YOURDATA.csv object.
5. Run the modified code below to import into R.

```{r importCSV, echo=FALSE}
temp.csv <- "/home/mwl04747/pHake_Lake/G0.csv" # change this to your path...
temp = read.csv(temp.csv) # This creates a data frame in R called temp

```

6. If the code works, add a comment line for each function to explain what it does. Comments begin with a #.
7. Optional: you can hide code and text that isn't necessary.

## Check the Data have been imported correctly

Run each line below to see if the data were read correctly!

```{r checkData, echo=FALSE, results='hide'}
ls() # list the object in the environment
names(temp) # list the column names
str(temp) # list the structure of the data frame
head(temp) # list the first 6 rows of the data frame

```

## Plot the Data

What is the equation for a line?

y = mx + b

y = function of x

m = slope, b = intercept, errors = e    

y ~ x



```{r plotData, echo=FALSE}
#plot(Temperature~Depth, data=temp)
#plot(-Depth~Temperature, data=temp)

#plot(-Depth~Temperature, data=temp, las=1, ty="l")

plot(Depth~Temperature, data=temp, las=1, ty="l", ylim=c(14,0), main="pHake Lake Temperature Profile (Jan 16, 2024)")
```
