---
title: "Regional Climate Narratives"
author: "Marc Los Huertos"
date: "EA030 · Compiled `r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    self_contained: true
    theme: null
    highlight: null
    toc: false
knit: (function(inputFile, encoding) {
    rmarkdown::render(
      inputFile,
      encoding = encoding,
      output_file = file.path(dirname(inputFile),
                              "../docs/Activities/A04_Regional_Climate_Narratives.html")
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  eval      = FALSE,   # handout code is display-only; nothing runs on knit
  warning   = FALSE,
  message   = FALSE,
  fig.width  = 7,
  fig.height = 4,
  fig.align  = "center"
)
```

```{r build_student_bundle, eval=TRUE, include=FALSE}
## ---------------------------------------------------------------------------
## Build the student install bundle during knit.
##
## 1. Finds the newest Regional_Climate_Narratives_<ver>.tar.gz near this Rmd
##    (searches the Rmd's folder, its parent, and parent/dist if present)
## 2. If the package SOURCE is also present (a DESCRIPTION file one level
##    up), rebuilds the tar.gz via R CMD build when it is older than R/,
##    man/, DESCRIPTION, or NAMESPACE. With no source, the existing
##    tar.gz is used as-is.
## 3. Stages the tar.gz plus install_package.R, example_workflow.R, and
##    the package reference into a folder
## 4. Zips them (flat, no wrapper folder) to docs/downloads/
##    Regional_Climate_Narratives--Student_Files.zip
##
## The handout links to the zip with a path relative to docs/Activities/,
## so the link works on the published site.
## ---------------------------------------------------------------------------

rmd_dir  <- dirname(knitr::current_input(dir = TRUE))
pkg_root <- normalizePath(file.path(rmd_dir, ".."))

## --- find the tarball ------------------------------------------------------
search_dirs <- unique(c(rmd_dir, pkg_root, file.path(pkg_root, "dist")))
search_dirs <- search_dirs[dir.exists(search_dirs)]

tarballs <- unlist(lapply(search_dirs, list.files,
                          pattern = "^Regional_Climate_Narratives_.*\\.tar\\.gz$",
                          full.names = TRUE))
if (length(tarballs) == 0 && !file.exists(file.path(pkg_root, "DESCRIPTION"))) {
  stop("No Regional_Climate_Narratives_*.tar.gz found in:\n  ",
       paste(search_dirs, collapse = "\n  "),
       "\nand no package source (DESCRIPTION) to build one from.")
}

## newest version wins if several are lying around (e.g., 0.3.0 and 0.3.1)
tb_versions <- sub("^Regional_Climate_Narratives_(.*)\\.tar\\.gz$", "\\1", basename(tarballs))
if (length(tarballs) > 0) {
  ord     <- order(as.numeric_version(tb_versions), decreasing = TRUE)
  tarball <- tarballs[ord[1]]
} else {
  tarball <- NULL
}

## --- rebuild from source, but only if source exists and tarball is stale ---
has_source <- file.exists(file.path(pkg_root, "DESCRIPTION"))
if (has_source) {
  src_files <- c(
    list.files(file.path(pkg_root, "R"),   full.names = TRUE),
    list.files(file.path(pkg_root, "man"), full.names = TRUE),
    file.path(pkg_root, c("DESCRIPTION", "NAMESPACE"))
  )
  src_files <- src_files[file.exists(src_files)]

  tarball_stale <- is.null(tarball) ||
    max(file.mtime(src_files)) > file.mtime(tarball)

  if (tarball_stale) {
    src_version <- unname(read.dcf(file.path(pkg_root, "DESCRIPTION"))[1, "Version"])
    owd <- setwd(pkg_root)                 # R CMD build writes to getwd()
    status <- system2("R", c("CMD", "build", "--no-build-vignettes",
                             shQuote(pkg_root)),
                      stdout = TRUE, stderr = TRUE)
    setwd(owd)
    tarball <- file.path(pkg_root,
                         sprintf("Regional_Climate_Narratives_%s.tar.gz", src_version))
    if (!file.exists(tarball)) {
      stop("R CMD build did not produce ", basename(tarball), ":\n",
           paste(status, collapse = "\n"))
    }
  }
}

tarball_name <- basename(tarball)

## True version comes from the DESCRIPTION inside the tarball, not the
## filename (a renamed file can disagree with its contents). Fall back to
## the filename version only if the archive cannot be read.
version_from_tarball <- function(tb) {
  contents <- tryCatch(utils::untar(tb, list = TRUE), error = function(e) NULL)
  descs <- grep("^[^/]+/DESCRIPTION$", contents, value = TRUE)
  if (length(descs) == 0) return(NA_character_)
  exd <- file.path(tempdir(), "tb_desc"); unlink(exd, recursive = TRUE)
  ok <- tryCatch({ utils::untar(tb, files = descs[1], exdir = exd); TRUE },
                 error = function(e) FALSE)
  if (!ok) return(NA_character_)
  unname(read.dcf(file.path(exd, descs[1]))[1, "Version"])
}

fname_version <- sub("^Regional_Climate_Narratives_(.*)\\.tar\\.gz$", "\\1", tarball_name)
pkg_version   <- version_from_tarball(tarball)
if (is.na(pkg_version)) pkg_version <- fname_version
if (!identical(pkg_version, fname_version)) {
  cat("\nNOTE [build_student_bundle]: tarball filename says v", fname_version,
      " but its DESCRIPTION says v", pkg_version,
      ".\nHandout will show v", pkg_version,
      " (the version that actually installs).\n",
      "Bump Version in DESCRIPTION and rebuild, or rename the file.\n\n",
      sep = "", file = stderr())
}

## --- stage the bundle ------------------------------------------------------
## companion files may sit next to the Rmd or at pkg_root
find_near <- function(fname) {
  hits <- file.path(search_dirs, fname)
  hits <- hits[file.exists(hits)]
  if (length(hits) > 0) hits[1] else NA_character_
}

zip_name <- "Regional_Climate_Narratives--Student_Files.zip"
stage    <- file.path(tempdir(), "student_bundle_stage")
unlink(stage, recursive = TRUE)
dir.create(stage)

bundle_files <- c(
  tarball,
  find_near("install_package.R"),
  find_near("example_workflow.R"),
  find_near("Regional_Climate_Narratives--Package_Reference.md")
)
missing_files <- is.na(bundle_files)
if (any(missing_files)) {
  warning("Bundle is missing: ",
          paste(c("install_package.R", "example_workflow.R",
                  "Regional_Climate_Narratives--Package_Reference.md")[missing_files[-1]],
                collapse = ", "))
}
invisible(file.copy(bundle_files[!missing_files], stage))

## --- zip into docs/downloads/ ----------------------------------------------
## NOTE: files sit at the zip ROOT (no wrapper folder). Students create
## ~/regional_climate_narratives themselves and upload the zip there; RStudio
## Server auto-extracts, so the files land directly in that folder.
downloads_dir <- file.path(pkg_root, "docs", "downloads")
dir.create(downloads_dir, recursive = TRUE, showWarnings = FALSE)
zip_path <- file.path(downloads_dir, zip_name)
unlink(zip_path)

owd <- setwd(stage)                          # zip from inside the stage so
utils::zip(zip_path,                         # there is no top-level folder
           files = list.files(stage),
           flags = "-9Xq")
setwd(owd)

if (!file.exists(zip_path)) stop("Failed to create ", zip_path)
zip_size_mb <- round(file.size(zip_path) / 1024^2, 1)
```

```{css, echo=FALSE, eval=TRUE}
@import url('https://fonts.googleapis.com/css2?family=Libre+Baskerville:ital,wght@0,400;0,700;1,400&family=Source+Code+Pro:wght@400;600&family=Lato:wght@300;400;700&display=swap');

*, *::before, *::after { box-sizing: border-box; margin: 0; padding: 0; }

:root {
  --forest:    #1e3a2f;
  --moss:      #2d5a3d;
  --fern:      #4a7c5e;
  --sage:      #8aad96;
  --mist:      #d4e6da;
  --cream:     #f7f3ed;
  --parchment: #ede8df;
  --ink:       #1a1a18;
  --charcoal:  #3a3a38;
  --dust:      #7a7a72;
  --amber:     #c8832a;
  --amber-lt:  #e8a84a;
  --white:     #ffffff;
  --card-shadow: 0 2px 10px rgba(30,58,47,.09);
}

html { scroll-behavior: smooth; }
body {
  font-family: 'Lato', sans-serif;
  background: var(--cream);
  color: var(--ink);
  font-size: 17px;
  line-height: 1.75;
  padding-left:  clamp(14px, 4vw, 48px);
  padding-right: clamp(14px, 4vw, 48px);
  padding-bottom: 80px;
}
.main-container { max-width: 920px !important; margin: 0 auto !important; padding: 0 clamp(10px, 2vw, 32px) 80px !important; }
.page-header {
  margin-left:  calc(-1 * clamp(14px, 4vw, 48px));
  margin-right: calc(-1 * clamp(14px, 4vw, 48px));
}
h1.title, h4.author, h4.date { display: none; }

.page-header {
  background-image:
    repeating-linear-gradient(120deg, transparent, transparent 60px, rgba(255,255,255,.018) 60px, rgba(255,255,255,.018) 61px),
    linear-gradient(160deg, #1e3a2f 0%, #2d5a3d 60%, #1a3d30 100%);
  color: #fff; padding: 44px 4vw 36px; margin: 0 -4vw 48px;
  border-bottom: 4px solid var(--amber); position: relative; overflow: hidden;
}
.page-header::after {
  content: ''; position: absolute; right: -40px; top: -40px;
  width: 260px; height: 260px; border-radius: 50%;
  background: radial-gradient(circle, rgba(138,173,150,.12) 0%, transparent 70%);
  pointer-events: none;
}
.page-header .module-tag {
  display: inline-block; font-family: 'Source Code Pro', monospace;
  font-size: .68rem; font-weight: 600; letter-spacing: .18em; text-transform: uppercase;
  color: var(--amber-lt); border: 1px solid rgba(200,131,42,.45);
  padding: 3px 10px; border-radius: 2px; margin-bottom: 12px;
}
.page-header h1 {
  font-family: 'Libre Baskerville', serif; font-size: clamp(1.6rem, 3vw, 2.3rem);
  font-weight: 700; color: #fff; line-height: 1.2; margin-bottom: 8px;
}
.page-header h1 span { color: var(--sage); }
.page-header .sub { font-size: .9rem; color: rgba(255,255,255,.65); font-weight: 300; }
.page-header .sub strong { color: rgba(255,255,255,.9); font-weight: 700; }
.back-link {
  display: inline-block; margin-top: 16px;
  font-family: 'Source Code Pro', monospace; font-size: .7rem; letter-spacing: .08em;
  color: var(--amber-lt); text-decoration: none;
  border: 1px solid rgba(200,131,42,.4); padding: 4px 11px; border-radius: 3px;
  transition: background .2s;
}
.back-link:hover { background: rgba(200,131,42,.18); }

.section-nav {
  display: flex; gap: 8px; flex-wrap: wrap; margin-bottom: 44px;
  padding-bottom: 20px; border-bottom: 1px solid var(--parchment);
}
.section-nav a {
  font-family: 'Source Code Pro', monospace; font-size: .7rem; font-weight: 600;
  letter-spacing: .08em; text-transform: uppercase;
  color: var(--fern); text-decoration: none;
  border: 1px solid var(--mist); padding: 4px 12px; border-radius: 3px;
  background: var(--white); transition: background .2s, color .2s;
}
.section-nav a:hover { background: var(--mist); color: var(--forest); }

h2 {
  font-family: 'Libre Baskerville', serif; font-size: 1.45rem; font-weight: 700;
  color: var(--forest); margin-top: 56px; margin-bottom: 6px;
  padding-bottom: 8px; border-bottom: 1px solid var(--parchment);
}
h3 {
  font-family: 'Libre Baskerville', serif; font-size: 1.1rem; font-weight: 700;
  color: var(--moss); margin-top: 28px; margin-bottom: 4px;
}
h4 {
  font-family: 'Source Code Pro', monospace; font-size: .9rem; font-weight: 600;
  color: var(--charcoal); margin-top: 20px; margin-bottom: 4px;
  letter-spacing: .04em; text-transform: uppercase;
}

p { margin-bottom: 1em; }
strong { color: var(--forest); }
ol, ul { padding-left: 1.5em; margin-bottom: 1em; }
li { margin-bottom: .35em; }
blockquote {
  border-left: 4px solid var(--sage); margin: 1.2em 0;
  padding: .5em 1em; background: var(--white);
  border-radius: 0 5px 5px 0; font-size: .94em; color: var(--charcoal);
}

code {
  font-family: 'Source Code Pro', monospace; font-size: .84em;
  background: var(--mist); color: var(--moss); padding: 1px 5px; border-radius: 3px;
}
pre {
  background: var(--ink); color: #e8e6df; border-radius: 5px;
  padding: 14px 18px; font-size: .83rem; line-height: 1.55;
  overflow-x: auto; border-left: 4px solid var(--amber); margin: 1.2em 0;
}
pre code { background: transparent; color: inherit; padding: 0; font-size: 1em; }

.callout {
  border-left: 5px solid; padding: .65em 1.1em;
  margin: 1.4em 0; border-radius: 0 5px 5px 0; font-size: .92em; line-height: 1.6;
}
.callout > strong:first-child { display: block; margin-bottom: .3em; }
.callout-concept { border-color: var(--fern);  background: #f0f7f2; }
.callout-tip     { border-color: #2e7d32;      background: #f1f8f1; }
.callout-warn    { border-color: #e65100;      background: #fff8f4; }
.callout-new     { border-color: #1565c0;      background: #f0f4ff; }
.callout-formula { border-color: var(--amber); background: #fdf8f0;
                   font-family: 'Source Code Pro', monospace; font-size: .88em; }

.outcomes-box {
  background: var(--white); border: 1px solid var(--mist);
  border-left: 5px solid var(--fern); border-radius: 0 6px 6px 0;
  padding: 18px 22px; margin: 0 0 2.5em;
}
.outcomes-box h4 {
  font-family: 'Libre Baskerville', serif; color: var(--forest);
  font-size: 1rem; margin-bottom: 10px; text-transform: none; letter-spacing: 0;
}
.outcomes-box ol { padding-left: 1.4em; }
.outcomes-box li { margin-bottom: .45em; font-size: .92em; }

.step-grid {
  display: grid; grid-template-columns: repeat(auto-fill, minmax(200px, 1fr));
  gap: 10px; margin: 1.4em 0 2em;
}
.step-card {
  background: var(--white); border: 1px solid var(--mist);
  border-top: 3px solid var(--fern); border-radius: 4px;
  padding: 12px 14px; box-shadow: var(--card-shadow);
}
.step-card.s2 { border-top-color: var(--amber); }
.step-card.s3 { border-top-color: #2e7d94; }
.step-card.s4 { border-top-color: #5d4037; }
.sc-num {
  font-family: 'Source Code Pro', monospace; font-size: .68rem; font-weight: 700;
  letter-spacing: .12em; text-transform: uppercase; color: var(--dust); margin-bottom: 4px;
}
.sc-title { font-family: 'Libre Baskerville', serif; font-size: .93rem;
            font-weight: 700; color: var(--forest); margin-bottom: 3px; }
.sc-desc  { font-size: .82rem; color: var(--charcoal); line-height: 1.4; }

table { font-size: .9rem; border-collapse: collapse; width: 100%; margin: 1.2em 0; }
thead th {
  background: var(--forest); color: #fff;
  font-family: 'Source Code Pro', monospace; font-size: .76rem;
  letter-spacing: .06em; text-transform: uppercase; padding: 8px 12px; text-align: left;
}
tbody tr:nth-child(even) { background: var(--mist); }
tbody td { padding: 7px 12px; border-bottom: 1px solid var(--parchment); vertical-align: top; }

.key-box {
  background: var(--forest); color: #fff; border-radius: 6px;
  padding: 20px 24px; margin: 2em 0; font-size: .9rem;
}
.key-box h4 {
  font-family: 'Libre Baskerville', serif; color: var(--sage);
  font-size: 1rem; margin-bottom: 10px; text-transform: none; letter-spacing: 0;
}
.key-box ul { padding-left: 1.3em; }
.key-box li { margin-bottom: 5px; color: rgba(255,255,255,.85); }
.key-box strong { color: var(--amber-lt); }
.key-box code { background: rgba(255,255,255,.12); color: var(--amber-lt); }

.discuss-box {
  background: var(--parchment); border: 1px solid var(--mist);
  border-radius: 6px; padding: 18px 22px; margin: 2em 0; font-size: .92em;
}
.discuss-box h4 {
  font-family: 'Libre Baskerville', serif; color: var(--moss);
  font-size: 1rem; margin-bottom: 10px; text-transform: none; letter-spacing: 0;
}
.discuss-box ol { padding-left: 1.4em; }
.discuss-box li { margin-bottom: .5em; }

.page-footer {
  background: var(--ink); color: rgba(255,255,255,.42); text-align: center;
  padding: 24px 4vw; margin: 60px -4vw 0;
  font-size: .76rem; font-family: 'Source Code Pro', monospace; letter-spacing: .04em;
}
.page-footer a { color: var(--sage); text-decoration: none; }
.page-footer .divider { margin: 0 8px; opacity: .3; }
```

<div class="page-header">
<div class="module-tag">EA030 &mdash; RTricks &mdash; Activity 04</div>
<h1>Regional Climate Narratives: <span>Mapping a Century of NOAA Station Trends</span></h1>
<p class="sub"><strong>Prof. Marc Los Huertos</strong> &nbsp;&bull;&nbsp; Pomona College &nbsp;&bull;&nbsp; Environmental Analysis</p>
<a class="back-link" href="../index.html">&#8592; Back to RTricks</a>
</div>

<div class="section-nav">
<a href="#download">Download</a>
<a href="#install">Install</a>
<a href="#data">Get Data</a>
<a href="#analysis">Analysis</a>
<a href="#maps">Heat Maps</a>
<a href="#results">Results</a>
<a href="#reference">Reference</a>
<a href="#trouble">Troubleshooting</a>
</div>

<div class="outcomes-box">
<h4>&#127919; Learning Outcomes</h4>
<ol>
<li>Install an R source package (tar.gz) and its dependencies on RStudio Server.</li>
<li>Download and manage a multi-decade NOAA GHCN daily weather dataset for an assigned state.</li>
<li>Explain climate normals, anomalies, and trends, and how a baseline period shapes each.</li>
<li>Compute per-station temperature and precipitation trends with <code>process_all_stations()</code>.</li>
<li>Create interpolated heat maps of annual and seasonal trends with <code>create_heatmap()</code>.</li>
<li>Interpret spatial and seasonal patterns of climate change to support a climate narrative.</li>
</ol>
</div>

> **What this activity covers.** Every weather station tells a hundred-year story, but a single thermometer cannot show *where* and *when* a state is changing fastest. In this activity you will install the **Regional Climate Narratives** package, `RegionalClimateNarratives` (v`r pkg_version`), download daily records for your assigned state's 50 longest-running NOAA stations, convert raw observations into normals, anomalies, and century-scale trends, and map those trends across your state. The maps become the evidence base for your climate narrative video.

<div class="step-grid">
<div class="step-card"><div class="sc-num">Part 1 &middot; once ever</div><div class="sc-title">Install</div><div class="sc-desc">Upload the zip, install dependencies and the package.</div></div>
<div class="step-card s2"><div class="sc-num">Part 2 &middot; once per state</div><div class="sc-title">Get Data</div><div class="sc-desc">Select 50 stations and download their NOAA records.</div></div>
<div class="step-card s3"><div class="sc-num">Part 3 &middot; every session</div><div class="sc-title">Analyze</div><div class="sc-desc">Reload data, compute trends, build spatial objects.</div></div>
<div class="step-card s4"><div class="sc-num">Part 3 &middot; deliverable</div><div class="sc-title">Map</div><div class="sc-desc">Make annual and seasonal heat maps for your narrative.</div></div>
</div>

---

## Download the Course Files {#download}

<div class="callout callout-new">
<strong>&#128230; Download first: <a href="../downloads/Regional_Climate_Narratives--Student_Files.zip">Regional_Climate_Narratives--Student_Files.zip</a> (`r zip_size_mb` MB)</strong>
The zip contains everything you need to upload to the RStudio Server:
<ul>
<li><code>`r tarball_name`</code> &ndash; the Regional Climate Narratives package installer</li>
<li><code>install_package.R</code> &ndash; a script that installs all required packages for you, with clear pass/fail reporting</li>
<li><code>example_workflow.R</code> &ndash; a complete worked example you can adapt</li>
<li><code>Regional_Climate_Narratives--Package_Reference.md</code> &ndash; reference for every function</li>
</ul>
Save the zip somewhere you can find it (Downloads is fine). In Part 1 you will create a <code>regional_climate_narratives</code> folder on the server and upload the zip there.
</div>

You also need:

- Access to RStudio Server (your instructor will provide the URL and login)
- Your assigned **two-letter state code** (e.g., `CA` for California, `TX` for Texas)
- An internet connection (the server needs it to download weather data from NOAA)

---

## Part 1: Install the Package {#install}

You only do Part 1 once, the very first time. After the package is installed, you never need to repeat these steps.

### Step 1 - Log in to RStudio Server

Open your web browser and go to the RStudio Server URL your instructor gave you. Log in with your username and password. You should see the RStudio interface with a Console pane on the left.

### Step 2 - Create a folder and upload the zip file

1. In the **Files** pane (bottom right), click the Home icon (the little house)
2. Click **New Folder** and name it exactly `regional_climate_narratives` (all lowercase, with underscores), then click **OK**
3. Click the `regional_climate_narratives` folder to open it
4. Click the **Upload** button (it has an upward arrow icon)
5. Click **Choose File**, select `Regional_Climate_Narratives--Student_Files.zip` from your computer, and click **OK**

RStudio Server automatically unzips the file for you, so the files land directly inside `regional_climate_narratives`.

<div class="callout callout-tip">
<strong>&#9989; Check</strong>
The <code>regional_climate_narratives</code> folder should now contain <code>`r tarball_name`</code>, <code>install_package.R</code>, <code>example_workflow.R</code>, and the package reference. If you instead see the files sitting in your Home folder, you uploaded before opening the folder in step 3; move them into <code>regional_climate_narratives</code> (check the boxes, then More &gt; Move).
</div>

### Step 3 - Point R at that folder

In the **Console** (bottom left pane), type:

```{r}
setwd("~/regional_climate_narratives")
```

Then verify the installer file is there:

```{r, code=sprintf('file.exists("%s")', tarball_name)}
```

This **must** print `TRUE`. If it prints `FALSE`:

- Did you upload the zip? (Go back to Step 2.)
- Are the files inside `regional_climate_narratives`? Check the Files pane (see the Step 2 check).
- Did you type the `setwd(...)` line exactly, including underscores? Run `getwd()` to see where R is looking.

<div class="callout callout-warn">
<strong>&#9888;&#65039; Do not continue until <code>file.exists(...)</code> returns <code>TRUE</code>.</strong>
Every installation problem in Step 5 traces back to this check being skipped.
</div>

### Step 4 - Install the required packages (dependencies)

The package needs several other R packages to work. Install them first. Copy and paste this entire block into the Console and press Enter:

```{r}
install.packages(c(
  "dplyr", "tidyr", "ggplot2", "lubridate",
  "sp", "sf", "gstat", "raster", "stars",
  "viridis", "maps", "mapdata", "patchwork"
))
```

This takes several minutes. You will see a lot of text scrolling by; that is normal. Wait until you see the `>` prompt again before continuing.

<div class="callout callout-tip">
<strong>&#9989; Check</strong>
After it finishes, look for any lines containing <code>had non-zero exit status</code>. That means a package failed to install. If you see it, note the package name and try installing it individually, e.g., <code>install.packages("sf")</code>. If a package keeps failing, tell your instructor; it may need a system library installed on the server.
</div>

### Step 5 - Install the Regional Climate Narratives package

Now install the actual package. Type:

```{r, code=sprintf('install.packages("%s", repos = NULL, type = "source")', tarball_name)}
```

Wait for it to finish. The last lines should include:

```
* DONE (RegionalClimateNarratives)
```

**If you see `ERROR` or `had non-zero exit status` instead:**
Read the error message. It usually names a missing package. Go back to Step 4 and install that package, then try Step 5 again.

### Step 6 - Verify it works

Load the package. Note the name here has no underscores: R package names cannot contain them, so the package is `RegionalClimateNarratives` even though the files use `Regional_Climate_Narratives`:

```{r}
library(RegionalClimateNarratives)
```

You should see this startup message:

```{r, echo=FALSE, eval=TRUE, results='asis'}
cat("```\n",
    "===========================================================\n",
    sprintf(" Regional Climate Narratives v%s\n", pkg_version),
    "===========================================================\n",
    " Regional Climate Trends Analysis and Visualization\n",
    " Type ?RegionalClimateNarratives for help\n",
    "===========================================================\n",
    "```\n", sep = "")
```

**If you see a `!!!` warning about missing packages instead**, run `check_dependencies()`, then try `library(RegionalClimateNarratives)` again.

<div class="callout callout-tip">
<strong>&#9989; Part 1 complete</strong>
If the startup message shows <code>v`r pkg_version`</code>, the package is installed permanently on the server. You will not need to repeat Part 1.
</div>

---

## Part 2: Download Your State's Data {#data}

Do this once per state.

### Step 7 - Initialize your project

```{r}
initialize_project("XX")
```

Replace `XX` with your state code. For example:

```{r}
initialize_project("CA")   # California
initialize_project("TX")   # Texas
initialize_project("NY")   # New York
```

**What this does:**

- Creates a project folder in your home directory (e.g., `~/Regional_Climate_Narratives_CA/`)
- Creates `Data/`, `Output/`, and `Figures/` subfolders inside it
- Downloads a list of available weather stations from NOAA
- Changes your working directory to the project folder (you do NOT need to call `setwd()` yourself)

You should see output ending with `Setup Complete!` and a count of potential stations.

### Step 8 - Select stations

```{r}
select_stations(n = 50)
```

This picks the 50 stations with the longest, most complete records in your state.

### Step 9 - Download data from NOAA

```{r}
download_stations()
```

<div class="callout callout-warn">
<strong>&#9200; This takes 10 to 30 minutes.</strong>
You will see progress for each station. You can keep working in another browser tab while it runs. If it gets interrupted (connection drops, you close your laptop, etc.), just run <code>download_stations()</code> again; it skips stations already downloaded.
</div>

### Step 10 - Save data in efficient format

```{r}
load_and_save_stations(cleanup = TRUE)
```

This converts the downloaded files into a single compressed file that loads much faster in the future. Setting `cleanup = TRUE` deletes the temporary files to save disk space.

**Part 2 is done.** Your data is saved to disk and ready for analysis.

---

## Part 3: Analysis {#analysis}

### What to do every time you open RStudio

When your R session ends (you log out, your session times out, or you restart R), all variables in memory are erased. Your files on disk are fine; they are saved permanently. But R needs to reload them.

<div class="key-box">
<h4>&#128273; Every new session starts with these three lines</h4>
<pre style="border-left-color: var(--amber); margin: 10px 0;"><code>library(RegionalClimateNarratives)
initialize_project("XX")     # your state code, e.g., "CA"
load_stations()</code></pre>
<ul>
<li><strong><code>library(RegionalClimateNarratives)</code></strong> loads the package functions (they are not available until you do this)</li>
<li><strong><code>initialize_project("XX")</code></strong> sets your working directory to <code>~/Regional_Climate_Narratives_XX/</code> and creates the path variables (<code>datafolder</code>, <code>figuresfolder</code>). This replaces any need to call <code>setwd()</code> yourself.</li>
<li><strong><code>load_stations()</code></strong> loads your previously saved weather data from the <code>Data/</code> folder back into memory</li>
</ul>
You do NOT need to call <code>setwd()</code>, re-download anything, or re-install the package.
</div>

### Step 11 - Calculate climate trends

```{r}
trends <- process_all_stations()
```

This runs the full analysis: cleaning dates, converting units, calculating anomalies, and fitting trend lines. It takes a minute or two.

**You may see messages like:**

- `"Station USC00045123: Normals period = 1971-2000"` - This is normal. That station did not have enough data in the standard 1961-1990 baseline period, so the package used a different 30-year window. The station is still included.
- `"Station USC00067890: Cannot compute normals... Skipping."` - That station did not have enough data for any baseline period. It is excluded. A few skipped stations is expected.

Check the results:

```{r}
cat("Stations analyzed:", nrow(trends), "\n")
summary(trends$annual_trend_TMAX)
```

### Step 12 - Create spatial objects

```{r}
create_spatial_objects(trends)
```

This converts your trend data into a spatial format that can be mapped.

---

## Part 3 (continued): Heat Maps {#maps}

### Step 13 - Make heat maps

**Annual maximum temperature trend:**

```{r}
map_tmax <- create_heatmap(
  trends_sp,
  trend_var = "annual_trend_TMAX",
  title = paste(my.state, "- Annual Maximum Temperature Trend"),
  subtitle = "Change in degrees C per 100 years",
  state = my.state,
  colors = "temp"
)
print(map_tmax)
```

**Save it as a high-resolution image:**

```{r}
ggsave(paste0(figuresfolder, "TMAX_annual_", my.state, ".png"),
       map_tmax, width = 10, height = 8, dpi = 300)
```

The file saves into your `Figures/` folder. Find it in the Files pane under `~/Regional_Climate_Narratives_XX/Figures/`.

**Annual minimum temperature trend:**

```{r}
map_tmin <- create_heatmap(
  trends_sp,
  trend_var = "annual_trend_TMIN",
  title = paste(my.state, "- Annual Minimum Temperature Trend"),
  subtitle = "Change in degrees C per 100 years",
  state = my.state,
  colors = "temp"
)
ggsave(paste0(figuresfolder, "TMIN_annual_", my.state, ".png"),
       map_tmin, width = 10, height = 8, dpi = 300)
```

**Precipitation trend** (uses a different color scheme):

```{r}
map_prcp <- create_heatmap(
  trends_sp,
  trend_var = "annual_trend_PRCP",
  title = paste(my.state, "- Annual Precipitation Trend"),
  subtitle = "Change in mm per 100 years",
  state = my.state,
  colors = "precip"
)
ggsave(paste0(figuresfolder, "PRCP_annual_", my.state, ".png"),
       map_prcp, width = 10, height = 8, dpi = 300)
```

### Step 14 - Seasonal comparison (4-panel map)

```{r}
library(patchwork)

winter_map <- create_heatmap(trends_sp, "winter_trend_TMAX",
              "Winter (Dec-Feb)", state = my.state, resolution = 0.15) +
  theme(legend.position = "none")

spring_map <- create_heatmap(trends_sp, "spring_trend_TMAX",
              "Spring (Mar-May)", state = my.state, resolution = 0.15) +
  theme(legend.position = "none")

summer_map <- create_heatmap(trends_sp, "summer_trend_TMAX",
              "Summer (Jun-Aug)", state = my.state, resolution = 0.15) +
  theme(legend.position = "right")

fall_map <- create_heatmap(trends_sp, "fall_trend_TMAX",
            "Fall (Sep-Nov)", state = my.state, resolution = 0.15) +
  theme(legend.position = "none")

seasonal_panel <- (winter_map | spring_map) / (summer_map | fall_map) +
  plot_annotation(
    title = paste("Seasonal TMAX Trends -", my.state),
    subtitle = "Change in degrees C per 100 years by season"
  )

ggsave(paste0(figuresfolder, "Seasonal_TMAX_", my.state, ".png"),
       seasonal_panel, width = 12, height = 10, dpi = 300)
```

### Quick-start template (after Parts 1 and 2 are done)

Copy this block, change the state code on line 2, and run it:

```{r}
library(RegionalClimateNarratives)
initialize_project("CA")     # <-- CHANGE THIS TO YOUR STATE
load_stations()

trends <- process_all_stations()
create_spatial_objects(trends)

map <- create_heatmap(trends_sp, "annual_trend_TMAX",
       paste(my.state, "Annual TMAX Trend"), state = my.state)
print(map)
ggsave(paste0(figuresfolder, "TMAX_", my.state, ".png"),
       map, width = 10, height = 8, dpi = 300)
```

---

## Understanding Your Results {#results}

<div class="callout callout-concept">
<strong>&#127777;&#65039; Normals, anomalies, and trends</strong>
A <strong>normal</strong> is the average value for each month over a 30-year baseline period; the standard is 1961&ndash;1990 (set by the World Meteorological Organization). If a station does not have enough data in that window, the package uses a later period and tells you. An <strong>anomaly</strong> is the observed value minus the normal: positive means warmer (or wetter) than the baseline average, negative means cooler (or drier). A <strong>trend</strong> is the slope of a line fitted through the anomalies over time, reported as change per century.
</div>

<div class="callout callout-formula">
<strong>Anomaly = Observed value &minus; Normal</strong>
TMAX trend of +1.5 degrees C/century = max temperatures increased ~1.5 degrees C over 100 years<br/>
PRCP trend of &minus;25 mm/century = annual precipitation decreased ~25 mm over 100 years
</div>

<div class="discuss-box">
<h4>&#128172; What to look for in your maps</h4>
<ol>
<li>Is your state warming uniformly, or are some regions warming faster?</li>
<li>Is nighttime warming (TMIN) different from daytime warming (TMAX)?</li>
<li>Do seasonal patterns differ? (Many states warm more in winter than summer.)</li>
<li>Is precipitation increasing, decreasing, or mixed across your state?</li>
</ol>
</div>

---

## Reference {#reference}

### Available heat map variables

| Variable | What it shows |
|----------|---------------|
| `annual_trend_TMAX` | Annual maximum temperature trend |
| `annual_trend_TMIN` | Annual minimum temperature trend |
| `annual_trend_PRCP` | Annual precipitation trend |
| `winter_trend_TMAX` | Winter (Dec-Feb) max temp trend |
| `spring_trend_TMAX` | Spring (Mar-May) max temp trend |
| `summer_trend_TMAX` | Summer (Jun-Aug) max temp trend |
| `fall_trend_TMAX` | Fall (Sep-Nov) max temp trend |

Replace `TMAX` with `TMIN` or `PRCP` for min temperature or precipitation versions.

### What happens when you close RStudio

| What | Saved to disk? | What to do next session |
|------|---------------|------------------------|
|  The `RegionalClimateNarratives` package | Yes, installed permanently | Run `library(RegionalClimateNarratives)` |
| Your downloaded weather data | Yes, in `~/Regional_Climate_Narratives_XX/Data/` | Run `load_stations()` |
| Your saved heat map images | Yes, in `~/Regional_Climate_Narratives_XX/Figures/` | Nothing; the PNG files are already saved |
| Variables in memory (`trends`, `my.state`, `map_tmax`, etc.) | No, R clears memory | Re-run `initialize_project()` and `load_stations()` |
| Your working directory setting | No, resets when session ends | `initialize_project()` sets it for you |

### Getting help

```{r}
?initialize_project          # Help for a specific function
?RegionalClimateNarratives           # Package overview
help(package = "RegionalClimateNarratives")   # List all functions
```

If you are stuck, show your instructor the **exact error message**. Copy the red text from the Console.

---

## Troubleshooting {#trouble}

### `file.exists(...)` returns FALSE in Step 3

The installer file is not in your current working directory. Check:

1. Did you upload the zip? (Part 1, Step 2)
2. Are the files inside `regional_climate_narratives`? Check the Files pane.
3. Did you run `setwd("~/regional_climate_narratives")`? Run `getwd()` to see where R is looking.

### "there is no package called ..."

A required package is missing. Install it:

```{r}
install.packages("sf")   # replace with the package name from the error
```

Or install all dependencies at once (copy-paste from Part 1, Step 4). Or run the included installer script, which handles everything with pass/fail reporting:

```{r}
setwd("~/regional_climate_narratives")
source("install_package.R")
```

### "Run initialize_project() first!"

You need to run `initialize_project("XX")` at the start of every R session. This is the most common error. It just means you started a new session and forgot the setup lines.

### "RData file not found"

Your data has not been downloaded yet, or the download did not finish. Go back to Part 2 (Steps 8-10).

### "could not find function 'process_all_stations'"

Either the package is not loaded, or an old version is installed. Try:

```{r}
library(RegionalClimateNarratives)
```

If that does not help, reinstall:

```{r, code=sprintf('setwd("~/regional_climate_narratives")\ninstall.packages("%s", repos = NULL, type = "source")', tarball_name)}
```

Then go to Session > Restart R in the RStudio menu, and try `library(RegionalClimateNarratives)` again.

### Download seems stuck

NOAA servers can be slow. For 50 stations, expect 10 to 30 minutes. If it stops with errors, run `download_stations()` again. It skips stations already downloaded.

### Messages about normals periods

`"Normals period = 1971-2000"` is informational. The station is still analyzed. This is not an error.

### Warning about skipped stations

`"Cannot compute normals for TMAX. Skipping."` means that station did not have enough data. A few skipped stations is expected.

### Heat map looks odd

- Need at least 20 stations (50 is better)
- Try `resolution = 0.15` for smaller states, `resolution = 0.08` for large states
- Make sure `trends_sp` exists: run `create_spatial_objects(trends)`

### Where are my files?

Run `getwd()` to see your current project folder. Your files are in `Data/` (weather data), `Figures/` (saved heat maps), and `Output/` (analysis results). Or navigate in the Files pane (bottom right of RStudio).

<div class="page-footer">
EA030 &middot; Environmental Analysis <span class="divider">|</span> Regional Climate Narratives v`r pkg_version` <span class="divider">|</span> <a href="../index.html">RTricks Home</a> <span class="divider">|</span> Compiled `r format(Sys.Date(), '%Y-%m-%d')`
</div>
