---
title: "EA030 Project Index"
author: "Marc Los Huertos"
date: "Last updated: `r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: null
    highlight: null
    toc: false
    self_contained: true
knit: (function(inputFile, encoding) {
    rmarkdown::render(
      inputFile,
      encoding = encoding,
      output_file = file.path(dirname(inputFile), "../docs/project_index.html")
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{css, echo=FALSE}
@import url('https://fonts.googleapis.com/css2?family=Libre+Baskerville:ital,wght@0,400;0,700;1,400&family=Source+Code+Pro:wght@400;600&family=Lato:wght@300;400;700&display=swap');

*, *::before, *::after { box-sizing: border-box; margin: 0; padding: 0; }

:root {
  --forest:    #1e3a2f;
  --moss:      #2d5a3d;
  --fern:      #4a7c5e;
  --sage:      #8aad96;
  --mist:      #d4e6da;
  --cream:     #f7f3ed;
  --parchment: #ede8df;
  --ink:       #1a1a18;
  --charcoal:  #3a3a38;
  --dust:      #7a7a72;
  --amber:     #c8832a;
  --amber-lt:  #e8a84a;
  --rust:      #8b3a2a;
  --white:     #ffffff;
  --card-shadow: 0 2px 12px rgba(30,58,47,.10), 0 1px 3px rgba(30,58,47,.07);
}

html { scroll-behavior: smooth; }

body {
  font-family: 'Lato', sans-serif;
  background: var(--cream);
  color: var(--ink);
  font-size: 16px;
  line-height: 1.7;
}

/* ── Suppress default rmarkdown chrome ── */
h1.title, h4.author, h4.date { display: none; }

/* ── Main content area ── */
.col-md-9 {
  padding: 0 !important;
}

.main-container {
  max-width: 100% !important;
  padding: 0 !important;
  margin: 0 !important;
}

/* ── Page header ── */
.page-header {
  background-image:
    repeating-linear-gradient(120deg, transparent, transparent 60px,
      rgba(255,255,255,.018) 60px, rgba(255,255,255,.018) 61px),
    linear-gradient(160deg, #1e3a2f 0%, #2d5a3d 60%, #1a3d30 100%);
  color: #fff;
  padding: 48px 48px 40px;
  border-bottom: 4px solid var(--amber);
  position: relative;
  overflow: hidden;
}
.page-header::after {
  content: ''; position: absolute; right: -60px; top: -60px;
  width: 300px; height: 300px; border-radius: 50%;
  background: radial-gradient(circle, rgba(138,173,150,.12) 0%, transparent 70%);
  pointer-events: none;
}
.page-header .course-tag {
  display: inline-block;
  font-family: 'Source Code Pro', monospace;
  font-size: 0.70rem; font-weight: 600;
  letter-spacing: 0.18em; text-transform: uppercase;
  color: var(--amber-lt);
  border: 1px solid rgba(200,131,42,.45);
  padding: 3px 10px; border-radius: 2px; margin-bottom: 14px;
}
.page-header h1 {
  font-family: 'Libre Baskerville', serif;
  font-size: clamp(1.7rem, 3vw, 2.4rem);
  font-weight: 700; line-height: 1.2; color: #fff;
  letter-spacing: -0.01em; margin-bottom: 8px;
}
.page-header h1 span { color: var(--sage); }
.page-header .header-sub {
  font-size: 0.93rem; font-weight: 300;
  color: rgba(255,255,255,.68);
}
.page-header .header-sub strong {
  font-weight: 700; color: rgba(255,255,255,.9);
}
.page-header .back-link {
  display: inline-block;
  margin-top: 16px;
  font-family: 'Source Code Pro', monospace;
  font-size: 0.72rem; letter-spacing: 0.08em;
  color: var(--amber-lt); text-decoration: none;
  border: 1px solid rgba(200,131,42,.4);
  padding: 4px 12px; border-radius: 3px;
  transition: background 0.2s;
}
.page-header .back-link:hover { background: rgba(200,131,42,.18); }

/* ── Body content wrapper ── */
.content-body {
  padding: 44px 48px 72px;
  max-width: 860px;
}

/* ── Section headings ── */
.content-body h2 {
  font-family: 'Libre Baskerville', serif;
  font-size: 1.35rem; font-weight: 700;
  color: var(--forest); letter-spacing: -0.01em;
  margin-top: 48px; margin-bottom: 4px;
  padding-bottom: 10px;
  border-bottom: 1px solid var(--parchment);
}
.content-body h2:first-of-type { margin-top: 0; }

.content-body h3 {
  font-family: 'Libre Baskerville', serif;
  font-size: 1rem; font-weight: 700;
  color: var(--charcoal); margin-top: 20px; margin-bottom: 4px;
}

/* ── Project cards ── */
.project-card {
  background: var(--white);
  border: 1px solid var(--mist);
  border-left: 4px solid var(--amber);
  border-radius: 0 6px 6px 0;
  padding: 20px 24px 18px;
  margin: 20px 0 32px;
  box-shadow: var(--card-shadow);
}

.project-card p { font-size: 0.92rem; color: var(--charcoal); margin-bottom: 10px; }
.project-card p:last-child { margin-bottom: 0; }

.project-card .status {
  display: inline-block;
  font-family: 'Source Code Pro', monospace;
  font-size: 0.68rem; letter-spacing: 0.1em; text-transform: uppercase;
  font-weight: 600; padding: 2px 8px; border-radius: 2px;
}
.status-soon    { background: var(--parchment); color: var(--dust); }
.status-ready   { background: var(--mist);      color: var(--fern); }

.project-card a {
  color: var(--fern); font-weight: 600; text-decoration: none;
  border-bottom: 1px solid var(--mist);
  transition: color 0.2s, border-color 0.2s;
}
.project-card a:hover { color: var(--moss); border-bottom-color: var(--sage); }

/* ── Intro blurb ── */
.intro-blurb {
  background: var(--white);
  border-left: 4px solid var(--fern);
  border-radius: 0 6px 6px 0;
  padding: 18px 22px;
  margin-bottom: 36px;
  font-size: 0.92rem; color: var(--charcoal);
  box-shadow: var(--card-shadow);
}

/* ── Notes box ── */
.notes-box {
  background: var(--forest);
  color: rgba(255,255,255,.82);
  border-radius: 6px;
  padding: 24px 28px;
  margin: 40px 0 20px;
  font-size: 0.88rem;
}
.notes-box h2 {
  color: var(--sage) !important;
  border-bottom-color: rgba(255,255,255,.12) !important;
  font-size: 1rem !important;
  margin-top: 0 !important;
}
.notes-box ul { padding-left: 1.3em; margin-top: 8px; }
.notes-box li { margin-bottom: 4px; }
.notes-box code {
  background: rgba(255,255,255,.1);
  color: var(--amber-lt);
  padding: 1px 5px; border-radius: 3px;
  font-family: 'Source Code Pro', monospace; font-size: 0.82em;
}

/* ── Inline code ── */
code {
  font-family: 'Source Code Pro', monospace;
  font-size: 0.84em;
  background: var(--mist);
  color: var(--moss);
  padding: 1px 5px; border-radius: 3px;
}

/* ── Footer ── */
.page-footer {
  background: var(--ink);
  color: rgba(255,255,255,.42);
  text-align: center;
  padding: 24px 48px;
  font-size: 0.76rem;
  font-family: 'Source Code Pro', monospace;
  letter-spacing: 0.04em;
}
.page-footer a { color: var(--sage); text-decoration: none; }
.page-footer a:hover { color: var(--amber-lt); }
.page-footer .divider { margin: 0 8px; opacity: .3; }

/* ── Horizontal rules ── */
hr { border: none; border-top: 1px solid var(--parchment); margin: 8px 0 0; }
```

<header class="page-header">
<div class="course-tag">EA030 &mdash; Environmental Analysis</div>
<h1>Project <span>Index</span></h1>
<p class="header-sub"><strong>Prof. Marc Los Huertos</strong> &nbsp;&bull;&nbsp; Pomona College</p>
<a class="back-link" href="index.html">&#8592; Back to RTricks</a>
</header>

<div class="content-body">

<div class="intro-blurb">
This page indexes EA030 Environmental Analysis projects for the current semester.
Each section will link to student work, datasets, reports, and interactive content
as those become available. Check back as the term progresses.
</div>

## Welcome

## Regional Climate Trends

<div class="project-card">
<p>A project exploring regional temperature, precipitation, and extreme weather patterns
using publicly available datasets (NOAA, PRISM, etc.).</p>
<p><span class="status status-soon">Coming soon</span></p>
</div>

## BFS / pHake Lake Stratification

<div class="project-card">
<p>Seasonal stratification patterns, temperature profiles, and dissolved oxygen
dynamics at pHake Lake in BFS</p>
<p><span class="status status-soon">Coming soon</span></p>
</div>

## Compost Maturity Analysis

<div class="project-card">
<p>Evaluation of compost temperature, moisture, C:N ratio, and stability indicators
to assess maturity and readiness for field application.</p>
<p><span class="status status-soon">Coming soon</span></p>
</div>

## Inland Empire Air Quality

<div class="project-card">
<p>EPA monitoring and PurpleAir sensor data used to analyze PM2.5 patterns across
the Inland Empire, including diurnal cycles, seasonal variability, and comparisons
with federal health standards.</p>
<p><strong>See also:</strong> <a href="Analyzing_EPA_PM25.html">Analyzing EPA PM2.5 Data</a></p>
<p><span class="status status-ready">Available</span></p>
</div>

## Surface Water Quality

<div class="project-card">
<p>Field sampling and lab-based analysis of nutrients, conductivity, turbidity, and
other indicators across local surface water bodies.</p>
<p><span class="status status-soon">Coming soon</span></p>
</div>

<div class="notes-box">

## Notes for Students

As each project develops, please submit:

- Final HTML report (knitted from your `.Rmd`)
- Key figures and data visualizations
- Analytical scripts
- A brief reflection or summary of findings

Links will be added here as work becomes available. Contact Prof. Los Huertos or
drop files in the shared course folder.

</div>

</div>

```{r footer, results='asis'}
cat(paste0(
'<footer class="page-footer">
  EA030 Project Index
  <span class="divider">|</span> Prof. Marc Los Huertos
  <span class="divider">|</span> Pomona College
  <span class="divider">|</span> Last updated: ', format(Sys.Date(), "%B %Y"), '
  <span class="divider">|</span> <a href="https://github.com/marclos/RTricks">GitHub</a>
</footer>'
))
```
