---
title: "Logistic Regression"
author: "Marc Los Huertos"
date: "EA030 · Compiled `r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    self_contained: true
    theme: null
    highlight: null
    toc: false
knit: (function(inputFile, encoding) {
    rmarkdown::render(
      inputFile,
      encoding = encoding,
      output_file = file.path(dirname(inputFile),
                              "../docs/basic_statistics/09_Logistic_Regression.html")
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 7,
  fig.height = 4
)
library(ggplot2)
library(dplyr)
```

```{css, echo=FALSE}
@import url('https://fonts.googleapis.com/css2?family=Libre+Baskerville:ital,wght@0,400;0,700;1,400&family=Source+Code+Pro:wght@400;600&family=Lato:wght@300;400;700&display=swap');

*, *::before, *::after { box-sizing: border-box; margin: 0; padding: 0; }

:root {
  --forest: #1e3a2f; --moss: #2d5a3d; --fern: #4a7c5e; --sage: #8aad96;
  --mist: #d4e6da; --cream: #f7f3ed; --parchment: #ede8df;
  --ink: #1a1a18; --charcoal: #3a3a38; --dust: #7a7a72;
  --amber: #c8832a; --amber-lt: #e8a84a; --white: #ffffff;
  --card-shadow: 0 2px 10px rgba(30,58,47,.09);
}

html { scroll-behavior: smooth; }
body { font-family: 'Lato', sans-serif; background: var(--cream); color: var(--ink); font-size: 17px; line-height: 1.75; padding-left: clamp(14px, 4vw, 48px); padding-right: clamp(14px, 4vw, 48px); padding-bottom: 80px; }
.main-container { max-width: 920px !important; margin: 0 auto !important; padding: 0 clamp(10px, 2vw, 32px) 80px !important; }
.page-header { margin-left: calc(-1 * clamp(14px, 4vw, 48px)); margin-right: calc(-1 * clamp(14px, 4vw, 48px)); }
h1.title, h4.author, h4.date { display: none; }

.page-header { background-image: repeating-linear-gradient(120deg, transparent, transparent 60px, rgba(255,255,255,.018) 60px, rgba(255,255,255,.018) 61px), linear-gradient(160deg, #1e3a2f 0%, #2d5a3d 60%, #1a3d30 100%); color: #fff; padding: 44px 4vw 36px; margin: 0 0 48px; border-bottom: 4px solid var(--amber); position: relative; overflow: hidden; }
.page-header::after { content: ''; position: absolute; right: -40px; top: -40px; width: 260px; height: 260px; border-radius: 50%; background: radial-gradient(circle, rgba(138,173,150,.12) 0%, transparent 70%); pointer-events: none; }
.page-header .module-tag { display: inline-block; font-family: 'Source Code Pro', monospace; font-size: .68rem; font-weight: 600; letter-spacing: .18em; text-transform: uppercase; color: var(--amber-lt); border: 1px solid rgba(200,131,42,.45); padding: 3px 10px; border-radius: 2px; margin-bottom: 12px; }
.page-header h1 { font-family: 'Libre Baskerville', serif; font-size: clamp(1.6rem, 3vw, 2.3rem); font-weight: 700; color: #fff; line-height: 1.2; margin-bottom: 8px; }
.page-header h1 span { color: var(--sage); }
.page-header .sub { font-size: .9rem; color: rgba(255,255,255,.65); font-weight: 300; }
.page-header .sub strong { color: rgba(255,255,255,.9); font-weight: 700; }
.back-link { display: inline-block; margin-top: 16px; font-family: 'Source Code Pro', monospace; font-size: .7rem; letter-spacing: .08em; color: var(--amber-lt); text-decoration: none; border: 1px solid rgba(200,131,42,.4); padding: 4px 11px; border-radius: 3px; transition: background .2s; }
.back-link:hover { background: rgba(200,131,42,.18); }

.section-nav { display: flex; gap: 8px; flex-wrap: wrap; margin-bottom: 44px; padding-bottom: 20px; border-bottom: 1px solid var(--parchment); }
.section-nav a { font-family: 'Source Code Pro', monospace; font-size: .7rem; font-weight: 600; letter-spacing: .08em; text-transform: uppercase; color: var(--fern); text-decoration: none; border: 1px solid var(--mist); padding: 4px 12px; border-radius: 3px; background: var(--white); transition: background .2s, color .2s; }
.section-nav a:hover { background: var(--mist); color: var(--forest); }

h2 { font-family: 'Libre Baskerville', serif; font-size: 1.45rem; font-weight: 700; color: var(--forest); margin-top: 56px; margin-bottom: 6px; padding-bottom: 8px; border-bottom: 1px solid var(--parchment); }
h3 { font-family: 'Libre Baskerville', serif; font-size: 1.1rem; font-weight: 700; color: var(--moss); margin-top: 28px; margin-bottom: 4px; }
p { margin-bottom: 1em; }
strong { color: var(--forest); }
code { font-family: 'Source Code Pro', monospace; font-size: .84em; background: var(--mist); color: var(--moss); padding: 1px 5px; border-radius: 3px; }
pre { background: var(--ink); color: #e8e6df; border-radius: 5px; padding: 14px 18px; font-size: .83rem; line-height: 1.55; overflow-x: auto; border-left: 4px solid var(--amber); margin: 1.2em 0; }
pre code { background: transparent; color: inherit; padding: 0; font-size: 1em; }

.callout { border-left: 5px solid; padding: .65em 1.1em; margin: 1.4em 0; border-radius: 0 5px 5px 0; font-size: .92em; line-height: 1.6; }
.callout > strong:first-child { display: block; margin-bottom: .3em; }
.callout-concept { border-color: var(--fern);  background: #f0f7f2; }
.callout-tip     { border-color: #2e7d32;      background: #f1f8f1; }
.callout-warn    { border-color: #e65100;      background: #fff8f4; }
.callout-formula { border-color: var(--amber); background: #fdf8f0; font-family: 'Source Code Pro', monospace; font-size: .88em; }

table { font-size: .9rem; border-collapse: collapse; width: 100%; margin: 1.2em 0; }
thead th { background: var(--forest); color: #fff; font-family: 'Source Code Pro', monospace; font-size: .76rem; letter-spacing: .06em; text-transform: uppercase; padding: 8px 12px; text-align: left; }
tbody tr:nth-child(even) { background: var(--mist); }
tbody td { padding: 7px 12px; border-bottom: 1px solid var(--parchment); }

.key-box { background: var(--forest); color: #fff; border-radius: 6px; padding: 20px 24px; margin: 2em 0; font-size: .9rem; }
.key-box h4 { font-family: 'Libre Baskerville', serif; color: var(--sage); font-size: 1rem; margin-bottom: 10px; }
.key-box ul { padding-left: 1.3em; }
.key-box li { margin-bottom: 5px; color: rgba(255,255,255,.85); }
.key-box code { background: rgba(255,255,255,.12); color: var(--amber-lt); }

.page-footer { background: var(--ink); color: rgba(255,255,255,.42); text-align: center; padding: 24px 4vw; margin: 60px calc(-1 * clamp(14px, 4vw, 48px)) 0; font-size: .76rem; font-family: 'Source Code Pro', monospace; letter-spacing: .04em; }
.page-footer a { color: var(--sage); text-decoration: none; }
.page-footer .divider { margin: 0 8px; opacity: .3; }
```

<div class="page-header">
<div class="module-tag">EA030 &mdash; Advanced Statistics &mdash; Module 09</div>
<h1>Logistic <span>Regression</span></h1>
<p class="sub"><strong>Prof. Marc Los Huertos</strong> &nbsp;&bull;&nbsp; Pomona College</p>
<a class="back-link" href="../index.html">&#8592; Back to RTricks</a>
</div>

<div class="section-nav">
<a href="#why">Why Logistic?</a>
<a href="#model">The Model</a>
<a href="#fitting">Fitting in R</a>
<a href="#output">Reading Output</a>
<a href="#oddsratios">Odds Ratios</a>
<a href="#predictions">Predictions</a>
<a href="#diagnostics">Diagnostics</a>
<a href="#roc">ROC & AUC</a>
</div>

> **What this module covers.** When your outcome is binary — present/absent,
> exceeded/not exceeded, survived/died — ordinary linear regression fails because
> it can predict probabilities outside [0, 1]. Logistic regression solves this
> by modeling the *log-odds* of the outcome, producing well-calibrated probabilities
> and interpretable odds ratios. This tutorial covers the full workflow: fitting
> with `glm()`, interpreting coefficients as odds ratios, assessing model fit,
> and evaluating predictive performance with ROC curves.

---

## 1. Why Not Just Use Linear Regression? {#why}

Suppose you want to predict whether a river site exceeds a pollution threshold
(1 = exceeds, 0 = does not) based on upstream land-use intensity. If you fit a
straight line through 0s and 1s, two problems arise immediately:

1. **Out-of-range predictions.** A line can predict probabilities below 0 or
   above 1, which are nonsensical.
2. **Non-constant variance.** The variance of a binary outcome is $p(1-p)$, which
   changes with p — violating the homoscedasticity assumption of linear regression.

Logistic regression maps the linear predictor to a probability using the
**logistic (sigmoid) function**, guaranteeing predictions in (0, 1).

```{r sigmoid-plot, echo=FALSE, fig.cap="The logistic (sigmoid) function maps any real-valued linear predictor to a probability between 0 and 1. The S-shape reflects how the response to increasing x is strongest in the middle of the probability range."}
x_seq <- seq(-6, 6, length.out = 300)
y_sig <- 1 / (1 + exp(-x_seq))

par(las = 1, mar = c(4.5, 4.5, 2.5, 1))
plot(x_seq, y_sig, type = "l", lwd = 2.5, col = "#1e3a2f",
     xlab = "Linear predictor (β₀ + β₁x)", ylab = "P(Y = 1)",
     main = "The Logistic Function: Linear Predictor → Probability",
     ylim = c(-0.05, 1.05))
abline(h = c(0, 0.5, 1), col = c("#c8832a", "#4a7c5e", "#c8832a"),
       lty = c(2, 3, 2), lwd = 1.2)
abline(v = 0, col = "#4a7c5e", lty = 3, lwd = 1.2)
legend("topleft",
       legend = c("Logistic function", "P = 0.5 threshold"),
       col = c("#1e3a2f", "#4a7c5e"), lty = c(1, 3), lwd = c(2.5, 1.2),
       bty = "n", cex = 0.85)
```

---

## 2. The Logistic Regression Model {#model}

The model links the probability of the outcome p to predictors through the **logit**
(log-odds) transformation:

$$\log\left(\frac{p}{1-p}\right) = \beta_0 + \beta_1 x_1 + \cdots + \beta_k x_k$$

Solving for p:

$$p = \frac{1}{1 + e^{-(\beta_0 + \beta_1 x)}} = \text{logistic}(\beta_0 + \beta_1 x)$$

The coefficients $\beta$ are estimated by **maximum likelihood** (not least squares).
The log-likelihood measures how well the model's predicted probabilities match the
observed 0/1 outcomes.

<div class="callout callout-formula">
<strong>Key relationships</strong>

- **Log-odds (logit):** log(p / (1-p)) — ranges from −∞ to +∞
- **Odds:** p / (1-p) — ranges from 0 to +∞
- **Probability:** p = logistic(log-odds) — ranges from 0 to 1
- **Odds ratio (OR):** exp(β) — the multiplicative change in odds per unit increase in x
</div>

---

## 3. Fitting Logistic Regression in R {#fitting}

We use a dataset of 189 low-birth-weight births from `MASS::birthwt`. The outcome is
whether birth weight was below 2.5 kg (`low`: 1 = low, 0 = normal).

```{r data}
library(MASS)
data(birthwt)

# Recode for clarity
bw <- birthwt |>
  mutate(
    low     = factor(low,   labels = c("Normal", "Low")),
    smoke   = factor(smoke, labels = c("Non-smoker", "Smoker")),
    race    = factor(race,  labels = c("White", "Black", "Other")),
    ht      = factor(ht,    labels = c("No hypertension", "Hypertension")),
    low_bin = as.integer(low == "Low")   # keep numeric version for glm
  )

table(bw$low)
```

### Explore predictors first

```{r explore, fig.cap="Birth weight by maternal smoking status. Low birth weight is more common among smokers — consistent with the known physiological pathway.", fig.height=3.5}
bw |>
  count(smoke, low) |>
  group_by(smoke) |>
  mutate(pct = n / sum(n)) |>
  filter(low == "Low") |>
  ggplot(aes(x = smoke, y = pct, fill = smoke)) +
  geom_col(width = 0.5) +
  scale_fill_manual(values = c("#4a7c5e", "#c8832a")) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = NULL, y = "Proportion with low birth weight",
       title = "Low birth weight by smoking status") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

### Fit the model

```{r fit-glm}
mod_logit <- glm(low_bin ~ smoke + age + lwt + race + ht,
                 data   = bw,
                 family = binomial(link = "logit"))
```

The key argument is `family = binomial(link = "logit")`. This tells `glm()` that:
(1) the outcome is binary (Bernoulli distribution), and (2) the link function is the
logit, connecting the linear predictor to the probability.

---

## 4. Reading `summary()` Output {#output}

```{r glm-summary}
summary(mod_logit)
```

The coefficients table has the same columns as `lm()` — Estimate, Std. Error, z value,
Pr(>|z|) — but the test statistic is a z-test (Wald test), not a t-test.

**Deviance.** The logistic model uses *deviance* (−2 × log-likelihood) instead of
residual sums of squares. The Null deviance is for a model with only an intercept;
the Residual deviance is for your model. A substantial drop indicates the predictors
explain something.

**AIC.** Akaike Information Criterion = deviance + 2k. Smaller is better; use AIC
to compare competing models.

<div class="callout callout-warn">
<strong>⚠️ Coefficients are on the log-odds scale</strong>
A coefficient of 0.72 for `smokerSmoker` does not mean smoking increases the
probability by 72 percentage points. It means the log-odds increase by 0.72. To
get the odds ratio, exponentiate: exp(0.72) ≈ 2.05. To get predicted probabilities,
use <code>predict(mod, type = "response")</code>.
</div>

---

## 5. Odds Ratios {#oddsratios}

Odds ratios are the most common effect size for logistic regression. OR > 1 means the
predictor increases the odds of the outcome; OR < 1 means it decreases the odds.

```{r odds-ratios}
# Exponentiate coefficients and confidence intervals
or_table <- exp(cbind(OR = coef(mod_logit), confint(mod_logit)))
round(or_table, 3)
```

```{r or-forest, fig.cap="Odds ratio plot with 95% confidence intervals. OR > 1 increases the odds of low birth weight; OR < 1 decreases them. The vertical dashed line at OR = 1 is the null hypothesis of no effect.", fig.height=4}
or_df <- as.data.frame(or_table[-1, ])   # drop intercept
or_df$term <- rownames(or_df)
names(or_df)[2:3] <- c("ci_lo", "ci_hi")

ggplot(or_df, aes(x = term, y = OR)) +
  geom_point(size = 3, color = "#4a7c5e") +
  geom_errorbar(aes(ymin = ci_lo, ymax = ci_hi),
                width = 0.2, color = "#4a7c5e") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "#c8832a") +
  coord_flip() +
  scale_y_log10() +
  labs(x = NULL, y = "Odds Ratio (log scale)",
       title = "Logistic Regression: Odds Ratios and 95% CI") +
  theme_minimal(base_size = 12)
```

---

## 6. Predicted Probabilities {#predictions}

Unlike linear regression, logistic regression predictions are probabilities, not raw
response values. Use `type = "response"` with `predict()`.

```{r predicted-probs, fig.cap="Predicted probability of low birth weight as a function of maternal weight at last menstrual period, holding other predictors at reference levels. Smoking status shifts the curve upward.", fig.height=4}
# Predictions across maternal weight range, by smoking status
new_data <- expand.grid(
  lwt   = seq(80, 250, by = 2),
  smoke = factor(c("Non-smoker", "Smoker"), levels = levels(bw$smoke)),
  age   = median(bw$age),
  race  = "White",
  ht    = "No hypertension"
)
new_data$pred_prob <- predict(mod_logit, newdata = new_data, type = "response")

ggplot(new_data, aes(x = lwt, y = pred_prob, color = smoke)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("#4a7c5e", "#c8832a")) +
  labs(x = "Maternal weight (lbs)", y = "P(low birth weight)",
       title = "Predicted probability of low birth weight",
       color = "Smoking status") +
  theme_minimal(base_size = 13)
```

---

## 7. Model Diagnostics {#diagnostics}

### Deviance goodness-of-fit

The likelihood ratio test compares your model to the null (intercept-only) model.

```{r lrt}
mod_null <- glm(low_bin ~ 1, data = bw, family = binomial)
anova(mod_null, mod_logit, test = "Chisq")
```

### Pseudo-R² (McFadden)

```{r pseudo-r2}
pseudo_r2 <- 1 - logLik(mod_logit) / logLik(mod_null)
```

McFadden pseudo-R² = **`r round(as.numeric(pseudo_r2), 3)`**.

McFadden values of 0.2–0.4 indicate good fit; above 0.4 is excellent. This is not
directly comparable to OLS R² — it measures relative improvement over the null model.

### Binned residual plot

```{r binned-residuals, fig.cap="Binned residual plot. Bins within the grey bands (±2 SE) suggest adequate fit. Points outside the bands may indicate model misspecification or influential observations.", fig.height=4}
# Manual binned residual plot (avoids arm package dependency)
pred   <- predict(mod_logit, type = "response")
resid  <- bw$low_bin - pred
n_bins <- 20
breaks <- quantile(pred, probs = seq(0, 1, length.out = n_bins + 1))
bin_id <- cut(pred, breaks = breaks, include.lowest = TRUE, labels = FALSE)

bin_df <- data.frame(pred, resid, bin_id) |>
  group_by(bin_id) |>
  summarise(mean_pred = mean(pred), mean_resid = mean(resid),
            se = sd(resid) / sqrt(n()), .groups = "drop")

ggplot(bin_df, aes(x = mean_pred, y = mean_resid)) +
  geom_ribbon(aes(ymin = -2 * se, ymax = 2 * se), alpha = 0.15, fill = "#4a7c5e") +
  geom_point(color = "#1e3a2f", size = 2.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "#c8832a") +
  labs(x = "Predicted probability", y = "Average residual",
       title = "Binned residual plot") +
  theme_minimal(base_size = 12)
```

---

## 8. ROC Curve and AUC {#roc}

The **receiver operating characteristic (ROC) curve** shows how well the model
discriminates between the two outcome classes across all possible probability
thresholds. The **area under the curve (AUC)** summarizes discrimination in a single
number: 0.5 = random guessing; 1.0 = perfect discrimination.

```{r roc, fig.cap="ROC curve for the low birth weight model. AUC > 0.7 indicates acceptable discrimination. The dashed diagonal is a model that guesses randomly (AUC = 0.5).", fig.height=4}
# Compute sensitivity and specificity at each threshold
thresholds <- seq(0, 1, length.out = 300)
pred_prob  <- predict(mod_logit, type = "response")
obs        <- bw$low_bin

roc_df <- data.frame(
  threshold = thresholds,
  sens = sapply(thresholds, function(t) mean((pred_prob >= t)[obs == 1])),
  spec = sapply(thresholds, function(t) mean((pred_prob <  t)[obs == 0]))
)

# AUC via trapezoidal rule
auc <- abs(sum(diff(1 - roc_df$spec) * roc_df$sens[-nrow(roc_df)]))

ggplot(roc_df, aes(x = 1 - spec, y = sens)) +
  geom_line(color = "#1e3a2f", linewidth = 1.2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              color = "#c8832a", linewidth = 0.8) +
  annotate("text", x = 0.7, y = 0.2,
           label = paste0("AUC = ", round(auc, 3)),
           color = "#1e3a2f", size = 4.5, fontface = "bold") +
  labs(x = "1 − Specificity (False Positive Rate)",
       y = "Sensitivity (True Positive Rate)",
       title = "ROC Curve — Low Birth Weight Model") +
  coord_equal() +
  theme_minimal(base_size = 12)
```

### Choosing a classification threshold

The default threshold of 0.5 is not always optimal. A confusion matrix at 0.5
shows the trade-off between sensitivity (catching true positives) and specificity
(avoiding false positives).

```{r confusion-matrix}
pred_class <- ifelse(pred_prob >= 0.5, 1, 0)
table(Predicted = pred_class, Observed = obs)
```

<div class="callout callout-tip">
<strong>💡 When to adjust the threshold</strong>
In public health contexts, false negatives (missing a high-risk case) may be more
costly than false positives. Lowering the threshold increases sensitivity at the
cost of specificity. The optimal threshold depends on the relative costs of the two
error types — a decision that requires domain knowledge, not statistics alone.
</div>

<div class="key-box">
<h4>🔑 Key takeaways</h4>
<ul>
<li>Use <code>glm(y ~ x, family = binomial)</code> for binary outcomes. Linear regression
will produce nonsensical probability predictions for extreme x values.</li>
<li>Coefficients are on the <strong>log-odds scale</strong>. Always exponentiate with
<code>exp(coef(mod))</code> to get interpretable <strong>odds ratios</strong>.</li>
<li>OR > 1 increases the odds; OR < 1 decreases them. An OR of 2 means the odds
double for a one-unit increase in x.</li>
<li>Use <code>predict(type = "response")</code> to get predicted <strong>probabilities</strong>.</li>
<li>Evaluate discrimination with <strong>AUC</strong>: above 0.7 is acceptable; above
0.8 is good. Check calibration with a binned residual plot.</li>
</ul>
</div>

---

### What's next?

- **[Survival Analysis](06_Survival_Analysis.html)** — extend to binary outcomes with
  a time dimension and censored observations
- **[Multiple Regression & GLMs](../advanced_stats/multiple_regression.html)** — count
  outcomes (Poisson), overdispersion, and model selection

```{r session-info, echo=FALSE}
# sessionInfo()
```

<footer class="page-footer">
EA030 Advanced Statistics — Module 09
<span class="divider">|</span> Prof. Marc Los Huertos
<span class="divider">|</span> Pomona College
<span class="divider">|</span> Last updated: `r format(Sys.Date(), "%B %Y")`
<span class="divider">|</span> <a href="https://github.com/marclos/RTricks">GitHub</a>
</footer>
