---
title: "Survival Analysis"
author: "Marc Los Huertos"
date: "EA030 · Compiled `r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    self_contained: true
    theme: null
    highlight: null
    toc: false
knit: (function(inputFile, encoding) {
    rmarkdown::render(
      inputFile,
      encoding = encoding,
      output_file = file.path(dirname(inputFile),
                              "../docs/basic_statistics/12_Survival_Analysis.html")
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 7,
  fig.height = 4
)
```

```{css, echo=FALSE}
@import url('https://fonts.googleapis.com/css2?family=Libre+Baskerville:ital,wght@0,400;0,700;1,400&family=Source+Code+Pro:wght@400;600&family=Lato:wght@300;400;700&display=swap');

*, *::before, *::after { box-sizing: border-box; margin: 0; padding: 0; }

:root {
  --forest:    #1e3a2f;
  --moss:      #2d5a3d;
  --fern:      #4a7c5e;
  --sage:      #8aad96;
  --mist:      #d4e6da;
  --cream:     #f7f3ed;
  --parchment: #ede8df;
  --ink:       #1a1a18;
  --charcoal:  #3a3a38;
  --dust:      #7a7a72;
  --amber:     #c8832a;
  --amber-lt:  #e8a84a;
  --white:     #ffffff;
  --card-shadow: 0 2px 10px rgba(30,58,47,.09);
}

html { scroll-behavior: smooth; }
body {
  font-family: 'Lato', sans-serif;
  background: var(--cream);
  color: var(--ink);
  font-size: 17px;
  line-height: 1.75;
  padding-left:  clamp(14px, 4vw, 48px);
  padding-right: clamp(14px, 4vw, 48px);
  padding-bottom: 80px;
}
.main-container { max-width: 920px !important; margin: 0 auto !important; padding: 0 clamp(10px, 2vw, 32px) 80px !important; }
.page-header {
  margin-left:  calc(-1 * clamp(14px, 4vw, 48px));
  margin-right: calc(-1 * clamp(14px, 4vw, 48px));
}
h1.title, h4.author, h4.date { display: none; }

/* Page header */
.page-header {
  background-image:
    repeating-linear-gradient(120deg, transparent, transparent 60px, rgba(255,255,255,.018) 60px, rgba(255,255,255,.018) 61px),
    linear-gradient(160deg, #1e3a2f 0%, #2d5a3d 60%, #1a3d30 100%);
  color: #fff; padding: 44px 4vw 36px; margin: 0 0 48px;
  border-bottom: 4px solid var(--amber); position: relative; overflow: hidden;
}
.page-header::after {
  content: ''; position: absolute; right: -40px; top: -40px;
  width: 260px; height: 260px; border-radius: 50%;
  background: radial-gradient(circle, rgba(138,173,150,.12) 0%, transparent 70%);
  pointer-events: none;
}
.page-header .module-tag {
  display: inline-block; font-family: 'Source Code Pro', monospace;
  font-size: .68rem; font-weight: 600; letter-spacing: .18em; text-transform: uppercase;
  color: var(--amber-lt); border: 1px solid rgba(200,131,42,.45);
  padding: 3px 10px; border-radius: 2px; margin-bottom: 12px;
}
.page-header h1 {
  font-family: 'Libre Baskerville', serif; font-size: clamp(1.6rem, 3vw, 2.3rem);
  font-weight: 700; color: #fff; line-height: 1.2; margin-bottom: 8px;
}
.page-header h1 span { color: var(--sage); }
.page-header .sub { font-size: .9rem; color: rgba(255,255,255,.65); font-weight: 300; }
.page-header .sub strong { color: rgba(255,255,255,.9); font-weight: 700; }
.back-link {
  display: inline-block; margin-top: 16px;
  font-family: 'Source Code Pro', monospace; font-size: .7rem; letter-spacing: .08em;
  color: var(--amber-lt); text-decoration: none;
  border: 1px solid rgba(200,131,42,.4); padding: 4px 11px; border-radius: 3px;
  transition: background .2s;
}
.back-link:hover { background: rgba(200,131,42,.18); }

/* Section nav pills */
.section-nav {
  display: flex; gap: 8px; flex-wrap: wrap; margin-bottom: 44px;
  padding-bottom: 20px; border-bottom: 1px solid var(--parchment);
}
.section-nav a {
  font-family: 'Source Code Pro', monospace; font-size: .7rem; font-weight: 600;
  letter-spacing: .08em; text-transform: uppercase;
  color: var(--fern); text-decoration: none;
  border: 1px solid var(--mist); padding: 4px 12px; border-radius: 3px;
  background: var(--white); transition: background .2s, color .2s;
}
.section-nav a:hover { background: var(--mist); color: var(--forest); }

/* Section headings */
h2 {
  font-family: 'Libre Baskerville', serif; font-size: 1.45rem; font-weight: 700;
  color: var(--forest); margin-top: 56px; margin-bottom: 6px;
  padding-bottom: 8px; border-bottom: 1px solid var(--parchment);
}
h3 {
  font-family: 'Libre Baskerville', serif; font-size: 1.1rem; font-weight: 700;
  color: var(--moss); margin-top: 28px; margin-bottom: 4px;
}

/* Body text */
p { margin-bottom: 1em; }
strong { color: var(--forest); }

/* Code */
code {
  font-family: 'Source Code Pro', monospace; font-size: .84em;
  background: var(--mist); color: var(--moss); padding: 1px 5px; border-radius: 3px;
}
pre {
  background: var(--ink); color: #e8e6df; border-radius: 5px;
  padding: 14px 18px; font-size: .83rem; line-height: 1.55;
  overflow-x: auto; border-left: 4px solid var(--amber); margin: 1.2em 0;
}
pre code { background: transparent; color: inherit; padding: 0; font-size: 1em; }

/* Callout boxes */
.callout {
  border-left: 5px solid; padding: .65em 1.1em;
  margin: 1.4em 0; border-radius: 0 5px 5px 0; font-size: .92em; line-height: 1.6;
}
.callout > strong:first-child { display: block; margin-bottom: .3em; }
.callout-concept { border-color: var(--fern);  background: #f0f7f2; }
.callout-tip     { border-color: #2e7d32;      background: #f1f8f1; }
.callout-warn    { border-color: #e65100;      background: #fff8f4; }
.callout-formula { border-color: var(--amber); background: #fdf8f0;
                   font-family: 'Source Code Pro', monospace; font-size: .88em; }

/* Math display */
.math-block {
  background: var(--white); border: 1px solid var(--parchment);
  border-radius: 5px; padding: 16px 20px; margin: 1.2em 0;
  text-align: center; box-shadow: var(--card-shadow);
}

/* Tables */
table { font-size: .9rem; border-collapse: collapse; width: 100%; margin: 1.2em 0; }
thead th {
  background: var(--forest); color: #fff;
  font-family: 'Source Code Pro', monospace; font-size: .76rem;
  letter-spacing: .06em; text-transform: uppercase; padding: 8px 12px; text-align: left;
}
tbody tr:nth-child(even) { background: var(--mist); }
tbody td { padding: 7px 12px; border-bottom: 1px solid var(--parchment); }

/* Key concept summary boxes */
.key-box {
  background: var(--forest); color: #fff; border-radius: 6px;
  padding: 20px 24px; margin: 2em 0; font-size: .9rem;
}
.key-box h4 {
  font-family: 'Libre Baskerville', serif; color: var(--sage);
  font-size: 1rem; margin-bottom: 10px;
}
.key-box ul { padding-left: 1.3em; }
.key-box li { margin-bottom: 5px; color: rgba(255,255,255,.85); }
.key-box code { background: rgba(255,255,255,.12); color: var(--amber-lt); }

/* Footer */
.page-footer {
  background: var(--ink); color: rgba(255,255,255,.42); text-align: center;
  padding: 24px 4vw; margin: 60px calc(-1 * clamp(14px, 4vw, 48px)) 0;
  font-size: .76rem; font-family: 'Source Code Pro', monospace; letter-spacing: .04em;
}
.page-footer a { color: var(--sage); text-decoration: none; }
.page-footer .divider { margin: 0 8px; opacity: .3; }
```

<div class="page-header">
<div class="module-tag">EA030 &mdash; Advanced Statistics &mdash; Module 06</div>
<h1>Survival Analysis: <span>Time-to-Event Methods</span></h1>
<p class="sub"><strong>Prof. Marc Los Huertos</strong> &nbsp;&bull;&nbsp; Pomona College</p>
<a class="back-link" href="../index.html">&#8592; Back to RTricks</a>
</div>

<div class="section-nav">
<a href="#why">Why Survival Analysis?</a>
<a href="#concepts">Core Concepts</a>
<a href="#data">Data Setup</a>
<a href="#rawcurves">Raw Curves</a>
<a href="#km">Kaplan–Meier</a>
<a href="#logrank">Log-Rank Test</a>
<a href="#cox">Cox Model</a>
<a href="#diagnostics">Diagnostics</a>
<a href="#reporting">Reporting</a>
</div>

> **What this module covers.** Survival analysis models *when* something happens, not
> just *whether* it happens. This tutorial introduces Kaplan–Meier curves, the log-rank
> test, and Cox proportional hazards regression — applied throughout to a 72-hour algae
> growth experiment under pollution stress. You will also learn to handle right-censored
> observations correctly, a problem that ANOVA and t-tests cannot address.

---

## 1. Why Survival Analysis? {#why}

Most introductory statistics courses focus on comparing group means (t-tests, ANOVA) or
modeling continuous outcomes (regression). But many environmental and biological
questions are fundamentally about **time**: How long before a population collapses?
When does a contaminant first exceed a threshold? How quickly do algae cultures stop
growing under different pollution treatments?

These are **time-to-event** questions, and they require a specialized toolkit for one
critical reason: **censoring**. An observation is *censored* when the event of interest
has not yet occurred by the time the study ends. If an algae culture is still growing at
hour 72 when the experiment stops, you do not know when — or whether — it would have
eventually declined. Discarding those observations (as a t-test would force you to do)
or treating them as "no event" introduces systematic bias.

Survival analysis handles censoring correctly by treating each culture's data as
informative up to its last measured time point, whether or not the event was observed.

### When the experiment defines the event

In the algae experiment that motivates this tutorial, three treatment groups
(Control, Low Pollution, High Pollution) were monitored over 72 hours. The "event" can
be defined in two ways:

1. **End-of-experiment censoring** — every culture is censored at hour 72, because the
   experiment ended before all cultures could show a clear biological endpoint. The
   Kaplan–Meier curves in this design show relative patterns of continued growth.

2. **Threshold crossing** — define a biologically meaningful absorbance level (e.g.,
   OD ≥ 0.6). Cultures that reach it are "events"; those that do not are censored at
   their last measurement. This approach requires a defensible threshold.

Both designs are valid. This tutorial implements the threshold approach, which produces
the most interpretable Kaplan–Meier curves and Cox model.

<div class="callout callout-concept">
<strong>📌 The six reasons to use survival analysis here</strong>

1. The experiment has a hard stop — censoring is structural, not accidental.
2. Some cultures never reach the threshold within 72 hours (right censoring).
3. Kaplan–Meier curves compare groups without assuming any particular growth model.
4. The log-rank test provides a formal test of group differences across the full timeline.
5. Cox regression yields hazard ratios — interpretable effect sizes for how much faster
   (or slower) one treatment reaches the threshold relative to another.
6. No data are discarded: all time points and censored observations contribute.
</div>

---

## 2. Core Concepts {#concepts}

### The survival function S(t)

The **survival function** S(t) gives the probability that the event has *not yet*
occurred by time t:

$$S(t) = P(T > t)$$

where T is the random variable representing time-to-event. S(0) = 1 (no events at
the start) and S(t) is non-increasing over time.

### Censoring

An observation is **right-censored** at time c if the event had not occurred by the
time the study ended or the subject left observation. The censored observation tells us
only that T > c — the event would have happened *eventually*, but we did not observe it.

In R's `Surv()` function, status = 1 indicates an observed event; status = 0 indicates
censoring.

### The hazard function h(t)

The **hazard** h(t) is the instantaneous rate at which events occur among those who have
not yet had the event. A Cox proportional hazards model estimates how covariates
(e.g., treatment group) scale the baseline hazard:

$$h(t \mid x) = h_0(t) \cdot \exp(\beta_1 x_1 + \beta_2 x_2 + \ldots)$$

The **hazard ratio** HR = exp(β) compares event rates between groups. HR > 1 means the
treatment group reaches the threshold faster; HR < 1 means it reaches it more slowly.

---

## 3. Packages and Data Setup {#data}

### Required packages

```{r packages}
library(survival)   # Surv(), survfit(), coxph(), cox.zph()
library(survminer)  # ggsurvplot(), pairwise_survdiff()
library(dplyr)      # data wrangling
library(tidyr)
library(broom)      # tidy() for Cox output
library(ggplot2)
```

### Simulated algae growth data

The data below simulate a 72-hour algae growth experiment with 10 biological replicates
per treatment. Growth follows a logistic curve with treatment-specific rates and noise.

```{r simulate-data}
set.seed(42)

# Simulate one replicate
sim_culture <- function(sample_id, treatment, times = seq(0, 72, by = 8)) {
  params <- list(
    A = list(K = 1.2, r = 0.08, n = 0.06),
    B = list(K = 1.0, r = 0.06, n = 0.09),
    C = list(K = 0.7, r = 0.04, n = 0.12)
  )[[treatment]]
  
  abs_val <- params$K / (1 + exp(-params$r * (times - 36))) +
             rnorm(length(times), 0, params$n)
  abs_val <- pmax(abs_val, 0)
  
  data.frame(sample = sample_id, treatment = treatment,
             time = times, absorbance = abs_val)
}

treatments <- c("A", "B", "C")
df_raw <- do.call(rbind, lapply(treatments, function(trt) {
  do.call(rbind, lapply(1:10, function(i) {
    sim_culture(paste0(trt, "_", i), trt)
  }))
}))
df_raw$treatment <- factor(df_raw$treatment, levels = c("A", "B", "C"))
```

### Define the event: time to reach threshold

Choose a biologically meaningful absorbance threshold. Here OD ≥ 0.6 represents the
point at which growth is considered well established. For each replicate, find the
*first* time that threshold is reached. Replicates that never reach it are right-censored
at their final measurement time.

```{r define-event}
threshold <- 0.6

time_to_event <- df_raw |>
  group_by(sample, treatment) |>
  arrange(time) |>
  summarise(
    event_time = {
      hit <- which(absorbance >= threshold)
      if (length(hit) == 0) NA_real_ else time[min(hit)]
    },
    last_time = max(time),
    .groups = "drop"
  ) |>
  mutate(
    status = if_else(is.na(event_time), 0L, 1L),  # 1 = event, 0 = censored
    time   = if_else(is.na(event_time), last_time, event_time)
  )

# Event count by treatment
time_to_event |> count(treatment, status)
```

---

## 4. Visualize Raw Growth Curves {#rawcurves}

Always plot the raw data before fitting any model. A spaghetti plot (individual traces
in the background, treatment mean in bold) shows whether the threshold makes biological
sense and reveals between-replicate variation.

```{r raw-curves, fig.cap="Individual growth curves (faint) and treatment means (bold) for 10 replicates per treatment over 72 hours. The dashed line marks the OD = 0.6 threshold that defines the survival event.", fig.height=5}
df_means <- df_raw |>
  group_by(treatment, time) |>
  summarise(mean_abs = mean(absorbance), .groups = "drop")

ggplot() +
  geom_line(data = df_raw,
            aes(time, absorbance, group = sample, color = treatment),
            alpha = 0.2, show.legend = FALSE) +
  geom_line(data = df_means,
            aes(time, mean_abs, color = treatment),
            linewidth = 1.2) +
  geom_hline(yintercept = threshold, linetype = "dashed", color = "#c8832a") +
  annotate("text", x = 68, y = threshold + 0.04,
           label = paste0("threshold = ", threshold),
           color = "#c8832a", size = 3.2, hjust = 1) +
  scale_color_manual(values = c(A = "#4a7c5e", B = "#c8832a", C = "#8b3a2a")) +
  labs(x = "Time (hours)", y = "Absorbance (OD)",
       title = "Raw growth curves by treatment",
       color = "Treatment") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```

<div class="callout callout-tip">
<strong>💡 What to look for</strong>
Treatment A cultures should reach the threshold earliest on average; treatment C should
reach it last or not at all within 72 hours. The spread of individual curves (grey
traces) shows how much between-replicate variability exists. A wide spread means the
treatment effect must overcome substantial biological noise to be detectable.
</div>

---

## 5. Kaplan–Meier Curves {#km}

### Fitting the model

The Kaplan–Meier estimator is a non-parametric method that estimates S(t) directly from
the data without assuming any underlying distribution. It updates at each observed event
time using:

$$\hat{S}(t) = \prod_{t_i \leq t} \left(1 - \frac{d_i}{n_i}\right)$$

where $d_i$ is the number of events and $n_i$ is the number still at risk at time $t_i$.

```{r km-fit}
km_fit <- survfit(Surv(time, status) ~ treatment, data = time_to_event)
summary(km_fit)
```

### Plot with risk table

```{r km-plot, fig.height=6.5, fig.cap="Kaplan–Meier survival curves showing the probability that each culture has *not yet* reached OD = 0.6. Shading = 95% CI. The risk table below shows how many replicates remain uncensored at each time point."}
ggsurvplot(
  km_fit,
  data        = time_to_event,
  pval        = TRUE,
  conf.int    = TRUE,
  risk.table  = TRUE,
  palette     = c("#4a7c5e", "#c8832a", "#8b3a2a"),
  ggtheme     = theme_minimal(base_size = 12),
  title       = "Kaplan–Meier: time to reach absorbance threshold",
  xlab        = "Time (hours)",
  ylab        = "Probability of not yet reaching threshold",
  legend.title = "Treatment",
  risk.table.height = 0.25
)
```

**Interpreting the curves.** S(t) is the probability that a replicate has *not yet*
reached OD 0.6 by time t. A curve that drops steeply early means the treatment
reached the threshold quickly; a flat curve means most replicates were still below
threshold (censored). Treatment A should show the steepest drop; treatment C the
flattest, because pollution suppresses growth.

<div class="callout callout-concept">
<strong>📌 Survival probability vs. event probability</strong>
The y-axis shows the *survival* probability — the fraction still below threshold.
To get the event probability (fraction that *has* crossed the threshold), compute
1 − S(t). Both framings are valid; choose the one that matches your biological question.
</div>

---

## 6. Log-Rank Test {#logrank}

The **log-rank test** is the standard global test for differences between survival
curves. It compares the observed and expected number of events in each group across all
time points, weighting each time point equally. A significant result means at least one
pair of curves differs; it does not identify which pair.

```{r logrank}
logrank <- survdiff(Surv(time, status) ~ treatment, data = time_to_event)
logrank

# Extract p-value
p_val <- 1 - pchisq(logrank$chisq, df = length(logrank$n) - 1)
```

The omnibus log-rank test gives χ² = `r round(logrank$chisq, 2)` on
`r length(logrank$n) - 1` df, **p = `r signif(p_val, 3)`**.

### Pairwise comparisons (adjusted)

If the omnibus test is significant, follow up with pairwise comparisons. The
Benjamini–Hochberg (BH) method controls the false discovery rate.

```{r pairwise}
pairwise_res <- pairwise_survdiff(
  Surv(time, status) ~ treatment,
  data = time_to_event,
  p.adjust.method = "BH"
)
pairwise_res
```

<div class="callout callout-warn">
<strong>⚠️ Multiple comparisons</strong>
With three treatments there are three pairwise tests. Always apply a correction (BH,
Bonferroni, or similar). Reporting raw p-values from multiple tests inflates the
probability of false positives.
</div>

---

## 7. Cox Proportional Hazards Model {#cox}

The Cox model provides **hazard ratios** — effect sizes that quantify how much faster
(or slower) one treatment group reaches the threshold relative to a reference group —
while accounting for the full time course and censored observations.

```{r cox-model}
# Set treatment A as reference
time_to_event <- time_to_event |>
  mutate(treatment = relevel(treatment, ref = "A"))

cox1 <- coxph(Surv(time, status) ~ treatment, data = time_to_event)
summary(cox1)
```

### Extracting and plotting hazard ratios

```{r cox-forest, fig.cap="Cox model hazard ratios (HR) with 95% confidence intervals. HR > 1 means the treatment reaches the threshold faster than treatment A; HR < 1 means slower."}
hr <- tidy(cox1, exponentiate = TRUE, conf.int = TRUE)
print(hr)

ggplot(hr, aes(x = term, y = estimate)) +
  geom_point(size = 3, color = "#4a7c5e") +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high),
                width = 0.15, color = "#4a7c5e") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "#c8832a") +
  coord_flip() +
  labs(y = "Hazard Ratio (HR)", x = "",
       title = "Cox model: HR and 95% CI (reference = Treatment A)") +
  theme_minimal(base_size = 13)
```

**Interpreting hazard ratios.** An HR of 2 for treatment C means that at any point in
time, a culture in treatment C is twice as likely to reach the threshold as one in
treatment A, conditional on not yet having done so. HR < 1 would indicate slower
progression.

---

## 8. Proportional Hazards Diagnostics {#diagnostics}

The Cox model assumes that the hazard ratio between any two groups is *constant over
time* (proportional hazards). Violations — where the treatment effect grows or shrinks
over the observation window — invalidate the hazard ratio interpretation.

```{r ph-test, fig.cap="Schoenfeld residuals against time. A flat line (no trend) supports the proportional hazards assumption. A significant p-value in the table indicates a possible violation."}
ph_test <- cox.zph(cox1)
ph_test

par(mfrow = c(1, ncol(ph_test$y)), mar = c(4, 4, 3, 1))
plot(ph_test)
par(mfrow = c(1, 1))
```

<div class="callout callout-tip">
<strong>💡 What to do if PH is violated</strong>
If `cox.zph()` returns a significant p-value for a covariate, consider: (1) stratifying
the Cox model on that covariate with `strata()`; (2) adding a time-interaction term
`tt(treatment)`; or (3) fitting a parametric survival model (Weibull, log-normal) with
`survreg()`, which does not require proportionality.
</div>

### Parametric alternative: Weibull model

```{r weibull, eval=TRUE}
weibull_mod <- survreg(Surv(time, status) ~ treatment,
                       data = time_to_event, dist = "weibull")
summary(weibull_mod)
```

---

## 9. Reporting Recommended Outputs {#reporting}

A complete survival analysis report should include the following elements, in this order:

1. **Raw growth curves** with the threshold annotated (Section 4).
2. **Time-to-event table** showing `sample`, `treatment`, `time`, and `status`
   (event or censored) for all replicates.
3. **Kaplan–Meier plot** with risk table and omnibus p-value (Section 5).
4. **Omnibus log-rank test** results with χ², degrees of freedom, and p-value, followed
   by pairwise BH-adjusted comparisons (Section 6).
5. **Cox model** coefficients, hazard ratios, 95% CIs, and concordance statistic
   (Section 7).
6. **Proportional hazards check** — report `cox.zph()` results and, if the assumption
   is satisfied, note this explicitly (Section 8).

### Complete time-to-event table

```{r tte-table}
knitr::kable(
  time_to_event[order(time_to_event$treatment, time_to_event$time), ],
  col.names = c("Sample", "Treatment", "Event time", "Last time", "Status", "Time used"),
  caption   = "Time-to-event data. Status: 1 = threshold reached, 0 = censored.",
  digits    = 1
)
```

<div class="key-box">
<h4>🔑 Key takeaways</h4>
<ul>
<li><strong>Survival analysis</strong> is necessary whenever observations can be
censored — i.e., the event did not occur by the end of the study window.</li>
<li>The <strong>Kaplan–Meier estimator</strong> is non-parametric and makes no
distributional assumptions; use it for visualization and the log-rank test for inference.</li>
<li><strong>Hazard ratios</strong> from the Cox model are the primary effect size:
HR > 1 means faster event occurrence relative to the reference group.</li>
<li>Always check the <strong>proportional hazards assumption</strong> with
<code>cox.zph()</code> before interpreting HRs.</li>
<li>Report the <strong>risk table</strong> alongside Kaplan–Meier curves so readers
can see how many replicates remain at each time point.</li>
</ul>
</div>

---

### What's next?

- **[Logistic Regression](09_Logistic_Regression.html)** — model binary outcomes
  with a generalized linear model when there is no time dimension
- **[Mixed-Effects Models](../advanced_stats/mixed_effects.html)** — extend survival
  frameworks to nested or repeated-measures designs

```{r session-info, echo=FALSE}
# sessionInfo()
```

<footer class="page-footer">
EA030 Advanced Statistics — Module 06
<span class="divider">|</span> Prof. Marc Los Huertos
<span class="divider">|</span> Pomona College
<span class="divider">|</span> Last updated: `r format(Sys.Date(), "%B %Y")`
<span class="divider">|</span> <a href="https://github.com/marclos/RTricks">GitHub</a>
</footer>
