---
title: "Course Time"
author: "Marc Los Huertos"
date: "`r Sys.Date()`"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## General Expections

Based on the "accepted" expectations, we should be a spending a minimum of 2 hours outside of class for every hour in class (2:1 ratio). Counting only the lecture and 1/2 of lab time that would be about atleast 6 hours outside of class. 

In general, I get worried if a class requires more than 8 hours outside of class for every hour in class on a consistent basis.

## Time Spent Outside of Class

```{r, echo=TRUE}

week2 <- c(1.5, 1, 15, 40, 3, 2, 10, 1.5, 
           0, 20, 2, 45, 7, 10, 2, 0, 2, 2, 2.5, 5, 0)
week3 <- c(0.5, 4, 2, 1.5, 6, 5, 3, 5, 3, 5, 3, 6, 2, 5, 5, 3, 3, 4.5, 5, 8)

```

This is an R Markdown presentation. But code can be added just like the PDFs we have been knitting.


## Skewed Distribution

- Outliers
- Data Entry Mistakes, maybe?
- Excessive Readings, likely!
```{r}
par(mfrow=c(1,2), las=1)
hist(week2, breaks=20, col="lightblue", border="black", xlim=c(0,50), xlab="Time (hours)", main="Time spent outside in a week", ylab="Frequency")
hist(week3, breaks=20, col="lightblue", border="black", xlim=c(0,10), xlab="Time (hours)", main="Time spent outside in a week", ylab="Frequency")

```

## t-test a hypothesis: Different than 6 hours

```{r ttest2, echo = TRUE}
t.test(week2, mu=6)
```

## t-test a hypothesis: Different than 6 hours (Week 3)

```{r test, echo = TRUE}
t.test(week3, mu=6)
```


## t-test: Greater than 6 hours

```{r testcompare, echo = TRUE}
week2.ttest =t.test(week2, mu=6, alternative="greater")
week3.ttest = t.test(week3, mu=6, alternative="less" )
# str(week2.ttest)

print(paste0("Week 2: ", round(week2.ttest$p.value,3), " (", week2.ttest$method, " Greater than mu)")
)
print(paste0("Week 3: ", round(week3.ttest$p.value,5), " (", week2.ttest$method, " (Less than mu)")
)
```

## Marc's Conclusions

- Make sure the quiz directions are clear.
- Reduced readings are working, perhaps too well!
- Offer additional support: If you find the readings to take more than 6-8 hours.
- Determine how lab work might be used to augment weekly hour workload. Project will take up more time. Will gauge next week.





