---
title: "Camera Trap Analysis: Wildlife Detection and Behavior"
subtitle: "EA030 — RTricks Module 15"
author: "Prof. Marc Los Huertos | Pomona College Environmental Analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    self_contained: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 8,
  fig.height = 5,
  fig.align  = "center"
)
```

---

# Introduction

Camera traps are one of the most powerful and non-invasive tools in wildlife
ecology. A motion-triggered camera records **who** was present, **when**, and
**what they were doing** — without a researcher ever being there. Over the
course of even a single week, 10 cameras can accumulate thousands of
individual images spanning dozens of taxa.

In this tutorial you will:

1. Load and clean a real one-week camera-trap dataset.
2. Define **independent detection events** to avoid pseudo-replication.
3. Produce **descriptive summaries** of species activity, behavior, and space
   use.
4. Apply four quantitative approaches that move beyond description:
   - **Chi-square / Fisher tests** — does behavior composition differ between
     day and night?
   - **Activity-overlap kernel density** — how much do two species' diel
     rhythms overlap?
   - **Species diversity and rarefaction** — are some cameras in richer
     habitat than others?
   - **Single-season occupancy modeling** — accounting for imperfect detection
     when estimating which cameras a species occupies.

> **New to R?** Don't worry if every line of code isn't immediately clear.
> Each section explains *what* the code does and *why* before showing you how
> to run it. Focus on understanding the logic; the syntax will become familiar
> with practice.

---

# Setup

## Packages

R's base functionality is extended by **packages** — collections of functions
written by the community. We need to install each package once
(`install.packages()`), then load it each session (`library()`).

```{r packages}
# Run the install lines once (remove the # to uncomment):
# install.packages(c("tidyverse", "lubridate", "janitor", "readxl",
#                    "vegan", "overlap", "unmarked"))

library(tidyverse)   # data wrangling + ggplot2
library(lubridate)   # date-time helpers
library(janitor)     # clean_names() standardises column headers
library(readxl)      # read Excel files
library(vegan)       # diversity indices and rarefaction
library(overlap)     # kernel-density activity-overlap (Ridout & Linkie 2009)
library(unmarked)    # occupancy models (MacKenzie et al. 2002)
```

---

# Data Import and Cleaning

## Loading the data

`read_excel()` reads a specific sheet from an `.xlsx` file into a **data
frame** — R's standard rectangular data structure (rows = observations,
columns = variables).

```{r load-data}
file_path <- "/home/mwl04747/RTricks/15_Camera_Traps/Camera Trap Data Entry.xlsx"

cam_raw <- read_excel(path = file_path, sheet = "Combined")

# Peek at structure
glimpse(cam_raw)
```

> **`glimpse()`** is your first-look friend. It shows every column name, its
> data type (`<chr>` = text, `<dbl>` = number, `<dttm>` = date-time), and the
> first few values.

## Standardising column names and parsing dates

Column names from Excel often contain spaces and mixed capitalisation.
`clean_names()` converts them all to `snake_case` (lower case, underscores
instead of spaces) so they are easier to type.

```{r clean}
cam_raw$datetime <- cam_raw$Datetime_corrected

cam <- cam_raw %>%
  clean_names() %>%
  mutate(
    date = as.Date(datetime),          # strip the time, keep only the date
    hour = hour(datetime)              # extract hour (0–23)
  )

# Sanity check: no missing datetimes
# stopifnot(!any(is.na(cam$datetime)))
summary(cam$date)
```

## Removing ambiguous taxa

Some records contain notes like "bird", "unknown", or combined entries
("squirrel/rat"). These are genuinely unidentifiable and should be excluded
before analysis.

```{r filter-taxa}
ambiguous_patterns <- c(
  " and ", " or ", "/",
  "unknown", "unidentified", "n/a", "^na$",
  "bird$", "small brown bird"
)

cam_clean <- cam %>%
  filter(
    !str_detect(
      str_to_lower(taxa_code),
      paste(ambiguous_patterns, collapse = "|")
    )
  )

cat("Taxa retained after filtering:\n")
unique(cam_clean$taxa_code)
```

> **Regular expressions** (`str_detect`) let you match text patterns.
> `paste(..., collapse = "|")` joins the pattern list with `|` (meaning "or"),
> so one call checks all ambiguous strings at once.

## Collapsing behaviors into functional types

Raw behavior codes (e.g., `"walk"`, `"run"`) are too granular for many
analyses. We group them into five **functional categories** using
`case_when()` — think of it as a multi-way if/else.

```{r behaviors}
cam <- cam_clean %>%
  mutate(
    behavior_function = case_when(
      str_detect(behavior_code, "walk|run|hop|fly|climb|crawl") ~ "movement",
      str_detect(behavior_code, "eat|dig|cache")                ~ "foraging",
      str_detect(behavior_code, "sniff")                        ~ "investigation",
      str_detect(behavior_code, "stare")                        ~ "vigilance",
      str_detect(behavior_code, "stop|stand|sit|lay")           ~ "stationary",
      TRUE                                                      ~ "other"
    )
  )

table(cam$behavior_function)
```

---

# Defining Independent Detection Events

## The 30-minute rule

A single animal may trigger the same camera dozens of times in a few minutes.
Treating each trigger as an independent observation would **pseudo-replicate**
our data — inflating sample sizes and biasing statistics.

The standard solution: any detections of the **same species at the same
camera** within 30 minutes of the previous detection are considered part of
the *same* event. Only new detections after a 30-minute gap count as a new
independent event.

```{r events}
cam_events <- cam %>%
  arrange(camera_id, taxa_code, datetime) %>%
  group_by(camera_id, taxa_code) %>%
  mutate(
    # Time gap in minutes between consecutive rows within each group
    minutes_since_last = as.numeric(
      difftime(datetime, lag(datetime), units = "mins")
    ),
    # A new event starts if gap > 30 min OR this is the first record
    new_event = if_else(
      is.na(minutes_since_last) | minutes_since_last > 30, 1L, 0L
    ),
    event_id = cumsum(new_event)   # running count of events per camera × species
  ) %>%
  ungroup() %>%
  mutate(diel = if_else(hour >= 6 & hour < 18, "day", "night"))
```

> **`lag()`** looks at the value in the *previous* row. Because we've grouped
> by camera × species, "previous" means the last detection of that species at
> that camera — exactly what we want.

---

# Part I — Descriptive Statistics

## 5.1 Detection event totals per taxon

```{r detection-summary}
event_summary <- cam_events %>%
  distinct(camera_id, taxa_code, event_id) %>%
  count(taxa_code, name = "detection_events") %>%
  arrange(desc(detection_events))

print(event_summary)

ggplot(event_summary, aes(x = reorder(taxa_code, detection_events),
                          y = detection_events)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  theme_bw() +
  labs(title = "Independent detection events by taxon",
       x = NULL, y = "Detection events")
```

## 5.2 Behavioral ethogram (stacked bar)

An **ethogram** documents the relative frequency of each behavior. Stacked
proportional bars make cross-species comparisons easy even when total sample
sizes differ.

```{r ethogram}
ethogram <- cam_events %>%
  count(taxa_code, behavior_function) %>%
  group_by(taxa_code) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

ggplot(ethogram, aes(x = taxa_code, y = prop, fill = behavior_function)) +
  geom_col() +
  scale_fill_brewer(palette = "Set2") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Behavioral composition by taxon",
       x = NULL, y = "Proportion of observations",
       fill = "Behavior")
```

## 5.3 Hourly activity heat map

A **heat map** encodes counts as colour intensity. Here each cell is a
taxon × hour combination — a quick way to spot nocturnal, diurnal, or
crepuscular patterns across all taxa simultaneously.

```{r heatmap}
taxon_hour <- cam_events %>%
  distinct(taxa_code, event_id, hour) %>%
  count(taxa_code, hour, name = "events")

ggplot(taxon_hour, aes(x = hour, y = taxa_code, fill = events)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(option = "C") +
  scale_x_continuous(breaks = seq(0, 23, 3)) +
  theme_bw() +
  labs(title = "Hourly detection patterns by taxon",
       x = "Hour of day (0–23)", y = NULL, fill = "Events")
```

## 5.4 Camera-level activity profiles

Different cameras are placed in different micro-habitats. This plot shows
whether behavioral composition varies by camera location.

```{r camera-profiles}
camera_behavior <- cam_events %>%
  distinct(camera_id, taxa_code, event_id, behavior_function) %>%
  count(camera_id, behavior_function, name = "events")

ggplot(camera_behavior,
       aes(x = camera_id, y = events, fill = behavior_function)) +
  geom_col() +
  scale_fill_brewer(palette = "Set2") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Camera activity profiles by behavior function",
       x = "Camera ID", y = "Detection events", fill = "Behavior")
```

---

# Part II — Hypothesis Testing: Behavior vs. Diel Period

## 6.1 Rationale

**Question:** Does the proportion of behaviors (e.g., foraging vs. vigilance)
differ between day and night?

We use a **chi-square test of independence** (or Fisher's exact test for small
counts) to test whether the distribution of behavior types is statistically
independent of diel period.

- **H₀:** Behavior composition is the same during the day and at night.
- **H₁:** At least one behavior type is more common at one time of day.

> **When to use Fisher vs. chi-square?** The rule of thumb: if any expected
> cell count is < 5, use Fisher's exact test; otherwise chi-square is fine.

## 6.2 Run tests per taxon

```{r chi-fisher}
# Build a helper function so we can apply it to each species
test_behavior_diel <- function(df) {
  tbl <- df %>%
    count(diel, behavior_function) %>%
    pivot_wider(names_from = behavior_function,
                values_from = n,
                values_fill = 0) %>%
    column_to_rownames("diel") %>%
    as.matrix()

  # Choose test based on minimum expected count
  exp_counts <- chisq.test(tbl)$expected
  if (any(exp_counts < 5)) {
    res <- fisher.test(tbl, simulate.p.value = TRUE, B = 2000)
    tibble(test = "Fisher", p_value = res$p.value)
  } else {
    res <- chisq.test(tbl)
    tibble(test = "Chi-square", statistic = res$statistic, p_value = res$p.value)
  }
}

behavior_tests <- cam_events %>%
  group_by(taxa_code) %>%
  group_modify(~ test_behavior_diel(.x)) %>%
  ungroup() %>%
  mutate(significant = p_value < 0.05)

print(behavior_tests)
```

## 6.3 Visualise day vs. night behavior shift

```{r diel-behavior-plot}
behavior_diel <- cam_events %>%
  count(taxa_code, diel, behavior_function) %>%
  group_by(taxa_code, diel) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

ggplot(behavior_diel,
       aes(x = diel, y = prop, fill = behavior_function)) +
  geom_col() +
  facet_wrap(~ taxa_code) +
  scale_fill_brewer(palette = "Set2") +
  theme_bw() +
  labs(title = "Behavioral composition: day vs. night",
       x = "Diel period", y = "Proportion",
       fill = "Behavior")
```

> **Interpreting results:** A significant p-value tells you *that* the
> behavior mix differs between day and night, but not *which* behaviors drive
> the difference. Inspect the stacked bars above to identify which functional
> types shift most.

---

# Part III — Activity Overlap (Kernel Density)

## 7.1 Rationale

Two species that are active at the same hours of the day are more likely to
interact — as competitors, predators, or prey. The **overlap coefficient Δ**
(Ridout & Linkie 2009) quantifies the similarity of two diel activity curves
on a 0–1 scale (0 = no overlap, 1 = identical schedules).

The `overlap` package fits smooth **kernel density** curves to the circular
distribution of detection times (circular because 23:00 and 01:00 are only
2 hours apart, not 22 hours apart).

## 7.2 Convert detection times to radians

Time of day on a 24-hour clock maps naturally onto a circle: midnight = 0 and
2π radians, noon = π radians.

```{r radians}
cam_events <- cam_events %>%
  mutate(time_rad = (hour(datetime) + minute(datetime) / 60) / 24 * 2 * pi)
```

## 7.3 Choose a species pair and estimate overlap

```{r activity-overlap, fig.height=5}
# Replace with your two most interesting species
sp1_code <- unique(cam_events$taxa_code)[1]
sp2_code <- unique(cam_events$taxa_code)[2]

t1 <- cam_events %>%
  filter(taxa_code == sp1_code) %>%
  pull(time_rad)

t2 <- cam_events %>%
  filter(taxa_code == sp2_code) %>%
  pull(time_rad)

# Estimate Δ (use "Dhat4" for n < 75, "Dhat1" for n ≥ 75 per species)
estimator <- if (min(length(t1), length(t2)) < 75) "Dhat4" else "Dhat1"

delta <- overlapEst(t1, t2, type = estimator)
cat(sprintf("Activity overlap (%s) between %s and %s: Δ = %.3f\n",
            estimator, sp1_code, sp2_code, delta[estimator]))

# Bootstrap 95% CI
set.seed(42)
bs <- bootstrap(t1, t2, nb = 999, type = estimator)
ci  <- bootCI(delta[estimator], bs)
cat(sprintf("95%% CI: [%.3f, %.3f]\n", ci["norm0", "lower"], ci["norm0", "upper"]))

# Plot both density curves
overlapPlot(t1, t2,
            rug    = TRUE,
            main   = paste("Activity overlap:", sp1_code, "vs.", sp2_code),
            xlab   = "Time of day",
            linecol = "steelblue", linecol2 = "tomato",
            linewidth = 2, linewidth2 = 2)
legend("topright",
       legend = c(sp1_code, sp2_code),
       col    = c("steelblue", "tomato"),
       lwd    = 2, bty = "n")
```

> **Interpretation guide:** Δ > 0.75 = high overlap; 0.50–0.75 = moderate;
> < 0.50 = low. If the CI excludes a reference value of interest, the
> difference is statistically meaningful. See Ridout & Linkie (2009) for
> formal guidance.

## 7.4 All pairwise overlaps (heatmap)

```{r pairwise-overlap, fig.height=6}
taxa_list <- unique(cam_events$taxa_code)
n_sp      <- length(taxa_list)

# Pre-compute one time vector per species
time_by_sp <- setNames(
  lapply(taxa_list, function(sp)
    cam_events %>% filter(taxa_code == sp) %>% pull(time_rad)),
  taxa_list
)

# Fill overlap matrix
overlap_mat <- matrix(NA, n_sp, n_sp,
                      dimnames = list(taxa_list, taxa_list))
for (i in seq_len(n_sp)) {
  for (j in seq_len(n_sp)) {
    if (i == j) {
      overlap_mat[i, j] <- 1
    } else if (!is.na(overlap_mat[j, i])) {
      overlap_mat[i, j] <- overlap_mat[j, i]
    } else {
      ti <- time_by_sp[[i]]; tj <- time_by_sp[[j]]
      est <- if (min(length(ti), length(tj)) < 75) "Dhat4" else "Dhat1"
      overlap_mat[i, j] <- overlapEst(ti, tj, type = est)[est]
    }
  }
}

overlap_df <- as.data.frame(overlap_mat) %>%
  rownames_to_column("sp1") %>%
  pivot_longer(-sp1, names_to = "sp2", values_to = "delta")

ggplot(overlap_df, aes(x = sp1, y = sp2, fill = delta)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(delta, 2)), size = 3) +
  scale_fill_gradient2(low = "white", mid = "steelblue", high = "navy",
                       midpoint = 0.5, limits = c(0, 1)) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Pairwise activity overlap (Δ)",
       x = NULL, y = NULL, fill = "Δ")
```

---

# Part IV — Species Diversity and Rarefaction

## 8.1 Rationale

Some cameras may be placed in richer habitats — near a water source, in denser
cover — and detect more species. But cameras also vary in **effort**: a busier
camera accumulates more detections simply by chance, which inflates apparent
diversity. **Rarefaction** corrects for this by asking: *how many species would
we expect to detect if all cameras had the same total number of events?*

The `vegan` package provides both diversity indices and rarefaction curves.

## 8.2 Build a site × species matrix

```{r site-sp-matrix}
# Count independent events per camera × species
site_sp <- cam_events %>%
  distinct(camera_id, taxa_code, event_id) %>%
  count(camera_id, taxa_code, name = "events") %>%
  pivot_wider(names_from  = taxa_code,
              values_from = events,
              values_fill = 0) %>%
  column_to_rownames("camera_id")

print(site_sp)
```

## 8.3 Diversity indices

```{r diversity}
div_df <- tibble(
  camera_id  = rownames(site_sp),
  richness   = specnumber(site_sp),           # S: raw species count
  shannon    = diversity(site_sp, "shannon"), # H': log-scale evenness
  simpson    = diversity(site_sp, "simpson"), # 1 - D: dominance-corrected
  total_events = rowSums(site_sp)
)

print(div_df)

div_df %>%
  pivot_longer(c(richness, shannon, simpson),
               names_to = "index", values_to = "value") %>%
  ggplot(aes(x = camera_id, y = value, fill = camera_id)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ index, scales = "free_y") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Diversity indices by camera",
       x = "Camera ID", y = "Index value")
```

> **Which index to report?**
>
> - **Richness (S)** counts species but ignores abundance; sensitive to
>   sampling effort.
> - **Shannon (H')** weights common and rare species; standard in ecology.
> - **Simpson (1–D)** focuses on dominant species; less sensitive to rare taxa.
>
> Report all three and discuss what differences reveal about the community.

## 8.4 Rarefaction curves

```{r rarefaction, fig.height=6}
# Minimum detection count sets the rarefaction target
min_n <- min(rowSums(site_sp))
cat("Rarefying to:", min_n, "detections per camera\n")

# rarecurve() plots and invisibly returns the curves
rarecurve(
  site_sp,
  step  = 1,
  col   = RColorBrewer::brewer.pal(nrow(site_sp), "Set1"),
  xlab  = "Detection events (sampling effort)",
  ylab  = "Expected species richness",
  main  = "Rarefaction curves by camera",
  label = TRUE
)
abline(v = min_n, lty = 2, col = "grey50")
text(min_n, 1, paste("Rarefaction\ntarget =", min_n),
     adj = c(-0.1, 0), cex = 0.8, col = "grey40")
```

> **Reading rarefaction curves:** If a curve has not levelled off (reached an
> asymptote) by the rarefaction target, that camera has not been sampled long
> enough to characterise the true local species pool — worth noting as a
> limitation.

---

# Part V — Single-Season Occupancy Modeling

## 9.1 Rationale and key insight

Occupancy modeling answers the question: *which cameras does species X truly
occupy?* The complication is **imperfect detection**: even when a species is
present, a camera won't always fire. Naively treating "no detection" as
"absent" underestimates occupancy.

The MacKenzie et al. (2002) model simultaneously estimates:

- **ψ (psi)** — the probability a site is occupied.
- **p** — the probability of detecting the species on a given survey
  occasion, *given* it is present.

We split the monitoring week into daily survey occasions. For each camera ×
day, the detection history is **1** (detected) or **0** (not detected).

## 9.2 Build detection histories

```{r det-history}
# Choose target species (replace with whichever interests you)
focal_sp <- unique(cam_events$taxa_code)[1]
cat("Focal species:", focal_sp, "\n")

all_dates   <- sort(unique(cam_events$date))
all_cameras <- sort(unique(cam_events$camera_id))

# 1 if species detected on that camera × day, 0 otherwise
det_wide <- cam_events %>%
  filter(taxa_code == focal_sp) %>%
  distinct(camera_id, date) %>%
  mutate(detected = 1) %>%
  complete(camera_id = all_cameras, date = all_dates,
           fill = list(detected = 0)) %>%
  pivot_wider(names_from = date, values_from = detected) %>%
  column_to_rownames("camera_id")

print(det_wide)
```

## 9.3 Fit null occupancy model

```{r occupancy}
# unmarkedFrameOccu expects: y = detection matrix (sites × occasions)
umf <- unmarkedFrameOccu(y = as.matrix(det_wide))
summary(umf)

# Null model: constant ψ and p (no covariates)
fit_null <- occu(~ 1 ~ 1, data = umf)
summary(fit_null)

# Back-transform from logit scale to probability
occ_est <- backTransform(fit_null, type = "state")
det_est <- backTransform(fit_null, type = "det")

cat(sprintf("\nEstimated occupancy  ψ = %.3f (95%% CI: %.3f – %.3f)\n",
            occ_est@estimate,
            confint(occ_est)[1], confint(occ_est)[2]))
cat(sprintf("Estimated detection  p = %.3f (95%% CI: %.3f – %.3f)\n",
            det_est@estimate,
            confint(det_est)[1], confint(det_est)[2]))
```

## 9.4 Covariate model: does vegetation density affect occupancy?

If you have a site-level covariate (e.g., canopy cover measured at each
camera), you can test whether occupancy varies with habitat quality.
The example below uses a hypothetical covariate; replace with your own.

```{r occupancy-covariate}
# ----- Replace this section with your real covariate data -----
set.seed(7)
site_covs <- data.frame(
  veg_density = scale(runif(nrow(det_wide), 0, 100))   # hypothetical %cover
)
# --------------------------------------------------------------

umf_cov <- unmarkedFrameOccu(
  y       = as.matrix(det_wide),
  siteCovs = site_covs
)

fit_cov <- occu(~ 1 ~ veg_density, data = umf_cov)
summary(fit_cov)

# Compare models with AIC
cat("\nAIC comparison:\n")
cat(sprintf("  Null model AIC:      %.2f\n", fit_null@AIC))
cat(sprintf("  Veg-density AIC:     %.2f\n", fit_cov@AIC))
cat(sprintf("  ΔAIC:                %.2f\n", fit_null@AIC - fit_cov@AIC))
```

> **Interpreting AIC:** A model with ΔAIC > 2 relative to the null model has
> meaningfully better support. A positive coefficient on `veg_density` on the
> occupancy (state) sub-model indicates the species is more likely to occupy
> densely vegetated sites.

## 9.5 Site-level occupancy predictions

```{r occ-predictions, fig.height=4}
# Predict ψ across the range of the covariate
newdat <- data.frame(
  veg_density = seq(min(site_covs$veg_density),
                    max(site_covs$veg_density),
                    length.out = 100)
)

pred <- predict(fit_cov, type = "state", newdata = newdat)

bind_cols(newdat, pred) %>%
  ggplot(aes(x = veg_density, y = Predicted)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, fill = "steelblue") +
  geom_line(color = "steelblue", linewidth = 1.2) +
  ylim(0, 1) +
  theme_bw() +
  labs(title = paste("Predicted occupancy for", focal_sp),
       x = "Vegetation density (standardised)",
       y = "Estimated occupancy probability (ψ)")
```

---

# Summary and Discussion Questions

This tutorial covered the full pipeline from raw camera-trap images to
publication-quality statistics:

| Stage | Key idea |
|---|---|
| Data cleaning | Remove ambiguous taxa; standardise columns |
| Independent events | 30-minute rule prevents pseudo-replication |
| Descriptive stats | Ethograms, heat maps, camera profiles |
| Chi-square / Fisher | Does behavior shift between day and night? |
| Activity overlap | Kernel density Δ quantifies temporal co-occurrence |
| Rarefaction | Controls for unequal sampling effort across cameras |
| Occupancy model | Separates true absence from non-detection |

**Discussion questions**

1. Which taxon showed the greatest day–night behavioral shift? What ecological
   mechanism might explain it?
2. Do species with high activity overlap (Δ) tend to be ecologically similar
   (same guild) or different? What does this suggest?
3. Look at the rarefaction curves. Which cameras appear under-sampled? How
   would you redesign the study to address this?
4. Compare the naïve detection rate for your focal species with the
   model-estimated occupancy ψ. How much does imperfect detection matter here?

---

# References

- MacKenzie, D. I., Nichols, J. D., Lachman, G. B., Droege, S., Royle, J. A.,
  & Langtimm, C. A. (2002). Estimating site occupancy rates when detection
  probabilities are less than one. *Ecology*, 83(8), 2248–2255.
- Ridout, M. S., & Linkie, M. (2009). Estimating overlap of daily activity
  patterns from camera trap data. *Journal of Agricultural, Biological, and
  Environmental Statistics*, 14(3), 322–337.
- Fiske, I., & Chandler, R. (2011). unmarked: An R package for fitting
  hierarchical models of wildlife occurrence and abundance. *Journal of
  Statistical Software*, 43(10), 1–23.
- Oksanen, J., et al. (2022). vegan: Community Ecology Package. R package
  version 2.6-4.
