---
title: "Water Quality Testing"
author: "Isa D'Amario-McShane"
date: "February 21, 2025"
output: pdf_document
---

Escherichia coli (E. coli) is a bacterium commonly found in the intestines of humans and animals, where most strains are harmless and play a role in digestion. However, the presence of E. coli in water is a major public health concern, as it serves as an indicator of fecal contamination and the potential presence of harmful pathogens. Certain strains can cause serious illnesses, including diarrhea, urinary tract infections, and even kidney failure. Contaminated water sources—whether from agricultural runoff, sewage leaks, or stormwater overflows—pose risks not only to drinking water supplies but also to recreational waters and agricultural irrigation. Ingestion of water containing pathogenic E. coli can lead to widespread outbreaks, highlighting the importance of routine water quality monitoring and treatment. Effective detection methods, such as the Colilert test, help identify E. coli contamination, allowing for timely interventions to protect public health and prevent the spread of waterborne diseases.

So this week in class we took water samples from three different places across campus - pHake lake, Seal Court (my group) and Pomona Bioswale - to test it for exactly these reasons. 

## Skipping this section

```{r, eval=FALSE}
# mlh None of this is my code, so I am going to take it offline...because I don't understand it...:-) this wouild cause major errors in the knitting process....

data = read.csv(data.csv)

# view(data) # MLH I don't know what this is supposed to do, but it doesn't work for me so  I commented it out.
head(data)
colnames(data)
summary(data)

unique(data$Dilution)
sum(is.na(data$Dilution))
sum(is.na(data$Yellow_Large_Wells))
sum(is.na(data$Yellow_Small_Wells)) 
sum(is.na(data$Fluorescence_Large_Wells))

library(dplyr)
colnames(data) <- gsub(" ", "_", colnames(data))

MPN_data.csv = "/home/idjm2024/Water_Quality_Ecoli_NTM./Ecoli and Total Coliform Data Entry - Data Entry.csv"
MPN_data <- read.csv(MPN_data.csv)
```


## Starting Here

NOTE: my guide says to download the cleaned up version...

```{r}
# MLH not my path, you can remove comment.
 
# MPN_data.csv = "/home/idjm2024/Water_Quality_Ecoli_NTM./Ecoli and Total Coliform Data Entry - Data Entry.csv"

MPN_data.csv = "/home/mwl04747/RTricks/06_Ecoli_NTM/Ecoli and Total Coliform Data Entry - Data Export (marc's cleaned version).csv"
MPN_data <- read.csv(MPN_data.csv)


MPN_data <- MPN_data[!grepl("Control", MPN_data$Site_ID),]
MPN_data$TC_MPN = MPN_data$TotalColiform
MPN_data$EC_MPN = MPN_data$Ecoli

MPN_data$TC_MPN[MPN_data$Dilution=="Diluted, 1:10"] <-
MPN_data$TotalColiform[MPN_data$Dilution=="Diluted, 1:10"] *10
MPN_data$Ecoli <- as.numeric(MPN_data$Ecoli)
MPN_data$Ecoli[is.na(MPN_data$Ecoli)] <- 0
MPN_data$EC_MPN[MPN_data$Dilution == "Diluted, 1:10"] <- MPN_data$Ecoli[MPN_data$Dilution == "Diluted, 1:10"] * 10
head(MPN_data)

```

## Not run... not sure what the goals are here, but I think not needed.

```{r, eval=FALSE}

head(MPN_data$Collection.Location) # MLH -- not sure your goal here
unique(MPN_data$Collection.Location) # good info, but not needed...
# sum(is.na(MPN_data$Collection.Location)) # I have code to remove that...


MPN_data <- MPN_data[!is.na(MPN_data$Collection.Location) & MPN_data$Collection.Location != "", ]
MPN_data$Site_ID <- as.factor(MPN_data$Collection.Location)
levels(MPN_data$Site_ID)

MPN_data$Site_ID <- MPN_data$`Collection.Location`

```

back to good stuff...

```{r}
aov_result <- aov(EC_MPN ~ Site_ID, data = MPN_data)
summary(aov_result)
aov_result_TC <- aov(TC_MPN ~ Site_ID, data = MPN_data)
summary(aov_result_TC)


```
Df Sum Sq Mean Sq F value Pr(>F)
Site_ID 2 114316 57158 5.095 0.0331*
Residuals 9 100971 11219
---
Signif. codes: 0 '***' 0.0001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

Since the Pr(>F) value (p-value) is less than 0.05 (as it's 0.03), it means that there is statistically significant evidence to reject the null hypothesis. The null hypothesis for the ANOVA test is that the means of the groups (in this case, the Site_ID locations) are all equal. A p-value less than 0.05 suggests that at least one of the sites has a significantly different mean for the EC_MPN (or TC_MPN) value, implying the locations are not the same with respect to the measured MPN (microbial contamination level).

