---
title: "Analyzing EPA PM2.5 Data"
author: "Marc Los Huertos"
date: "Spring 2026"
output:
  tufte::tufte_html:
    toc: true
    number_sections: true
bibliography: skeleton.bib
link-citations: yes
---

```{r setup, include=FALSE}
library(tufte)
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  tidy    = FALSE,
  fig.width  = 7,
  fig.height = 4
)
```

## Part II: Raspberry Pi Sensor vs. EPA Standards

*Updated: `r format(Sys.time(), "%Y-%m-%d %H:%M")`*

---

# How to Use This Document

This page is a **reference and guide, not a template.** Read through it before writing
any code. Where you see a chunk with `___` blanks, those are arguments you must supply
yourself. The output you are aiming for appears just above each scaffold in a hidden
chunk. Use `?function_name` in the R console to look up argument details.

---

# Introduction

Low-cost optical particle counters like the PMS5003 sensor (used in the Raspberry Pi
kits deployed in this class) make continuous monitoring accessible to community
scientists and researchers. However, these sensors are not regulatory-grade instruments.
They may respond differently to humidity, temperature, and aerosol composition compared
to Federal Reference Method (FRM) samplers [@Pope2006].

This exercise asks two questions about your group's deployment data:

1. Does the mean PM2.5 at your site significantly exceed the EPA annual standard
   of 12 ug/m3?
2. On what fraction of days did your sensor record concentrations above the
   24-hour standard of 35 ug/m3?

These are simultaneously statistical and environmental justice questions: the
communities where these sensors were deployed often bear the heaviest air pollution
burden.

## Sensor Hardware

Your kit uses a **Pimoroni Enviro+ board** mounted on a Raspberry Pi Zero W. The
primary sensors for this exercise are:

| Sensor | What it measures |
|---|---|
| PMS5003 optical particle counter | PM1, PM2.5, PM10 (ug/m3) |
| BME280 | Temperature (deg C), pressure (hPa), humidity (%) |
| LTR559 | Ambient light (lux) |
| MICS6814 gas array | Oxidising, reducing, NH3 (kOhm resistance) |

Data are logged every 10 minutes. The column names in the CSV are:

| Column | Units | Description |
|---|---|---|
| `timestamp` | YYYY-MM-DD HH:MM:SS | Date and time |
| `temperature_C` | deg C | Air temperature (CPU-heat corrected) |
| `pressure_hPa` | hPa | Barometric pressure |
| `humidity_pct` | % | Relative humidity |
| `light_lux` | lux | Ambient light |
| `oxidised_kO` | kOhm | Oxidising gas resistance |
| `reduced_kO` | kOhm | Reducing gas resistance |
| `nh3_kO` | kOhm | Ammonia gas resistance |
| `pm1_ug_m3` | ug/m3 | PM1 mass concentration |
| `pm25_ug_m3` | ug/m3 | PM2.5 mass concentration |
| `pm10_ug_m3` | ug/m3 | PM10 mass concentration |

## Regulatory Standards

- **Annual standard: 12 ug/m3** --- the three-year mean of annual averages must stay
  below this value. Chronic exposure protection.
- **24-hour standard: 35 ug/m3** --- no more than one exceedance per year on a
  three-year rolling average. Acute exposure protection.

## What Your Submission Should Include

- Data loaded and quality-controlled, with zero-readings removed and daily means
  computed, and a brief description of the deployment period and observation count.
- A time-series plot of daily mean PM2.5 with both NAAQS threshold lines annotated.
- A one-sample t-test against the annual standard (12 ug/m3), with a sentence stating
  the null hypothesis, the result, and your interpretation.
- An exceedance count and histogram for the 24-hour standard (35 ug/m3).
- A discussion of 2--4 paragraphs addressing local sources, meteorology, sensor
  limitations, and environmental justice, citing 2--8 sources.

---

# Before You Begin

## Project Directory Setup

`r newthought("You are working on a shared RStudio Server.")` Keep Part II in its
own folder, separate from Part I. Open the RStudio **Terminal** tab and run:

```bash
mkdir -p ~/EJnPi/data
```

Then upload `skeleton.bib` to `~/EJnPi/` using the **Files** pane (Upload
button) and create your `.Rmd` there. Your final structure:

```
~/EJnPi/
  Analyzing_RPi_Enviroplus_YourLastName.Rmd
  skeleton.bib
  data/
    RPi_sensor.csv    <- downloaded from Canvas and renamed
```

## Course Functions

This assignment uses R functions written for this course that handle
downloading and cleaning the EPA data, so you can focus on the analysis.
They are maintained in the course GitHub repository.

**Browse the code:** [https://github.com/marclos/EJnPi/tree/master/docs](https://github.com/marclos/EJnPi/tree/master/docs)

You do not need to download or copy anything. Your Rmd loads them
automatically with this line (already in the setup section below):

```r
source("https://raw.githubusercontent.com/marclos/EJnPi/master/docs/EJnPi_functions.R")
```

If you are curious how the functions work, click the link above, open
`EJnPi_functions.R`, and read the comments. You are not expected to
modify them.

## Install Required Packages

Run this **once** in the RStudio Console before knitting:

```{r install-packages, eval=FALSE}
install.packages(c("tufte", "dplyr", "lubridate"))
```

Answer **yes** if prompted about a personal library. Do not put `install.packages()`
inside a chunk that knits.

## Download Your Data from Canvas

1. Log in to Canvas and navigate to course Files.
2. Download your group's CSV (named something like `PiZ03_ea30_sp26_v1.csv`).
3. In the RStudio **Files** pane, navigate to `~/EJnPi/data/`.
4. Click **Upload**, select the CSV, and **rename it `RPi_sensor.csv`** after uploading.

`r margin_note("All analysis chunks below include eval = rpi_exists and are silently skipped if the file is missing. The document will knit cleanly while you are waiting for the data.")`

---

# Load Libraries

Also set the working directory in your setup chunk so knitr resolves file paths
correctly on the server:

```{r load-functions-and-libraries}
knitr::opts_knit$set(root.dir = "~/EJnPi")   # adjust to your actual folder name

# Load shared course functions from GitHub
source("https://raw.githubusercontent.com/marclos/EJnPi/master/docs/EJnPi_functions.R")

library(dplyr)       # group_by(), summarise(), mutate()
library(lubridate)   # ymd_hms(), as.Date(), hour(), month()
```

`r margin_note("root.dir tells knitr where to look for files during knitting. Without it, 'data/RPi_sensor.csv' may not resolve even if the Console can find it.")`

---

# Check for Data File

Before loading anything, check whether the file actually exists. The variable
`rpi_exists` is used as the `eval` flag on every subsequent chunk: if `FALSE`,
those chunks are silently skipped so the document still knits.

```{r rpi-check}
rpi_file   <- "data/RPi_sensor.csv"
rpi_exists <- file.exists(rpi_file)

if (!rpi_exists) {
  message("Data file not found at: ", rpi_file)
  message("Download your CSV from Canvas, rename it RPi_sensor.csv, and place it in data/")
} else {
  message("Found: ", rpi_file)
}
```

---

# Load the Raw Data

`read.csv()` loads the file into a dataframe; each row is one 10-minute observation.
The `timestamp` column arrives as a character string. We parse it immediately into a
proper datetime object with `ymd_hms()`, then extract `Date`, `Hour`, and `Month` as
separate columns for later grouping.

`r margin_note("ymd_hms() is from lubridate and parses strings in the format 'YYYY-MM-DD HH:MM:SS' into a POSIXct datetime. as.Date() strips the time component, keeping only the calendar date.")`

```{r rpi-load, eval=rpi_exists}
RPi          <- read.csv(rpi_file)
RPi$Datetime <- ymd_hms(RPi$timestamp)   # parse character -> POSIXct datetime
RPi$Date     <- as.Date(RPi$Datetime)    # keep date only (drop time)
RPi$Hour     <- hour(RPi$Datetime)       # integer 0-23
RPi$Month    <- month(RPi$Datetime)      # integer 1-12
```

---

# Quality Control

## Inspect the Raw Data

Look at the first few rows and the structure before doing anything else. Confirm
that column names match the table in the Introduction and that numeric columns
are numeric (not character).

```{r rpi-inspect, eval=rpi_exists}
str(RPi)
```

```{r rpi-head, eval=rpi_exists}
head(RPi[, c("timestamp", "temperature_C", "humidity_pct",
             "pm1_ug_m3", "pm25_ug_m3", "pm10_ug_m3")])
```

## Remove Zero Readings

The PMS5003 occasionally fails to return a measurement and logs `0` rather than `NA`.
Zero is physically implausible for outdoor PM2.5 (background is typically 1--5 ug/m3)
and would pull the mean downward artificially. We filter these rows before any analysis.

```{r remove-zeros, eval=rpi_exists}
n_before <- nrow(RPi)
RPi      <- subset(RPi, pm25_ug_m3 > 0)
n_after  <- nrow(RPi)

message(n_before - n_after, " zero-PM2.5 rows removed (", n_before, " -> ", n_after, " remaining)")
```

## Summarize the Deployment

```{r deployment-summary, eval=rpi_exists}
message("Deployment: ", min(RPi$Date), " to ", max(RPi$Date))
message("Total 10-min readings after QC: ", nrow(RPi))
```

---

# Aggregate to Daily Means

`r newthought("EPA standards are defined on a 24-hour basis,")` so we must aggregate
the 10-minute readings to daily means before any threshold comparison. A day with
only one or two readings (due to sensor restart or power outage) is not representative
of a full 24-hour period. We count readings per day and flag low-coverage days.

`r margin_note("A full day of 10-minute readings produces 144 rows. Requiring N_readings >= 6 means at least one hour of data per day --- a conservative minimum.")`

```{r rpi-daily, eval=rpi_exists}
RPi.daily <- RPi %>%
  group_by(Date) %>%
  summarise(
    N_readings = n(),                                       # 10-min observations this day
    PM25_daily = mean(pm25_ug_m3, na.rm = TRUE),
    PM1_daily  = mean(pm1_ug_m3,  na.rm = TRUE),
    PM10_daily = mean(pm10_ug_m3, na.rm = TRUE),
    Temp_C     = mean(temperature_C, na.rm = TRUE),
    RH_pct     = mean(humidity_pct,  na.rm = TRUE),
    .groups    = "drop"
  )
```

```{r daily-coverage, eval=rpi_exists}
message("Total days with any data:              ", nrow(RPi.daily))
message("Days with >= 6 readings (at least 1h): ", sum(RPi.daily$N_readings >= 6))
```

---

# Time-Series Plot

`r newthought("Always plot your data before running statistics.")` Look for trends,
episodic spikes (possible wildfire smoke or combustion events), and whether the record
appears continuous and complete. Two horizontal reference lines mark both NAAQS thresholds.

Here is the output your code should produce:

```{r ts-output, echo=FALSE, eval=rpi_exists, fig.cap="Daily mean PM2.5 from the Raspberry Pi sensor. Red dashed line = 24-hr NAAQS (35 ug/m3). Orange dashed line = annual NAAQS (12 ug/m3). Days above red are acute exceedance events; prolonged elevation above orange implies chronic exposure concern."}
par(las = 1)
plot(PM25_daily ~ Date, data = RPi.daily,
     type = "b", pch = 16, cex = 0.7,
     xlab = "Date", ylab = "PM2.5 (ug/m3)",
     main = "Raspberry Pi -- Daily Mean PM2.5",
     ylim = c(0, max(RPi.daily$PM25_daily, 40) * 1.1))
abline(h = 35, col = "red",    lty = 2, lwd = 1.5)
abline(h = 12, col = "orange", lty = 2, lwd = 1.5)
legend("topleft",
       legend = c("24-hr NAAQS (35 ug/m3)", "Annual NAAQS (12 ug/m3)"),
       col    = c("red", "orange"), lty = 2, bty = "n")
```

`r margin_note("type='b' draws both points and connecting lines ('b' = both). type='l' gives lines only, type='p' points only. See ?plot.default for all options.")`

Scaffold:

```{r ts-scaffold, eval=FALSE}
par(las = ___)
plot(PM25_daily ~ Date, data = RPi.daily,
     type = ___,
     pch  = ___,
     cex  = ___,
     xlab = ___,
     ylab = ___,
     main = ___,
     ylim = c(0, max(RPi.daily$PM25_daily, ___) * 1.1))

abline(h = ___, col = "red",    lty = ___, lwd = ___)   # 24-hr standard
abline(h = ___, col = "orange", lty = ___, lwd = ___)   # annual standard

legend(___, legend = c(___, ___), col = c(___, ___), lty = ___, bty = "n")
```

---

# Summary Statistics

Before running the t-test, compute basic descriptive statistics. Examining the 95%
CI here gives you an informal preview of whether the interval overlaps 12 ug/m3.

```{r rpi-summary, eval=rpi_exists}
RPi.summary <- RPi.daily %>%
  summarise(
    N_days = n(),
    median = median(PM25_daily, na.rm = TRUE),
    mean   = mean(PM25_daily,   na.rm = TRUE),
    sd     = sd(PM25_daily,     na.rm = TRUE)
  ) %>%
  mutate(
    UCL95 = mean + qnorm(0.975) * sd / sqrt(N_days),
    LCL95 = mean - qnorm(0.975) * sd / sqrt(N_days)
  )

print(RPi.summary)
```

If the 95% CI lies entirely above 12 ug/m3, there is already strong descriptive evidence
that the deployment site exceeds the annual standard. The formal test follows.

---

# Hypothesis Test: Annual Standard

`r newthought("A one-sample t-test")` compares your sample mean to a known reference
value. Here the reference is the EPA annual PM2.5 standard of 12 ug/m3.

State the null hypothesis explicitly in your document:

$$H_0: \mu_{\text{site}} = 12 \; \text{ug/m}^3$$
$$H_A: \mu_{\text{site}} \neq 12 \; \text{ug/m}^3$$

A p-value below 0.05 means we reject $H_0$ at the 5% significance level. The direction
of the difference (above or below 12) is shown by the sign of the t-statistic and by
which side of 12 the confidence interval falls on.

```{r ttest-annual, eval=rpi_exists}
t_annual <- t.test(RPi.daily$PM25_daily, mu = 12)
print(t_annual)
```

**Reading the output:**

- `t` --- how many standard errors the sample mean is from 12.
- `df` --- degrees of freedom (N - 1 for a one-sample test).
- `p-value` --- probability of observing a result this extreme if $H_0$ were true.
- `95 percent confidence interval` --- range of plausible true means. If 12 is not
  inside this interval, the result is statistically significant.
- `mean of x` --- your sample mean daily PM2.5 in ug/m3.

`r margin_note("Regulatory determination requires a three-year mean. This test uses your deployment period only and should be interpreted with that limitation stated explicitly in your discussion.")`

---

# Exceedance: 24-Hour Standard

`r newthought("The 24-hour standard is not evaluated by a t-test")` but by counting
days above 35 ug/m3. Regulatorily, no more than one exceedance per year is permitted
on a three-year rolling average. We report the raw rate from your deployment.

```{r exceedance-count, eval=rpi_exists}
n_exceed   <- sum(RPi.daily$PM25_daily > 35, na.rm = TRUE)
pct_exceed <- 100 * n_exceed / nrow(RPi.daily)

message(sprintf("Days exceeding 35 ug/m3:  %d of %d  (%.1f%%)",
                n_exceed, nrow(RPi.daily), pct_exceed))
```

`r margin_note("sprintf() is a formatted string function. %d inserts an integer, %.1f inserts a decimal rounded to 1 place. See ?sprintf.")`

## Exceedance Histogram

A histogram of daily means shows the full distribution relative to both thresholds.
Days in the right tail represent acute exposure events.

Here is the output your code should produce:

```{r hist-output, echo=FALSE, eval=rpi_exists, fig.cap="Distribution of daily mean PM2.5 across the full deployment period. Red = 24-hr NAAQS (35 ug/m3); orange = annual NAAQS (12 ug/m3). Days to the right of the red line are acute exceedances."}
hist(RPi.daily$PM25_daily,
     breaks = 20, col = "lightblue", border = "white",
     main = "Distribution of Daily Mean PM2.5 (RPi)",
     xlab = "PM2.5 (ug/m3)")
abline(v = 35, col = "red",    lty = 2, lwd = 2)
abline(v = 12, col = "orange", lty = 2, lwd = 2)
legend("topright",
       legend = c("24-hr NAAQS (35)", "Annual NAAQS (12)"),
       col    = c("red", "orange"), lty = 2, bty = "n")
```

`r margin_note("For a histogram, use abline(v = ...) for vertical threshold lines rather than abline(h = ...). The v= argument draws a vertical line at that x-axis value.")`

Scaffold:

```{r hist-scaffold, eval=FALSE}
hist(RPi.daily$PM25_daily,
     breaks = ___,
     col    = ___,
     border = ___,
     main   = ___,
     xlab   = ___)

abline(v = ___, col = "red",    lty = ___, lwd = ___)   # 24-hr standard (vertical line)
abline(v = ___, col = "orange", lty = ___, lwd = ___)   # annual standard

legend(___, legend = c(___, ___), col = c(___, ___), lty = ___, bty = "n")
```

---

# Discussion

`r newthought("Write 2--4 paragraphs in your own words,")` citing 2--8 peer-reviewed
sources. Write flowing prose that connects ideas across the themes below --- not a
numbered list.

**Local sources and meteorology.** Where was your sensor deployed? What sources ---
traffic, cooking, industrial activity, outdoor burning --- are present nearby? How
does local meteorology (wind patterns, building geometry, proximity to open land)
affect what your sensor recorded [@Kleeman2005]?

**Comparison to EPA standards.** Report your t-test result and exceedance rate in
plain language. Are exceedances clustered in time (suggesting episodic events like
wildfires [@Liu2017]) or distributed throughout the deployment?

**Sensor limitations.** The PMS5003 is an optical particle counter, not a gravimetric
FRM sampler. It may over- or under-read in high humidity or when aerosol composition
differs from its factory calibration. How confident are you in the absolute values?
What would a co-location study with an FRM monitor tell you [@Sax2004]?

**Environmental justice.** Who lives near your deployment site? Look up census data
for the area. Do your results suggest that residents face disproportionate exposure?
What policy or community actions could reduce that burden?

---

# Session Information {-}

`r newthought("sessionInfo() records the exact computational environment")` used to
produce this document: the R version, operating system, and the version number of
every package that was loaded. It appears at the bottom of every well-documented R
analysis for one practical reason --- reproducibility.

If your output looks different from a classmate's, or your code throws an error that
works fine on another machine, the first diagnostic question is always: are you running
the same versions? This output answers that without anyone having to ask.

You do not need to interpret or discuss it. Just include it at the end of your
document and let it run.

```{r session-info}
sessionInfo()
```

# References
