---
title: "Decay Function of CO2"
author: "Marc Los Huertos"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setting Up CO2 Decay Function

The decay of CO2 in the atmosphere is a function of time. The IPCC AR5 report provides the following equation for the decay of CO2 over time:

\begin{equation}
CO2 (t) = a_0 + a_1 e^{-t/\tau_1} + a_2 e^{-t/\tau_2} + a_3 e^{-t/\tau_3}
\end{equation}

where $a_0 = 0.217$, $a_1 = 0.259$, $a_2 = 0.338$, $a_3 = 0.186$, $\tau_1 = 172.9$, $\tau_2 = 18.51$, and $\tau_3 = 1.186$.


```{r cars}
```{r}
# Function to calculate the decay of CO2 over time
decay_function_co2 <- function(t) {
  # IPCC AR5 parameters for CO2
  a0 <- 0.217
  a1 <- 0.259
  a2 <- 0.338
  a3 <- 0.186
  tau1 <- 172.9
  tau2 <- 18.51
  tau3 <- 1.186
  
  decay <- a0 + a1 * exp(-t / tau1) + a2 * exp(-t / tau2) + a3 * exp(-t / tau3)
  return(decay)
}

```


```{r}
timeframe = 0:100
```



```{r, echo=TRUE}
plot(timeframe, decay_function_co2(timeframe), 
     type = "l", 
     xlab = "Time (years)", 
     ylab = "Decay", 
     main = "Decay Estimates for CO2")
```


```{r, echo=TRUE}
par(las=1)
plot(timeframe, decay_function_co2(timeframe)*100, 
     type = "l", 
     xlab = "Time (years)", 
     ylab = "Percent CO2 Remaining", 
     main = "Decay Estimates for CO2")
```

# 500 Years
```{r}
timeframe = 0:500
```

```{r, echo=FALSE}
par(las=1)
plot(timeframe, decay_function_co2(timeframe)*100, 
     type = "l", 
     xlab = "Time (years)", 
     ylab = "Percent CO2 Remaining", 
     main = "Decay Estimates for CO2 (500 years)")
```

# 1000 Years
```{r, echo=FALSE}
timeframe = 0:1000
```

```{r, echo=FALSE}
par(las=1)
plot(timeframe, decay_function_co2(timeframe)*100, type = "l", xlab = "Time (years)", ylab = "Percent CO2 Remaining", main = "Decay Estimates for CO2 (1000 years)")
```