---
title: "Keeling Curve Demo Code"
author: "Marc Los Huertos"
date: "2024-01-23"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here) # used to help move between github users, and with knitr sessions
```

# Today's Goal: Create a Compelling Plot using Mauna Loa CO2 Data


What is the equation for a line?

$y = mx + b$, where

m = slope, b = intercept

In statistics, we usually write it as a function where y is a function of x or: 

$y =f(x)$

that includes things like the slope, intercept and error, which are generally 
referred to as coefficents, and each have associated error, since real data 
never fall on the line. 

## Read Data

```{r}
#filepath = here("02_keeling","Mauna_Loa.csv")
filename = "/home/mwl04747/RTricks/02_keeling/Mauna_Loa.csv"
maunaloa <- read.csv(filename)
```






































## Plot Data

```{r}
plot(average~decimal.date, data=maunaloa, ty="l", xlab="Year", ylab="CO2 Concentration (ppm)")

maunaloa.lm=lm(average~decimal.date, data=maunaloa)
abline(maunaloa.lm, col="red")
```























## Check Assumptions
```{r}
par(mfrow=c(2,2))
plot(maunaloa.lm)
```





















## Create Loess Plots

```{r}
plot(average~decimal.date, data=maunaloa, ty="l", xlab="Year", ylab="CO2 Concentration (ppmv)", las=1, main="Carbon Dioxide Concentration at Mauna Loa Observatory", cex.lab=1.3)

fitted <- loess(average~decimal.date, data=maunaloa, span=1/3)
lines(maunaloa$decimal.date, fitted$fitted, col="blue", lwd=2)
```

See handout for more details and more code and examples.
