---
title: "Keeling Curve Demo Code"
author: "Marc Los Huertos"
date: "2024-01-23"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here) # used to help move between github users, and with knitr sessions
```

# Today's Goal: Create a Compelling Plot using Mauna Loa CO2 Data


What is the equation for a line?

$y = mx + b$, where

m = slope, b = intercept

In statistics, we usually write it as a function where y is a function of x or: 

$y =f(x)$

that includes things like the slope, intercept and error, which are generally 
referred to as coefficents, and each have associated error, since real data 
never fall on the line. 

## Read Data

```{r}
#filepath = here("02_keeling","Mauna_Loa.csv")
filename = "/home/mwl04747/RTricks/03_Keeling_GWP/Mauna_Loa.csv"
maunaloa <- read.csv(filename)
```









# Generate code to use RF to estimate GWP for Carbon dioxide and methane

Approximate RF values for common greenhouse gases:
Carbon dioxide (CO2): 2.1 W/m² 100 years
Methane (CH4): 0.84 W/m²   7-12 years
Nitrous oxide (N2O): 0.34 W/m²  >100 years

calculate_gwp Function: This function takes radiative forcing values and a time horizon as inputs and computes the GWP for that time horizon. The decay rate is an example value and should be set according to the specific gas being analyzed.

Radiative Forcing Values: Replace the radiative_forcing vector with the actual published radiative forcing values for the gas of interest.

Time Horizons: The time_horizons vector contains the different time horizons for which the GWP is to be calculated.

GWP Calculation: The sapply function is used to apply the calculate_gwp function to each time horizon.

Output: The GWP values for the different time horizons are printed in a data frame.

Make sure to replace the example radiative forcing values with the actual published values for your specific analysis. Additionally, adjust the decay rate as needed based on the specific properties of the gas being studied.

```{r}
# Function to calculate GWP
calculate_gwp <- function(radiative_forcing, time_horizon) {
  # Constants
  decay_rate <- 0.5  # Example decay rate (this would depend on the specific gas)
  
  # Calculate GWP for the given time horizon
  gwp <- sum(radiative_forcing * exp(-decay_rate * (0:time_horizon)))
  
  return(gwp)
}

# Example radiative forcing values (W/m^2)
# These values should be replaced with actual published radiative forcing values
radiative_forcing <- c(1.2, 1.1, 1.05, 1.0, 0.95, 0.9, 0.85, 0.8, 0.75, 0.7)

# Time horizons
time_horizons <- c(10, 20, 50, 100, 500)

# Calculate GWP for each time horizon
gwp_values <- sapply(time_horizons, function(th) calculate_gwp(radiative_forcing, th))

# Print GWP values
gwp_data <- data.frame(
  Time_Horizon = time_horizons,
  GWP = gwp_values
)

print(gwp_data)
```


```{r}
# Read in the data
# Function to calculate GWP

# radiative forcing for methane from fossil fuels
radiative_forcing <- c(5.7*10^-4); radiative_forcing # W/m^2/ppb
radiative_forcingco2 <- c(1.33^-5); radiative_forcingco2 # W/m^2/ppb


radiative_forcing_kg <- radiative_forcing * 1e9; radiative_forcing_kg # W/m^2/kg
radiative_forcing_kg = 1.3e-13

# atmospheric lifetime for methane
lifetime <- c(11.8) # years
lifetimeco2 <- c(1) # years
# Calculate decay rate
decay_rate <- 1 / lifetime; decay_rate
decay_rateco2 <- 1 / lifetimeco2; decay_rateco2


# Calculate GWP for the given time horizon (100 years)
gwp <- radiative_forcing * exp(-decay_rate * (0:100)); gwp
gwpco2 <- radiative_forcingco2 * exp(-decay_rateco2 * (0:100)); gwpco2


plot(0:100, gwp, type = "l", xlab = "Time (years)", ylab = "GWP", main = "GWP for Methane")

# Calculate GWP relative for carbon dixode
gwp_relative <- gwp / gwpco2; gwp_relative

print(sum(gwp_relative))

gwp_relative <- sum(gwp) / sum(gwpco2); gwp_relative


print(gwp)
```


```{r}



calculate_gwp <- function(radiative_forcing, time_horizon, decay_rate = 0.5 ) {
  # Constants
  # decay_rate <- 0.5  # Example decay rate (this would depend on the specific gas)
  
  # Calculate GWP for the given time horizon
  gwp <- sum(radiative_forcing * exp(-decay_rate * (0:time_horizon)))
  return(gwp)
}

```


```{r}

# Example radiative forcing values (W/m^2)
# These values should be replaced with actual published radiative forcing values
RF <- data.frame(Gas = c("CO2", "CH4", "N2O"), RF = c(2.1, 0.84, .34)); RF
radiative_forcing = 2.1

# Time horizons
time_horizons <- c(10, 20, 50, 100, 500)

# Calculate GWP for each time horizon
gwp_values <- sapply(time_horizons, function(th) calculate_gwp(radiative_forcing, th))

# Print GWP values
gwp_data <- data.frame(
  Time_Horizon = time_horizons,
  GWP = gwp_values
)

print(gwp_data)

```


```{r}
# Function to calculate GWP for methane
calculate_gwp_methane <- function(time_horizon) {
  # IPCC AR5 parameters for methane
  radiative_efficiency <- 5.7e-4  # W/m²/ppb
  lifetime <- 11.8  # years
  
  # CO2 radiative efficiency (reference gas)
  radiative_efficiency_co2 <- 1.33e-5  # W/m²/ppb
  
  # Function to calculate the time-dependent decay of CH4
  decay_function_ch4 <- function(t) {
    exp(-t / lifetime)
  }
  
  # Integrate the decay function over the time horizon to get the integrated RF
  integrate_rf_ch4 <- integrate(decay_function_ch4, 0, time_horizon)$value
  
  # Calculate the GWP for methane
  gwp_ch4 <- (radiative_efficiency / radiative_efficiency_co2) * integrate_rf_ch4
  
  return(gwp_ch4)
}

# Time horizons
time_horizons <- c(20, 100, 500)

# Calculate GWP for each time horizon
gwp_values_ch4 <- sapply(time_horizons, calculate_gwp_methane)

# Print GWP values
gwp_data_ch4 <- data.frame(
  Time_Horizon = time_horizons,
  GWP_CH4 = gwp_values_ch4
)

print(gwp_data_ch4)
```


```{r}
# Function to calculate the decay of CO2 over time
decay_function_co2 <- function(t) {
  # IPCC AR5 parameters for CO2
  a0 <- 0.217
  a1 <- 0.259
  a2 <- 0.338
  a3 <- 0.186
  tau1 <- 172.9
  tau2 <- 18.51
  tau3 <- 1.186
  
  decay <- a0 + a1 * exp(-t / tau1) + a2 * exp(-t / tau2) + a3 * exp(-t / tau3)
  return(decay)
}

timeframe = 0:100

plot(timeframe, decay_function_co2(timeframe), type = "l", xlab = "Time (years)", ylab = "Decay", main = "Decay Estimates for CO2")



decay_function_co2 <- function(t) {
  lifetime_co2 <- 100  # years
  decay <- exp(-t / lifetime_co2)
  return(decay)
}

# Function to calculate the decay of CH4 over time
decay_function_ch4 <- function(t) {
  lifetime_ch4 <- 11.8  # years
  decay <- exp(-t / lifetime_ch4)
  return(decay)
}

# Function to calculate GWP for methane
calculate_gwp_methane <- function(time_horizon) {
  # IPCC AR5 parameters for methane
  radiative_efficiency_ch4 <- 9.7e-4  # W/m²/ppb
  
  # Radiative efficiency for CO2 (reference gas)
  radiative_efficiency_co2 <- 1.33e-5  # W/m²/ppb
  
  # Integrate the decay function over the time horizon for CH4 and CO2
  integrate_rf_ch4 <- integrate(decay_function_ch4, 0, time_horizon)$value
  integrate_rf_co2 <- integrate(decay_function_co2, 0, time_horizon)$value
  
  # Calculate the GWP for methane
  gwp_ch4 <- (radiative_efficiency_ch4 * integrate_rf_ch4) / (radiative_efficiency_co2 * integrate_rf_co2)
  
  return(gwp_ch4)
}

# Time horizons
time_horizons <- c(20, 100, 500)

# Calculate GWP for each time horizon
gwp_values_ch4 <- sapply(time_horizons, calculate_gwp_methane)

# Print GWP values
gwp_data_ch4 <- data.frame(
  Time_Horizon = time_horizons,
  GWP_CH4 = gwp_values_ch4
)

print(gwp_data_ch4)
```



```{r}
# Function to calculate the decay of CO2 over time
decay_function_co2 <- function(t) {
  # IPCC AR5 parameters for CO2
  a0 <- 0.217
  a1 <- 0.259
  a2 <- 0.338
  a3 <- 0.186
  tau1 <- 172.9
  tau2 <- 18.51
  tau3 <- 1.186
  
  decay <- a0 + a1 * exp(-t / tau1) + a2 * exp(-t / tau2) + a3 * exp(-t / tau3)
  return(decay)
}

# Function to calculate the decay of CH4 over time
decay_function_ch4 <- function(t) {
  lifetime_ch4 <- 11.8  # years
  decay <- exp(-t / lifetime_ch4)
  return(decay)
}

# Function to calculate GWP for methane
calculate_gwp_methane <- function(time_horizon) {
  # IPCC AR5 parameters for methane
  radiative_efficiency_ch4 <- 5.7e-4  # W/m²/ppb
  
  # Radiative efficiency for CO2 (reference gas)
  radiative_efficiency_co2 <- 1.33e-5  # W/m²/ppb
  
  # Integrate the decay function over the time horizon for CH4 and CO2
  integrate_rf_ch4 <- integrate(decay_function_ch4, 0, time_horizon)$value
  integrate_rf_co2 <- integrate(decay_function_co2, 0, time_horizon)$value
  
  # Calculate the GWP for methane
  gwp_ch4 <- (radiative_efficiency_ch4 * integrate_rf_ch4) / (radiative_efficiency_co2 * integrate_rf_co2)
  
  return(gwp_ch4)
}

# Time horizons
time_horizons <- c(20, 100, 500)

# Calculate GWP for each time horizon
gwp_values_ch4 <- sapply(time_horizons, calculate_gwp_methane)

# Print GWP values
gwp_data_ch4 <- data.frame(
  Time_Horizon = time_horizons,
  GWP_CH4 = round(gwp_values_ch4, 2)  # Round the results to 2 decimal places for clarity
)

print(gwp_data_ch4)
```


```{r}
# Function to calculate the decay of CO2 over time
decay_function_co2 <- function(t) {
  lifetime_co2 <- 100  # years
  decay <- exp(-t / lifetime_co2)
  return(decay)
}


decay_function_co2 <- function(t) {
  # IPCC AR5 parameters for CO2
  a0 <- 0.217
  a1 <- 0.259
  a2 <- 0.338
  a3 <- 0.186
  tau1 <- 172.9
  tau2 <- 18.51
  tau3 <- 1.186
  
  decay <- a0 + a1 * exp(-t / tau1) + a2 * exp(-t / tau2) + a3 * exp(-t / tau3)
  return(decay)
}

# Function to calculate the decay of CH4 over time
decay_function_ch4 <- function(t) {
  lifetime_ch4 <- 11.8  # years
  decay <- exp(-t / lifetime_ch4)
  return(decay)
}

# Function to calculate GWP for methane
calculate_gwp_methane <- function(time_horizon) {
  # IPCC AR5 parameters for methane
  radiative_efficiency_ch4 <- 5.7e-4  # W/m²/ppb
  molecular_weight_ch4 <- 16.04  # g/mol
  
  # IPCC AR5 parameters for CO2
  radiative_efficiency_co2 <- 1.33e-5  # W/m²/ppb
  molecular_weight_co2 <- 44.01  # g/mol
  
  # Number of moles in 1 ton (1,000,000 grams) of each gas
  moles_ch4 <- 1e6 / molecular_weight_ch4
  moles_co2 <- 1e6 / molecular_weight_co2
  
  # Convert moles to ppb equivalents
  ppb_ch4 <- moles_ch4 * 1e9
  ppb_co2 <- moles_co2 * 1e9
  
  # Integrate the decay function over the time horizon for CH4 and CO2
  integrate_rf_ch4 <- integrate(function(t) decay_function_ch4(t) * radiative_efficiency_ch4 * ppb_ch4, 0, time_horizon)$value
  integrate_rf_co2 <- integrate(function(t) decay_function_co2(t) * radiative_efficiency_co2 * ppb_co2, 0, time_horizon)$value
  
  # Calculate the GWP for methane per ton
  gwp_ch4 <- integrate_rf_ch4 / integrate_rf_co2
  
  return(gwp_ch4)
}

# Time horizons
time_horizons <- c(20, 100, 500)

# Calculate GWP for each time horizon
gwp_values_ch4 <- sapply(time_horizons, calculate_gwp_methane)

# Print GWP values
gwp_data_ch4 <- data.frame(
  Time_Horizon = time_horizons,
  GWP_CH4 = round(gwp_values_ch4, 2)  # Round the results to 2 decimal places for clarity
)

print(gwp_data_ch4)
```




















## Plot Data

```{r}
plot(average~decimal.date, data=maunaloa, ty="l", xlab="Year", ylab="CO2 Concentration (ppm)")

maunaloa.lm=lm(average~decimal.date, data=maunaloa)
abline(maunaloa.lm, col="red")
```























## Check Assumptions
```{r}
par(mfrow=c(2,2))
plot(maunaloa.lm)
```





















## Create Loess Plots

```{r}
plot(average~decimal.date, data=maunaloa, ty="l", xlab="Year", ylab="CO2 Concentration (ppmv)", las=1, main="Carbon Dioxide Concentration at Mauna Loa Observatory", cex.lab=1.3)

fitted <- loess(average~decimal.date, data=maunaloa, span=1/3)
lines(maunaloa$decimal.date, fitted$fitted, col="blue", lwd=2)
```

See handout for more details and more code and examples.
